### 2022-06-24 class-info/orig-latex-template.tex
Template files and a makefile

% $Id: template.tex 11 2007-04-03 22:25:53Z jpeltier $

\documentclass{vgtc}                          % final (conference style)
%\documentclass[review]{vgtc}                 % review
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint]{vgtc}               % preprint
%\documentclass[electronic]{vgtc}             % electronic version

%% Uncomment one of the lines above depending on where your paper is
%% in the conference process. ``review'' and ``widereview'' are for review
%% submission, ``preprint'' is for pre-publication, and the final version
%% doesn't use a specific qualifier. Further, ``electronic'' includes
%% hyperreferences for more convenient online viewing.

%% Please use one of the ``review'' options in combination with the
%% assigned online id (see below) ONLY if your paper uses a double blind
%% review process. Some conferences, like IEEE Vis and InfoVis, have NOT
%% in the past.

%% Figures should be in CMYK or Grey scale format, otherwise, colour 
%% shifting may occur during the printing process.

%% These few lines make a distinction between latex and pdflatex calls and they
%% bring in essential packages for graphics and font handling.
%% Note that due to the \DeclareGraphicsExtensions{} call it is no longer necessary
%% to provide the the path and extension of a graphics file:
%% \includegraphics{diamondrule} is completely sufficient.
%%
\ifpdf%                                % if we use pdflatex
  \pdfoutput=1\relax                   % create PDFs from pdfLaTeX
  \pdfcompresslevel=9                  % PDF Compression
  \pdfoptionpdfminorversion=7          % create PDF 1.7
  \ExecuteOptions{pdftex}
  \usepackage{graphicx}                % allow us to embed graphics files
  \DeclareGraphicsExtensions{.pdf,.png,.jpg,.jpeg} % for pdflatex we expect .pdf, .png, or .jpg files
\else%                                 % else we use pure latex
  \ExecuteOptions{dvips}
  \usepackage{graphicx}                % allow us to embed graphics files
  \DeclareGraphicsExtensions{.eps}     % for pure latex we expect eps files
\fi%

%% it is recomended to use ``\autoref{sec:bla}'' instead of ``Fig.~\ref{sec:bla}''
\graphicspath{{figures/}{pictures/}{images/}{./}} % where to search for the images

\usepackage{microtype}                 % use micro-typography (slightly more compact, better to read)
\PassOptionsToPackage{warn}{textcomp}  % to address font issues with \textrightarrow
\usepackage{textcomp}                  % use better special symbols
\usepackage{mathptmx}                  % use matching math font
\usepackage{times}                     % we use Times as the main font
\renewcommand*\ttdefault{txtt}         % a nicer typewriter font
\usepackage{cite}                      % needed to automatically sort the references
\usepackage{tabu}                      % only used for the table example
\usepackage{booktabs}                  % only used for the table example
%% We encourage the use of mathptmx for consistent usage of times font
%% throughout the proceedings. However, if you encounter conflicts
%% with other math-related packages, you may want to disable it.

%% If you are submitting a paper to a conference for review with a double
%% blind reviewing process, please replace the value ``0'' below with your
%% OnlineID. Otherwise, you may safely leave it at ``0''.
\onlineid{0}

%% declare the category of your paper, only shown in review mode
\vgtccategory{Research}

%% allow for this line if you want the electronic option to work properly
\vgtcinsertpkg

%% In preprint mode you may define your own headline. If not, the default IEEE copyright message will appear in preprint mode.
%\preprinttext{To appear in an IEEE VGTC sponsored conference.}

%% This adds a link to the version of the paper on IEEEXplore
%% Uncomment this line when you produce a preprint version of the article 
%% after the article receives a DOI for the paper from IEEE
%\ieeedoi{xx.xxxx/TVCG.201x.xxxxxxx}

%% Paper title.

\title{Global Illumination for Fun and Profit}

%% This is how authors are specified in the conference style

%% Author and Affiliation (single author).
%%\author{Roy G. Biv\thanks{e-mail: roy.g.biv@aol.com}}
%%\affiliation{\scriptsize Allied Widgets Research}

%% Author and Affiliation (multiple authors with single affiliations).
%%\author{Roy G. Biv\thanks{e-mail: roy.g.biv@aol.com} %
%%\and Ed Grimley\thanks{e-mail:ed.grimley@aol.com} %
%%\and Martha Stewart\thanks{e-mail:martha.stewart@marthastewart.com}}
%%\affiliation{\scriptsize Martha Stewart Enterprises \\ Microsoft Research}

%% Author and Affiliation (multiple authors with multiple affiliations)
\author{Roy G. Biv\thanks{e-mail: roy.g.biv@aol.com}\\ %
        \scriptsize Starbucks Research %
\and Ed Grimley\thanks{e-mail: ed.grimley@aol.com}\\ %
     \scriptsize Grimley Widgets, Inc. %
\and Martha Stewart\thanks{e-mail: martha.stewart@marthastewart.com}\\ %
     \parbox{1.4in}{\scriptsize \centering Martha Stewart Enterprises \\ Microsoft Research}}

%% A teaser figure can be included as follows
\teaser{
  \centering
  \includegraphics[width=\linewidth]{CypressView}
  \caption{In the Clouds: Vancouver from Cypress Mountain. Note that the teaser may not be wider than the abstract block.}
  \label{fig:teaser}
}

%% Abstract section.
\abstract{Duis autem vel eum iriure dolor in hendrerit in vulputate
velit esse molestie consequat, vel illum dolore eu feugiat nulla
facilisis at vero eros et accumsan et iusto odio dignissim qui blandit
praesent luptatum zzril delenit augue duis dolore te feugait nulla
facilisi. Lorem ipsum dolor sit amet, consectetuer adipiscing elit,
sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna
aliquam erat volutpat. Ut wisi enim ad minim veniam, quis nostrud exerci tation ullamcorper
suscipit lobortis nisl ut aliquip ex ea commodo consequat. Duis autem
vel eum iriure dolor in hendrerit in vulputate velit esse molestie
consequat, vel illum dolore eu feugiat nulla facilisis at vero eros et
accumsan et iusto odio dignissim qui blandit praesent luptatum zzril
delenit augue duis dolore te feugait nulla facilisi.%
} % end of abstract

%% ACM Computing Classification System (CCS). 
%% See <http://www.acm.org/about/class> for details.
%% We recommend the 2012 system <http://www.acm.org/about/class/class/2012>
%% For the 2012 system use the ``\CCScatTwelve'' which command takes four arguments.
%% The 1998 system <http://www.acm.org/about/class/class/2012> is still possible
%% For the 1998 system use the ``\CCScat'' which command takes four arguments.
%% In both cases the last two arguments (1998) or last three (2012) can be empty.

\CCScatlist{
  \CCScatTwelve{Human-centered computing}{Visu\-al\-iza\-tion}{Visu\-al\-iza\-tion techniques}{Treemaps};
  \CCScatTwelve{Human-centered computing}{Visu\-al\-iza\-tion}{Visualization design and evaluation methods}{}
}

%\CCScatlist{
  %\CCScat{H.5.2}{User Interfaces}{User Interfaces}{Graphical user interfaces (GUI)}{};
  %\CCScat{H.5.m}{Information Interfaces and Presentation}{Miscellaneous}{}{}
%}

%% Copyright space is enabled by default as required by guidelines.
%% It is disabled by the 'review' option or via the following command:
% \nocopyrightspace

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%% START OF THE PAPER %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% The ``\maketitle'' command must be the first command after the
%% ``\begin{document}'' command. It prepares and prints the title block.

%% the only exception to this rule is the \firstsection command
\firstsection{Introduction}

\maketitle

%% \section{Introduction} %for journal use above \firstsection{..} instead
This template is for papers of VGTC-sponsored conferences which are \emph{\textbf{not}} published in a special issue of TVCG.

\section{Using the Style Template}

\begin{itemize}
\item If you receive compilation errors along the lines of ``\texttt{Package ifpdf Error: Name clash, \textbackslash ifpdf is already defined}'' then please add a new line ``\texttt{\textbackslash let\textbackslash ifpdf\textbackslash relax}'' right after the ``\texttt{\textbackslash documentclass[journal]\{vgtc\}}'' call. Note that your error is due to packages you use that define ``\texttt{\textbackslash ifpdf}'' which is obsolete (the result is that \texttt{\textbackslash ifpdf} is defined twice); these packages should be changed to use ifpdf package instead.
\item The style uses the hyperref package, thus turns references into internal links. We thus recommend to make use of the ``\texttt{\textbackslash autoref\{reference\}}'' call (instead of ``\texttt{Figure\~{}\textbackslash ref\{reference\}}'' or similar) since ``\texttt{\textbackslash autoref\{reference\}}'' turns the entire reference into an internal link, not just the number. Examples: \autoref{fig:sample} and \autoref{tab:vis_papers}.
\item The style automatically looks for image files with the correct extension (eps for regular \LaTeX; pdf, png, and jpg for pdf\LaTeX), in a set of given subfolders (figures/, pictures/, images/). It is thus sufficient to use ``\texttt{\textbackslash includegraphics\{CypressView\}}'' (instead of ``\texttt{\textbackslash includegraphics\{pictures/CypressView.jpg\}}'').
\item For adding hyperlinks and DOIs to the list of references, you can use ``\texttt{\textbackslash bibliographystyle\{abbrv-doi-hyperref-narrow\}}'' (instead of ``\texttt{\textbackslash bibliographystyle\{abbrv\}}''). It uses the doi and url fields in a bib\TeX\ entry and turns the entire reference into a link, giving priority to the doi. The doi can be entered with or without the ``\texttt{http://dx.doi.org/}'' url part. See the examples in the bib\TeX\ file and the bibliography at the end of this template.\\[1em]
\textbf{Note 1:} occasionally (for some \LaTeX\ distributions) this hyper-linked bib\TeX\ style may lead to \textbf{compilation errors} (``\texttt{pdfendlink ended up in different nesting level ...}'') if a reference entry is broken across two pages (due to a bug in hyperref). In this case make sure you have the latest version of the hyperref package (i.\,e., update your \LaTeX\ installation/packages) or, alternatively, revert back to ``\texttt{\textbackslash bibliographystyle\{abbrv-doi-narrow\}}'' (at the expense of removing hyperlinks from the bibliography) and try ``\texttt{\textbackslash bibliographystyle\{abbrv-doi-hyperref-narrow\}}'' again after some more editing.\\[1em]
\textbf{Note 2:} the ``\texttt{-narrow}'' versions of the bibliography style use the font ``PTSansNarrow-TLF'' for typesetting the DOIs in a compact way. This font needs to be available on your \LaTeX\ system. It is part of the \href{https://www.ctan.org/pkg/paratype}{``paratype'' package}, and many distributions (such as MikTeX) have it automatically installed. If you do not have this package yet and want to use a ``\texttt{-narrow}'' bibliography style then use your \LaTeX\ system's package installer to add it. If this is not possible you can also revert to the respective bibliography styles without the ``\texttt{-narrow}'' in the file name.\\[1em]
DVI-based processes to compile the template apparently cannot handle the different font so, by default, the template file uses the \texttt{abbrv-doi} bibliography style but the compiled PDF shows you the effect of the \texttt{abbrv-doi-hyperref-narrow} style.
\end{itemize}

\section{Bibliography Instructions}

\begin{itemize}
\item Sort all bibliographic entries alphabetically but the last name of the first author. This \LaTeX/bib\TeX\ template takes care of this sorting automatically.
\item Merge multiple references into one; e.\,g., use \cite{Max-1995-OMF,Kitware-2003} (not \cite{Kitware-2003}\cite{Max-1995-OMF}). Within each set of multiple references, the references should be sorted in ascending order. This \LaTeX/bib\TeX\ template takes care of both the merging and the sorting automatically.
\item Verify all data obtained from digital libraries, even ACM's DL and IEEE Xplore  etc.\ are sometimes wrong or incomplete.
\item Do not trust bibliographic data from other services such as Mendeley.com, Google Scholar, or similar; these are even more likely to be incorrect or incomplete.
\item Articles in journal---items to include:
  \begin{itemize}
  \item author names
	\item title
	\item journal name
	\item year
	\item volume
	\item number
	\item month of publication as variable name (i.\,e., \{jan\} for January, etc.; month ranges using \{jan \#\{/\}\# feb\} or \{jan \#\{-{}-\}\# feb\})
  \end{itemize}
\item use journal names in proper style: correct: ``IEEE Transactions on Visualization and Computer Graphics'', incorrect: ``Visualization and Computer Graphics, IEEE Transactions on''
\item Papers in proceedings---items to include:
  \begin{itemize}
  \item author names
	\item title
	\item abbreviated proceedings name: e.\,g., ``Proc.\textbackslash{} CONF\_ACRONYNM'' without the year; example: ``Proc.\textbackslash{} CHI'', ``Proc.\textbackslash{} 3DUI'', ``Proc.\textbackslash{} Eurographics'', ``Proc.\textbackslash{} EuroVis''
	\item year
	\item publisher
	\item town with country of publisher (the town can be abbreviated for well-known towns such as New York or Berlin)
  \end{itemize}
\item article/paper title convention: refrain from using curly brackets, except for acronyms/proper names/words following dashes/question marks etc.; example:
\begin{itemize}
	\item paper ``Marching Cubes: A High Resolution 3D Surface Construction Algorithm''
	\item should be entered as ``\{M\}arching \{C\}ubes: A High Resolution \{3D\} Surface Construction Algorithm'' or  ``\{M\}arching \{C\}ubes: A high resolution \{3D\} surface construction algorithm''
	\item will be typeset as ``Marching Cubes: A high resolution 3D surface construction algorithm''
\end{itemize}
\item for all entries
\begin{itemize}
	\item DOI can be entered in the DOI field as plain DOI number or as DOI url; alternative: a url in the URL field
	\item provide full page ranges AA-{}-BB
\end{itemize}
\item when citing references, do not use the reference as a sentence object; e.\,g., wrong: ``In \cite{Lorensen-1987-MCA} the authors describe \dots'', correct: ``Lorensen and Cline \cite{Lorensen-1987-MCA} describe \dots''
\end{itemize}

\section{Example Section}

Lorem\marginpar{\small You can use the margins for comments while editing the submission, but please remove the marginpar comments for submission.} ipsum dolor sit amet, consetetur sadipscing elitr, sed diam
nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
dolore magna aliquyam erat, sed diam
voluptua~\cite{Kitware-2003,Max-1995-OMF}. At vero eos et accusam et
justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit
amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor
invidunt ut labore et dolore magna aliquyam erat, sed diam
voluptua. At vero eos et accusam et justo duo dolores et ea
rebum. Stet clita kasd gubergren, no sea takimata sanctus est.

\section{Exposition}

Duis autem vel eum iriure dolor in hendrerit in vulputate velit esse
molestie consequat, vel illum dolore eu feugiat nulla facilisis at
vero eros et accumsan et iusto odio dignissim qui blandit praesent
luptatum zzril delenit augue duis dolore te feugait nulla
facilisi. Lorem ipsum dolor sit amet, consectetuer adipiscing elit,
sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna
aliquam erat volutpat~\cite{Kindlmann-1999-SAG}.

\begin{equation}
\sum_{j=1}^{z} j = \frac{z(z+1)}{2}
\end{equation}

Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam
nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
takimata sanctus est Lorem ipsum dolor sit amet.

\subsection{Lorem ipsum}

Lorem ipsum dolor sit amet (see \autoref{tab:vis_papers}), consetetur sadipscing elitr, sed diam
nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit
amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor
invidunt ut labore et dolore magna aliquyam erat, sed diam
voluptua. At vero eos et accusam et justo duo dolores et ea
rebum.

\begin{table}[tb]
  \caption{VIS/VisWeek accepted/presented papers: 1990--2016.}
  \label{tab:vis_papers}
  \scriptsize%
	\centering%
  \begin{tabu}{%
	r%
	*{7}{c}%
	*{2}{r}%
	}
  \toprule
   year & \rotatebox{90}{Vis/SciVis} &   \rotatebox{90}{SciVis conf} &   \rotatebox{90}{InfoVis} &   \rotatebox{90}{VAST} &   \rotatebox{90}{VAST conf} &   \rotatebox{90}{TVCG @ VIS} &   \rotatebox{90}{CG\&A @ VIS} &   \rotatebox{90}{VIS/VisWeek} \rotatebox{90}{incl. TVCG/CG\&A}   &   \rotatebox{90}{VIS/VisWeek} \rotatebox{90}{w/o TVCG/CG\&A}   \\
  \midrule
	2016 & 30 &   & 37 & 33 & 15 & 23 & 10 & 148 & 115 \\
  2015 & 33 & 9 & 38 & 33 & 14 & 17 & 15 & 159 & 127 \\
  2014 & 34 &   & 45 & 33 & 21 & 20 &   & 153 & 133 \\
  2013 & 31 &   & 38 & 32 &   & 20 &   & 121 & 101 \\
  2012 & 42 &   & 44 & 30 &   & 23 &   & 139 & 116 \\
  2011 & 49 &   & 44 & 26 &   & 20 &   & 139 & 119 \\
  2010 & 48 &   & 35 & 26 &   &   &   & 109 & 109 \\
  2009 & 54 &   & 37 & 26 &   &   &   & 117 & 117 \\
  2008 & 50 &   & 28 & 21 &   &   &   & 99 & 99 \\
  2007 & 56 &   & 27 & 24 &   &   &   & 107 & 107 \\
  2006 & 63 &   & 24 & 26 &   &   &   & 113 & 113 \\
  2005 & 88 &   & 31 &   &   &   &   & 119 & 119 \\
  2004 & 70 &   & 27 &   &   &   &   & 97 & 97 \\
  2003 & 74 &   & 29 &   &   &   &   & 103 & 103 \\
  2002 & 78 &   & 23 &   &   &   &   & 101 & 101 \\
  2001 & 74 &   & 22 &   &   &   &   & 96 & 96 \\
  2000 & 73 &   & 20 &   &   &   &   & 93 & 93 \\
  1999 & 69 &   & 19 &   &   &   &   & 88 & 88 \\
  1998 & 72 &   & 18 &   &   &   &   & 90 & 90 \\
  1997 & 72 &   & 16 &   &   &   &   & 88 & 88 \\
  1996 & 65 &   & 12 &   &   &   &   & 77 & 77 \\
  1995 & 56 &   & 18 &   &   &   &   & 74 & 74 \\
  1994 & 53 &   &   &   &   &   &   & 53 & 53 \\
  1993 & 55 &   &   &   &   &   &   & 55 & 55 \\
  1992 & 53 &   &   &   &   &   &   & 53 & 53 \\
  1991 & 50 &   &   &   &   &   &   & 50 & 50 \\
  1990 & 53 &   &   &   &   &   &   & 53 & 53 \\
  \midrule
  \textbf{sum} & \textbf{1545} & \textbf{9} & \textbf{632} & \textbf{310} & \textbf{50} & \textbf{123} & \textbf{25} & \textbf{2694} & \textbf{2546} \\
  \bottomrule
  \end{tabu}%
\end{table}

\subsection{Mezcal Head}

Lorem ipsum dolor sit amet (see \autoref{fig:sample}), consetetur sadipscing elitr, sed diam
nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
takimata sanctus est Lorem ipsum dolor sit amet.

\subsubsection{Duis Autem}

Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam
nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit
amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor
invidunt ut labore et dolore magna aliquyam erat, sed diam
voluptua. At vero eos et accusam et justo duo dolores et ea
rebum. Stet clita kasd gubergren, no sea takimata sanctus est. Lorem
ipsum dolor sit amet.

\begin{figure}[tb]
 \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
 \includegraphics[width=\columnwidth]{paper-count-w-2015-new}
 \caption{A visualization of the 1990--2015 data from \autoref{tab:vis_papers}. The image is from \cite{Isenberg:2017:VMC} and is in the public domain.}
 \label{fig:sample}
\end{figure}

\subsubsection{Ejector Seat Reservation}

Duis autem~\cite{Lorensen-1987-MCA}\footnote{The algorithm behind
Marching Cubes \cite{Lorensen-1987-MCA} had already been
described by Wyvill et al. \cite{Wyvill:1986:DSS} a year
earlier.} vel eum iriure dolor in hendrerit
in vulputate velit esse molestie consequat,\footnote{Footnotes
appear at the bottom of the column.} vel illum dolore eu
feugiat nulla facilisis at vero eros et accumsan et iusto odio
dignissim qui blandit praesent luptatum zzril delenit augue duis
dolore te feugait nulla facilisi. Lorem ipsum dolor sit amet,
consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt
ut laoreet dolore magna aliquam erat volutpat.

\paragraph{Confirmed Ejector Seat Reservation}

Ut wisi enim ad minim veniam, quis nostrud exerci tation ullamcorper
suscipit lobortis nisl ut aliquip ex ea commodo
consequat~\cite{Nielson:1991:TAD}. Duis autem vel eum iriure dolor in
hendrerit in vulputate velit esse molestie consequat, vel illum dolore
eu feugiat nulla facilisis at vero eros et accumsan et iusto odio
dignissim qui blandit praesent luptatum zzril delenit augue duis
dolore te feugait nulla facilisi.

\paragraph{Rejected Ejector Seat Reservation}

Ut wisi enim ad minim veniam, quis nostrud exerci tation ullamcorper
suscipit lobortis nisl ut aliquip ex ea commodo consequat. Duis autem
vel eum iriure dolor in hendrerit in vulputate velit esse molestie

\section{Conclusion}

Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam
nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit
amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor
invidunt ut labore et dolore magna aliquyam erat, sed diam
voluptua. At vero eos et accusam et justo duo dolores et ea
rebum.

%% if specified like this the section will be committed in review mode
\acknowledgments{
The authors wish to thank A, B, and C. This work was supported in part by
a grant from XYZ.}

%\bibliographystyle{abbrv}
\bibliographystyle{abbrv-doi}
%\bibliographystyle{abbrv-doi-narrow}
%\bibliographystyle{abbrv-doi-hyperref}
%\bibliographystyle{abbrv-doi-hyperref-narrow}

\bibliography{template}
\end{document}

### 2022-06-24 class-info/template.qmd
Template files and a makefile

---
title: Short Paper
author-list:
  - name: Alice Anonymous
    affiliation: Some Institute of Technology
    email: alice@example.com
  - name: Bob Security
    affiliation: Another University
    email: bob@example.com

abstract: |
  Duis autem vel eum iriure dolor in hendrerit in vulputate
  velit esse molestie consequat, vel illum dolore eu feugiat nulla
  facilisis at vero eros et accumsan et iusto odio dignissim qui blandit
  praesent luptatum zzril delenit augue duis dolore te feugait nulla
  facilisi. Lorem ipsum dolor sit amet, consectetuer adipiscing elit,
  sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna
  aliquam erat volutpat. Ut wisi enim ad minim veniam, quis nostrud exerci tation ullamcorper
  suscipit lobortis nisl ut aliquip ex ea commodo consequat. Duis autem
  vel eum iriure dolor in hendrerit in vulputate velit esse molestie
  consequat, vel illum dolore eu feugiat nulla facilisis at vero eros et
  accumsan et iusto odio dignissim qui blandit praesent luptatum zzril
  delenit augue duis dolore te feugait nulla facilisi.

category: Research
bib: template.bib
bibstyle: abbrv-doi.bst
csl: ieee-transactions-on-visualization-and-computer-graphics.csl
ieee-keywords: 
  - \CCScatTwelve{Human-centered computing}{Visu\-al\-iza\-tion}{Visu\-al\-iza\-tion techniques}{Treemaps}
  - \CCScatTwelve{Human-centered computing}{Visu\-al\-iza\-tion}{Visualization design and evaluation methods}{}
firstsection: Introduction

link-citations: false

format: 
  latex:
    template-partials: [partials/doc-class.tex, partials/title.tex, partials/before-body.tex, partials/after-body.tex]

pdf-engine: pdflatex

date: last-modified
---

This template is for papers of VGTC-sponsored conferences which are \emph{\textbf{not}} published in a special issue of TVCG.

# Using the Style Template

- If you receive compilation errors along the lines of `\texttt{Package ifpdf Error: Name clash, \textbackslash ifpdf is already defined}` then please add a new line `\texttt{\textbackslash let\textbackslash ifpdf\textbackslash relax}` right after the `\texttt{\textbackslash documentclass[journal]\{vgtc\}}` call. Note that your error is due to packages you use that define `\texttt{\textbackslash ifpdf}` which is obsolete (the result is that `\texttt{\textbackslash ifpdf}` is defined twice); these packages should be changed to use ifpdf package instead.
- The style uses the hyperref package, thus turns references into internal links. We thus recommend to make use of the `\texttt{\textbackslash autoref\{reference\}}` call (instead of `\texttt{Figure\~{}\textbackslash ref\{reference\}}` or similar) since `\texttt{\textbackslash autoref\{reference\}}` turns the entire reference into an internal link, not just the number. Example: @fig-sample.
- The style automatically looks for image files with the correct extension (eps for regular \LaTeX; pdf, png, and jpg for pdf\LaTeX), in a set of given subfolders (figures/, pictures/, images/). It is thus sufficient to use `\texttt{\textbackslash includegraphics\{CypressView\}}` (instead of `\texttt{\textbackslash includegraphics\{pictures/CypressView.jpg\}}`).
- For adding hyperlinks and DOIs to the list of references, you can use `\texttt{\textbackslash bibliographystyle\{abbrv-doi-hyperref-narrow\}}` (instead of `\texttt{\textbackslash bibliographystyle\{abbrv\}}`). It uses the doi and url fields in a bib\TeX\ entry and turns the entire reference into a link, giving priority to the doi. The doi can be entered with or without the `\texttt{http://dx.doi.org/}` url part. See the examples in the bib\TeX\ file and the bibliography at the end of this template.\\[1em]
\textbf{Note 1:} occasionally (for some \LaTeX\ distributions) this hyper-linked bib\TeX\ style may lead to \textbf{compilation errors} (`\texttt{pdfendlink ended up in different nesting level ...}`) if a reference entry is broken across two pages (due to a bug in hyperref). In this case make sure you have the latest version of the hyperref package (i.\,e., update your \LaTeX\ installation/packages) or, alternatively, revert back to `\texttt{\textbackslash bibliographystyle\{abbrv-doi-narrow\}}` (at the expense of removing hyperlinks from the bibliography) and try `\texttt{\textbackslash bibliographystyle\{abbrv-doi-hyperref-narrow\}}` again after some more editing.\\[1em]
\textbf{Note 2:} the `\texttt{-narrow}` versions of the bibliography style use the font `PTSansNarrow-TLF` for typesetting the DOIs in a compact way. This font needs to be available on your \LaTeX\ system. It is part of the [`paratype` package](https://www.ctan.org/pkg/paratype), and many distributions (such as MikTeX) have it automatically installed. If you do not have this package yet and want to use a `\texttt{-narrow}` bibliography style then use your \LaTeX\ system's package installer to add it. If this is not possible you can also revert to the respective bibliography styles without the `\texttt{-narrow}` in the file name.\\[1em]
DVI-based processes to compile the template apparently cannot handle the different font so, by default, the template file uses the \texttt{abbrv-doi} bibliography style but the compiled PDF shows you the effect of the \texttt{abbrv-doi-hyperref-narrow} style.

# Bibliography Instructions

- Sort all bibliographic entries alphabetically but the last name of the first author. This \LaTeX/bib\TeX\ template takes care of this sorting automatically.
- Merge multiple references into one; e.\,g., use \cite{Max-1995-OMF,Kitware-2003} (not \cite{Kitware-2003}\cite{Max-1995-OMF}). Within each set of multiple references, the references should be sorted in ascending order. This \LaTeX/bib\TeX\ template takes care of both the merging and the sorting automatically.
- Verify all data obtained from digital libraries, even ACM's DL and IEEE Xplore  etc.\ are sometimes wrong or incomplete.
- Do not trust bibliographic data from other services such as Mendeley.com, Google Scholar, or similar; these are even more likely to be incorrect or incomplete.
- Articles in journal---items to include:
  - author names
	- title
	- journal name
	- year
	- volume
	- number
	- month of publication as variable name (i.\,e., \{jan\} for January, etc.; month ranges using \{jan \#\{/\}\# feb\} or \{jan \#\{-{}-\}\# feb\})

- use journal names in proper style: correct: "IEEE Transactions on Visualization and Computer Graphics", incorrect: "Visualization and Computer Graphics, IEEE Transactions on"
- Papers in proceedings---items to include:
  - author names
	- title
	- abbreviated proceedings name: e.\,g., "Proc.\textbackslash{} CONF\_ACRONYNM" without the year; example: "Proc.\textbackslash{} CHI", "Proc.\textbackslash{} 3DUI", "Proc.\textbackslash{} Eurographics", "Proc.\textbackslash{} EuroVis"
	- year
	- publisher
	- town with country of publisher (the town can be abbreviated for well-known towns such as New York or Berlin)

- article/paper title convention: refrain from using curly brackets, except for acronyms/proper names/words following dashes/question marks etc.; example:
	- paper "Marching Cubes: A High Resolution 3D Surface Construction Algorithm"
	- should be entered as "\{M\}arching \{C\}ubes: A High Resolution \{3D\} Surface Construction Algorithm" or  "\{M\}arching \{C\}ubes: A high resolution \{3D\} surface construction algorithm"
	- will be typeset as "Marching Cubes: A high resolution 3D surface construction algorithm"

- for all entries
	- DOI can be entered in the DOI field as plain DOI number or as DOI url; alternative: a url in the URL field
	- provide full page ranges AA-{}-BB

- when citing references, do not use the reference as a sentence object; e.\,g., wrong: "In \cite{Lorensen-1987-MCA} the authors describe \dots", correct: "Lorensen and Cline \cite{Lorensen-1987-MCA} describe \dots"

# Example Section

Lorem\marginpar{\small You can use the margins for comments while editing the submission, but please remove the marginpar comments for submission.} ipsum dolor sit amet, consetetur sadipscing elitr, sed diam
nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
dolore magna aliquyam erat, sed diam
voluptua [@Kitware-2003;Max-1995-OMF]. At vero eos et accusam et
justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit
amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor
invidunt ut labore et dolore magna aliquyam erat, sed diam
voluptua. At vero eos et accusam et justo duo dolores et ea
rebum. Stet clita kasd gubergren, no sea takimata sanctus est.

# Exposition

Duis autem vel eum iriure dolor in hendrerit in vulputate velit esse
molestie consequat, vel illum dolore eu feugiat nulla facilisis at
vero eros et accumsan et iusto odio dignissim qui blandit praesent
luptatum zzril delenit augue duis dolore te feugait nulla
facilisi. Lorem ipsum dolor sit amet, consectetuer adipiscing elit,
sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna
aliquam erat volutpat~\cite{Kindlmann-1999-SAG}.

\begin{equation}
\sum_{j=1}^{z} j = \frac{z(z+1)}{2}
\end{equation}

Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam
nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
takimata sanctus est Lorem ipsum dolor sit amet.

## Lorem ipsum

Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam
nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit
amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor
invidunt ut labore et dolore magna aliquyam erat, sed diam
voluptua. At vero eos et accusam et justo duo dolores et ea
rebum.

## Mezcal Head

Lorem ipsum dolor sit amet (see @fig-sample), consetetur sadipscing elitr, sed diam
nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
takimata sanctus est Lorem ipsum dolor sit amet.

### Duis Autem

Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam
nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit
amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor
invidunt ut labore et dolore magna aliquyam erat, sed diam
voluptua. At vero eos et accusam et justo duo dolores et ea
rebum. Stet clita kasd gubergren, no sea takimata sanctus est. Lorem
ipsum dolor sit amet.

![A visualization of the 1990--2015 data. The image is from @Isenberg-2017-VMC and is in the public domain.](paper-count-w-2015-new){#fig-sample}

### Ejector Seat Reservation

Duis autem \cite{Lorensen-1987-MCA}^[The algorithm behind
Marching Cubes \cite{Lorensen-1987-MCA} had already been
described by Wyvill et al. \cite{Wyvill-1986-DSS} a year
earlier.] vel eum iriure dolor in hendrerit
in vulputate velit esse molestie consequat,^[Footnotes
appear at the bottom of the column.] vel illum dolore eu
feugiat nulla facilisis at vero eros et accumsan et iusto odio
dignissim qui blandit praesent luptatum zzril delenit augue duis
dolore te feugait nulla facilisi. Lorem ipsum dolor sit amet,
consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt
ut laoreet dolore magna aliquam erat volutpat.

#### Confirmed Ejector Seat Reservation

Ut wisi enim ad minim veniam, quis nostrud exerci tation ullamcorper
suscipit lobortis nisl ut aliquip ex ea commodo
consequat [@Nielson-1991-TAD]. Duis autem vel eum iriure dolor in
hendrerit in vulputate velit esse molestie consequat, vel illum dolore
eu feugiat nulla facilisis at vero eros et accumsan et iusto odio
dignissim qui blandit praesent luptatum zzril delenit augue duis
dolore te feugait nulla facilisi.

#### Rejected Ejector Seat Reservation

Ut wisi enim ad minim veniam, quis nostrud exerci tation ullamcorper
suscipit lobortis nisl ut aliquip ex ea commodo consequat. Duis autem
vel eum iriure dolor in hendrerit in vulputate velit esse molestie

# Conclusion

Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam
nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit
amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor
invidunt ut labore et dolore magna aliquyam erat, sed diam
voluptua. At vero eos et accusam et justo duo dolores et ea
rebum.

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this-

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

### 2022-06-24 index.qmd
Template files and a makefile

---
title: Testing Statistical Graphics
author-list:
  - name: Emily Robinson
    dept: Statistics Department
    affiliation: CalPoly
    email: earobinson@huskers.unl.edu
  - name: Heike Hofmann
    dept: Statistics Department
    affiliation: Iowa State University
    email: hofmann@iastate.edu
  - name: Susan Vanderplas
    dept: Statistics Department
    affiliation: University of Nebraska, Lincoln
    email: svanderplas2@unl.edu

abstract: |
  We outline our process and common pitfalls for developing and testing human perception of statistical graphics.

category: Research
bib: refs.bib
bibstyle: class-info/abbrv-doi.bst
# csl: ieee-transactions-on-visualization-and-computer-graphics.csl
ieee-keywords: 
  - \CCScatTwelve{Human-centered computing}{Visu\-al\-iza\-tion}{Visualization design and evaluation methods}{}
  - \CCScatTwelve{Human-centered computing}{Visu\-al\-iza\-tion}{Empirical studies in visualization}{}
  - \CCScatTwelve{Human-centered computing}{Visu\-al\-iza\-tion}{Visualization application domains}{Scientific visualization}
  - \CCScatTwelve{Applied computing}{Physical sciences and engineering}{Mathematics and statistics}{}
firstsection: Introduction

link-citations: false

format: 
  pdf:
    keep-tex: true
    template-partials: [partials/doc-class.tex, partials/title.tex, partials/before-body.tex, partials/after-body.tex]

pdf-engine: pdflatex

date: last-modified
---

Goal: Integrate information from the design of \cite{clusters,sineillusion,emily-diss,power}.

### 2022-06-24 partials/title.tex
Template files and a makefile

$if(title)$
  \title{$title$}
$endif$
$if(subtitle)$
  \subtitle{$subtitle$}
$endif$

\author{%
$for(author-list)$%
  $it.name$\thanks{email: $it.email$}\\\scriptsize $it.dept$\\\scriptsize $it.affiliation$%
  $sep$\and%
$endfor$%
}

$if(date)$
  \date{$date$}
$endif$

$if(teaserimg)$
\teaser{
  \centering
  \includegraphics[width=\linewidth]{$teaser-img$}
  \caption{$teaser-caption$}
  \label{fig:teaser}
}
$endif$

$if(abstract)$
\abstract{$abstract$}
$endif$

\vgtccategory{$category$}

$if(ieee-keywords)$
\CCScatlist{
$for(ieee-keywords)$
$it$
$sep$;
$endfor$
}
$endif$

### 2022-06-24 index.qmd
Add outline

@@
   We outline our process and common pitfalls for developing and testing human perception of statistical graphics. 
+  Goal: Integrate information from the design of \cite{clusters,sineillusion,emily-diss,power}.
 
@@
 
-Goal: Integrate information from the design of \cite{clusters,sineillusion,emily-diss,power}.
+Importance of testing statistical graphics
+
+# Testing Methods and Engagement
+
+Lineups
+You Draw It
+Direct Assessment
+
+# Developing a Model
+This is the single longest part of the entire testing process
+
+## The Role of Simulation
+
+## Selecting Parameters
+
+# Pilot Testing and Quality Assurance
+
+Test early, test often, and make sure you look at all of your plots!!!
+If you're generating new data for each participant, be sure to save that data to a database so that you can at least see where/how things went wrong!
+
+# Gathering and Analyzing Additional Information
+
+Calculator observations
+Scratchpad
+Explanation
+Confidence Level
+User Interactions
+Browser fingerprints
+
+# Building the Application
+Shiny is awesome
+
+# Analyzing the Data
+Be prepared for surprises

### 2022-06-25 index.qmd
Adding references and a few bits of intro

@@
 
-Importance of testing statistical graphics
+<!-- Importance of testing statistical graphics -->
+
+Charts and data visualizations are primary methods of communicating scientific information to the public in what we typically hope is an easy-to-digest, visually attractive package.
+There are many strategies for creating charts and graphs, from Tufte-esque minimalism\cite{tufte} to charts designed with extra imagery and aesthetic appeal that draw the viewer's attention and persist in memory\cite[Ch 3]{cairoFunctionalArtIntroduction2012}.
+For a specified type of data, there are also usually many different chart forms to display that data: for instance, if we have a set of categorical data and we wish to show the relative proportions of each category, we could do so using a stacked bar chart or the polar equivalent, a pie chart.
+There have been several attempts to list out all of the types of charts\cite{ribeccaSearchChartsData}, create a taxonomy of charts\cite{bertin1983semiology}, or even to classify charts using a domain-specific grammar; the grammar of graphics\cite{wilkinsonGrammarGraphics1999}.
+With all of the different design choices available, how are chart creators to know what is the best approach for communicating data to the appropriate audience?
 
@@
 
-# Pilot Testing and Quality Assurance
-
-Test early, test often, and make sure you look at all of your plots!!!
-If you're generating new data for each participant, be sure to save that data to a database so that you can at least see where/how things went wrong!
-
 # Gathering and Analyzing Additional Information
@@
 
+# Pilot Testing and Quality Assurance
+
+Test early, test often, and make sure you look at all of your plots!!!
+If you're generating new data for each participant, be sure to save that data to a database so that you can at least see where/how things went wrong!
+
+# Monitoring the Data
+
+Use of pilot data to set up analysis documents, github actions and/or shiny dashboard to monitor incoming results.
+
+Not necessarily important for small tests of limited duration, because data from e.g. prolific comes in so quickly. Server load is actually a much greater concern.
+
 # Building the Application

### 2022-06-25 index.qmd
Importance of testing statistical graphics

@@
 
+There are heuristics, rules-of-thumb, and various guidelines\cite{asme-standards-graphics,clevelandShapeParameterTwoVariable1988,allenVisualizingScientificData2016} for creating useful and visually attractive charts, but as scientists, we believe empirical evidence is essential for making correct decisions.
+The best way to establish the efficacy of various design decisions, then, is to test our charts by having them evaluated by people under controlled conditions and gathering appropriate data\cite{clevelandGraphicalPerceptionGraphical1985}.
+It can be extremely challenging to design studies which strike the right combination of internal and external validity. 
+Simply asking people to read quantities off of a graph may not generalize beyond the questions asked or the data used in the chart\cite{eellsRelativeMeritsCircles1926,croxtonBarChartsCircle1927,vonhuhnFurtherStudiesGraphic1927,croxtonGraphicComparisonsBars1932}, but designing a study that is sufficiently robust to those issues requires manipulation of so many factors as to be functionally impossible to complete.
+In this paper, we attempt to distill lessons learned from several different sets of experimental graphics experiments, discussing useful testing methods for statistical graphics, the process of designing a graphical experiment, and analysis of the resulting empirical data.
+It is our hope that by writing this paper, we will save other researchers some hard lessons in what to do (and what not to do), and thereby lower the barriers that exist for conducting research involving empirical tests of statistical graphics.
+
 # Testing Methods and Engagement
@@
 Be prepared for surprises
+
+\clearpage

### 2022-06-25 index.qmd
Screenshots and text+refs

--- 2022-06-25 index.qmd
@@
 For a specified type of data, there are also usually many different chart forms to display that data: for instance, if we have a set of categorical data and we wish to show the relative proportions of each category, we could do so using a stacked bar chart or the polar equivalent, a pie chart.
-There have been several attempts to list out all of the types of charts\cite{ribeccaSearchChartsData}, create a taxonomy of charts\cite{bertin1983semiology}, or even to classify charts using a domain-specific grammar; the grammar of graphics\cite{wilkinsonGrammarGraphics1999}.
+There have been several attempts to list out all of the types of charts\cite{ribeccaSearchChartsData}, create a taxonomy of charts\cite{bertin1983semiology,desnoyersTaxonomyVisualsScience2011}, and even to classify charts using a domain-specific grammar of graphics\cite{wilkinsonGrammarGraphics1999}.
 With all of the different design choices available, how are chart creators to know what is the best approach for communicating data to the appropriate audience?
 
+There are heuristics, rules-of-thumb, and various guidelines \cite{asme-standards-graphics,allenVisualizingScientificData2016,haemerDoubleScalesAre1948,fewInformationDashboardDesign2006,kosslynGraphDesignEye2006} for creating useful and visually attractive charts, but as scientists, we believe empirical evidence is essential for making correct decisions.
+The best way to establish the efficacy of various design decisions, then, is to test our charts by having them evaluated by people under controlled conditions and gathering appropriate data\cite{clevelandShapeParameterTwoVariable1988,clevelandGraphicalPerceptionGraphical1985}.
+It can be extremely challenging to design studies which strike the right combination of internal and external validity. 
+Simply asking people to read quantities off of a graph may not generalize beyond the questions asked or the data used in the chart\cite{eellsRelativeMeritsCircles1926,croxtonBarChartsCircle1927,vonhuhnFurtherStudiesGraphic1927,croxtonGraphicComparisonsBars1932}, but designing a study that is sufficiently robust to those issues requires manipulation of so many factors as to be functionally impossible to complete.
+In this paper, we attempt to distill lessons learned from several different sets of experimental graphics experiments\cite{clusters,sineillusion,emily-diss,power}, discussing the use of several different testing methods for statistical graphics\cite{vanderplasTestingStatisticalCharts2020}, the process of designing a graphical experiment, and analysis of the resulting empirical data.
+It is our hope that by writing this paper, we will prevent other researchers from repeating hard lessons in what to do (and what not to do), lowering the barriers that exist for conducting research involving empirical tests of statistical graphics.
+
 # Testing Methods and Engagement
+<!-- Lineups, You Draw It, Direct Assessment -->
 
-Lineups
-You Draw It
-Direct Assessment
+We will start with an overview of different testing methods that we have used to empirically assess graphics. 
+While it is certainly possible to use equipment such as eye-tracking devices, and we have conducted such studies, here we will primarily focus on studies which can be conducted over the internet, because these experiments have the lowest overhead, offer relatively fast data collection, and provide useful results that can be very clear if the experiment is designed correctly.
+Our toolkit is based around R\cite{R}, and includes ggplot2\cite{ggplot2} and Shiny\cite{shiny} as primary components, as well as  JavaScript (including D3\cite{d3}) to customize the Shiny interface for many of the experiments.
+
+One primary concern when testing graphics is the level of user engagement required. 
+For instance, if we are testing whether someone can detect an effect such as a trend line in noisy data, this is a *perceptual* question best addressed using methods which allow the user to interact with the data on a basic visual level.
+Sometimes, it may be possible to set up a scenario where the user adjusts the plot directly, as in our experiment testing the strength of the sine illusion experiment\cite{sineillusion}: users adjusted the strength of the illusion correction until the lines appeared to be the same length, providing a direct measure of the magnitude of the sine illusion's effect, as in @fig-sine-illusion.
+
+![Direct adjustment of a plot in a perceptual task](sine_illusion_screenshot.png){#fig-sine-illusion}
+
+Statistical lineups\cite{bujaStatisticalInferenceExploratory2009,wickhamGraphicalInferenceInfovis2010} are a more generally applicable testing tool for perceptual questions such as "which chart displays this data more clearly"\cite{power}.
+As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot instead of any of the 19 null plots, then the target plot is said to be visually significant\cite{majumderValidationVisualStatistical2013,vanderplasStatisticalSignificanceCalculations2021} and a "see" value (the visual analogue of a p-value for a statistical test) can be calculated\cite{niladriroychowdhurySeeValueApp2020}.
+In another variation of statistical lineup procedure, two models are compared, with target plots from each model embedded in the array of 20 total plots; the remaining null plots are constructed from a mixture model that blends the two competing models.
+This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data\cite{clusters}.
+Statistical lineups do not require the experimenter to provide any contextual information for the data: all of the necessary information is embedded in the choice of a null statistical model, which is convenient for testing purposes, but does not require engagement at a level that would allow for testing the viewer's understanding of the information shown in the chart.
+An example of a statistical lineup for testing the perception of exponential growth is shown in @fig-lineup-log.
+
+![A lineup experiment testing whether users can perceive differences in rates of exponential growth on linear and log scales.
+In addition to asking users to select the plot which is the most different, we also asked for basic reasoning and the user's level of confidence in their answer.](Lineup_log_screenshot.png){#fig-lineup-log}
+
+To assess the viewer's *understanding* of information shown in a chart, we must ask questions and allow the user to provide feedback.
+User feedback may be collected on a numerical scale or through the use of written comments, recorded "think-aloud" processes, and other more qualitative interaction methods.
+We have conducted studies asking users to provide us with a written description of what the chart shows within a wider story-driven scenario (@fig-estimation-describe), and we have also asked for written feedback about why a certain panel of a statistical lineup was chosen; the insights from participant responses to these questions have been rich and insightful (and will be discussed more in the analysis section).
+While we have not (to date) recorded users talking out loud about what they are seeing, it is entirely feasible that so-called "think aloud" processes could be used within a Shiny application, with audio recordings saved to the server for transcription and analysis\cite{dunbar1995scientists,traftonTurningPicturesNumbers2000,kirschenbaum2003comparative}.
+
+![This question asks users to write out a description of how the population of Ewoks changes over time, without any further cues, to determine whether participants default to multiplicative or additive descriptions.](Estimation_describe_plot_crop.png){#fig-estimation-describe}
+
+Of course, as we have confined ourselves to the discussion of online, asynchronous testing methods, every user interaction with the testing materials (typically hosted on a web page) can also be recorded along with time stamps, mouse positions, and so on.
+While we have not used this information heavily in our experimental analyses thus far, we do collect some time stamp data in order to assess how long participants spend on each question; it could almost certainly be more informative in the right setting.
+In addition, we have also provided participants with supportive tools, such as "scratch pads" and limited function calculators built into the Shiny application that allow us to gain additional insight into the estimation and visual comparison process when more complex calculations are required to answer the numerical estimation question (@fig-estimation-calc).
+These supportive tools have provided a wealth of information that, while not easy to analyze and generally messy, provides real insight into how participants were looking at the data, what strategies they tried and discarded for reading the chart, and what visual estimation methods were used.
+
+![This question asks participants for a numerical estimate, but provides a basic calculator and scratchpad.
+All user interactions with the calculator and scratchpad are logged, providing insight into the user's thought process.](Estimation_numerical_screenshot_crop.png){#fig-estimation-calc}
+
+In our experience, one of the most difficult components of designing an experiment which asks users to directly estimate information from a chart is that the wording of such questions must be extremely carefully determined, and even when great care is taken with the construction of the question, participant answer accuracy is fundamentally limited by the fact that many participants do not read and interpret the question with the care and precision that it was written.
+Questions that ask participants to e.g.
+estimate the multiplicative difference in a quantity at two timepoints may be misunderstood as asking for an estimate of the additive difference, and the resulting estimates are then one or more orders of magnitude off of the correct answer.
+This is one area where lineup tests are definitely superior - they do not depend on participants to understand the nuances of language or scenarios built around the chart under investigation.
+
+In some cases, we want participants to *engage directly* with an interactive visualization. 
+We may want to record the way in which they interact with the chart to see what types of information they seek out, but we may also want to ask them to interact with the chart in a specific way and record the result.
+In one recent experiment, we asked participants to forecast an exponential trend, with data presented on either a linear or log scale. 
+Using some JavaScript modified from New York Times interactive graphics "You Draw It" features\cite{katzYouDrawIt2017}, we had users draw trend lines and make forecasts directly on interactive charts, with the predictions and trend lines recorded to our database for later analysis^[See a gif of testing with You Draw It [here](https://i.imgur.com/GM5YSen.gif)].
+With interactive graphics rendered using JavaScript (or other web libraries), the only limit to the types of questions you can ask in testing graphics is in your ability to write code to interact with the visualization library.
+This type of testing method can be extremely natural for participants, but it also is hard to generalize when discussing testing methods because of the potential range of applications where it might be employed.
+
+Whichever testing method is chosen should be appropriate to the type of question under investigation.
+Lineups require a great amount of work in developing a null sampling model which replicates the important visual features of the data, but model specification is extremely important when asking users questions about any simulated data set or even when examining user interactions directly.
+It  is important to carefully calibrate the model parameters to allow for some variability, but not too much.
+This Goldilocks-style problem is the focus of the next section.
 

### 2022-06-25 index.qmd
A few more references and fleshing out the outline for the next few sections

@@
 # Developing a Model
-This is the single longest part of the entire testing process
+
+This is the single longest part of the entire testing process, in part because choosing a model that replicates important visual features of the data is an extremely complex process\cite{hullmanDesigningInteractiveExploratory2021,cookFoundationAvailableThinking2021,vanderplasDesigningGraphicsRequires2021}.
+
+- Importance of starting somewhere
 
@@
 
+Visual assessment is key, but narrowing the parameter space down to something that can be statistically determined is useful for providing a sort of magnitude of difficulty assessment numerically.
+
+Feature engineering is a key part of why this works - understanding the visual calculations that are being performed (via introspection) can help with identifying a useful summary statistic.
+Gini inequality for clustering, R^2 for linear regression, lack-of-fit statistic for difference between exponential curves.
+
 ## Selecting Parameters
 
-# Gathering and Analyzing Additional Information
+One experimenter develops the model and selects the broad range of parameters, the other calibrates those parameters.
+
+## Visual Assessment is Critical
+
+Always look at an exhaustive amount of your models... several times, with fresh eyes. Do think-aloud pilot testing.
+
+Explain the mishap with feature hierarchy - too many plots (because giant factorial experiment) meant that we missed cluster size imbalance.
+When we corrected that imbalance, though, we ended up with clusters that were oddly shaped.
+Sometimes, there's just no way to win, and we might have needed to rethink our sampling model.
+
+# Experimental Design Considerations
+
+- Attention span
+- Sensory overload - too many lineups in a row is hard
+- Comparing across participants - importance of random effects for participant skill level
+- Consider asking both demographic questions and questions about mathematical skill level.
+We would have been wise to ask about country of primary/secondary education in our log study, as logarithms are introduced at different levels in different countries.
+But we didn't think of that when recruiting participants, so it would have been hard to manipulate that factor.
+There are always missed opportunities.
+
+- We've had lots of success with complete and incomplete block designs
+
+- Importance of randomizing trial order to combat order effects
+
+# Pilot Testing and Quality Assurance
+
+Test early, test often, and make sure you look at all of your plots!!!
+If you're generating new data for each participant, be sure to save that data to a database so that you can at least see where/how things went wrong!
+
+# Data Collection and Monitoring
+
+Prolific, Amazon Turk, testing with interested subgroups.
+
+Use of pilot data to set up analysis documents, github actions and/or shiny dashboard to monitor incoming results.
+
+Not necessarily important for small tests of limited duration, because data from e.g. prolific comes in so quickly. Server load is actually a much greater concern.
+
+# Analyzing the Data
+Be prepared for surprises
+
+Use of linear models to test lineup data - see-value approach is great for single plots, but when we're interested in more concrete effects over many plots, it can be better to aggregate effects and just think of a multinomial model.
+
+Direct estimation can be tricky to analyze because of how messy the data is.
+Additional supplemental information can be critical to provide context, but graphical analysis of participant responses was more helpful than modeling for our experiment.
+
+Useful graphical analysis methods: density plots + rug plots (so that you can see rounding effects and common errors), wordclouds, estimation sketches + spaghetti plots
+
+## Analyzing Supplemental Information
 
@@
 User Interactions
-Browser fingerprints
 
-# Pilot Testing and Quality Assurance
-
-Test early, test often, and make sure you look at all of your plots!!!
-If you're generating new data for each participant, be sure to save that data to a database so that you can at least see where/how things went wrong!
-
-# Monitoring the Data
-
-Use of pilot data to set up analysis documents, github actions and/or shiny dashboard to monitor incoming results.
-
-Not necessarily important for small tests of limited duration, because data from e.g. prolific comes in so quickly. Server load is actually a much greater concern.
-
-# Building the Application
-Shiny is awesome
-
-# Analyzing the Data
-Be prepared for surprises
+# Conclusion
 

### 2022-06-26 index.qmd
Information on parameter selection

@@
 We will start with an overview of different testing methods that we have used to empirically assess graphics. 
-While it is certainly possible to use equipment such as eye-tracking devices, and we have conducted such studies, here we will primarily focus on studies which can be conducted over the internet, because these experiments have the lowest overhead, offer relatively fast data collection, and provide useful results that can be very clear if the experiment is designed correctly.
-Our toolkit is based around R\cite{R}, and includes ggplot2\cite{ggplot2} and Shiny\cite{shiny} as primary components, as well as  JavaScript (including D3\cite{d3}) to customize the Shiny interface for many of the experiments.
+While it is certainly possible to use equipment such as eye-tracking devices, here we will primarily focus on studies which can be conducted over the internet, because these experiments have the lowest overhead, offer relatively fast data collection, and provide useful results that can be very clear if the experiment is designed correctly.
+Our toolkit is R-based\cite{R}, and includes ggplot2\cite{ggplot2} and Shiny\cite{shiny} as primary components, as well as  JavaScript (including D3\cite{d3}) to customize the Shiny interface for many of the experiments.
 
@@
 
-This is the single longest part of the entire testing process, in part because choosing a model that replicates important visual features of the data is an extremely complex process\cite{hullmanDesigningInteractiveExploratory2021,cookFoundationAvailableThinking2021,vanderplasDesigningGraphicsRequires2021}.
+Once you have decided what graphical task you want to explore or test, it is necessary to develop a model which can be used to test the graphical features you want to explore in a precise manner. 
+This is the single longest part of the entire experimental design and execution process, in part because choosing a model that replicates important visual features of the data is an extremely complex process \cite{hullmanDesigningInteractiveExploratory2021,cookFoundationAvailableThinking2021,vanderplasDesigningGraphicsRequires2021}.
 
-- Importance of starting somewhere
+There are two main options when developing a statistical model for graphical testing: you can either start with a large dataset and sample from that dataset\cite{power}, or you can start from a model and sample data from that model generating process\cite{sineillusion,clusters,emily-diss}.
+This decision is largely determined by the availability of a large dataset containing the requisite features of interest and the qualities being manipulated in the experiment.
+For instance, Hofmann et al.
+\cite{power} used samples of different sizes from a pre-existing dataset to manipulate the amount of signal in each comparison; with a small sample, there is less signal and the same amount of noise, making the true plot harder to spot.
+In many situations, though, a convenient dataset with the right properties is harder to acquire, and it becomes necessary to develop a sampling model to generate data for user evaluation.
 
-## The Role of Simulation
+The tools we discuss in the remainder of this section can be applied both to pre-existing data sets and to model-based sampling methods. 
+We will do our best to keep this discussion general, but we will provide details in some situations to ensure that our general descriptions are not hopelessly vague.
 
-Visual assessment is key, but narrowing the parameter space down to something that can be statistically determined is useful for providing a sort of magnitude of difficulty assessment numerically.
+## Screening Parameters with Simulation
 
-Feature engineering is a key part of why this works - understanding the visual calculations that are being performed (via introspection) can help with identifying a useful summary statistic.
-Gini inequality for clustering, R^2 for linear regression, lack-of-fit statistic for difference between exponential curves.
+Visual assessment is key, but narrowing the parameter space down to something that can be statistically determined is useful for providing a sort of numerical magnitude of difficulty assessment. 
+If using model based sampling methods, it is important to iteratively assess the sampling procedure through simulation: Start with a sampling method and define a parameter space that is appropriate for the model, and then simulate many data sets from each possible parameter combinations using a grid search.
+When sampling from a large data set, it is still important to assess the sample size and any potential stratification methods employed in order to ensure that the visual effect is represented.
 
-## Selecting Parameters
+In both cases, it can help to employ some general feature engineering - what numerical statistic provides an estimation of the basic concept which is being visually evaluated?
+For instance, we have used
 
-One experimenter develops the model and selects the broad range of parameters, the other calibrates those parameters.
+- $R^2$ as a measure of the strength of a linear relationship
+- Gini inequality as a measure of the strength of clustering
+- lack-of-fit statistics to assess the amount of curvature in an exponential relationship (shown in @fig-lof-density-curves)
+
+Then, a wide range of potential combinations of parameter values or sampling strategies can be explored and summarized graphically; if the numerical statistic cannot differentiate between the null and target under a condition, it is reasonable to expect that a visual inspection of the data may also not show significant results.
+As with any measure, it is important that difficulty levels span a range from easy to hard; we do not learn anything from finding out that everyone can distinguish all of the combinations.
+
+```{r lof-density-curves, echo = F, include = T, message = F, warning = F}
+#| label: fig-lof-density-curves
+#| fig-cap: "Density plot of the lack of fit statistic showing separation of selected difficulty levels: High (obvious curvature), Medium (noticable curvature), and Low (almost linear).
+Each density plot is the result of 1000 simulations from a model $y_i = \\alpha\\cdot e^{\\beta\\cdot x_i + \\epsilon_i} + \\theta$, where $\\epsilon \\sim N(0, \\sigma^2)$.
+$\\alpha$ and $\\theta$ were selected after manipulation of $\\beta$ and $\\sigma$ to ensure that all data generated had similar $y$ ranges so as not to provide visual cues about model differences outside of the plot curvature."
+#| out.width: "\\columnwidth"
+#| fig.height: 2.5
+#| fig.width: 5
+
+library(tidyverse)
+lofData <- read.csv(file = "https://raw.githubusercontent.com/earobinson95/EmilyARobinson-UNL-dissertation/main/data/01-lineups/lineup-lof-data.csv")
+lofPlot_curvature <- lofData %>%
+  mutate(Curvature = factor(Curvature, levels = c("Obvious Curvature", "Noticeable Curvature", "Almost Linear"), labels = c("High Curvature", "Medium Curvature", "Low Curvature"))) %>%
+  mutate(Variability = factor(Variability, levels = c("Low"))) %>%
+  ggplot(aes(x = statistic, fill = Curvature, color = Curvature)) +
+  geom_density(alpha = 0.7) +
+  scale_fill_manual("Difficulty", values = c("#004400", "#116611", "#55aa55")) +
+  scale_color_manual("Difficulty", values = c("#004400", "#116611", "#55aa55")) +
+  theme_bw(base_size = 14) +
+  theme(legend.position = "bottom",
+        axis.text    = element_text(size = 10),
+        axis.title   = element_text(size = 10),
+        legend.title = element_text(size = 10),
+        legend.text  = element_text(size = 10),
+        legend.key.size = unit(0.5, "line")
+        ) + 
+  scale_x_continuous("Lack of Fit Statistic") +
+  scale_y_continuous("Density") + 
+  theme(legend.position = c(1, 1), legend.justification = c(1.01, 1.01), legend.box.background = element_rect(fill = "white", color = "black"))
+lofPlot_curvature
+```
+
+## Fine-Tuning Parameter Choices
+
+Once an appropriate set of parameters are identified using the numerical screening method, then it is important to calibrate these parameter selections visually; after all, no numerical statistic is a perfect measure of what we actually see; at best, they are approximations of what we might potentially see.
+We have found it to be useful to have one experimenter calibrate the model parameters at a gross level, and then have another experimenter narrow in on the parameters which are visually reasonable within the selected range.
+Then, both examiners visually inspect a large number of plots generated using those parameters to get a sense for how difficult the task at hand is (this strategy is also described by Lu et al.
+\cite{luModelingJustNoticeable2022}).
+At some point, all experimenters become so visually saturated with the nuances of the data generating mechanism that it may become necessary to "sanity check" the protocol with family members, friends, and colleagues.
+These informal surveys provide extremely useful feedback and can help to counteract the visual saturation of being immersed in the design of a visualization experiment for months at a time.
 
@@
 
-Always look at an exhaustive amount of your models... several times, with fresh eyes. Do think-aloud pilot testing.
+We cannot overstate the importance of visual assessment of your model stimuli, preferrably with fresh eyes.
+We highly recommend performing a series of think-aloud pilot testing before deploying an experiment.
+In support of this assessment, we offer up a cautionary tale of our own experience: that of \cite{clusters}, where we designed an experiment to test which plot aesthetics promoted discovery of linear trends and/or clusters.
 
-Explain the mishap with feature hierarchy - too many plots (because giant factorial experiment) meant that we missed cluster size imbalance.
-When we corrected that imbalance, though, we ended up with clusters that were oddly shaped.
-Sometimes, there's just no way to win, and we might have needed to rethink our sampling model.
+The experiment was a full 2x3x3 factorial exploration of three data generating parameters, with 3 replicates at each parameter combination (54 datasets) and 10 aesthetic combinations (for a total of 540 lineups).
+Each lineup had 20 different sub-panels, so we should have carefully visually inspected some 10800 different panels.
+As is evident from the fact that we're telling this story as a cautionary tale, we missed a critical problem with our data-generating mechanism: when clusters were assigned to randomly generated data after the fact, we didn't control the cluster size, leading to clusters of one or two points in relatively few sub-panels.
+This became particularly noticeable when bounding ellipses were added to the plot, as the method used to generate those ellipses required at least 3 points in the cluster. 
+Many sub-plots were generated with only 2 or 4 ellipses (with 3 or 5 total clusters), which escaped our notice during the stimuli proof-reading phase of the experiment, but did not escape the notice of our participants, who only needed to examine about 10 lineups each (around 200 panels).
+An example of one of the problematic lineups is shown in @fig-lineup-problems: many participants selected panel 16 because of the missing ellipse; not a wrong choice, but certainly not the visual effect we intended to test.
+
+![A lineup from Vanderplas & Hofmann \cite{clusters}.
+Panel 10 shows the clustered target data and panel 17 shows the target data with a strong linear relationship; either of these target panels was the expected choice.
+Unfortunately, panel 16 has only two bounding ellipses shown, which is an unintentional difference that resulted from a faulty method for assigning clusters to null plots; many participants selected this panel instead of one of the target panels.](lineup-missing-ellipse.png){#fig-lineup-problems}
+
+One reason why it is so difficult to generate sampling models for visual explorations is that our visual systems are very, very good at finding differences and unexpected results. 
+We re-ran the experiment using a different clustering method, and found that instead of cuing in on the number of ellipses, when that variable was removed participants instead cued in on the differences in size and shape of the ellipses formed using k-means clustering after the data generating procedure.
+That is, participants could still detect the artificial nature of the induced clusters using other features. 
+While it can be difficult to get the data generating method right, it is essential to conducting visual experiments that generalize well beyond the effects shown in a single data set or phenomenon. 
+The time and effort invested in this step at the outset of the experiment pays dividends when it allows for clear generalization of the experimental results to an entire statistical concept rather than a single data set.
 

### 2022-06-26 index.qmd
Experimental design

@@
 Simply asking people to read quantities off of a graph may not generalize beyond the questions asked or the data used in the chart\cite{eellsRelativeMeritsCircles1926,croxtonBarChartsCircle1927,vonhuhnFurtherStudiesGraphic1927,croxtonGraphicComparisonsBars1932}, but designing a study that is sufficiently robust to those issues requires manipulation of so many factors as to be functionally impossible to complete.
-In this paper, we attempt to distill lessons learned from several different sets of experimental graphics experiments\cite{clusters,sineillusion,emily-diss,power}, discussing the use of several different testing methods for statistical graphics\cite{vanderplasTestingStatisticalCharts2020}, the process of designing a graphical experiment, and analysis of the resulting empirical data.
+In this paper, we attempt to distill lessons learned from several different sets of experimental graphics experiments\cite{clusters,sineillusion,emily-diss,power,vanderplasFramedReproducingRevisiting2019}, discussing the use of several different testing methods for statistical graphics\cite{vanderplasTestingStatisticalCharts2020}, the process of designing a graphical experiment, and analysis of the resulting empirical data.
 It is our hope that by writing this paper, we will prevent other researchers from repeating hard lessons in what to do (and what not to do), lowering the barriers that exist for conducting research involving empirical tests of statistical graphics.
@@
 
-Of course, as we have confined ourselves to the discussion of online, asynchronous testing methods, every user interaction with the testing materials (typically hosted on a web page) can also be recorded along with time stamps, mouse positions, and so on.
-While we have not used this information heavily in our experimental analyses thus far, we do collect some time stamp data in order to assess how long participants spend on each question; it could almost certainly be more informative in the right setting.
+Of course, as we have confined ourselves to the discussion of online, asynchronous testing methods, every user interaction with the testing materials (typically hosted on a web page) can also be recorded along with time stamps, mouse positions, browser size and screen resolution, and other information.
+While we have not used this type of information heavily in our experimental analyses thus far, we do collect some time stamp data in order to assess how long participants spend on each question; it could almost certainly be more informative in the right setting.
 In addition, we have also provided participants with supportive tools, such as "scratch pads" and limited function calculators built into the Shiny application that allow us to gain additional insight into the estimation and visual comparison process when more complex calculations are required to answer the numerical estimation question (@fig-estimation-calc).
@@
 
-In our experience, one of the most difficult components of designing an experiment which asks users to directly estimate information from a chart is that the wording of such questions must be extremely carefully determined, and even when great care is taken with the construction of the question, participant answer accuracy is fundamentally limited by the fact that many participants do not read and interpret the question with the care and precision that it was written.
+In our experience, one of the most difficult components of designing an experiment which asks users to directly estimate information from a chart using a full scenario (background information, etc.
+as well as contextual details from the chart) is that the wording of such questions must be extremely carefully determined, and even when great care is taken with the construction of the question, participant answer accuracy is fundamentally limited by the fact that many participants do not read and interpret the question with the care and precision that it was written.
 Questions that ask participants to e.g.
 estimate the multiplicative difference in a quantity at two timepoints may be misunderstood as asking for an estimate of the additive difference, and the resulting estimates are then one or more orders of magnitude off of the correct answer.
-This is one area where lineup tests are definitely superior - they do not depend on participants to understand the nuances of language or scenarios built around the chart under investigation.
+This is one area where lineup tests are convenient - they do not depend on participants to understand the nuances of language or scenarios built around the chart under investigation.
+However, in some situations it may be sufficient to ask participants to estimate direct numerical quantities that have little contextual information, as we did in \cite{vanderplasFramedReproducingRevisiting2019} when assessing the accuracy of framed plots re-created from the Statistical Atlas.
 
-In some cases, we want participants to *engage directly* with an interactive visualization. 
-We may want to record the way in which they interact with the chart to see what types of information they seek out, but we may also want to ask them to interact with the chart in a specific way and record the result.
+It is also possible to design a task that requires participants to *engage directly* with an interactive visualization. 
+This can be useful in a directed task, where users are asked to interact with the chart in a specific way and the result is recorded, but it is also possible to use interactive visualizations in an open-ended task, recording how users engage with the graphic in an exploratory (as opposed to goal-directed) manner.
 In one recent experiment, we asked participants to forecast an exponential trend, with data presented on either a linear or log scale. 
-Using some JavaScript modified from New York Times interactive graphics "You Draw It" features\cite{katzYouDrawIt2017}, we had users draw trend lines and make forecasts directly on interactive charts, with the predictions and trend lines recorded to our database for later analysis^[See a gif of testing with You Draw It [here](https://i.imgur.com/GM5YSen.gif)].
+Using JavaScript code modified from New York Times interactive graphics "You Draw It" features\cite{katzYouDrawIt2017}, we had users draw trend lines and make forecasts directly on interactive charts, with the data and user-drawn predictions recorded to our database.
 With interactive graphics rendered using JavaScript (or other web libraries), the only limit to the types of questions you can ask in testing graphics is in your ability to write code to interact with the visualization library.
@@
 
-Whichever testing method is chosen should be appropriate to the type of question under investigation.
+Whichever testing method is chosen should be appropriate to the type of question under investigation and the level of visual engagement required to answer that question.
+While lineups are excellent tools for assessing perceptual questions, they cannot address questions aimed at understanding how people use charts within the wider context of a story or practical task; this requires more direct methods with higher ecological validity.
+
 Lineups require a great amount of work in developing a null sampling model which replicates the important visual features of the data, but model specification is extremely important when asking users questions about any simulated data set or even when examining user interactions directly.
-It  is important to carefully calibrate the model parameters to allow for some variability, but not too much.
+It  is important to carefully calibrate the model parameters to allow for some variability, but not too much, and to ensure that participants can succeed at the task and do not feel like they are being made to analyze random noise.
 This Goldilocks-style problem is the focus of the next section.
@@
 
-- Attention span
-- Sensory overload - too many lineups in a row is hard
-- Comparing across participants - importance of random effects for participant skill level
-- Consider asking both demographic questions and questions about mathematical skill level.
-We would have been wise to ask about country of primary/secondary education in our log study, as logarithms are introduced at different levels in different countries.
-But we didn't think of that when recruiting participants, so it would have been hard to manipulate that factor.
-There are always missed opportunities.
+It would be difficult to develop a full data generating model without some idea of the experimental design: the basic structure of the parameters which are to be manipulated, how the users will be tested, and so on.
+These experimental design factors are fairly natural for scientists to accumulate over the course of imagining and planning an experiment.
+When conducting graphical tests, however, there are additional considerations beyond those taught in a standard experimental design course.
 
-- We've had lots of success with complete and incomplete block designs
+The primary human design factor to be aware of is that visual tasks and assessing statistical graphics can be extremely cognitively taxing.
+In our experience, it is difficult to expect participants to evaluate more than about 15 statistical lineups in one sitting.
+It is easy to understand how this arises when we consider the concept of decision fatigue: each standard lineup has 20 panels, so there are 190 pairwise comparisons to evaluate between the plots of just a single lineup (though it is unlikely that participants actually sequentially evaluate each comparison in a combinatorial fashion).
+Tasks which are more interactive, such as you-draw-it, may be somewhat easier for participants, but it is unlikely that participants would be willing to complete more than about 20 tasks in one sitting even with tasks that require fewer decisions and more engagement.
+At some point, participants' interest in accuracy will decline, and the researcher is better off fitting the task within the window where participants are attentive and engaged.
 
-- Importance of randomizing trial order to combat order effects
+It can be extremely helpful to include "practice" demonstrations of the task to show the basic process, logic, and reasoning.
+While it is tempting to make these tasks fully representative of the type of judgement which will be required of participants, practice tasks which are too close to the experimental task may bias participants; we have found that it works well to have a relatively easy practice task which utilizes a slightly different type of plot and/or type of data than what will be tested in the experiment.
+In cases which require interactivity, gif animations of the task being carried out are useful, as are additional visual cues, such as the yellow box used in the You-Draw-It task[^1] to indicate that there were points which were not completed.
+
+While studies have found some relationship between lineup performance and completion of calculus\cite{vanderplasSpatialReasoningData2016a}, these differences are relatively small. 
+Studies have also not found a strong relationship between visual task performance and participant recruitment method\cite{vanderplasFramedReproducingRevisiting2019}, perhaps in part because demographics which use social media sites and demographics completing online research tasks for pay overlap heavily.
+What is clear, however, is that some participants are very good at evaluating graphics and some participants are not; as a result, it can be extremely helpful to use random effects models with an effect for individual participants.
+
+When we conduct visual graphics studies, we typically ask for basic demographic information from participants in order to assess how representative our sample is of the wider population; we typically do not analyze results by demographic variables in part because past studies have found so few differences in performance.
+It may be useful to ask a few more demographic questions about STEM education level for studies which ask more of participants from a mathematical standpoint; while lineup studies have not found strong associations with those variables, lineup studies also do not require participants to engage with the data presented in a chart in a way that requires higher-order mathematical reasoning.
+This has allowed us to make the argument to IRB that our research is exempt, as we do not collect enough demographic information to identify participants, however, this has occasionally come at some cost.
+Our recent study examining the use of log scales and exponential data was conducted using Prolific, which recruits participants from around the world; we required only that participants were fluent in English to participate.
+It was only after the experiment was completed that we realized that different countries introduce logarithms as a concept at different points during primary and secondary education; it might be that individuals in some countries have much more experience with log scales than those educated in the United States.
+We mention this only to point out that while every experiment contains a few missed opportunities, it is worth giving careful though to the demographic questions asked of participants and what information may be helpful during the analysis stage.
+
+In statistical design terms, most of our studies involve some type of balanced incomplete block design, where participants are assigned to a subset of experimental conditions which allow for estimation of the full range of effects specified in the model.
+The particular structure of these designs depends heavily on the factorial structure of the study, but we typically arrange participants' trials to ensure that participants see the same data set only once (where possible) and see as many different experimental conditions as possible.
+It is also important to reduce the impact of order effects using either e.g.
+latin square designs or randomization where possible, but we recognize that this is not always feasible due to the need to maintain participant naivete during some portions of the experiment.
+
+We have conducted visualization experiments using a wide variety of tools: custom web servers running experiments using PHP black magic, Qualtrics surveys for static graphics, and writing full Shiny applications that control every part of the experimental process from instructions to providing completion codes for payment.
+In our research, Shiny has provided the right balance between control over the experimental setting, procedure, etc.
+and the intricate details of web server management, however, this balance is likely different for every group and potentially for every experiment.
+Likewise, we have recruited participants using Amazon Mechanical Turk, Prolific, Reddit and other social media sites, and email; each recruitment method has trade-offs between cost, convenience, and control over demographic variables.
+
+[^1]: See a gif of testing with You Draw It [here](https://i.imgur.com/GM5YSen.gif)
 

### 2022-06-26 index.qmd
Pilot testing/QA section

@@
 
-Test early, test often, and make sure you look at all of your plots!!!
-If you're generating new data for each participant, be sure to save that data to a database so that you can at least see where/how things went wrong!
+Once you have your experiment designed and your charts to test, the next step is an extensive round of pilot testing. 
+The goal with pilot testing is to ensure that the experiment is set up precisely the way you want it, to ensure that your participants know what they are supposed to be doing and that any vague instructions are fixed, to provide time estimates for how long it will take to complete the task, and to provide you with some early data to examine for red flags.
+Our studies usually go through 2-3 rounds of pilot testing, with at least one of those rounds involving any relatives we can find that are less computer literate and will invariably figure out how to crash the testing application, and another round involving any and all coworkers and friends who may be available for 10-15 minutes during the work day.
+Pilot study samples do not need to be representative of any particular population, though we would caution against using exclusively visualization researchers or statisticians in your pilot sample because some issues are less likely to show up with knowledgeable participants.
+In some studies, each participant may see a new set of data, depending on how the study is designed.
+In these cases, we highly recommend saving all generated data to a database as well, so that it is possible to go back and examine exactly what happened and/or how things went wrong.
+Hard drive space is historically cheap relative to almost any other cost in an experiment; saving all of the data is a sensible measure.
 
-# Data Collection and Monitoring
-
-Prolific, Amazon Turk, testing with interested subgroups.
-
-Use of pilot data to set up analysis documents, github actions and/or shiny dashboard to monitor incoming results.
-
-Not necessarily important for small tests of limited duration, because data from e.g. prolific comes in so quickly. Server load is actually a much greater concern.
-
+One highly useful (but not strictly essential) component of an experiment that can be set up during the pilot testing stage is a basic analysis script which summarizes all data collected to date visually.
+We have used such scripts in the past to produce automatically updating dashboards or web pages, allowing for real time or near-real time monitoring of data collection efforts. 
+This provides an easy way to summarize completion of the experiment so that individuals can receive credit (if using services like MTurk or Prolific), but also allows interested participants to see the data if recruiting participants who are more likely to be interested in the results, as sometimes happens on social media sites.
+As data collection online can also happen extremely quickly (300 participants in <2h in our most recent Prolific experiment), this can provide an illusion of control over the deluge of data, allowing any issues to be spotted and resolved relatively quickly.
+If server load is a potential issue, it may also help to release batches of trials over a longer period of time in order to minimize the chance of having to make participants wait for others to complete the task before the server can handle additional connections^[This is the one major drawback to our preferred solution of self-hosting a Shiny server to handle data collection: the free version of Shiny server is limited to about 15 connections at any given time.].
+\
 # Analyzing the Data

### 2022-06-26 index.qmd
Fix formatting
Add mahbub's arxiv paper about demographics

@@
 
-While studies have found some relationship between lineup performance and completion of calculus\cite{vanderplasSpatialReasoningData2016a}, these differences are relatively small. 
+While studies have found some relationship between lineup performance and demographic factors\cite{majumderHumanFactorsInfluencing2014, vanderplasSpatialReasoningData2016a}, these differences are relatively small and unlikely to be important.
 Studies have also not found a strong relationship between visual task performance and participant recruitment method\cite{vanderplasFramedReproducingRevisiting2019}, perhaps in part because demographics which use social media sites and demographics completing online research tasks for pay overlap heavily.
@@
 If server load is a potential issue, it may also help to release batches of trials over a longer period of time in order to minimize the chance of having to make participants wait for others to complete the task before the server can handle additional connections^[This is the one major drawback to our preferred solution of self-hosting a Shiny server to handle data collection: the free version of Shiny server is limited to about 15 connections at any given time.].
-\
+
 # Analyzing the Data
+
 Be prepared for surprises

### 2022-06-27 index.qmd
Roudning refs and analysis methods

@@
 pdf-engine: pdflatex
+
+execute: 
+  echo: false
+  out-width: "\\columnwidth"
 
@@
 
-Be prepared for surprises
+One constant with data analysis of these types of experiments is that no matter how carefully the experiment is planned, designed, and executed, there will be surprises. 
+This is the dual curse and blessing of studying human perception: the visual system never quite works the way that we expect that it will, which provides endless fodder for science and occasionally complicates the data analysis.
 
-Use of linear models to test lineup data - see-value approach is great for single plots, but when we're interested in more concrete effects over many plots, it can be better to aggregate effects and just think of a multinomial model.
+**Generalized Linear Mixed Models** 
+ The see-value approach \cite{niladriroychowdhurySeeValueApp2020,vanderplasStatisticalSignificanceCalculations2021,majumderValidationVisualStatistical2013} is extremely useful for single lineups, but when a series of lineups that are part of a designed experiment are used, generalized linear mixed models are a much simpler way to summarize the overall effect of various manipulated factors\cite{power,clusters,emily-diss}.
+This approach also works extremely well for psychometric experiments \cite{sineillusion}, as psychometric models can be easily fit into the framework of a generalized linear mixed model.
+In addition, similar model structures can be used to model accuracy, response time, and confidence, if all three types of information are collected from participants during the experiment.
 
-Direct estimation can be tricky to analyze because of how messy the data is.
-Additional supplemental information can be critical to provide context, but graphical analysis of participant responses was more helpful than modeling for our experiment.
+**Numerical Estimation** There are additional considerations that should be expected when asking participants to estimate numerical quantities.
+Anchoring and rounding cause participant responses to cluster in ways that can bias statistical estimators, requiring methods designed for these types of data\cite{ushakovStatisticalAnalysisRounded2017,tourangeau_rips_rasinski_2000,heitjanIgnorabilityCoarseData1991}.
+An alternative approach is to analyze the data graphically using e.g.
+a combination of density and rug geoms for point estimates (as shown in @fig-density-rug), or spaghetti plots for individual response curves, with annotations to show anchor points.
+The advantage of this approach is that it can accommodate extremely messy data without requiring the extensive data cleaning and elimination of nonsensical responses that might be necessary to fit a statistical model.
 
-Useful graphical analysis methods: density plots + rug plots (so that you can see rounding effects and common errors), wordclouds, estimation sketches + spaghetti plots
+```{r estimation-data, message=FALSE, warning=FALSE, echo = F, cache = F}
+library(tidyverse)
+estimation_model_data <- read_csv("https://github.com/earobinson95/EmilyARobinson-UNL-dissertation/raw/main/data/03-estimation/estimation-model-data.csv")
+q0_text_summary <- read_csv("https://raw.githubusercontent.com/earobinson95/EmilyARobinson-UNL-dissertation/main/data/03-estimation/q0-text-summary.csv")
+estimation_simulated_data <- read_csv("https://github.com/earobinson95/EmilyARobinson-UNL-dissertation/raw/main/data/03-estimation/estimation-simulated-data.csv") %>%
+  mutate(x = x - 3000)
+estimation_scenario_text <- read_csv("https://github.com/earobinson95/EmilyARobinson-UNL-dissertation/raw/main/data/03-estimation/estimation-scenario-text.csv")
+estimation_parameters <- read_csv("https://github.com/earobinson95/EmilyARobinson-UNL-dissertation/raw/main/data/03-estimation/estimation-parameters.csv")
+estimation_questions <- read_csv("https://github.com/earobinson95/EmilyARobinson-UNL-dissertation/raw/main/data/03-estimation/estimation-questions.csv")
+population_estimates_data <- read_csv("https://github.com/earobinson95/EmilyARobinson-UNL-dissertation/raw/main/data/03-estimation/first-level-population-estimates.csv")
 
-## Analyzing Supplemental Information
+grid_lines_data <- tibble(scale = c(rep("linear", 12), rep("log2", 10)), 
+                          grid_lines = c(seq(0,55000, 5000),
+                                         2^seq(7,16))
+) %>%
+  expand_grid(dataset = c("dataset1", "dataset2"))
 
-Calculator observations
-Scratchpad
-Explanation
-Confidence Level
-User Interactions
+qe2_data <- estimation_model_data %>% 
+  filter(q_id == "QE2") %>%
+  mutate(response = as.numeric(response))
+```
+
+```{r density-plot, include = F, echo = F}
+densityPlot <- function(data, datasetID, estimate, xlabel = "Estimate", x_limits = c(0,70000), zoom = F, scalesx = T, zoom_limits = c(NA,NA), gridlines = T, rugjitter = 0.25) {
+
+  estPlot <- data %>%
+    filter(dataset %in% datasetID) %>%
+    ggplot(aes_string(x = estimate, fill = "scale", color = "scale")) +
+    geom_density(alpha = 0.5, color = NA) +
+    geom_rug(aes(y = -Inf), alpha = 0.6, show.legend = F, position = position_jitter(width = rugjitter, height = 0)) +
+    geom_vline(aes(xintercept = true_value, linetype = "a")) +
+    geom_vline(aes(xintercept = closest_pt_value, linetype = "b"))
+
+  if (gridlines) {
+  estPlot <- estPlot + 
+    geom_vline(data = grid_lines_data %>% filter(dataset == "dataset1", grid_lines >= x_limits[1], grid_lines <= x_limits[2]),
+               aes(xintercept = grid_lines, color = scale, linetype = "c"))
+  }
+
+  estPlot <- estPlot +
+    theme_bw() +
+    theme(aspect.ratio = 0.5) +
+    scale_color_manual("Scale", values = c("steelblue", "orange3")) +
+    scale_fill_manual("Scale", values = c("steelblue", "orange3")) +
+    scale_linetype_manual("", labels = c("Underlying Model Value", "Closest Point Value", "Grid Line Breaks"), values = c("solid", "dashed", "dotted")) +
+    scale_y_continuous("Density", labels = scales::comma)
+
+  if (zoom) {
+  estPlot <- estPlot +
+    facet_zoom(xlim = zoom_limits)
+  }
+
+  if (scalesx) {
+    estPlot <- estPlot +
+    scale_x_continuous(xlabel, limits = x_limits)
+  }
+
+  return(estPlot)
+
+}
+```
+
+```{r, qe2-density-plot, message=FALSE, warning=FALSE, echo = F}
+#| label: fig-density-rug
+#| fig-cap: "Density of participant estimates for the year in which the population reaches 4000.
+Colors are associated to scale - linear (blue) and log (orange) - and vertical lines indicate the true value based on the underlying model equation (black solid) and closest point value based on the simulated data set (black dashed).
+A jittered rug plot along the $x$-axis shows where participant estimates were made.
+The plot shows anchoring occured to the closest point as shown by an increase in density around the dashed line.
+Density peaks occurred at whole values indicating rounding errors."
+#| fig-width: 6
+#| fig-height: 3
+#| out-width: "\\columnwidth"
+
+library(patchwork)
+# DATA SET 1
+qe2_density_1 <- densityPlot(data = qe2_data, datasetID = "dataset1", "response", xlabel = "Estimated Year", x_limits = c(20,35), zoom = F, scalesx = T, gridlines = F, rugjitter = 0.25) +
+  ggtitle("Estimation of X given a Y value for Exponential Data") + 
+  guides(linetype = guide_legend(title = NULL)) + 
+  theme(legend.position = c(1,1), legend.justification = c(1,1), 
+        legend.background = element_rect(fill = "white", color = "black"),
+        legend.box.just = "right")
+
+qe2_density_1 
+```
+
+**Direct Interactions** If participants are making predictions and/or fitting visual statistics, we have had success analyzing these responses by comparing the responses to results from a statistical model to determine how visual statistics differ from the numerical quantities derived mathematically.
+For instance, we calculate the deviation between participant responses and the linear regression in You-Draw-It experiments, then fitted generalized additive mixed models to summarize the results across different experimental conditions to assess how user-drawn predictions deviated from the statistical estimates.
+In other direct interactions, it may be useful to compare participant selections or annotations to closest points on the chart to assess anchoring behavior; for discrete selections, methods discussed in numerical estimation may also be useful.
+
+**Qualitative Responses** In many cases, it is helpful to combine participants' qualitative reasoning with their quantitative responses to designed graphical experiments.
+This approach provides useful context as to what participants use to make their decisions, and can be useful when assessing why unexpected responses occurred.
+We have used wordclouds to show overall themes in participant responses successfully in \cite{clusters}; when paired with an appropriate linear model it became clear that participants were fixating on unequal cluster size as a visual cue.
+In cases where participants are provided with additional utilities such as calculators and scratchpads, it can be useful to select responses from individual participants which illustrate the different types of calculations performed (but analyzing this data quantitatively can be difficult).
 

### 2022-06-27 index.qmd
I really hate writing conclusions.

@@
 # Conclusion
+Testing features of visualization graphically using online platforms provides an incredibly powerful and efficient way to establish empirical guidelines for statistical graphics and visualization.
+There are nearly endless ways to combine web graphics, user interactions, and data collection to get insight into perception and use of graphics in practical settings.
+We have been continually surprised at the richness of the data collected in these experiments and the ability to combine qualitative and quantitative assessment to support conclusions that are both nuanced and of practical use when deciding how to design and present data using visualizations.
 

### 2022-06-27 index.qmd
rephrasing

@@
 
-Visual assessment is key, but narrowing the parameter space down to something that can be statistically determined is useful for providing a sort of numerical magnitude of difficulty assessment. 
+Visual assessment is key, but narrowing the parameter space down to something that can be statistically determined is useful for providing a sort of numerical summary estimate of potential visual difficulty.
 If using model based sampling methods, it is important to iteratively assess the sampling procedure through simulation: Start with a sampling method and define a parameter space that is appropriate for the model, and then simulate many data sets from each possible parameter combinations using a grid search.

### 2022-06-27 index.qmd
initial read through at face value with minor edits.

@@
     dept: Statistics Department
-    affiliation: CalPoly
-    email: earobinson@huskers.unl.edu
+    affiliation: California Polytechnic State University
+    email: erobin17@calpoly.edu
   - name: Heike Hofmann
@@
 
-# Testing Methods and Engagement
+# Testing Methods and \ear{User} Engagement
 <!-- Lineups, You Draw It, Direct Assessment -->
@@
 As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot instead of any of the 19 null plots, then the target plot is said to be visually significant\cite{majumderValidationVisualStatistical2013,vanderplasStatisticalSignificanceCalculations2021} and a "see" value (the visual analogue of a p-value for a statistical test) can be calculated\cite{niladriroychowdhurySeeValueApp2020}.
-In another variation of statistical lineup procedure, two models are compared, with target plots from each model embedded in the array of 20 total plots; the remaining null plots are constructed from a mixture model that blends the two competing models.
+In another variation of the statistical lineup procedure, two models are compared, with target plots from each model embedded in the array of 20 total plots; the remaining null plots are constructed from a mixture model that blends the two competing models.
 This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data\cite{clusters}.
@@
 
-![This question asks users to write out a description of how the population of Ewoks changes over time, without any further cues, to determine whether participants default to multiplicative or additive descriptions.](Estimation_describe_plot_crop.png){#fig-estimation-describe}
+![This question asks users to write out a description of how the population of Ewoks changes over time, without any further cues, to determine whether participants default to multiplicative or additive \ear{language} descriptions.](Estimation_describe_plot_crop.png){#fig-estimation-describe}
 
@@
 ![This question asks participants for a numerical estimate, but provides a basic calculator and scratchpad.
-All user interactions with the calculator and scratchpad are logged, providing insight into the user's thought process.](Estimation_numerical_screenshot_crop.png){#fig-estimation-calc}
+All user interactions with the calculator and scratchpad are logged, providing insight into the user's thought process \ear{and estimation strategy}.](Estimation_numerical_screenshot_crop.png){#fig-estimation-calc}
 
@@
 Questions that ask participants to e.g.
-estimate the multiplicative difference in a quantity at two timepoints may be misunderstood as asking for an estimate of the additive difference, and the resulting estimates are then one or more orders of magnitude off of the correct answer.
+estimate the multiplicative difference\ear{change} in a quantity at two time points may be misunderstood as asking for an estimate of the additive difference, and the resulting estimates are then one or more orders of magnitude off of the correct answer.
 This is one area where lineup tests are convenient - they do not depend on participants to understand the nuances of language or scenarios built around the chart under investigation.
@@
 
-We cannot overstate the importance of visual assessment of your model stimuli, preferrably with fresh eyes.
+We cannot overstate the importance of visual assessment of your model stimuli, preferably with fresh eyes.
 We highly recommend performing a series of think-aloud pilot testing before deploying an experiment.
@@
 It is easy to understand how this arises when we consider the concept of decision fatigue: each standard lineup has 20 panels, so there are 190 pairwise comparisons to evaluate between the plots of just a single lineup (though it is unlikely that participants actually sequentially evaluate each comparison in a combinatorial fashion).
-Tasks which are more interactive, such as you-draw-it, may be somewhat easier for participants, but it is unlikely that participants would be willing to complete more than about 20 tasks in one sitting even with tasks that require fewer decisions and more engagement.
+Tasks which are more interactive, such as \'You Draw It'}you-draw-it, may be somewhat easier for participants, but it is unlikely that participants would be willing to complete more than about 20 tasks in one sitting even with tasks that require fewer decisions and more engagement.
 At some point, participants' interest in accuracy will decline, and the researcher is better off fitting the task within the window where participants are attentive and engaged.
@@
 While it is tempting to make these tasks fully representative of the type of judgement which will be required of participants, practice tasks which are too close to the experimental task may bias participants; we have found that it works well to have a relatively easy practice task which utilizes a slightly different type of plot and/or type of data than what will be tested in the experiment.
-In cases which require interactivity, gif animations of the task being carried out are useful, as are additional visual cues, such as the yellow box used in the You-Draw-It task[^1] to indicate that there were points which were not completed.
+In cases which require interactivity, gif animations of the task being carried out are useful, as are additional visual cues, such as the yellow box used in the \ear{'You Draw It'} task[^1] to indicate that there were points which were not completed.
 
@@
 
-We have conducted visualization experiments using a wide variety of tools: custom web servers running experiments using PHP black magic, Qualtrics surveys for static graphics, and writing full Shiny applications that control every part of the experimental process from instructions to providing completion codes for payment.
+We have conducted visualization experiments using a wide variety of tools: custom web servers running experiments using PHP \ear{explain acronym} black magic, Qualtrics surveys for static graphics, and writing full Shiny applications that control every part of the experimental process from instructions to providing completion codes for payment.
 In our research, Shiny has provided the right balance between control over the experimental setting, procedure, etc.
@@
 
-[^1]: See a gif of testing with You Draw It [here](https://i.imgur.com/GM5YSen.gif)
+[^1]: See a gif of testing with 'You Draw It' [here](https://i.imgur.com/GM5YSen.gif)
 
@@
 **Direct Interactions** If participants are making predictions and/or fitting visual statistics, we have had success analyzing these responses by comparing the responses to results from a statistical model to determine how visual statistics differ from the numerical quantities derived mathematically.
-For instance, we calculate the deviation between participant responses and the linear regression in You-Draw-It experiments, then fitted generalized additive mixed models to summarize the results across different experimental conditions to assess how user-drawn predictions deviated from the statistical estimates.
+For instance, we calculate the deviation between participant responses and the linear regression in 'You Draw It' experiments, then fitted generalized additive mixed models to summarize the results across different experimental conditions to assess how user-drawn predictions deviated from the statistical estimates.
 In other direct interactions, it may be useful to compare participant selections or annotations to closest points on the chart to assess anchoring behavior; for discrete selections, methods discussed in numerical estimation may also be useful.

### 2022-06-28 index.qmd
add in color names/text
a couple more read throughs and comments.

@@
 
+header-includes:
+   - \usepackage[dvipsnames]{xcolor} % colors
+   - \newcommand{\ear}[1]{{\textcolor{blue}{#1}}}
+   - \newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
+   - \newcommand{\hh}[1]{{\textcolor{Green}{#1}}}
+
 date: last-modified
@@
 In another variation of the statistical lineup procedure, two models are compared, with target plots from each model embedded in the array of 20 total plots; the remaining null plots are constructed from a mixture model that blends the two competing models.
-This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data\cite{clusters}.
+This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data \ear{and assign a hiearchy of visual attributes}\cite{clusters}.
 Statistical lineups do not require the experimenter to provide any contextual information for the data: all of the necessary information is embedded in the choice of a null statistical model, which is convenient for testing purposes, but does not require engagement at a level that would allow for testing the viewer's understanding of the information shown in the chart.
@@
 In our experience, one of the most difficult components of designing an experiment which asks users to directly estimate information from a chart using a full scenario (background information, etc.
-as well as contextual details from the chart) is that the wording of such questions must be extremely carefully determined, and even when great care is taken with the construction of the question, participant answer accuracy is fundamentally limited by the fact that many participants do not read and interpret the question with the care and precision that it was written.
+as well as contextual details from the chart) is that the \ear{selection and} wording of such questions must be extremely carefully determined.
+\ear{Mathematics education researchers have provided guidelines for selecting different levels of questioning in order to assess graph comprehension: literal reading of the data, reading between the data, and reading beyond the data \cite{wood1968objectives, curcio1987comprehension, friel2001making, glazer2011challenges}.
+In a recent study, we identified questions based on this framework for evaluating direct estimates and extending those estimates to make comparisons between two points.}
+<!-- , and e -->
+\ear{E}ven when great care is taken with the construction of the question, participant answer accuracy is fundamentally limited by the fact that many participants do not read and interpret the question with the care and precision that it was written.
 Questions that ask participants to e.g.
-estimate the multiplicative difference\ear{change} in a quantity at two time points may be misunderstood as asking for an estimate of the additive difference, and the resulting estimates are then one or more orders of magnitude off of the correct answer.
+estimate the multiplicative \ear{change} in a quantity at two time points may be misunderstood as asking for an estimate of the additive difference, and the resulting estimates are then one or more orders of magnitude off of the correct answer.
 This is one area where lineup tests are convenient - they do not depend on participants to understand the nuances of language or scenarios built around the chart under investigation.
@@
 In one recent experiment, we asked participants to forecast an exponential trend, with data presented on either a linear or log scale. 
-Using JavaScript code modified from New York Times interactive graphics "You Draw It" features\cite{katzYouDrawIt2017}, we had users draw trend lines and make forecasts directly on interactive charts, with the data and user-drawn predictions recorded to our database.
+Using JavaScript code modified from New York Times interactive graphics "You Draw It" features\cite{katzYouDrawIt2017}, we had users draw trend lines \ear{with their computer mouse} and make forecasts directly on interactive charts, with the data and user-drawn predictions recorded to our database.
 With interactive graphics rendered using JavaScript (or other web libraries), the only limit to the types of questions you can ask in testing graphics is in your ability to write code to interact with the visualization library.
@@
 
-Whichever testing method is chosen should be appropriate to the type of question under investigation and the level of visual engagement required to answer that question.
+Whichever testing method is chosen should be appropriate to the type of question under investigation and the level of visual \ear{and cognitive} engagement required to answer that question.
 While lineups are excellent tools for assessing perceptual questions, they cannot address questions aimed at understanding how people use charts within the wider context of a story or practical task; this requires more direct methods with higher ecological validity.
 
+\ear{kind of a sudden jump into this paragraph, but I'm not sure how to transition into it right now.}
 Lineups require a great amount of work in developing a null sampling model which replicates the important visual features of the data, but model specification is extremely important when asking users questions about any simulated data set or even when examining user interactions directly.
@@
 It is easy to understand how this arises when we consider the concept of decision fatigue: each standard lineup has 20 panels, so there are 190 pairwise comparisons to evaluate between the plots of just a single lineup (though it is unlikely that participants actually sequentially evaluate each comparison in a combinatorial fashion).
-Tasks which are more interactive, such as \'You Draw It'}you-draw-it, may be somewhat easier for participants, but it is unlikely that participants would be willing to complete more than about 20 tasks in one sitting even with tasks that require fewer decisions and more engagement.
+Tasks which are more interactive, such as \ear{`You Draw It'}, may be somewhat easier for participants, but it is unlikely that participants would be willing to complete more than about 20 tasks in one sitting even with tasks that require fewer decisions and more engagement.
 At some point, participants' interest in accuracy will decline, and the researcher is better off fitting the task within the window where participants are attentive and engaged.
@@
 While it is tempting to make these tasks fully representative of the type of judgement which will be required of participants, practice tasks which are too close to the experimental task may bias participants; we have found that it works well to have a relatively easy practice task which utilizes a slightly different type of plot and/or type of data than what will be tested in the experiment.
-In cases which require interactivity, gif animations of the task being carried out are useful, as are additional visual cues, such as the yellow box used in the \ear{'You Draw It'} task[^1] to indicate that there were points which were not completed.
+In cases which require interactivity, gif animations of the task being carried out are useful, as are additional visual cues, such as the yellow box used in the \ear{`You Draw It'} task[^1] to indicate that there were points which were not completed.
 
@@
 
-We have conducted visualization experiments using a wide variety of tools: custom web servers running experiments using PHP \ear{explain acronym} black magic, Qualtrics surveys for static graphics, and writing full Shiny applications that control every part of the experimental process from instructions to providing completion codes for payment.
+We have conducted visualization experiments using a wide variety of tools: custom web servers running experiments using PHP black magic, Qualtrics surveys for static graphics, and writing full Shiny applications that control every part of the experimental process from instructions to providing completion codes for payment.
 In our research, Shiny has provided the right balance between control over the experimental setting, procedure, etc.
@@
 
-[^1]: See a gif of testing with 'You Draw It' [here](https://i.imgur.com/GM5YSen.gif)
+[^1]: See a gif of testing with `You Draw It' [here](https://i.imgur.com/GM5YSen.gif)
 
@@
 **Direct Interactions** If participants are making predictions and/or fitting visual statistics, we have had success analyzing these responses by comparing the responses to results from a statistical model to determine how visual statistics differ from the numerical quantities derived mathematically.
-For instance, we calculate the deviation between participant responses and the linear regression in 'You Draw It' experiments, then fitted generalized additive mixed models to summarize the results across different experimental conditions to assess how user-drawn predictions deviated from the statistical estimates.
+For instance, we calculate the deviation between participant responses and the linear regression in `You Draw It' experiments, then fitted generalized additive mixed models to summarize the results across different experimental conditions to assess how user-drawn predictions deviated from the statistical estimates.
 In other direct interactions, it may be useful to compare participant selections or annotations to closest points on the chart to assess anchoring behavior; for discrete selections, methods discussed in numerical estimation may also be useful.

### 2022-06-29 index.qmd
Accept Emily's changes

@@
 
-# Testing Methods and \ear{User} Engagement 
+# Testing Methods and User Engagement 
 <!-- Lineups, You Draw It, Direct Assessment -->
@@
 In another variation of the statistical lineup procedure, two models are compared, with target plots from each model embedded in the array of 20 total plots; the remaining null plots are constructed from a mixture model that blends the two competing models.
-This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data \ear{and assign a hiearchy of visual attributes}\cite{clusters}.
+This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data and assign a hiearchy of visual attributes\cite{clusters}.
 Statistical lineups do not require the experimenter to provide any contextual information for the data: all of the necessary information is embedded in the choice of a null statistical model, which is convenient for testing purposes, but does not require engagement at a level that would allow for testing the viewer's understanding of the information shown in the chart.
@@
 
-![This question asks users to write out a description of how the population of Ewoks changes over time, without any further cues, to determine whether participants default to multiplicative or additive \ear{language} descriptions.](Estimation_describe_plot_crop.png){#fig-estimation-describe}
+![This question asks users to write out a description of how the population of Ewoks changes over time, without any further cues, to determine whether participants default to multiplicative or additive language descriptions.](Estimation_describe_plot_crop.png){#fig-estimation-describe}
 
@@
 ![This question asks participants for a numerical estimate, but provides a basic calculator and scratchpad.
-All user interactions with the calculator and scratchpad are logged, providing insight into the user's thought process \ear{and estimation strategy}.](Estimation_numerical_screenshot_crop.png){#fig-estimation-calc}
+All user interactions with the calculator and scratchpad are logged, providing insight into the user's thought process and estimation strategy.](Estimation_numerical_screenshot_crop.png){#fig-estimation-calc}
 
 In our experience, one of the most difficult components of designing an experiment which asks users to directly estimate information from a chart using a full scenario (background information, etc.
-as well as contextual details from the chart) is that the \ear{selection and} wording of such questions must be extremely carefully determined.
-\ear{Mathematics education researchers have provided guidelines for selecting different levels of questioning in order to assess graph comprehension: literal reading of the data, reading between the data, and reading beyond the data \cite{wood1968objectives, curcio1987comprehension, friel2001making, glazer2011challenges}.
-In a recent study, we identified questions based on this framework for evaluating direct estimates and extending those estimates to make comparisons between two points.}
-<!-- , and e -->
-\ear{E}ven when great care is taken with the construction of the question, participant answer accuracy is fundamentally limited by the fact that many participants do not read and interpret the question with the care and precision that it was written.
+as well as contextual details from the chart) is that the selection and wording of such questions must be extremely carefully determined.
+Mathematics education researchers have provided guidelines for selecting different levels of questioning in order to assess graph comprehension: literal reading of the data, reading between the data, and reading beyond the data \cite{wood1968objectives, curcio1987comprehension, friel2001making, glazer2011challenges}.
+In a recent study, we identified questions based on this framework for evaluating direct estimates and extending those estimates to make comparisons between two points.
+
+Even when great care is taken with the construction of the question, participant answer accuracy is fundamentally limited by the fact that many participants do not read and interpret the question with the care and precision that it was written.
 Questions that ask participants to e.g.
-estimate the multiplicative \ear{change} in a quantity at two time points may be misunderstood as asking for an estimate of the additive difference, and the resulting estimates are then one or more orders of magnitude off of the correct answer.
+estimate the multiplicative change in a quantity at two time points may be misunderstood as asking for an estimate of the additive difference, and the resulting estimates are then one or more orders of magnitude off of the correct answer.
 This is one area where lineup tests are convenient - they do not depend on participants to understand the nuances of language or scenarios built around the chart under investigation.
@@
 In one recent experiment, we asked participants to forecast an exponential trend, with data presented on either a linear or log scale. 
-Using JavaScript code modified from New York Times interactive graphics "You Draw It" features\cite{katzYouDrawIt2017}, we had users draw trend lines \ear{with their computer mouse} and make forecasts directly on interactive charts, with the data and user-drawn predictions recorded to our database.
+Using JavaScript code modified from New York Times interactive graphics "You Draw It" features\cite{katzYouDrawIt2017}, we had users draw trend lines with their computer mouse and make forecasts directly on interactive charts, with the data and user-drawn predictions recorded to our database.
 With interactive graphics rendered using JavaScript (or other web libraries), the only limit to the types of questions you can ask in testing graphics is in your ability to write code to interact with the visualization library.
@@
 
-Whichever testing method is chosen should be appropriate to the type of question under investigation and the level of visual \ear{and cognitive} engagement required to answer that question.
+Whichever testing method is chosen should be appropriate to the type of question under investigation and the level of visual and cognitive engagement required to answer that question.
 While lineups are excellent tools for assessing perceptual questions, they cannot address questions aimed at understanding how people use charts within the wider context of a story or practical task; this requires more direct methods with higher ecological validity.
 
-\ear{kind of a sudden jump into this paragraph, but I'm not sure how to transition into it right now.}
+<!-- \ear{kind of a sudden jump into this paragraph, but I'm not sure how to transition into it right now.} -->
 Lineups require a great amount of work in developing a null sampling model which replicates the important visual features of the data, but model specification is extremely important when asking users questions about any simulated data set or even when examining user interactions directly.
@@
 It is easy to understand how this arises when we consider the concept of decision fatigue: each standard lineup has 20 panels, so there are 190 pairwise comparisons to evaluate between the plots of just a single lineup (though it is unlikely that participants actually sequentially evaluate each comparison in a combinatorial fashion).
-Tasks which are more interactive, such as \ear{`You Draw It'}, may be somewhat easier for participants, but it is unlikely that participants would be willing to complete more than about 20 tasks in one sitting even with tasks that require fewer decisions and more engagement.
+Tasks which are more interactive, such as `You Draw It', may be somewhat easier for participants, but it is unlikely that participants would be willing to complete more than about 20 tasks in one sitting even with tasks that require fewer decisions and more engagement.
 At some point, participants' interest in accuracy will decline, and the researcher is better off fitting the task within the window where participants are attentive and engaged.
@@
 While it is tempting to make these tasks fully representative of the type of judgement which will be required of participants, practice tasks which are too close to the experimental task may bias participants; we have found that it works well to have a relatively easy practice task which utilizes a slightly different type of plot and/or type of data than what will be tested in the experiment.
-In cases which require interactivity, gif animations of the task being carried out are useful, as are additional visual cues, such as the yellow box used in the \ear{`You Draw It'} task[^1] to indicate that there were points which were not completed.
+In cases which require interactivity, gif animations of the task being carried out are useful, as are additional visual cues, such as the yellow box used in the `You Draw It' task[^1] to indicate that there were points which were not completed.
 

### 2022-06-29 index.qmd
A real abstract

@@
 abstract: |
-  We outline our process and common pitfalls for developing and testing human perception of statistical graphics. 
-  Goal: Integrate information from the design of \cite{clusters,sineillusion,emily-diss,power}.
+  In this paper, we discuss considerations and methods for experimentally testing visualizations.
+We discuss levels of user engagement with graphics, common issues when developing a sampling or data generation model, the importance of pilot testing, and data analysis methods.
+Along the way, we also share hard-learned lessons on what not to do and some of the unique pitfalls of human testing in statistical and visualization research.
 

### 2022-06-29 index.qmd
I added a sentence to the abstract; I'm not sure how long to make it.

--- 2022-06-29 index.qmd
@@
 abstract: |
-  We outline our process and common pitfalls for developing and testing human perception of statistical graphics. 
-  Goal: Integrate information from the design of \cite{clusters,sineillusion,emily-diss,power}.
+   Through experimentation, graphical testing methods allow researchers to conduct studies geared at understanding human ability to conduct tasks related to the perception of statistical charts such as differentiation, prediction, estimation, and extrapolation.
+In this paper, we outline our process and common pitfalls for developing and testing human perception of statistical graphics by integrating information from the design of \cite{clusters,sineillusion,emily-diss,power}.
 

### 2022-07-05 index.qmd
couple of comments

--- 2022-06-29 index.qmd
@@
 It can be extremely challenging to design studies which strike the right combination of internal and external validity. 
-Simply asking people to read quantities off of a graph may not generalize beyond the questions asked or the data used in the chart\cite{eellsRelativeMeritsCircles1926,croxtonBarChartsCircle1927,vonhuhnFurtherStudiesGraphic1927,croxtonGraphicComparisonsBars1932}, but designing a study that is sufficiently robust to those issues requires manipulation of so many factors as to be functionally impossible to complete.
+Simply asking people to read quantities off of a graph may not generalize beyond the questions asked or the data used in the chart\cite{eellsRelativeMeritsCircles1926,croxtonBarChartsCircle1927,vonhuhnFurtherStudiesGraphic1927,croxtonGraphicComparisonsBars1932}, but designing a study that is sufficiently robust to those issues requires manipulation of so many factors as to be functionally close to impossible to complete.
 In this paper, we attempt to distill lessons learned from several different sets of experimental graphics experiments\cite{clusters,sineillusion,emily-diss,power,vanderplasFramedReproducingRevisiting2019}, discussing the use of several different testing methods for statistical graphics\cite{vanderplasTestingStatisticalCharts2020}, the process of designing a graphical experiment, and analysis of the resulting empirical data.
@@
 
-Statistical lineups\cite{bujaStatisticalInferenceExploratory2009,wickhamGraphicalInferenceInfovis2010} are a more generally applicable testing tool for perceptual questions such as "which chart displays this data more clearly"\cite{power}.
-As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot instead of any of the 19 null plots, then the target plot is said to be visually significant\cite{majumderValidationVisualStatistical2013,vanderplasStatisticalSignificanceCalculations2021} and a "see" value (the visual analogue of a p-value for a statistical test) can be calculated\cite{niladriroychowdhurySeeValueApp2020}.
+Statistical lineups\cite{bujaStatisticalInferenceExploratory2009,wickhamGraphicalInferenceInfovis2010} are a more generally applicable testing tool for perceptual questions such as "which chart displays this data more clearly"\cite{power} while simultaneously assessing the significance of the graphical finding in the chart.
+As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot at a higher rate than the 19 null plots, then the target plot is said to be visually significant\cite{majumderValidationVisualStatistical2013,vanderplasStatisticalSignificanceCalculations2021} and a "see" value (the visual analogue of a $p$-value for a statistical test) can be calculated\cite{niladriroychowdhurySeeValueApp2020}.
+<!-- Beyond the quantitative assessment of the signal with $p$- and see-values, the human involvement also allows us to collect information on the qualitative nature of *what* a viewer sees.--> 
 In another variation of the statistical lineup procedure, two models are compared, with target plots from each model embedded in the array of 20 total plots; the remaining null plots are constructed from a mixture model that blends the two competing models.
@@
 Of course, as we have confined ourselves to the discussion of online, asynchronous testing methods, every user interaction with the testing materials (typically hosted on a web page) can also be recorded along with time stamps, mouse positions, browser size and screen resolution, and other information.
-While we have not used this type of information heavily in our experimental analyses thus far, we do collect some time stamp data in order to assess how long participants spend on each question; it could almost certainly be more informative in the right setting.
+While we have not used this type of information heavily in our experimental analyses thus far, we do collect some time stamp data in order to assess how long participants spend on each question.
+What we typically see, is that the first round of test questions takes the longest for participants to complete.
+Additional experiments do not affect acuracy (i.e.
+there is no immediate learning effect) until after `too many' tests cognitive fatigue proves to be detrimental to accuracy\cite{chowdhury2018}.
+This sweet point between user replicates and fatigue depends on the cognitive burden in each test and should factor into setting the experiment.
+<!--While we have not used this type of information heavily in our experimental analyses thus far, we do collect some time stamp data in order to assess how long participants spend on each question; it could almost certainly be more informative in the right setting.-->
 In addition, we have also provided participants with supportive tools, such as "scratch pads" and limited function calculators built into the Shiny application that allow us to gain additional insight into the estimation and visual comparison process when more complex calculations are required to answer the numerical estimation question (@fig-estimation-calc).
@@
 Using JavaScript code modified from New York Times interactive graphics "You Draw It" features\cite{katzYouDrawIt2017}, we had users draw trend lines with their computer mouse and make forecasts directly on interactive charts, with the data and user-drawn predictions recorded to our database.
-With interactive graphics rendered using JavaScript (or other web libraries), the only limit to the types of questions you can ask in testing graphics is in your ability to write code to interact with the visualization library.
+With interactive graphics rendered using JavaScript (or other web libraries), the only limit to the types of questions one can ask in testing graphics is one's ability to write code to interact with the visualization library.
 This type of testing method can be extremely natural for participants, but it also is hard to generalize when discussing testing methods because of the potential range of applications where it might be employed.

### 2022-07-05 index.qmd
Edit title and fix spelling, trying to get things into the page limit

@@
 ---
-title: Testing Statistical Graphics
+title: A Guide to Designing Experiments to Test Statistical Graphics
 author-list:
@@
   - \CCScatTwelve{Human-centered computing}{Visu\-al\-iza\-tion}{Empirical studies in visualization}{}
-  - \CCScatTwelve{Human-centered computing}{Visu\-al\-iza\-tion}{Visualization application domains}{Scientific visualization}
-  - \CCScatTwelve{Applied computing}{Physical sciences and engineering}{Mathematics and statistics}{}
 firstsection: Introduction
@@
 In another variation of the statistical lineup procedure, two models are compared, with target plots from each model embedded in the array of 20 total plots; the remaining null plots are constructed from a mixture model that blends the two competing models.
-This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data and assign a hiearchy of visual attributes\cite{clusters}.
+This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data and assign a hierarchy of visual attributes\cite{clusters}.
 Statistical lineups do not require the experimenter to provide any contextual information for the data: all of the necessary information is embedded in the choice of a null statistical model, which is convenient for testing purposes, but does not require engagement at a level that would allow for testing the viewer's understanding of the information shown in the chart.

### 2022-07-05 index.qmd
tightening up wording

@@
 Many sub-plots were generated with only 2 or 4 ellipses (with 3 or 5 total clusters), which escaped our notice during the stimuli proof-reading phase of the experiment, but did not escape the notice of our participants, who only needed to examine about 10 lineups each (around 200 panels).
-An example of one of the problematic lineups is shown in @fig-lineup-problems: many participants selected panel 16 because of the missing ellipse; not a wrong choice, but certainly not the visual effect we intended to test.
+An example of one of the problematic lineups is shown in @fig-lineup-problems: many participants selected panel 16 because of the missing ellipse; not a wrong choice, but certainly not the effect we intended to test.
 
@@
 One reason why it is so difficult to generate sampling models for visual explorations is that our visual systems are very, very good at finding differences and unexpected results. 
-We re-ran the experiment using a different clustering method, and found that instead of cuing in on the number of ellipses, when that variable was removed participants instead cued in on the differences in size and shape of the ellipses formed using k-means clustering after the data generating procedure.
+We re-ran the experiment using a different clustering method, and found that instead of noticing the number of ellipses, when that variable was removed participants instead cued in on the differences in size and shape of the ellipses formed using k-means clustering after the data generating procedure.
 That is, participants could still detect the artificial nature of the induced clusters using other features. 

### 2022-07-05 index.qmd
Tightening language

@@
 
-In our experience, one of the most difficult components of designing an experiment which asks users to directly estimate information from a chart using a full scenario (background information, etc.
-as well as contextual details from the chart) is that the selection and wording of such questions must be extremely carefully determined.
-Mathematics education researchers have provided guidelines for selecting different levels of questioning in order to assess graph comprehension: literal reading of the data, reading between the data, and reading beyond the data \cite{wood1968objectives, curcio1987comprehension, friel2001making, glazer2011challenges}.
-In a recent study, we identified questions based on this framework for evaluating direct estimates and extending those estimates to make comparisons between two points.
+One of the most difficult components of designing an experiment which asks users to directly estimate information from a chart using a full scenario (background information, etc.
+as well as contextual details from the chart) is that the questions must be extremely carefully constructed.
+Mathematics education researchers provide guidelines for selecting different levels of questioning in order to assess graph comprehension: literal reading of the data, reading between the data, and reading beyond the data \cite{wood1968objectives, curcio1987comprehension, friel2001making, glazer2011challenges}.
+In a recent study, we identified questions based on this framework to evaluate direct estimates and extend those estimates to make comparisons between two points.
 

### 2022-07-05 index.qmd
mini changes

--- 2022-07-05 index.qmd
@@
 Each lineup had 20 different sub-panels, so we should have carefully visually inspected some 10800 different panels.
-As is evident from the fact that we're telling this story as a cautionary tale, we missed a critical problem with our data-generating mechanism: when clusters were assigned to randomly generated data after the fact, we didn't control the cluster size, leading to clusters of one or two points in relatively few sub-panels.
-This became particularly noticeable when bounding ellipses were added to the plot, as the method used to generate those ellipses required at least 3 points in the cluster. 
-Many sub-plots were generated with only 2 or 4 ellipses (with 3 or 5 total clusters), which escaped our notice during the stimuli proof-reading phase of the experiment, but did not escape the notice of our participants, who only needed to examine about 10 lineups each (around 200 panels).
+As is evident from the fact that we're telling this story as a cautionary tale, we missed a critical problem with our data-generating mechanism: when clusters were assigned to randomly generated data after the fact, we didn't control for the cluster size, leading to clusters of one or two points in some of the sub-panels.
+This became particularly noticeable when bounding ellipses were added to the plot, as the method used to generate those ellipses requires at least 3 points in the cluster.
+The corresponding sub-plots then missed one of the bounding ellipses.
+This problem escaped our notice during the stimuli proof-reading phase of the experiment, but did not escape the notice of our participants, who only needed to examine about 10 lineups each (around 200 panels).
 An example of one of the problematic lineups is shown in @fig-lineup-problems: many participants selected panel 16 because of the missing ellipse; not a wrong choice, but certainly not the visual effect we intended to test.
@@
 
-One reason why it is so difficult to generate sampling models for visual explorations is that our visual systems are very, very good at finding differences and unexpected results. 
+One reason why it is so difficult to generate sampling models for visual explorations is that our visual system is very, very good at finding differences and unexpected results. 
 We re-ran the experiment using a different clustering method, and found that instead of cuing in on the number of ellipses, when that variable was removed participants instead cued in on the differences in size and shape of the ellipses formed using k-means clustering after the data generating procedure.

### 2022-07-05 index.qmd
spell check

--- 2022-07-05 index.qmd
@@
 What we typically see, is that the first round of test questions takes the longest for participants to complete.
-Additional experiments do not affect acuracy (i.e.
+Additional experiments do not affect accuracy (i.e.
 there is no immediate learning effect) until after `too many' tests cognitive fatigue proves to be detrimental to accuracy\cite{chowdhury2018}.
-This sweet point between user replicates and fatigue depends on the cognitive burden in each test and should factor into setting the experiment.
+This sweet spot between user replicates and fatigue depends on the cognitive burden in each test and should factor into designing the experiment.
 <!--While we have not used this type of information heavily in our experimental analyses thus far, we do collect some time stamp data in order to assess how long participants spend on each question; it could almost certainly be more informative in the right setting.-->
@@
 #| label: fig-lof-density-curves
-#| fig-cap: "Density plot of the lack of fit statistic showing separation of selected difficulty levels: High (obvious curvature), Medium (noticable curvature), and Low (almost linear).
+#| fig-cap: "Density plot of the lack of fit statistic showing separation of selected difficulty levels: High (obvious curvature), Medium (noticeable curvature), and Low (almost linear).
 Each density plot is the result of 1000 simulations from a model $y_i = \\alpha\\cdot e^{\\beta\\cdot x_i + \\epsilon_i} + \\theta$, where $\\epsilon \\sim N(0, \\sigma^2)$.
@@
 
-The experiment was a full 2x3x3 factorial exploration of three data generating parameters, with 3 replicates at each parameter combination (54 datasets) and 10 aesthetic combinations (for a total of 540 lineups).
+The experiment was a full 2x3x3 factorial exploration of three data generating parameters, with 3 replicates at each parameter combination (54 data sets) and 10 aesthetic combinations (for a total of 540 lineups).
 Each lineup had 20 different sub-panels, so we should have carefully visually inspected some 10800 different panels.
@@
 It is also important to reduce the impact of order effects using either e.g.
-latin square designs or randomization where possible, but we recognize that this is not always feasible due to the need to maintain participant naivete during some portions of the experiment.
+Latin square designs or randomization where possible, but we recognize that this is not always feasible due to the need to maintain participant naivete during some portions of the experiment.
 
@@
 A jittered rug plot along the $x$-axis shows where participant estimates were made.
-The plot shows anchoring occured to the closest point as shown by an increase in density around the dashed line.
+The plot shows anchoring occurred to the closest point as shown by an increase in density around the dashed line.
 Density peaks occurred at whole values indicating rounding errors."
@@
 This approach provides useful context as to what participants use to make their decisions, and can be useful when assessing why unexpected responses occurred.
-We have used wordclouds to show overall themes in participant responses successfully in \cite{clusters}; when paired with an appropriate linear model it became clear that participants were fixating on unequal cluster size as a visual cue.
+We have used word clouds to show overall themes in participant responses successfully in \cite{clusters}; when paired with an appropriate linear model it became clear that participants were fixating on unequal cluster size as a visual cue.
 In cases where participants are provided with additional utilities such as calculators and scratchpads, it can be useful to select responses from individual participants which illustrate the different types of calculations performed (but analyzing this data quantitatively can be difficult).

### 2022-07-05 index.qmd

--- 2022-07-05 index.qmd
@@
 Each lineup had 20 different sub-panels, so we should have carefully visually inspected some 10800 different panels.
+
 As is evident from the fact that we're telling this story as a cautionary tale, we missed a critical problem with our data-generating mechanism: when clusters were assigned to randomly generated data after the fact, we didn't control the cluster size, leading to clusters of one or two points in relatively few sub-panels.
 This became particularly noticeable when bounding ellipses were added to the plot, as the method used to generate those ellipses required at least 3 points in the cluster.
-Many sub-plots were generated with only 2 or 4 ellipses (with 3 or 5 total clusters), which escaped our notice during the stimuli proof-reading phase of the experiment, but did not escape the notice of our participants, who only needed to examine about 10 lineups each (around 200 panels).
+The missing boundary ellipse in the corresponding sub-panels escaped our notice during the stimuli proof-reading phase of the experiment, but did not escape the notice of our participants, who only needed to examine about 10 lineups each (around 200 panels).
 An example of one of the problematic lineups is shown in @fig-lineup-problems: many participants selected panel 16 because of the missing ellipse; not a wrong choice, but certainly not the effect we intended to test.
@@
 
-One reason why it is so difficult to generate sampling models for visual explorations is that our visual systems are very, very good at finding differences and unexpected results. 
+One reason why it is so difficult to generate sampling models for visual explorations is that our visual system is very, very good at finding differences and unexpected results. 
 We re-ran the experiment using a different clustering method, and found that instead of noticing the number of ellipses, when that variable was removed participants instead cued in on the differences in size and shape of the ellipses formed using k-means clustering after the data generating procedure.

### 2022-07-05 index.qmd

--- 2022-07-05 index.qmd
@@
 
-The experiment was a full 2x3x3 factorial exploration of three data generating parameters, with 3 replicates at each parameter combination (54 data sets) and 10 aesthetic combinations (for a total of 540 lineups).
+The experiment was a full 2x3x3 factorial exploration of three data generating parameters, with 3 replicates at each parameter combination (54 datasets) and 10 aesthetic combinations (for a total of 540 lineups).
 Each lineup had 20 different sub-panels, so we should have carefully visually inspected some 10800 different panels.
@@
 This became particularly noticeable when bounding ellipses were added to the plot, as the method used to generate those ellipses required at least 3 points in the cluster.
-Many sub-plots were generated with only 2 or 4 ellipses (with 3 or 5 total clusters), which escaped our notice during the stimuli proof-reading phase of the experiment, but did not escape the notice of our participants, who only needed to examine about 10 lineups each (around 200 panels).
+The missing boundary ellipse in the corresponding sub-panels escaped our notice during the stimuli proof-reading phase of the experiment, but did not escape the notice of our participants, who only needed to examine about 10 lineups each (around 200 panels).
 An example of one of the problematic lineups is shown in @fig-lineup-problems: many participants selected panel 16 because of the missing ellipse; not a wrong choice, but certainly not the effect we intended to test.

### 2024-09-08 index.qmd
Updates to work with WIRE standards. Still to do: update refs with clickable DOIs, look for related wire articles, minimize lineup discussion in favor of other considerations, add introduction overview

@@
 title: A Guide to Designing Experiments to Test Statistical Graphics
-author-list:
+author:
   - name: Emily Robinson
@@
 We discuss levels of user engagement with graphics, common issues when developing a sampling or data generation model, the importance of pilot testing, and data analysis methods.
-Along the way, we also share hard-learned lessons on what not to do and some of the unique pitfalls of human testing in statistical and visualization research.
+Along the way, we also provide recommendations of how to avoid some of the unique pitfalls of human testing in statistical and visualization research.
 
-category: Research
-bib: refs.bib
-bibstyle: class-info/abbrv-doi.bst
-# csl: ieee-transactions-on-visualization-and-computer-graphics.csl
-ieee-keywords: 
-  - \CCScatTwelve{Human-centered computing}{Visu\-al\-iza\-tion}{Visualization design and evaluation methods}{} 
-  - \CCScatTwelve{Human-centered computing}{Visu\-al\-iza\-tion}{Empirical studies in visualization}{}
-firstsection: Introduction
+bibliography: refs.bib
+csl: apa.csl
 
-link-citations: false
-
-format:  
-  pdf:
-    keep-tex: true
-    template-partials: [partials/doc-class.tex, partials/title.tex, partials/before-body.tex, partials/after-body.tex]
-
-pdf-engine: pdflatex
+mainfont: Helvetica
+monofont: Roboto
+number-depth: 2
+number-sections: true
 
@@
   echo: false
-  out-width: "\\columnwidth"
 
@@
 date: last-modified
+
+format:  
+  pdf:
+    keep-tex: true
+    fontsize: 10pt
 ---
@@
 
+# Introduction
+
 Charts and data visualizations are primary methods of communicating scientific information to the public in what we typically hope is an easy-to-digest, visually attractive package.
-There are many strategies for creating charts and graphs, from Tufte-esque minimalism\cite{tufte} to charts designed with extra imagery and aesthetic appeal that draw the viewer's attention and persist in memory\cite[Ch 3]{cairoFunctionalArtIntroduction2012}.
+There are many strategies for creating charts and graphs, from Tufte-esque minimalism [@tufte] to charts designed with extra imagery and aesthetic appeal that draw the viewer's attention and persist in memory [@cairoFunctionalArtIntroduction2012 Ch 3].
 For a specified type of data, there are also usually many different chart forms to display that data: for instance, if we have a set of categorical data and we wish to show the relative proportions of each category, we could do so using a stacked bar chart or the polar equivalent, a pie chart.
-There have been several attempts to list out all of the types of charts\cite{ribeccaSearchChartsData}, create a taxonomy of charts\cite{bertin1983semiology,desnoyersTaxonomyVisualsScience2011}, and even to classify charts using a domain-specific grammar of graphics\cite{wilkinsonGrammarGraphics1999}.
+There have been several attempts to list out all of the types of charts [@ribeccaSearchChartsData], create a taxonomy of charts [@bertin1983semiology;@desnoyersTaxonomyVisualsScience2011], and even to classify charts using a domain-specific grammar of graphics [@wilkinsonGrammarGraphics1999].
 With all of the different design choices available, how are chart creators to know what is the best approach for communicating data to the appropriate audience?
 
-There are heuristics, rules-of-thumb, and various guidelines \cite{asme-standards-graphics,allenVisualizingScientificData2016,haemerDoubleScalesAre1948,fewInformationDashboardDesign2006,kosslynGraphDesignEye2006} for creating useful and visually attractive charts, but as scientists, we believe empirical evidence is essential for making correct decisions.
-The best way to establish the efficacy of various design decisions, then, is to test our charts by having them evaluated by people under controlled conditions and gathering appropriate data\cite{clevelandShapeParameterTwoVariable1988,clevelandGraphicalPerceptionGraphical1985}.
+While there are heuristics, rules-of-thumb, and various guidelines [@asme-standards-graphics;@allenVisualizingScientificData2016;@haemerDoubleScalesAre1948;@fewInformationDashboardDesign2006;@kosslynGraphDesignEye2006] for creating useful and visually attractive charts, the best way to establish the efficacy of various design decisions is to test charts, evaluate them under controlled conditions, and gather appropriate data [@clevelandShapeParameterTwoVariable1988;@clevelandGraphicalPerceptionGraphical1985].
 It can be extremely challenging to design studies which strike the right combination of internal and external validity. 
-Simply asking people to read quantities off of a graph may not generalize beyond the questions asked or the data used in the chart\cite{eellsRelativeMeritsCircles1926,croxtonBarChartsCircle1927,vonhuhnFurtherStudiesGraphic1927,croxtonGraphicComparisonsBars1932}, but designing a study that is sufficiently robust to those issues requires manipulation of so many factors as to be functionally close to impossible to complete.
-In this paper, we attempt to distill lessons learned from several different sets of experimental graphics experiments\cite{clusters,sineillusion,emily-diss,power,vanderplasFramedReproducingRevisiting2019}, discussing the use of several different testing methods for statistical graphics\cite{vanderplasTestingStatisticalCharts2020}, the process of designing a graphical experiment, and analysis of the resulting empirical data.
-It is our hope that by writing this paper, we will prevent other researchers from repeating hard lessons in what to do (and what not to do), lowering the barriers that exist for conducting research involving empirical tests of statistical graphics.
+Simply asking people to read quantities off of a graph may not generalize beyond the questions asked or the data used in the chart [@eellsRelativeMeritsCircles1926;@croxtonBarChartsCircle1927;@vonhuhnFurtherStudiesGraphic1927;@croxtonGraphicComparisonsBars1932], but designing a study that is sufficiently robust to those issues requires manipulation of so many factors as to be functionally close to impossible to complete.
+In this paper, we attempt to distill the experience gained from conducting several different types of graphics experiments [@clusters;@sineillusion;@emily-diss;@power;@vanderplasFramedReproducingRevisiting2019], discussing the use of different testing methods [@vanderplasTestingStatisticalCharts2020], the process of designing a graphical experiment, and analysis of the resulting empirical data.
+It is our hope that this paper will lower the barriers that exist for conducting empirical graphics research and reduce the probability of costly mistakes.
 
@@
 
-We will start with an overview of different testing methods that we have used to empirically assess graphics. 
-While it is certainly possible to use equipment such as eye-tracking devices, here we will primarily focus on studies which can be conducted over the internet, because these experiments have the lowest overhead, offer relatively fast data collection, and provide useful results that can be very clear if the experiment is designed correctly.
-Our toolkit is R-based\cite{R}, and includes ggplot2\cite{ggplot2} and Shiny\cite{shiny} as primary components, as well as  JavaScript (including D3\cite{d3}) to customize the Shiny interface for many of the experiments.
+There are many different testing methods used to empirically assess graphics. 
+While it is certainly possible to use equipment such as eye-tracking devices, this paper focuses primarily focus on studies which can be conducted over the internet, because these experiments have the lowest overhead, offer relatively fast data collection, and provide useful results that can be very clear if the experiment is designed correctly.
+The toolkit used here is R-based [@R], and includes ggplot2 [@ggplot2] and Shiny [@shiny] as primary components, as well as  JavaScript (including D3[@d3]) to customize the Shiny interface for many of the experiments.
 
 One primary concern when testing graphics is the level of user engagement required. 
-For instance, if we are testing whether someone can detect an effect such as a trend line in noisy data, this is a *perceptual* question best addressed using methods which allow the user to interact with the data on a basic visual level.
-Sometimes, it may be possible to set up a scenario where the user adjusts the plot directly, as in our experiment testing the strength of the sine illusion experiment\cite{sineillusion}: users adjusted the strength of the illusion correction until the lines appeared to be the same length, providing a direct measure of the magnitude of the sine illusion's effect, as in @fig-sine-illusion.
+For instance, testing whether someone can detect an effect such as a trend line in noisy data is a *perceptual* question best addressed using methods which allow the user to interact with the data on a basic visual level.
+Sometimes, it may be possible to set up a scenario where the user adjusts the plot directly, as in an experiment testing the strength of the sine illusion experiment [@sineillusion]: users adjusted the strength of the illusion correction until the lines appeared to be the same length, providing a direct measure of the magnitude of the sine illusion's effect, as in @fig-sine-illusion.
 
@@
 
-Statistical lineups\cite{bujaStatisticalInferenceExploratory2009,wickhamGraphicalInferenceInfovis2010} are a more generally applicable testing tool for perceptual questions such as "which chart displays this data more clearly"\cite{power} while simultaneously assessing the significance of the graphical finding in the chart.
-As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot at a higher rate than the 19 null plots, then the target plot is said to be visually significant\cite{majumderValidationVisualStatistical2013,vanderplasStatisticalSignificanceCalculations2021} and a "see" value (the visual analogue of a $p$-value for a statistical test) can be calculated\cite{niladriroychowdhurySeeValueApp2020}.
+Statistical lineups [@bujaStatisticalInferenceExploratory2009;@wickhamGraphicalInferenceInfovis2010] are a more generally applicable testing tool for perceptual questions such as "which chart displays this data more clearly" [@power] while simultaneously assessing the significance of the graphical finding in the chart.
+As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot at a higher rate than the 19 null plots, then the target plot is said to be visually significant [@majumderValidationVisualStatistical2013;vanderplasStatisticalSignificanceCalculations2021] and a "see" value (the visual analogue of a $p$-value for a statistical test) can be calculated [@niladriroychowdhurySeeValueApp2020].
 <!-- Beyond the quantitative assessment of the signal with $p$- and see-values, the human involvement also allows us to collect information on the qualitative nature of *what* a viewer sees.--> 
 In another variation of the statistical lineup procedure, two models are compared, with target plots from each model embedded in the array of 20 total plots; the remaining null plots are constructed from a mixture model that blends the two competing models.
-This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data and assign a hierarchy of visual attributes\cite{clusters}.
+This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data and assign a hierarchy of visual attributes [@clusters].
 Statistical lineups do not require the experimenter to provide any contextual information for the data: all of the necessary information is embedded in the choice of a null statistical model, which is convenient for testing purposes, but does not require engagement at a level that would allow for testing the viewer's understanding of the information shown in the chart.
-An example of a statistical lineup for testing the perception of exponential growth is shown in @fig-lineup-log.
+<!-- An example of a statistical lineup for testing the perception of exponential growth is shown in @fig-lineup-log. -->
 
-![A lineup experiment testing whether users can perceive differences in rates of exponential growth on linear and log scales.
-In addition to asking users to select the plot which is the most different, we also asked for basic reasoning and the user's level of confidence in their answer.](Lineup_log_screenshot.png){#fig-lineup-log}
+<!-- ![A lineup experiment testing whether users can perceive differences in rates of exponential growth on linear and log scales.
+In addition to asking users to select the plot which is the most different, we also asked for basic reasoning and the user's level of confidence in their answer.](Lineup_log_screenshot.png){#fig-lineup-log} -->
 
@@
 User feedback may be collected on a numerical scale or through the use of written comments, recorded "think-aloud" processes, and other more qualitative interaction methods. 
-We have conducted studies asking users to provide us with a written description of what the chart shows within a wider story-driven scenario (@fig-estimation-describe), and we have also asked for written feedback about why a certain panel of a statistical lineup was chosen; the insights from participant responses to these questions have been rich and insightful (and will be discussed more in the analysis section).
-While we have not (to date) recorded users talking out loud about what they are seeing, it is entirely feasible that so-called "think aloud" processes could be used within a Shiny application, with audio recordings saved to the server for transcription and analysis\cite{dunbar1995scientists,traftonTurningPicturesNumbers2000,kirschenbaum2003comparative}.
+In some studies, asking users to interpret a chart within a larger scenario can be effective, as in @fig-estimation-describe, while in others it is more helpful to ask users to explain answers. 
+In lineup studies, asking users why a specific panel was chosen can provide rich insight into otherwise confusing numerical results. 
+While we have not (to date) recorded users talking out loud about what they are seeing, think aloud methods could be implemented within a Shiny application, with audio recordings saved to the server for transcription and analysis
+[@dunbar1995scientists;@traftonTurningPicturesNumbers2000;@kirschenbaum2003comparative]; it is even possible that these recordings could be automatically transcribed using text-to-speech functions of large language models.
 
@@
 
-Of course, as we have confined ourselves to the discussion of online, asynchronous testing methods, every user interaction with the testing materials (typically hosted on a web page) can also be recorded along with time stamps, mouse positions, browser size and screen resolution, and other information.
-While we have not used this type of information heavily in our experimental analyses thus far, we do collect some time stamp data in order to assess how long participants spend on each question.
-What we typically see, is that the first round of test questions takes the longest for participants to complete.
-Additional experiments do not affect accuracy (i.e.
-there is no immediate learning effect) until after `too many' tests cognitive fatigue proves to be detrimental to accuracy\cite{chowdhury2018}.
-This sweet spot between user replicates and fatigue depends on the cognitive burden in each test and should factor into designing the experiment.
+Of course, in an online, asynchronous experiment, every user interaction with the testing materials (typically hosted on a web page) can also be recorded along with time stamps, mouse positions, browser size and screen resolution, and other information.
+While we have not used this type of information heavily in our experimental analyses thus far, in most experiments, we collect at least some time stamp data in order to assess how long participants spend on each question.
+Typically, the first round of test questions takes the longest for participants to complete.
+Additional replicates do not affect accuracy (i.e.
+there is no immediate learning effect) until after `too many' tests cognitive fatigue proves to be detrimental to accuracy [@chowdhury2018].
+This sweet spot between replicates and fatigue depends on the cognitive burden in each test and should factor into designing the experiment.
 <!--While we have not used this type of information heavily in our experimental analyses thus far, we do collect some time stamp data in order to assess how long participants spend on each question; it could almost certainly be more informative in the right setting.-->
-In addition, we have also provided participants with supportive tools, such as "scratch pads" and limited function calculators built into the Shiny application that allow us to gain additional insight into the estimation and visual comparison process when more complex calculations are required to answer the numerical estimation question (@fig-estimation-calc).
-These supportive tools have provided a wealth of information that, while not easy to analyze and generally messy, provides real insight into how participants were looking at the data, what strategies they tried and discarded for reading the chart, and what visual estimation methods were used.
+In some experiments, we have provided participants with supportive tools, such as "scratch pads" and calculators built into the Shiny application to support the complex calculations  required to answer higher-level numerical estimation questions (@fig-estimation-calc).
+These supportive tools reduce participant cognitive load, but also provide a wealth of (messy and difficult to analyze) information support real insight into how participants were looking at the data, what strategies they tried and discarded for reading the chart, and what visual estimation methods were used.
 
@@
 as well as contextual details from the chart) is that the questions must be extremely carefully constructed.
-Mathematics education researchers provide guidelines for selecting different levels of questioning in order to assess graph comprehension: literal reading of the data, reading between the data, and reading beyond the data \cite{wood1968objectives, curcio1987comprehension, friel2001making, glazer2011challenges}.
+Mathematics education researchers provide guidelines for selecting different levels of questioning in order to assess graph comprehension: literal reading of the data, reading between the data, and reading beyond the data [@wood1968objectives;@curcio1987comprehension;@friel2001making;@glazer2011challenges].
 In a recent study, we identified questions based on this framework to evaluate direct estimates and extend those estimates to make comparisons between two points.
@@
 estimate the multiplicative change in a quantity at two time points may be misunderstood as asking for an estimate of the additive difference, and the resulting estimates are then one or more orders of magnitude off of the correct answer.
-This is one area where lineup tests are convenient - they do not depend on participants to understand the nuances of language or scenarios built around the chart under investigation.
-However, in some situations it may be sufficient to ask participants to estimate direct numerical quantities that have little contextual information, as we did in \cite{vanderplasFramedReproducingRevisiting2019} when assessing the accuracy of framed plots re-created from the Statistical Atlas.
+This is one area where lineup methods are convenient - they do not depend on participants to understand the nuances of language or scenarios built around the chart under investigation.
+However, in some situations it may be sufficient to ask participants to estimate direct numerical quantities that have little contextual information, as we did in [@vanderplasFramedReproducingRevisiting2019] when assessing the accuracy of framed plots re-created from the Statistical Atlas.
 
@@
 In one recent experiment, we asked participants to forecast an exponential trend, with data presented on either a linear or log scale. 
-Using JavaScript code modified from New York Times interactive graphics "You Draw It" features\cite{katzYouDrawIt2017}, we had users draw trend lines with their computer mouse and make forecasts directly on interactive charts, with the data and user-drawn predictions recorded to our database.
+Using JavaScript code modified from New York Times interactive graphics "You Draw It" features [@katzYouDrawIt2017], we had users draw trend lines with their computer mouse and make forecasts directly on interactive charts, with the data and user-drawn predictions recorded to our database.
 With interactive graphics rendered using JavaScript (or other web libraries), the only limit to the types of questions one can ask in testing graphics is one's ability to write code to interact with the visualization library.
@@
 
-Once you have decided what graphical task you want to explore or test, it is necessary to develop a model which can be used to test the graphical features you want to explore in a precise manner. 
-This is the single longest part of the entire experimental design and execution process, in part because choosing a model that replicates important visual features of the data is an extremely complex process \cite{hullmanDesigningInteractiveExploratory2021,cookFoundationAvailableThinking2021,vanderplasDesigningGraphicsRequires2021}.
+Once the graphical task has been identified, it is necessary to develop a model which can be used to explore the graphical features of interest in a precise manner. 
+This is the single longest part of the entire experimental design and execution process, in part because choosing a model that replicates important visual features of the data is  extremely complex [@hullmanDesigningInteractiveExploratory2021;@cookFoundationAvailableThinking2021;@vanderplasDesigningGraphicsRequires2021].
 
-There are two main options when developing a statistical model for graphical testing: you can either start with a large dataset and sample from that dataset\cite{power}, or you can start from a model and sample data from that model generating process\cite{sineillusion,clusters,emily-diss}.
-This decision is largely determined by the availability of a large dataset containing the requisite features of interest and the qualities being manipulated in the experiment.
-For instance, Hofmann et al.
-\cite{power} used samples of different sizes from a pre-existing dataset to manipulate the amount of signal in each comparison; with a small sample, there is less signal and the same amount of noise, making the true plot harder to spot.
-In many situations, though, a convenient dataset with the right properties is harder to acquire, and it becomes necessary to develop a sampling model to generate data for user evaluation.
+There are two main options when developing a statistical model for graphical testing: start with a large data set and sample from that data set [@power], or start from a model and sample data from that model generating process [@sineillusion;@clusters;@emily-diss].
+This decision is largely determined by the availability of a large data set containing the requisite features of interest and the qualities being manipulated in the experiment. 
+For instance, @power used samples of different sizes from a pre-existing data set to manipulate the amount of signal in each comparison; with a small sample, there is less signal and the same amount of noise, making the true plot harder to spot.
+In many situations, though, a convenient data set with the right properties is harder to acquire, and it becomes necessary to develop a sampling model to generate data for user evaluation.
 
 The tools we discuss in the remainder of this section can be applied both to pre-existing data sets and to model-based sampling methods.
-We will do our best to keep this discussion general, but we will provide details in some situations to ensure that our general descriptions are not hopelessly vague.
 
@@
 Then, both examiners visually inspect a large number of plots generated using those parameters to get a sense for how difficult the task at hand is (this strategy is also described by Lu et al.
-\cite{luModelingJustNoticeable2022}).
+[@luModelingJustNoticeable2022]).
 At some point, all experimenters become so visually saturated with the nuances of the data generating mechanism that it may become necessary to "sanity check" the protocol with family members, friends, and colleagues.
@@
 We highly recommend performing a series of think-aloud pilot testing before deploying an experiment.
-In support of this assessment, we offer up a cautionary tale of our own experience: that of \cite{clusters}, where we designed an experiment to test which plot aesthetics promoted discovery of linear trends and/or clusters.
+In support of this assessment, we offer up a cautionary tale of our own experience: that of [@clusters], where we designed an experiment to test which plot aesthetics promoted discovery of linear trends and/or clusters.
 
-The experiment was a full 2x3x3 factorial exploration of three data generating parameters, with 3 replicates at each parameter combination (54 datasets) and 10 aesthetic combinations (for a total of 540 lineups).
+The experiment was a full 2x3x3 factorial exploration of three data generating parameters, with 3 replicates at each parameter combination (54 data sets) and 10 aesthetic combinations (for a total of 540 lineups).
 Each lineup had 20 different sub-panels, so we should have carefully visually inspected some 10800 different panels.
@@
 
-![A lineup from Vanderplas & Hofmann \cite{clusters}.
+![A lineup from Vanderplas & Hofmann [@clusters].
 Panel 10 shows the clustered target data and panel 17 shows the target data with a strong linear relationship; either of these target panels was the expected choice.
@@
 
-While studies have found some relationship between lineup performance and demographic factors\cite{majumderHumanFactorsInfluencing2014, vanderplasSpatialReasoningData2016a}, these differences are relatively small and unlikely to be important.
-Studies have also not found a strong relationship between visual task performance and participant recruitment method\cite{vanderplasFramedReproducingRevisiting2019}, perhaps in part because demographics which use social media sites and demographics completing online research tasks for pay overlap heavily.
+While studies have found some relationship between lineup performance and demographic factors [@majumderHumanFactorsInfluencing2014;@vanderplasSpatialReasoningData2016a], these differences are relatively small and unlikely to be important.
+Studies have also not found a strong relationship between visual task performance and participant recruitment method [@vanderplasFramedReproducingRevisiting2019], perhaps in part because demographics which use social media sites and demographics completing online research tasks for pay overlap heavily.
 What is clear, however, is that some participants are very good at evaluating graphics and some participants are not; as a result, it can be extremely helpful to use random effects models with an effect for individual participants.
@@
 **Generalized Linear Mixed Models** 
- The see-value approach \cite{niladriroychowdhurySeeValueApp2020,vanderplasStatisticalSignificanceCalculations2021,majumderValidationVisualStatistical2013} is extremely useful for single lineups, but when a series of lineups that are part of a designed experiment are used, generalized linear mixed models are a much simpler way to summarize the overall effect of various manipulated factors\cite{power,clusters,emily-diss}.
-This approach also works extremely well for psychometric experiments \cite{sineillusion}, as psychometric models can be easily fit into the framework of a generalized linear mixed model.
+ The see-value approach [@niladriroychowdhurySeeValueApp2020;@vanderplasStatisticalSignificanceCalculations2021;@majumderValidationVisualStatistical2013] is extremely useful for single lineups, but when a series of lineups that are part of a designed experiment are used, generalized linear mixed models are a much simpler way to summarize the overall effect of various manipulated factors [@power;@clusters;@emily-diss].
+This approach also works extremely well for psychometric experiments [@sineillusion], as psychometric models can be easily fit into the framework of a generalized linear mixed model.
 In addition, similar model structures can be used to model accuracy, response time, and confidence, if all three types of information are collected from participants during the experiment.
@@
 **Numerical Estimation** There are additional considerations that should be expected when asking participants to estimate numerical quantities.
-Anchoring and rounding cause participant responses to cluster in ways that can bias statistical estimators, requiring methods designed for these types of data\cite{ushakovStatisticalAnalysisRounded2017,tourangeau_rips_rasinski_2000,heitjanIgnorabilityCoarseData1991}.
+Anchoring and rounding cause participant responses to cluster in ways that can bias statistical estimators, requiring methods designed for these types of data [@ushakovStatisticalAnalysisRounded2017;@tourangeau_rips_rasinski_2000;@heitjanIgnorabilityCoarseData1991].
 An alternative approach is to analyze the data graphically using e.g.
@@
 ) %>%
-  expand_grid(dataset = c("dataset1", "dataset2"))
+  expand_grid(data set = c("data set1", "data set2"))
 
@@
 ```{r density-plot, include = F, echo = F}
-densityPlot <- function(data, datasetID, estimate, xlabel = "Estimate", x_limits = c(0,70000), zoom = F, scalesx = T, zoom_limits = c(NA,NA), gridlines = T, rugjitter = 0.25) {
+densityPlot <- function(data, data setID, estimate, xlabel = "Estimate", x_limits = c(0,70000), zoom = F, scalesx = T, zoom_limits = c(NA,NA), gridlines = T, rugjitter = 0.25) {
 
   estPlot <- data %>%
-    filter(dataset %in% datasetID) %>%
+    filter(data set %in% data setID) %>%
     ggplot(aes_string(x = estimate, fill = "scale", color = "scale")) +
@@
   estPlot <- estPlot + 
-    geom_vline(data = grid_lines_data %>% filter(dataset == "dataset1", grid_lines >= x_limits[1], grid_lines <= x_limits[2]),
+    geom_vline(data = grid_lines_data %>% filter(data set == "data set1", grid_lines >= x_limits[1], grid_lines <= x_limits[2]),
                aes(xintercept = grid_lines, color = scale, linetype = "c"))
@@
 # DATA SET 1
-qe2_density_1 <- densityPlot(data = qe2_data, datasetID = "dataset1", "response", xlabel = "Estimated Year", x_limits = c(20,35), zoom = F, scalesx = T, gridlines = F, rugjitter = 0.25) +
+qe2_density_1 <- densityPlot(data = qe2_data, data setID = "data set1", "response", xlabel = "Estimated Year", x_limits = c(20,35), zoom = F, scalesx = T, gridlines = F, rugjitter = 0.25) +
   ggtitle("Estimation of X given a Y value for Exponential Data") + 
@@
 This approach provides useful context as to what participants use to make their decisions, and can be useful when assessing why unexpected responses occurred.
-We have used word clouds to show overall themes in participant responses successfully in \cite{clusters}; when paired with an appropriate linear model it became clear that participants were fixating on unequal cluster size as a visual cue.
+We have used word clouds to show overall themes in participant responses successfully in [@clusters]; when paired with an appropriate linear model it became clear that participants were fixating on unequal cluster size as a visual cue.
 In cases where participants are provided with additional utilities such as calculators and scratchpads, it can be useful to select responses from individual participants which illustrate the different types of calculations performed (but analyzing this data quantitatively can be difficult).

### 2024-09-09 index.qmd
Update author formatting

@@
     dept: Statistics Department
-    affiliation: California Polytechnic State University
+    affiliation: 
     email: erobin17@calpoly.edu
+    orcid: 0000-0001-9800-7304
+    affiliations: 
+      - name: Statistics Department, California Polytechnic State University
+        address: Faculty Offices East, Building 25, Room 107D
+        city: San Luis Obispo
+        state: CA
+        postal-code: 93407-0405
   - name: Heike Hofmann
-    dept: Statistics Department 
-    affiliation: Iowa State University
-    email: hofmann@iastate.edu
+    orcid: 0000-0001-6216-5183
+    email: hhofmann4@unl.edu    
+    affiliations: 
+      - name: Statistics Department, University of Nebraska - Lincoln
+        address: 340 Hardin Hall North Wing
+        city: Lincoln
+        state: NE
+        postal-code: 68503
   - name: Susan Vanderplas
-    dept: Statistics Department
-    affiliation: University of Nebraska, Lincoln
     email: svanderplas2@unl.edu
+    orcid: 0000-0002-3803-0972
+    affiliations: 
+      - name: Statistics Department, University of Nebraska - Lincoln
+        address: 340 Hardin Hall North Wing
+        city: Lincoln
+        state: NE
+        postal-code: 68503
 

### 2024-09-09 index.qmd
Cite wire articles

@@
 Statistical lineups [@bujaStatisticalInferenceExploratory2009;@wickhamGraphicalInferenceInfovis2010] are a more generally applicable testing tool for perceptual questions such as "which chart displays this data more clearly" [@power] while simultaneously assessing the significance of the graphical finding in the chart.
-As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot at a higher rate than the 19 null plots, then the target plot is said to be visually significant [@majumderValidationVisualStatistical2013;vanderplasStatisticalSignificanceCalculations2021] and a "see" value (the visual analogue of a $p$-value for a statistical test) can be calculated [@niladriroychowdhurySeeValueApp2020].
+As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot at a higher rate than the 19 null plots, then the target plot is said to be visually significant [@loyDiagnosticToolsHierarchical2013;@majumderValidationVisualStatistical2013;vanderplasStatisticalSignificanceCalculations2021] and a "see" value (the visual analogue of a $p$-value for a statistical test) can be calculated [@niladriroychowdhurySeeValueApp2020].
 <!-- Beyond the quantitative assessment of the signal with $p$- and see-values, the human involvement also allows us to collect information on the qualitative nature of *what* a viewer sees.--> 

### 2024-09-15 index.qmd
Revisions and adding additional references

@@
 
-Charts and data visualizations are primary methods of communicating scientific information to the public in what we typically hope is an easy-to-digest, visually attractive package.
-There are many strategies for creating charts and graphs, from Tufte-esque minimalism [@tufte] to charts designed with extra imagery and aesthetic appeal that draw the viewer's attention and persist in memory [@cairoFunctionalArtIntroduction2012 Ch 3].
-For a specified type of data, there are also usually many different chart forms to display that data: for instance, if we have a set of categorical data and we wish to show the relative proportions of each category, we could do so using a stacked bar chart or the polar equivalent, a pie chart.
+Charts and data visualizations are primary methods of communicating scientific information to the public in what creators hope is an easy-to-digest, visually attractive package.
+There are many strategies for creating charts and graphs, from Tufte-esque minimalism [@tufte] to charts designed with extra imagery and aesthetic appeal that draw the viewer's attention and persist in memory [@cairoFunctionalArtIntroduction2012].
+For a specific type of data, there are also usually many different chart forms to display that data: for instance, if we have a set of categorical data and we wish to show the relative proportions of each category, we could do so using a stacked bar chart or the polar equivalent, a pie chart.
 There have been several attempts to list out all of the types of charts [@ribeccaSearchChartsData], create a taxonomy of charts [@bertin1983semiology;@desnoyersTaxonomyVisualsScience2011], and even to classify charts using a domain-specific grammar of graphics [@wilkinsonGrammarGraphics1999].
@@
 
-While there are heuristics, rules-of-thumb, and various guidelines [@asme-standards-graphics;@allenVisualizingScientificData2016;@haemerDoubleScalesAre1948;@fewInformationDashboardDesign2006;@kosslynGraphDesignEye2006] for creating useful and visually attractive charts, the best way to establish the efficacy of various design decisions is to test charts, evaluate them under controlled conditions, and gather appropriate data [@clevelandShapeParameterTwoVariable1988;@clevelandGraphicalPerceptionGraphical1985].
+While there are heuristics, rules-of-thumb, and various guidelines [@asme-standards-graphics;@allenVisualizingScientificData2016;@haemerDoubleScalesAre1948;@fewInformationDashboardDesign2006;@kosslynGraphDesignEye2006] for creating useful and visually attractive charts, the best way to establish the efficacy of various design decisions is to test charts: evaluate them under controlled conditions, and gather appropriate data [@clevelandShapeParameterTwoVariable1988;@clevelandGraphicalPerceptionGraphical1985] to compare different representations of the same data.
 It can be extremely challenging to design studies which strike the right combination of internal and external validity. 
-Simply asking people to read quantities off of a graph may not generalize beyond the questions asked or the data used in the chart [@eellsRelativeMeritsCircles1926;@croxtonBarChartsCircle1927;@vonhuhnFurtherStudiesGraphic1927;@croxtonGraphicComparisonsBars1932], but designing a study that is sufficiently robust to those issues requires manipulation of so many factors as to be functionally close to impossible to complete.
-In this paper, we attempt to distill the experience gained from conducting several different types of graphics experiments [@clusters;@sineillusion;@emily-diss;@power;@vanderplasFramedReproducingRevisiting2019], discussing the use of different testing methods [@vanderplasTestingStatisticalCharts2020], the process of designing a graphical experiment, and analysis of the resulting empirical data.
+Simply asking people to read quantities off of a graph may not generalize beyond the questions asked or the data used in the chart [@eellsRelativeMeritsCircles1926;@croxtonBarChartsCircle1927;@vonhuhnFurtherStudiesGraphic1927;@croxtonGraphicComparisonsBars1932], but designing a study that is sufficiently robust to those issues requires manipulation of so many factors as to be functionally close to impossible to control or manipulate in a single experiment.
+In this paper, we attempt to distill the experience gained from conducting several different types of graphics experiments [@clusters;@sineillusion;@emily-diss;@power;@vanderplasFramedReproducingRevisiting2019;@vanderplasEscapingFlatlandGraphics2024], discussing the use of different testing methods [@vanderplasTestingStatisticalCharts2020], the process of designing a graphical experiment, and analysis of the resulting empirical data.
 It is our hope that this paper will lower the barriers that exist for conducting empirical graphics research and reduce the probability of costly mistakes.
 
-# Testing Methods and User Engagement 
+@sec-testing-methods discusses different methods for testing graphics, and which methods best address different levels of user engagement. 
+In @sec-model-dev, we discuss the process of developing a data-generating model that can be used to control the statistical features of data used in the tested visualizations. 
+Model development is a nuanced and iterative process that ultimately determines the success and generalizability of the experimental results. 
+We explore different experimental design considerations in @sec-exp-design and then move to the importance of pilot testing in @sec-pilot-test. 
+Finally, we  provide some common analysis strategies in @sec-analysis, including strategies for handling the unexpected data features which are so common in graphical testing experiments.
+
+# Testing Methods and User Engagement {#sec-testing-methods}
 <!-- Lineups, You Draw It, Direct Assessment -->
@@
 There are many different testing methods used to empirically assess graphics. 
-While it is certainly possible to use equipment such as eye-tracking devices, this paper focuses primarily focus on studies which can be conducted over the internet, because these experiments have the lowest overhead, offer relatively fast data collection, and provide useful results that can be very clear if the experiment is designed correctly.
-The toolkit used here is R-based [@R], and includes ggplot2 [@ggplot2] and Shiny [@shiny] as primary components, as well as  JavaScript (including D3[@d3]) to customize the Shiny interface for many of the experiments.
+While it is certainly possible to use equipment such as eye-tracking devices, this paper focuses primarily focus on studies which can be conducted without testing equipment, and primarily online, because these experiments have the lowest overhead, offer relatively fast data collection, and provide useful results that can be very clear if the experiment is designed correctly.
+The toolkit used for these experiments is R-based [@R], and includes ggplot2 [@ggplot2] and Shiny [@shiny] as primary components, as well as  JavaScript (including D3[@d3]) to customize the Shiny interface for many of the experiments.
+While we prefer this set of tools, most of the observations described in this paper apply to a wide variety of different workflows for graphical experimentation.
 
@@
 Sometimes, it may be possible to set up a scenario where the user adjusts the plot directly, as in an experiment testing the strength of the sine illusion experiment [@sineillusion]: users adjusted the strength of the illusion correction until the lines appeared to be the same length, providing a direct measure of the magnitude of the sine illusion's effect, as in @fig-sine-illusion.
+In other situations, it may be possible to have the user directly interact with the plot, as in @vanderplasEscapingFlatlandGraphics2024, where users could rotate and interact with a 3D rendered bar chart; the application then can record these user interactions to provide insight into the visual comparisons the user may have been performing to supplement the estimates which the user explicitly provided.
 
@@
 
-Statistical lineups [@bujaStatisticalInferenceExploratory2009;@wickhamGraphicalInferenceInfovis2010] are a more generally applicable testing tool for perceptual questions such as "which chart displays this data more clearly" [@power] while simultaneously assessing the significance of the graphical finding in the chart.
+Statistical lineups [@bujaStatisticalInferenceExploratory2009;@wickhamGraphicalInferenceInfovis2010] are another useful testing tool for perceptual questions such as "which chart displays this data more clearly" [@power] while simultaneously assessing the significance of the graphical finding in the chart.
 As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot at a higher rate than the 19 null plots, then the target plot is said to be visually significant [@loyDiagnosticToolsHierarchical2013;@majumderValidationVisualStatistical2013;vanderplasStatisticalSignificanceCalculations2021] and a "see" value (the visual analogue of a $p$-value for a statistical test) can be calculated [@niladriroychowdhurySeeValueApp2020].
@@
 This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data and assign a hierarchy of visual attributes [@clusters].
-Statistical lineups do not require the experimenter to provide any contextual information for the data: all of the necessary information is embedded in the choice of a null statistical model, which is convenient for testing purposes, but does not require engagement at a level that would allow for testing the viewer's understanding of the information shown in the chart.
+Statistical lineups do not require the experimenter to provide any contextual information for the data: all of the necessary information is embedded in the choice of a null statistical model, which is convenient for testing purposes, but does not allow for testing the viewer's understanding of the information shown in the chart.
 <!-- An example of a statistical lineup for testing the perception of exponential growth is shown in @fig-lineup-log. -->
@@
 Of course, in an online, asynchronous experiment, every user interaction with the testing materials (typically hosted on a web page) can also be recorded along with time stamps, mouse positions, browser size and screen resolution, and other information.
-While we have not used this type of information heavily in our experimental analyses thus far, in most experiments, we collect at least some time stamp data in order to assess how long participants spend on each question.
+While we have not used this type of information heavily in our experimental analyses thus far, in most experiments, we collect time stamp data in order to assess how long participants spend on each question.
 Typically, the first round of test questions takes the longest for participants to complete. 
-Additional replicates do not affect accuracy (i.e.
-there is no immediate learning effect) until after `too many' tests cognitive fatigue proves to be detrimental to accuracy [@chowdhury2018].
+Additional replicates do not usually affect accuracy (i.e. there is no immediate learning effect) until after `too many' tests cognitive fatigue proves to be detrimental to accuracy [@chowdhury2018]. 
 This sweet spot between replicates and fatigue depends on the cognitive burden in each test and should factor into designing the experiment.
@@
 In some experiments, we have provided participants with supportive tools, such as "scratch pads" and calculators built into the Shiny application to support the complex calculations required to answer higher-level numerical estimation questions (@fig-estimation-calc).
-These supportive tools reduce participant cognitive load, but also provide a wealth of (messy and difficult to analyze) information support real insight into how participants were looking at the data, what strategies they tried and discarded for reading the chart, and what visual estimation methods were used.
+These supportive tools reduce participant cognitive load, but also record a wealth of messy information that provide real insight into how participants were looking at the data, what strategies they tried and discarded for reading the chart, and what visual estimation methods were used.
+While systematic analysis and modeling of this messy data is difficult, the insights provided can be extremely useful.
 
@@
 
-It is also possible to design a task that requires participants to *engage directly* with an interactive visualization. 
+Another useful measurement strategy is to require participants to *engage directly* with an interactive visualization. 
 This can be useful in a directed task, where users are asked to interact with the chart in a specific way and the result is recorded, but it is also possible to use interactive visualizations in an open-ended task, recording how users engage with the graphic in an exploratory (as opposed to goal-directed) manner.
@@
 <!-- \ear{kind of a sudden jump into this paragraph, but I'm not sure how to transition into it right now.} -->
-Lineups require a great amount of work in developing a null sampling model which replicates the important visual features of the data, but model specification is extremely important when asking users questions about any simulated data set or even when examining user interactions directly.
-It  is important to carefully calibrate the model parameters to allow for some variability, but not too much, and to ensure that participants can succeed at the task and do not feel like they are being made to analyze random noise.
+All of the testing methods described here require significant work to develop a strategy for data generation appropriate for testing the underlying question. 
+For instance, when testing the perception of exponential growth, we had to develop a model which would generate data with varying growth rates, but where the data had a pre-specified domain and range.
+The data generating model is particularly critical when using lineups, as the null sampling model must replicate the important visual features in the data. 
+Each testing method has specific requirements, but it is important to carefully calibrate the model parameters to allow for some variability, but not too much, and to ensure that participants can succeed at the task and do not feel like they are being made to analyze random noise.
 This Goldilocks-style problem is the focus of the next section.
 
-# Developing a Model
+# Developing a Model {#sec-model-dev}
 
@@
 
-Once an appropriate set of parameters are identified using the numerical screening method, then it is important to calibrate these parameter selections visually; after all, no numerical statistic is a perfect measure of what we actually see; at best, they are approximations of what we might potentially see.
+Once an appropriate set of parameters are identified using the numerical screening method, then it is important to calibrate these parameter selections visually. 
+No numerical statistic is a perfect measure of what we actually see: at best, they are approximations of what we might potentially see. 
 We have found it to be useful to have one experimenter calibrate the model parameters at a gross level, and then have another experimenter narrow in on the parameters which are visually reasonable within the selected range.
@@
 We cannot overstate the importance of visual assessment of your model stimuli, preferably with fresh eyes. 
-We highly recommend performing a series of think-aloud pilot testing before deploying an experiment.
+We highly recommend performing several rounds of think-aloud pilot testing before deploying an experiment. 
 In support of this assessment, we offer up a cautionary tale of our own experience: that of [@clusters], where we designed an experiment to test which plot aesthetics promoted discovery of linear trends and/or clusters.
@@
 The primary human design factor to be aware of is that visual tasks and assessing statistical graphics can be extremely cognitively taxing. 
-In our experience, it is difficult to expect participants to evaluate more than about 15 statistical lineups in one sitting.
-It is easy to understand how this arises when we consider the concept of decision fatigue: each standard lineup has 20 panels, so there are 190 pairwise comparisons to evaluate between the plots of just a single lineup (though it is unlikely that participants actually sequentially evaluate each comparison in a combinatorial fashion).
+In our experience, it is difficult to expect participants to evaluate more than about 15 charts in one sitting. 
+If users are asked to deeply engage and answer multiple questions about each chart, this limit may be lower (8-10), but even with a relatively simple task, as in lineup methods, it is hard for participants to accurately evaluate more than about 15 charts.
 Tasks which are more interactive, such as `You Draw It', may be somewhat easier for participants, but it is unlikely that participants would be willing to complete more than about 20 tasks in one sitting even with tasks that require fewer decisions and more engagement.
-At some point, participants' interest in accuracy will decline, and the researcher is better off fitting the task within the window where participants are attentive and engaged.
+At some point, participants' interest in accuracy will decline, and the researcher is better off using a larger number of participants completing fewer tasks within the window where participants are attentive and engaged.
 
@@
 
-While studies have found some relationship between lineup performance and demographic factors [@majumderHumanFactorsInfluencing2014;@vanderplasSpatialReasoningData2016a], these differences are relatively small and unlikely to be important.
+While studies have found some relationship between lineup performance and demographic factors [@majumderHumanFactorsInfluencing2014;@vanderplasSpatialReasoningData2016a], these differences are relatively small when participants are recruited from online testing platforms like Amazon Mechanical Turk or Prolific, or when participants are recruited from a university student population.
 Studies have also not found a strong relationship between visual task performance and participant recruitment method [@vanderplasFramedReproducingRevisiting2019], perhaps in part because demographics which use social media sites and demographics completing online research tasks for pay overlap heavily.
-What is clear, however, is that some participants are very good at evaluating graphics and some participants are not; as a result, it can be extremely helpful to use random effects models with an effect for individual participants.
+However, when participants are recruited using a statistically representative sample of the wider population, education becomes an extremely useful predictor of ability to effectively read and draw conclusions about a chart [@riceTestingPerceptualAccuracy2024].
+In almost every situation, though, we find that some participants are very good at evaluating graphics and some participants are not; as a result, it can be extremely helpful to use random effects models with an effect for individual participants.
 
-When we conduct visual graphics studies, we typically ask for basic demographic information from participants in order to assess how representative our sample is of the wider population; we typically do not analyze results by demographic variables in part because past studies have found so few differences in performance.
 It may be useful to ask a few more demographic questions about STEM education level for studies which ask more of participants from a mathematical standpoint; while lineup studies have not found strong associations with those variables, lineup studies also do not require participants to engage with the data presented in a chart in a way that requires higher-order mathematical reasoning.
@@
 Likewise, we have recruited participants using Amazon Mechanical Turk, Prolific, Reddit and other social media sites, and email; each recruitment method has trade-offs between cost, convenience, and control over demographic variables.
+@riceTestingPerceptualAccuracy2024 recruited participants using NORC panel samples and found that demographics in a fully representative sample are extremely important.
+While using representative panel-based sampling methods is considerably more expensive, there are generalizability concerns with online recruitment methods such as Prolific and Mechanical Turk. 
+People who participate in studies via online platforms are more technologically sophisticated and educated than the general population, and this bias may significantly impact the conclusions.
 
@@
 
-# Pilot Testing and Quality Assurance
+# Pilot Testing and Quality Assurance {#sec-pilot-test}
 
-Once you have your experiment designed and your charts to test, the next step is an extensive round of pilot testing. 
-The goal with pilot testing is to ensure that the experiment is set up precisely the way you want it, to ensure that your participants know what they are supposed to be doing and that any vague instructions are fixed, to provide time estimates for how long it will take to complete the task, and to provide you with some early data to examine for red flags.
-Our studies usually go through 2-3 rounds of pilot testing, with at least one of those rounds involving any relatives we can find that are less computer literate and will invariably figure out how to crash the testing application, and another round involving any and all coworkers and friends who may be available for 10-15 minutes during the work day.
+Once the data generating model is set, the experiment is designed, and the charts have been developed, the next step is an extensive round of pilot testing. 
+The goal of pilot testing is to ensure that the experiment is set up properly and that no issues have been overlooked.
+Pilot testing also provides an opportunity to ensure that directions are clear, participants know what they are supposed to be doing, and to estimate task completion time (which determines how much participant will be paid on many online testing platforms).
+Our studies usually go through 2-3 rounds of pilot testing, with at least one of those rounds involving any relatives and friends who are less technologically savvy. 
+We also purposely include talented individuals who can accidentally crash any testing applications. 
+A final round of testing usually includes any and all coworkers and friends who may be available for 10-15 minutes during the work day. 
 Pilot study samples do not need to be representative of any particular population, though we would caution against using exclusively visualization researchers or statisticians in your pilot sample because some issues are less likely to show up with knowledgeable participants.
@@
 In these cases, we highly recommend saving all generated data to a database as well, so that it is possible to go back and examine exactly what happened and/or how things went wrong. 
-Hard drive space is historically cheap relative to almost any other cost in an experiment; saving all of the data is a sensible measure.
+Hard drive space is extremely cheap relative to almost any other cost in an experiment; saving all of the data is a sensible measure.
 
@@
 As data collection online can also happen extremely quickly (300 participants in <2h in our most recent Prolific experiment), this can provide an illusion of control over the deluge of data, allowing any issues to be spotted and resolved relatively quickly.
-If server load is a potential issue, it may also help to release batches of trials over a longer period of time in order to minimize the chance of having to make participants wait for others to complete the task before the server can handle additional connections^[This is the one major drawback to our preferred solution of self-hosting a Shiny server to handle data collection: the free version of Shiny server is limited to about 15 connections at any given time.].
+If server load is a potential issue, it may also help to release batches of trials over a longer period of time in order to minimize the chance of having to make participants wait for others to complete the task before the server can handle additional connections^[This is the one major drawback to our preferred solution of self-hosting a Shiny server to handle data collection: the free version of Shiny server is limited to about 15 connections at any given time.
+Prolific has recently added rate-limiting functions to the experimental control  platform, which makes controlling the number of active jobs much easier.].
 
-# Analyzing the Data
+# Analyzing the Data {#sec-analysis}
 
@@
 The see-value approach [@niladriroychowdhurySeeValueApp2020;@vanderplasStatisticalSignificanceCalculations2021;@majumderValidationVisualStatistical2013] is extremely useful for single lineups, but when a series of lineups that are part of a designed experiment are used, generalized linear mixed models are a much simpler way to summarize the overall effect of various manipulated factors [@power;@clusters;@emily-diss].
-This approach also works extremely well for psychometric experiments [@sineillusion], as psychometric models can be easily fit into the framework of a generalized linear mixed model.
+This approach also works extremely well for psychometric experiments [@sineillusion], as psychometric models can be easily fit into the framework of a generalized linear mixed model [@juOneModelThat2024] that has higher power than the Rasch models which were historically recommended for these experiments.
 In addition, similar model structures can be used to model accuracy, response time, and confidence, if all three types of information are collected from participants during the experiment.
@@
 ) %>%
-  expand_grid(data set = c("data set1", "data set2"))
+  expand_grid(dataset = c("data set1", "data set2"))
 
@@
 ```{r density-plot, include = F, echo = F}
-densityPlot <- function(data, data setID, estimate, xlabel = "Estimate", x_limits = c(0,70000), zoom = F, scalesx = T, zoom_limits = c(NA,NA), gridlines = T, rugjitter = 0.25) {
+densityPlot <- function(data, datasetID, estimate, xlabel = "Estimate", x_limits = c(0,70000), zoom = F, scalesx = T, zoom_limits = c(NA,NA), gridlines = T, rugjitter = 0.25) {
 
   estPlot <- data %>%
-    filter(data set %in% data setID) %>%
+    filter(dataset %in% datasetID) %>%
     ggplot(aes_string(x = estimate, fill = "scale", color = "scale")) +
@@
   estPlot <- estPlot + 
-    geom_vline(data = grid_lines_data %>% filter(data set == "data set1", grid_lines >= x_limits[1], grid_lines <= x_limits[2]),
+    geom_vline(data = grid_lines_data %>% filter(dataset == "dataset1", grid_lines >= x_limits[1], grid_lines <= x_limits[2]),
                aes(xintercept = grid_lines, color = scale, linetype = "c"))
@@
 # DATA SET 1
-qe2_density_1 <- densityPlot(data = qe2_data, data setID = "data set1", "response", xlabel = "Estimated Year", x_limits = c(20,35), zoom = F, scalesx = T, gridlines = F, rugjitter = 0.25) +
+qe2_density_1 <- densityPlot(data = qe2_data, datasetID = "dataset1", "response", xlabel = "Estimated Year", x_limits = c(20,35), zoom = F, scalesx = T, gridlines = F, rugjitter = 0.25) +
   ggtitle("Estimation of X given a Y value for Exponential Data") + 

### 2024-09-16 index.qmd
Update format

@@
 format:  
-  pdf:
+  wordcount-pdf:
     keep-tex: true
@@
 
-# Experimental Design Considerations
+# Experimental Design Considerations {#sec-experimental-design}
 
@@
 
-While studies have found some relationship between lineup performance and demographic factors [@majumderHumanFactorsInfluencing2014;@vanderplasSpatialReasoningData2016a], these differences are relatively small when participants are recruited from online testing platforms like Amazon Mechanical Turk or Prolific, or when participants are recruited from a university student population.
+While studies have found some relationship between lineup performance and demographic factors [@vanderplasSpatialReasoningData2016a], these differences are relatively small when participants are recruited from online testing platforms like Amazon Mechanical Turk or Prolific, or when participants are recruited from a university student population.
 Studies have also not found a strong relationship between visual task performance and participant recruitment method [@vanderplasFramedReproducingRevisiting2019], perhaps in part because demographics which use social media sites and demographics completing online research tasks for pay overlap heavily.
@@
 
-\clearpage
+# References

### 2024-09-20 index.qmd
Emily changes/comments/fixes

@@
 While it is certainly possible to use equipment such as eye-tracking devices, this paper focuses primarily focus on studies which can be conducted without testing equipment, and primarily online, because these experiments have the lowest overhead, offer relatively fast data collection, and provide useful results that can be very clear if the experiment is designed correctly.
-The toolkit used for these experiments is R-based [@R], and includes ggplot2 [@ggplot2] and Shiny [@shiny] as primary components, as well as  JavaScript (including D3[@d3]) to customize the Shiny interface for many of the experiments.
+The toolkit used for these experiments is R-based [@R], and includes ggplot2 [@ggplot2] and Shiny [@shiny] as primary components, as well as  JavaScript (including D3 [@d3]) to customize the Shiny interface for many of the experiments.
 While we prefer this set of tools, most of the observations described in this paper apply to a wide variety of different workflows for graphical experimentation.
@@
 Statistical lineups [@bujaStatisticalInferenceExploratory2009;@wickhamGraphicalInferenceInfovis2010] are another useful testing tool for perceptual questions such as "which chart displays this data more clearly" [@power] while simultaneously assessing the significance of the graphical finding in the chart.
-As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot at a higher rate than the 19 null plots, then the target plot is said to be visually significant [@loyDiagnosticToolsHierarchical2013;@majumderValidationVisualStatistical2013;vanderplasStatisticalSignificanceCalculations2021] and a "see" value (the visual analogue of a $p$-value for a statistical test) can be calculated [@niladriroychowdhurySeeValueApp2020].
+As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot at a higher rate than the 19 null plots, then the target plot is said to be visually significant [@loyDiagnosticToolsHierarchical2013;@majumderValidationVisualStatistical2013;@vanderplasStatisticalSignificanceCalculations2021] and a "see" value (the visual analogue of a $p$-value for a statistical test) can be calculated [@niladriroychowdhurySeeValueApp2020].
 <!-- Beyond the quantitative assessment of the signal with $p$- and see-values, the human involvement also allows us to collect information on the qualitative nature of *what* a viewer sees.--> 
@@
 
-# Experimental Design Considerations {#sec-experimental-design}
+# Experimental Design Considerations {#sec-exp-design}
 
@@
 It was only after the experiment was completed that we realized that different countries introduce logarithms as a concept at different points during primary and secondary education; it might be that individuals in some countries have much more experience with log scales than those educated in the United States.
-We mention this only to point out that while every experiment contains a few missed opportunities, it is worth giving careful though to the demographic questions asked of participants and what information may be helpful during the analysis stage.
+We mention this only to point out that while every experiment contains a few missed opportunities, it is worth giving careful thought to the demographic questions asked of participants and what information may be helpful during the analysis stage.
 
@@
 
-**Generalized Linear Mixed Models** 
+## Generalized Linear Mixed Models
 The see-value approach [@niladriroychowdhurySeeValueApp2020;@vanderplasStatisticalSignificanceCalculations2021;@majumderValidationVisualStatistical2013] is extremely useful for single lineups, but when a series of lineups that are part of a designed experiment are used, generalized linear mixed models are a much simpler way to summarize the overall effect of various manipulated factors [@power;@clusters;@emily-diss].
@@
 
-**Numerical Estimation** There are additional considerations that should be expected when asking participants to estimate numerical quantities. 
+## Numerical Estimation
+There are additional considerations that should be expected when asking participants to estimate numerical quantities. 
 Anchoring and rounding cause participant responses to cluster in ways that can bias statistical estimators, requiring methods designed for these types of data [@ushakovStatisticalAnalysisRounded2017;@tourangeau_rips_rasinski_2000;@heitjanIgnorabilityCoarseData1991].
@@
 
-**Direct Interactions** If participants are making predictions and/or fitting visual statistics, we have had success analyzing these responses by comparing the responses to results from a statistical model to determine how visual statistics differ from the numerical quantities derived mathematically.
+## Direct Interactions
+If participants are making predictions and/or fitting visual statistics, we have had success analyzing these responses by comparing the responses to results from a statistical model to determine how visual statistics differ from the numerical quantities derived mathematically.
 For instance, we calculate the deviation between participant responses and the linear regression in `You Draw It' experiments, then fitted generalized additive mixed models to summarize the results across different experimental conditions to assess how user-drawn predictions deviated from the statistical estimates.
@@
 
-**Qualitative Responses** In many cases, it is helpful to combine participants' qualitative reasoning with their quantitative responses to designed graphical experiments. 
+## Qualitative Responses
+In many cases, it is helpful to combine participants' qualitative reasoning with their quantitative responses to designed graphical experiments. 
 This approach provides useful context as to what participants use to make their decisions, and can be useful when assessing why unexpected responses occurred. 

### 2024-09-20 index.qmd
just suggestions

@@
 
-Charts and data visualizations are primary methods of communicating scientific information to the public in what creators hope is an easy-to-digest, visually attractive package.
+\hh{How do you call the tested audience? viewer: 5  user: 32  mostly in the user engagement section, participant: 50+ times.
+I'm not sure that I love the word 'user', but I can see how it is more active than participant. }
+
+Charts and data visualizations are primary methods of communicating scientific information to the public in what creators hope is an easy-to-digest, visually attractive package.\hh{form?
+method?
+or am I overthinking packages?}
 There are many strategies for creating charts and graphs, from Tufte-esque minimalism [@tufte] to charts designed with extra imagery and aesthetic appeal that draw the viewer's attention and persist in memory [@cairoFunctionalArtIntroduction2012].
@@
 While there are heuristics, rules-of-thumb, and various guidelines [@asme-standards-graphics;@allenVisualizingScientificData2016;@haemerDoubleScalesAre1948;@fewInformationDashboardDesign2006;@kosslynGraphDesignEye2006] for creating useful and visually attractive charts, the best way to establish the efficacy of various design decisions is to test charts: evaluate them under controlled conditions, and gather appropriate data [@clevelandShapeParameterTwoVariable1988;@clevelandGraphicalPerceptionGraphical1985] to compare different representations of the same data.
-It can be extremely challenging to design studies which strike the right combination of internal and external validity. 
+It is extremely challenging to design studies which strike the right combination of internal and external validity. 
 Simply asking people to read quantities off of a graph may not generalize beyond the questions asked or the data used in the chart [@eellsRelativeMeritsCircles1926;@croxtonBarChartsCircle1927;@vonhuhnFurtherStudiesGraphic1927;@croxtonGraphicComparisonsBars1932], but designing a study that is sufficiently robust to those issues requires manipulation of so many factors as to be functionally close to impossible to control or manipulate in a single experiment.
@@
 @sec-testing-methods discusses different methods for testing graphics, and which methods best address different levels of user engagement. 
-In @sec-model-dev, we discuss the process of developing a data-generating model that can be used to control the statistical features of data used in the tested visualizations. 
+In @sec-model-dev, we discuss the process of developing a data-generating model used to control the statistical features of data in the tested visualizations. 
 Model development is a nuanced and iterative process that ultimately determines the success and generalizability of the experimental results. 
@@
 There are many different testing methods used to empirically assess graphics. 
-While it is certainly possible to use equipment such as eye-tracking devices, this paper focuses primarily focus on studies which can be conducted without testing equipment, and primarily online, because these experiments have the lowest overhead, offer relatively fast data collection, and provide useful results that can be very clear if the experiment is designed correctly.
+While it is certainly possible to use equipment such as eye-tracking devices, this paper focuses primarily focus on studies conducted without testing equipment, and primarily online, because these experiments have the lowest overhead, offer relatively fast data collection, and provide useful results \hh{for well-designed experiments.}
 The toolkit used for these experiments is R-based [@R], and includes ggplot2 [@ggplot2] and Shiny [@shiny] as primary components, as well as  JavaScript (including D3 [@d3]) to customize the Shiny interface for many of the experiments.
-While we prefer this set of tools, most of the observations described in this paper apply to a wide variety of different workflows for graphical experimentation.
+While we prefer this set of tools, most of the observations described \hh{here} apply to a wide variety of different workflows for graphical experimentation.
 
@@
 For instance, testing whether someone can detect an effect such as a trend line in noisy data is a *perceptual* question best addressed using methods which allow the user to interact with the data on a basic visual level.
-Sometimes, it may be possible to set up a scenario where the user adjusts the plot directly, as in an experiment testing the strength of the sine illusion experiment [@sineillusion]: users adjusted the strength of the illusion correction until the lines appeared to be the same length, providing a direct measure of the magnitude of the sine illusion's effect, as in @fig-sine-illusion.
-In other situations, it may be possible to have the user directly interact with the plot, as in @vanderplasEscapingFlatlandGraphics2024, where users could rotate and interact with a 3D rendered bar chart; the application then can record these user interactions to provide insight into the visual comparisons the user may have been performing to supplement the estimates which the user explicitly provided.
+Sometimes, it is possible to set up a scenario where the user adjusts the plot directly, as in an experiment testing the strength of the sine illusion experiment [@sineillusion]: users adjusted the strength of the illusion correction until the lines appeared to be the same length, providing a direct measure of the magnitude of the sine illusion's effect, as in @fig-sine-illusion.
+In other situations, it may be possible to have the user directly interact with the plot, as in @vanderplasEscapingFlatlandGraphics2024, where  \hh{ participants were asked to}
+rotate and interact with a 3D rendered bar chart; the application then records \hh{all relevant} user interactions to provide insight into the visual comparisons the user may have been performing \hh{as a} supplement to the explicitly provided estimates.
 
@@
 Statistical lineups [@bujaStatisticalInferenceExploratory2009;@wickhamGraphicalInferenceInfovis2010] are another useful testing tool for perceptual questions such as "which chart displays this data more clearly" [@power] while simultaneously assessing the significance of the graphical finding in the chart.
-As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot at a higher rate than the 19 null plots, then the target plot is said to be visually significant [@loyDiagnosticToolsHierarchical2013;@majumderValidationVisualStatistical2013;@vanderplasStatisticalSignificanceCalculations2021] and a "see" value (the visual analogue of a $p$-value for a statistical test) can be calculated [@niladriroychowdhurySeeValueApp2020].
+As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise.
+If viewers consistently pick the target plot at a higher rate than \hh{any of} the 19 null plots, the target plot is said to be visually significant [@loyDiagnosticToolsHierarchical2013;@majumderValidationVisualStatistical2013] and a "see" value (the visual analogue of a $p$-value for a statistical test) can be calculated [@niladriroychowdhurySeeValueApp2020; @vanderplasStatisticalSignificanceCalculations2021].
 <!-- Beyond the quantitative assessment of the signal with $p$- and see-values, the human involvement also allows us to collect information on the qualitative nature of *what* a viewer sees.--> 
@@
 In some experiments, we have provided participants with supportive tools, such as "scratch pads" and calculators built into the Shiny application to support the complex calculations required to answer higher-level numerical estimation questions (@fig-estimation-calc).
-These supportive tools reduce participant cognitive load, but also record a wealth of messy information that provide real insight into how participants were looking at the data, what strategies they tried and discarded for reading the chart, and what visual estimation methods were used.
+These supportive tools \hh{XXX In order to be supportive, the tools must be easy to use - otherwise they just add an addititional burden and are a source of frustration.} reduce participant cognitive load, but also record a wealth of messy information that provide real insight into how participants were looking at the data, what strategies they tried and discarded for reading the chart, and what visual estimation methods were used.
 While systematic analysis and modeling of this messy data is difficult, the insights provided can be extremely useful.
@@
 This is one area where lineup methods are convenient - they do not depend on participants to understand the nuances of language or scenarios built around the chart under investigation.
-However, in some situations it may be sufficient to ask participants to estimate direct numerical quantities that have little contextual information, as we did in [@vanderplasFramedReproducingRevisiting2019] when assessing the accuracy of framed plots re-created from the Statistical Atlas.
+However, in some situations it may be sufficient to ask participants to estimate direct numerical quantities that have little contextual information, as done in [@vanderplasFramedReproducingRevisiting2019] when assessing the accuracy of framed plots re-created from the Statistical Atlas.
 
 Another useful measurement strategy is to require participants to *engage directly* with an interactive visualization. 
-This can be useful in a directed task, where users are asked to interact with the chart in a specific way and the result is recorded, but it is also possible to use interactive visualizations in an open-ended task, recording how users engage with the graphic in an exploratory (as opposed to goal-directed) manner.
+This is useful in a directed task, where users are asked to interact with the chart in a specific way and the result is recorded, but it is also possible to use interactive visualizations in an open-ended task, recording how users engage with the graphic in an exploratory (as opposed to goal-directed) manner.
 In one recent experiment, we asked participants to forecast an exponential trend, with data presented on either a linear or log scale. 
@@
 
-Visual assessment is key, but narrowing the parameter space down to something that can be statistically determined is useful for providing a sort of numerical summary estimate of potential visual difficulty.
+\hh{The choice of the tested space is crucial to gain insight from a study without putting too much burden on participants with overlong studies.
+Choosing an appropriate space for testing parameters is a well-known problem in psychometric testing [@shah]: the space considered should cover the area between 'only some activation' to 'almost full activation' of an appropriate psycho-metric function.
+}
+\hh{When testing charts, }
+visual assessment is \hh{obviously} key, but \hh{we can make use of statistical indices related to the testing condition to} narrow the parameter space down.
+\hh{These indices then serve as quantitative proxies} of the visual difficulty.
+
 If using model based sampling methods, it is important to iteratively assess the sampling procedure through simulation: Start with a sampling method and define a parameter space that is appropriate for the model, and then simulate many data sets from each possible parameter combinations using a grid search.
@@
 
-Once an appropriate set of parameters are identified using the numerical screening method, then it is important to calibrate these parameter selections visually. 
+Once an appropriate set of parameters are identified using the numerical screening method,  it is important to calibrate these parameter selections visually. 
 No numerical statistic is a perfect measure of what we actually see: at best, they are approximations of what we might potentially see. 
@@
 The experiment was a full 2x3x3 factorial exploration of three data generating parameters, with 3 replicates at each parameter combination (54 data sets) and 10 aesthetic combinations (for a total of 540 lineups).
-Each lineup had 20 different sub-panels, so we should have carefully visually inspected some 10800 different panels. 
+Each lineup had 20 different sub-panels, so we should have carefully visually inspected some 10,800 different panels. 
 As is evident from the fact that we're telling this story as a cautionary tale, we missed a critical problem with our data-generating mechanism: when clusters were assigned to randomly generated data after the fact, we didn't control the cluster size, leading to clusters of one or two points in relatively few sub-panels.

### 2024-11-01 index.qmd
Accept Heike's changes ;)

@@
 
-\hh{How do you call the tested audience? viewer: 5  user: 32  mostly in the user engagement section, participant: 50+ times.
-I'm not sure that I love the word 'user', but I can see how it is more active than participant. }
+<!-- \hh{How do you call the tested audience? viewer: 5  user: 32  mostly in the user engagement section, participant: 50+ times. -->
+<!-- I'm not sure that I love the word 'user', but I can see how it is more active than participant. } -->
 
-Charts and data visualizations are primary methods of communicating scientific information to the public in what creators hope is an easy-to-digest, visually attractive package.\hh{form?
-method?
-or am I overthinking packages?}
+Charts and data visualizations are primary methods of communicating scientific information to the public in what creators hope is an easy-to-digest, visually attractive form.
 There are many strategies for creating charts and graphs, from Tufte-esque minimalism [@tufte] to charts designed with extra imagery and aesthetic appeal that draw the viewer's attention and persist in memory [@cairoFunctionalArtIntroduction2012].
@@
 There are many different testing methods used to empirically assess graphics. 
-While it is certainly possible to use equipment such as eye-tracking devices, this paper focuses primarily focus on studies conducted without testing equipment, and primarily online, because these experiments have the lowest overhead, offer relatively fast data collection, and provide useful results \hh{for well-designed experiments.}
+While it is certainly possible to use equipment such as eye-tracking devices, this paper focuses primarily focus on studies conducted without additional equipment, and primarily online, because these experiments have the lowest overhead, offer relatively fast data collection, and provide useful results, for well-designed experiments.
 The toolkit used for these experiments is R-based [@R], and includes ggplot2 [@ggplot2] and Shiny [@shiny] as primary components, as well as  JavaScript (including D3 [@d3]) to customize the Shiny interface for many of the experiments.
-While we prefer this set of tools, most of the observations described \hh{here} apply to a wide variety of different workflows for graphical experimentation.
+While we prefer this set of tools, most of the observations described here apply to a wide variety of different workflows for graphical experimentation.
 
@@
 Sometimes, it is possible to set up a scenario where the user adjusts the plot directly, as in an experiment testing the strength of the sine illusion experiment [@sineillusion]: users adjusted the strength of the illusion correction until the lines appeared to be the same length, providing a direct measure of the magnitude of the sine illusion's effect, as in @fig-sine-illusion.
-In other situations, it may be possible to have the user directly interact with the plot, as in @vanderplasEscapingFlatlandGraphics2024, where  \hh{ participants were asked to}
-rotate and interact with a 3D rendered bar chart; the application then records \hh{all relevant} user interactions to provide insight into the visual comparisons the user may have been performing \hh{as a} supplement to the explicitly provided estimates.
+In other situations, it may be possible to have the user directly interact with the plot, as in @vanderplasEscapingFlatlandGraphics2024, where  participants were asked to rotate and interact with a 3D rendered bar chart; the application then records all relevant user interactions to provide insight into the visual comparisons the user may have been performing as a supplement to the explicitly provided estimates.
 
-![Direct adjustment of a plot in a perceptual task](sine_illusion_screenshot.png){#fig-sine-illusion}
+![Direct adjustment of a plot in a perceptual task.](sine_illusion_screenshot.png){#fig-sine-illusion}
 
@@
 As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise.
-If viewers consistently pick the target plot at a higher rate than \hh{any of} the 19 null plots, the target plot is said to be visually significant [@loyDiagnosticToolsHierarchical2013;@majumderValidationVisualStatistical2013] and a "see" value (the visual analogue of a $p$-value for a statistical test) can be calculated [@niladriroychowdhurySeeValueApp2020; @vanderplasStatisticalSignificanceCalculations2021].
+If viewers consistently pick the target plot at a higher rate than any of the 19 null plots, the target plot is said to be visually significant [@loyDiagnosticToolsHierarchical2013;@majumderValidationVisualStatistical2013] and a "see" value (the visual analogue of a $p$-value for a statistical test) can be calculated [@niladriroychowdhurySeeValueApp2020; @vanderplasStatisticalSignificanceCalculations2021].
 <!-- Beyond the quantitative assessment of the signal with $p$- and see-values, the human involvement also allows us to collect information on the qualitative nature of *what* a viewer sees.--> 
@@
 In some experiments, we have provided participants with supportive tools, such as "scratch pads" and calculators built into the Shiny application to support the complex calculations required to answer higher-level numerical estimation questions (@fig-estimation-calc).
-These supportive tools \hh{XXX In order to be supportive, the tools must be easy to use - otherwise they just add an addititional burden and are a source of frustration.} reduce participant cognitive load, but also record a wealth of messy information that provide real insight into how participants were looking at the data, what strategies they tried and discarded for reading the chart, and what visual estimation methods were used.
-While systematic analysis and modeling of this messy data is difficult, the insights provided can be extremely useful.
+In order to be supportive, the tools must be easy to use, but assuming this bar is met, the tools can reduce participant cognitive load while  recording a wealth of information.
+This information provides real insight into how participants were looking at the data, what strategies they tried and discarded for reading the chart, and what visual estimation methods were used. 
+While systematic analysis and modeling of this messy data may be difficult, the insights provided can be extremely useful.
 
@@
 
-\hh{The choice of the tested space is crucial to gain insight from a study without putting too much burden on participants with overlong studies.
-Choosing an appropriate space for testing parameters is a well-known problem in psychometric testing [@shah]: the space considered should cover the area between 'only some activation' to 'almost full activation' of an appropriate psycho-metric function.
-}
-\hh{When testing charts, }
-visual assessment is \hh{obviously} key, but \hh{we can make use of statistical indices related to the testing condition to} narrow the parameter space down.
-\hh{These indices then serve as quantitative proxies} of the visual difficulty.
+The choice of the tested space is crucial to gain insight from a study without putting too much burden on participants with overlong studies.
+Choosing an appropriate space for testing parameters is a well-known problem in psychometric testing [@shah]: the space considered should cover the area between 'only some activation' to 'almost full activation' of an appropriate psychometric function.
+When testing charts, visual assessment is obviously key, but researchers can make use of statistical indices related to the testing condition to narrow the parameter space.  
+These indices then serve as quantitative proxies of visual difficulty.
 
-If using model based sampling methods, it is important to iteratively assess the sampling procedure through simulation: Start with a sampling method and define a parameter space that is appropriate for the model, and then simulate many data sets from each possible parameter combinations using a grid search.
+If using model based sampling methods, it is important to iteratively assess the sampling procedure through simulation.
+Start with a sampling method and define a parameter space that is appropriate for the model, and then simulate many data sets from each possible parameter combinations using a grid search.
 When sampling from a large data set, it is still important to assess the sample size and any potential stratification methods employed in order to ensure that the visual effect is represented.

### 2025-02-21 index.qmd
Some edits

@@
 
-Charts and data visualizations are primary methods of communicating scientific information to the public in what creators hope is an easy-to-digest, visually attractive form.
+Data visualizations are a critically important tool for communicating scientific information to the public in what creators hope is an easy-to-digest, visually attractive form.
 There are many strategies for creating charts and graphs, from Tufte-esque minimalism [@tufte] to charts designed with extra imagery and aesthetic appeal that draw the viewer's attention and persist in memory [@cairoFunctionalArtIntroduction2012].
@@
 There have been several attempts to list out all of the types of charts [@ribeccaSearchChartsData], create a taxonomy of charts [@bertin1983semiology;@desnoyersTaxonomyVisualsScience2011], and even to classify charts using a domain-specific grammar of graphics [@wilkinsonGrammarGraphics1999].
+One extremely useful reference is from Data to Viz (https://www.data-to-viz.com/), which uses a decision tree to show different visualizations compatible with the data; R, python, D3.js, and React code are provided to demonstrate how to create those visualizations.
 With all of the different design choices available, how are chart creators to know what is the best approach for communicating data to the appropriate audience?
 
-While there are heuristics, rules-of-thumb, and various guidelines [@asme-standards-graphics;@allenVisualizingScientificData2016;@haemerDoubleScalesAre1948;@fewInformationDashboardDesign2006;@kosslynGraphDesignEye2006] for creating useful and visually attractive charts, the best way to establish the efficacy of various design decisions is to test charts: evaluate them under controlled conditions, and gather appropriate data [@clevelandShapeParameterTwoVariable1988;@clevelandGraphicalPerceptionGraphical1985] to compare different representations of the same data.
-It is extremely challenging to design studies which strike the right combination of internal and external validity. 
-Simply asking people to read quantities off of a graph may not generalize beyond the questions asked or the data used in the chart [@eellsRelativeMeritsCircles1926;@croxtonBarChartsCircle1927;@vonhuhnFurtherStudiesGraphic1927;@croxtonGraphicComparisonsBars1932], but designing a study that is sufficiently robust to those issues requires manipulation of so many factors as to be functionally close to impossible to control or manipulate in a single experiment.
+While there are heuristics, rules-of-thumb, and guidelines [@asme-standards-graphics;@allenVisualizingScientificData2016;@haemerDoubleScalesAre1948;@fewInformationDashboardDesign2006;@kosslynGraphDesignEye2006] for creating useful and visually attractive data displays, the best way to establish the efficacy of various design decisions is to test the visualization on humans, evaluating different variants under controlled conditions [@clevelandShapeParameterTwoVariable1988;@clevelandGraphicalPerceptionGraphical1985].
+Empirical assessments of visualizations, when carefully designed, allow statisticians to determine which representation of the same data is most effective along one or more dimension(s) of interest: estimation or prediction accuracy, within or between group comparisons, response time, and ability to draw real-world decisions are common goals for charts.
+
+It is extremely challenging to design studies which strike the right balance between experimental control (i.e. internal validity) and generalizability to a wider context (i.e. external validity). 
+Simply asking people to read quantities off of a graph may not generalize beyond the questions asked or the data used in the chart [@eellsRelativeMeritsCircles1926;@croxtonBarChartsCircle1927;@vonhuhnFurtherStudiesGraphic1927;@croxtonGraphicComparisonsBars1932], but designing a study that is sufficiently robust to those issues requires manipulation or control of so many factors that the amount of participants and trials quickly becomes daunting or unaffordable.
 In this paper, we attempt to distill the experience gained from conducting several different types of graphics experiments [@clusters;@sineillusion;@emily-diss;@power;@vanderplasFramedReproducingRevisiting2019;@vanderplasEscapingFlatlandGraphics2024], discussing the use of different testing methods [@vanderplasTestingStatisticalCharts2020], the process of designing a graphical experiment, and analysis of the resulting empirical data.
@@
 
-There are many different testing methods used to empirically assess graphics. 
-While it is certainly possible to use equipment such as eye-tracking devices, this paper focuses primarily focus on studies conducted without additional equipment, and primarily online, because these experiments have the lowest overhead, offer relatively fast data collection, and provide useful results, for well-designed experiments.
+There are many different testing methods used to empirically assess statistical graphics. 
+While it is certainly possible to use equipment such as eye-tracking devices, this paper focuses primarily focus on studies conducted online  without additional equipment.
+These experiments have the lowest overhead, offer relatively fast data collection, and provide useful results for well-designed experiments.
 The toolkit used for these experiments is R-based [@R], and includes ggplot2 [@ggplot2] and Shiny [@shiny] as primary components, as well as  JavaScript (including D3 [@d3]) to customize the Shiny interface for many of the experiments.
@@
 
-One primary concern when testing graphics is the level of user engagement required. 
-For instance, testing whether someone can detect an effect such as a trend line in noisy data is a *perceptual* question best addressed using methods which allow the user to interact with the data on a basic visual level.
-Sometimes, it is possible to set up a scenario where the user adjusts the plot directly, as in an experiment testing the strength of the sine illusion experiment [@sineillusion]: users adjusted the strength of the illusion correction until the lines appeared to be the same length, providing a direct measure of the magnitude of the sine illusion's effect, as in @fig-sine-illusion.
-In other situations, it may be possible to have the user directly interact with the plot, as in @vanderplasEscapingFlatlandGraphics2024, where  participants were asked to rotate and interact with a 3D rendered bar chart; the application then records all relevant user interactions to provide insight into the visual comparisons the user may have been performing as a supplement to the explicitly provided estimates.
+It is important to consider the level of user engagement which is necessary to complete a particular visual or graphical task. 
+For instance, testing whether someone can detect an effect such as a linear trend in noisy data is a perceptual question. 
+Perceptual questions are often examined experimentally using methods which allow the user to interact with the data on a basic visual level: users are presented with a visual stimulus and answer yes/no questions to indicate whether the effect is detected.
+Numerical estimation is another common task when testing graphics: in these experiments, the participant views a chart, estimates the requested numerical quantity, and enters the estimate into the application through a numerical input, slider, or other form element.
+Sometimes, it is possible to set up a scenario where the user adjusts the plot using a set of controls designed to provide a fixed set of interactive operations. 
+This type of user engagement was used to assess the strength of the sine illusion [@sineillusion]: users adjusted the strength of a transformation designed to correct the illusion until the lines appeared to be the same length, as shown in @fig-sine-illusion, providing a direct measure of the magnitude of the sine illusion's effect.
+In other situations, it may be preferable to have the user directly interact with the visual stimulus. 
+In @vanderplasEscapingFlatlandGraphics2024, participants were asked to rotate and interact with a 3D rendered bar chart; the application recorded user interactions and corresponding rotation matrices, providing insight into the visual comparisons the user may have been performing.
+This information was used as a supplement to the explicitly provided estimates, providing some contextual information as well as the ability to identify the level of participant engagement with the questions.
+When experiments are conducted as part of classroom experiential learning, it is sometimes helpful to be able to separate the low-effort participants from those who were fully intellectually engaged in the task.
+Interactive graphics provide another level of user engagement that can be much more open-ended. 
+With interactive graphics, researchers can ask participants to directly annotate plots, toggle aesthetics, and highlight groups and plot features.
+Careful implementation of the experiment application may allow for each of these interactions to be recorded and analyzed, producing a rich, if messy, set of data that may allow researchers to tease apart visual estimation error from common shortcuts such as rounding used during direct numerical estimation.
 
-![Direct adjustment of a plot in a perceptual task.](sine_illusion_screenshot.png){#fig-sine-illusion}
+![Direct adjustment of a plot in a perceptual task.
+In this experiment, designed to assess the strength of the sine illusion, the user adjusts the plot using - and + buttons, which control the strength of a transformation designed to correct the effect of the sine illusion.
+When the user is satisfied that the lines are of equal length, they select the 'Finished' button to move to the next task.
+The experiment used a psychophysics experimental design, the method of adjustment, but leveraged the interactive Shiny interface to record the entire sequence of adjustments made by the user for each trial.
+A demo version of this application can be found at https://shiny.srvanderplas.com/sine-illusion/.
+](sine_illusion_screenshot.png){#fig-sine-illusion}
 
-Statistical lineups [@bujaStatisticalInferenceExploratory2009;@wickhamGraphicalInferenceInfovis2010] are another useful testing tool for perceptual questions such as "which chart displays this data more clearly" [@power] while simultaneously assessing the significance of the graphical finding in the chart.
-As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise.
-If viewers consistently pick the target plot at a higher rate than any of the 19 null plots, the target plot is said to be visually significant [@loyDiagnosticToolsHierarchical2013;@majumderValidationVisualStatistical2013] and a "see" value (the visual analogue of a $p$-value for a statistical test) can be calculated [@niladriroychowdhurySeeValueApp2020; @vanderplasStatisticalSignificanceCalculations2021].
+Visual inference [@bujaStatisticalInferenceExploratory2009;@wickhamGraphicalInferenceInfovis2010] is another useful testing tool for perceptual questions such as "which chart displays this data more clearly" [@power] while simultaneously assessing the significance of the graphical finding in the chart.
+Visual inference charts are often called "lineups" in analogy to the criminal procedure where the suspect is placed in a line with several other individuals with similar characteristics. 
+In a graphical lineup procedure, there is a target plot containing the real data, embedded in an array of innocent "null" plots generated through resampling or simulation. 
+If viewers consistently pick the target plot at a higher rate than any of the null plots, the target plot is said to be visually significant [@loyDiagnosticToolsHierarchical2013;@majumderValidationVisualStatistical2013] and a "see" value (the visual analogue of a $p$-value [@niladriroychowdhurySeeValueApp2020]) can be calculated using the `vinference` R package or the process described in @vanderplasStatisticalSignificanceCalculations2021.
+The details of this calculation are beyond the scope of this broader discussion of how to test charts, but more detail on visual inference is provided in <insert citation to visual inference WIRE article under development>.
 <!-- Beyond the quantitative assessment of the signal with $p$- and see-values, the human involvement also allows us to collect information on the qualitative nature of *what* a viewer sees.-->
-In another variation of the statistical lineup procedure, two models are compared, with target plots from each model embedded in the array of 20 total plots; the remaining null plots are constructed from a mixture model that blends the two competing models.
-This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data and assign a hierarchy of visual attributes [@clusters].
-Statistical lineups do not require  the experimenter to provide any contextual information for the data: all of the necessary information is embedded in the choice of a null statistical model, which is convenient for testing purposes, but does not allow for testing the viewer's understanding of the information shown in the chart.
+
+In another variation of the statistical lineup procedure, data generated from two models are compared, with target plots from each model embedded in the array of $K$ total plots. 
+The $K-2$ null plots are constructed from a mixture model that blends the two competing models [@clusters]. 
+Viewers are asked to select the panel(s) which are most different, and the primary source of information are trials in which viewers selected the target from one model but not the other, indicating that the display method used allowed viewers to differentiate one model's data (but not the other) from the nulls created through a mixture model.
+This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data [@clusters].
+
+One advantage of the visual inference technique is that the experimenter can ask a very general question, such as "which of these plots is the most different?", rather than a specific question about the displayed data which may require more quantitative sophistication.
+All of the necessary information to make the decision is embedded in the choice of the model used to generate the null plots. 
+This feature is extremely convenient when conducting the experiment and even allows small children to complete the task. 
+The downside is that as a result, visual  inference experiments do not allow experimenters to assess the viewer's understanding of the information shown in the chart. 
+In most cases, visual inference experiments remove any contextual information from the charts, including axis labels and values, plot titles, and so on, in order to encourage participants to make decisions based solely on the graphical presentation.
+This lack of context is a double-edged sword: visual inference can involve participants who do not have any mathematical training or instincts (including children), but researchers also cannot use this technique to assess higher levels of engagement with a chart, such as estimation, prediction, or reasoning based on displayed information.
 <!-- An example of a statistical lineup for testing the perception of exponential growth is shown in @fig-lineup-log. -->
@@
 In some studies, asking users to interpret a chart within a larger scenario can be effective, as in @fig-estimation-describe, while in others it is more helpful to ask users to explain answers. 
-In lineup studies, asking users why a specific panel was chosen can provide rich insight into otherwise confusing numerical results. 
-While we have not (to date) recorded users talking out loud about what they are seeing, think aloud methods could be implemented within a Shiny application, with audio recordings saved to the server for transcription and analysis
-[@dunbar1995scientists;@traftonTurningPicturesNumbers2000;@kirschenbaum2003comparative]; it is even possible that these recordings could be automatically transcribed using text-to-speech functions of large language models.
+In visual inference studies, asking users why a specific panel was chosen has been demonstrated to provide rich insight into otherwise confusing numerical results [@clusters].
+
+Think-aloud methods ask the viewer to narrate their internal thought process, either during or after completing a task [@thinkaloud]. 
+These recordings (or transcripts) can provide valuable insights into conscious cognition, and are often used when conducting usability studies. 
+While we have not (to date) recorded users talking out loud about what they are seeing, think aloud methods could easily be implemented within a Shiny application, with audio recordings saved to the server for transcription and analysis
+[@dunbar1995scientists; @traftonTurningPicturesNumbers2000; @kirschenbaum2003comparative]; it is even possible that these recordings could be automatically transcribed using the speech-to-text functions of large language models.
 
@@
 We have found it to be useful to have one experimenter calibrate the model parameters at a gross level, and then have another experimenter narrow in on the parameters which are visually reasonable within the selected range.
-Then, both examiners visually inspect a large number of plots generated using those parameters to get a sense for how difficult the task at hand is (this strategy is also described by Lu et al.
-[@luModelingJustNoticeable2022]).
+Then, both examiners visually inspect a large number of plots generated using those parameters to get a sense for how difficult the task at hand is (this strategy is also described by @luModelingJustNoticeable2022).
 At some point, all experimenters become so visually saturated with the nuances of the data generating mechanism that it may become necessary to "sanity check" the protocol with family members, friends, and colleagues.
@@
 
-One reason why it is so difficult to generate sampling models for visual explorations is that our visual system is very, very good at finding differences and unexpected results. 
-We re-ran the experiment using a different clustering method, and found that instead of noticing the number of ellipses, when that variable was removed participants instead cued in on the differences in size and shape of the ellipses formed using k-means clustering after the data generating procedure.
+One reason why it is so difficult to generate sampling models for visual explorations is that our visual system is optimized for identifying differences between groups.
+This ability can interfere with the natural  to use the null sampling models that might be used in equivalent numerical tests when running experiments that use visualizations. 
+We re-ran the experiment using a different clustering method that controlled the number of points in each group. 
+Instead of noticing the number of ellipses, participants instead used the differences in size and shape of the ellipses formed when clustering after the data generating procedure. 
 That is, participants could still detect the artificial nature of the induced clusters using other features. 
@@
 Anchoring and rounding cause participant responses to cluster in ways that can bias statistical estimators, requiring methods designed for these types of data [@ushakovStatisticalAnalysisRounded2017;@tourangeau_rips_rasinski_2000;@heitjanIgnorabilityCoarseData1991].
-An alternative approach is to analyze the data graphically using e.g.
-a combination of density and rug geoms for point estimates (as shown in @fig-density-rug), or spaghetti plots for individual response curves, with annotations to show anchor points.
+An alternative approach is to analyze the data graphically, as shown in @fig-density-rug, which uses a density plot with rug annotations for point estimates.
+When individual response curves are generated, a spaghetti plot with annotations to show anchor points may be useful. 
 The advantage of this approach is that it can accommodate extremely messy data without requiring the extensive data cleaning and elimination of nonsensical responses that might be necessary to fit a statistical model.

### 2025-03-14 index.qmd
Review modifications

@@
 For a specific type of data, there are also usually many different chart forms to display that data: for instance, if we have a set of categorical data and we wish to show the relative proportions of each category, we could do so using a stacked bar chart or the polar equivalent, a pie chart.
-There have been several attempts to list out all of the types of charts [@ribeccaSearchChartsData], create a taxonomy of charts [@bertin1983semiology;@desnoyersTaxonomyVisualsScience2011], and even to classify charts using a domain-specific grammar of graphics [@wilkinsonGrammarGraphics1999].
+There have been several attempts to list out all of the types of charts [@ribeccaSearchChartsData], create a taxonomy of charts [@bertin1983semiology;@desnoyersTaxonomyVisualsScience2011], and even to create charts using a domain-specific grammar of graphics [@wilkinsonGrammarGraphics1999] that is also useful for classification.
 One extremely useful reference is from Data to Viz (https://www.data-to-viz.com/), which uses a decision tree to show different visualizations compatible with the data; R, python, D3.js, and React code are provided to demonstrate how to create those visualizations.
@@
 There are many different testing methods used to empirically assess statistical graphics. 
-While it is certainly possible to use equipment such as eye-tracking devices, this paper focuses primarily focus on studies conducted online  without additional equipment.
-These experiments have the lowest overhead, offer relatively fast data collection, and provide useful results for well-designed experiments.
-The toolkit used for these experiments is R-based [@R], and includes ggplot2 [@ggplot2] and Shiny [@shiny] as primary components, as well as  JavaScript (including D3 [@d3]) to customize the Shiny interface for many of the experiments.
-While we prefer this set of tools, most of the observations described here apply to a wide variety of different workflows for graphical experimentation.
+This paper uses studies conducted online  without additional equipment as primary examples, though many of the same considerations apply to in-person experiments conducted using additional equipment, including 3D printed charts, eye-tracking equipment, and interactive data displays.
+Online experiments have lower overhead, offer relatively fast data collection, and provide useful results for well-designed experiments.
+The toolkit used for these experiments is R-based [@R], and includes ggplot2 [@ggplot2] and Shiny [@shiny] as primary components.
+In many experiments, we customized the Shiny interface with JavaScript and d3 [@d3], enabling interactive graphics, use of svgs, and other useful extensions.
+While we prefer this set of tools, most of the observations described here apply to a wide variety of different workflows for graphical experimentation, including in-person experiments.
 
@@
 
-Visual inference [@bujaStatisticalInferenceExploratory2009;@wickhamGraphicalInferenceInfovis2010] is another useful testing tool for perceptual questions such as "which chart displays this data more clearly" [@power] while simultaneously assessing the significance of the graphical finding in the chart.
+Visual inference [@bujaStatisticalInferenceExploratory2009;@wickhamGraphicalInferenceInfovis2010] is another useful testing tool for perceptual questions such as "which chart displays this data more clearly" [@power] while simultaneously assessing the statistical significance of the graphical finding in a chart.
 Visual inference charts are often called "lineups" in analogy to the criminal procedure where the suspect is placed in a line with several other individuals with similar characteristics. 
 In a graphical lineup procedure, there is a target plot containing the real data, embedded in an array of innocent "null" plots generated through resampling or simulation. 
-If viewers consistently pick the target plot at a higher rate than any of the null plots, the target plot is said to be visually significant [@loyDiagnosticToolsHierarchical2013;@majumderValidationVisualStatistical2013] and a "see" value (the visual analogue of a $p$-value [@niladriroychowdhurySeeValueApp2020]) can be calculated using the `vinference` R package or the process described in @vanderplasStatisticalSignificanceCalculations2021.
-The details of this calculation are beyond the scope of this broader discussion of how to test charts, but more detail on visual inference is provided in <insert citation to visual inference WIRE article under development>.
+If viewers consistently pick the target plot at a higher rate than any of the null plots, the target plot is said to be visually significant [@loyDiagnosticToolsHierarchical2013;@majumderValidationVisualStatistical2013] and a "see" value, the visual analogue of a $p$-value [@niladriroychowdhurySeeValueApp2020], can be calculated using the `vinference` R package or the process described in @vanderplasStatisticalSignificanceCalculations2021.
+The details of this calculation are beyond the scope of this broader discussion of how to test charts, but more detail on visual inference is provided in \<insert citation to visual inference WIRE article under development\>.
 <!-- Beyond the quantitative assessment of the signal with $p$- and see-values, the human involvement also allows us to collect information on the qualitative nature of *what* a viewer sees.-->
@@
 These recordings (or transcripts) can provide valuable insights into conscious cognition, and are often used when conducting usability studies. 
-While we have not (to date) recorded users talking out loud about what they are seeing, think aloud methods could easily be implemented within a Shiny application, with audio recordings saved to the server for transcription and analysis
-[@dunbar1995scientists; @traftonTurningPicturesNumbers2000; @kirschenbaum2003comparative]; it is even possible that these recordings could be automatically transcribed using the speech-to-text functions of large language models.
+While we have not to date recorded users talking out loud about what they are seeing during a study, think aloud methods could easily be implemented within a Shiny application, with audio recordings saved to the server for transcription and analysis
+[@dunbar1995scientists; @traftonTurningPicturesNumbers2000; @kirschenbaum2003comparative].
+It is even possible that these recordings could be automatically transcribed using speech-to-text models.
+We have used think-aloud methods informally during pilot studies to "harden" graphical experiments and verify the selection of parameters used in an experiment. 
+The success of this approach, combined with the few studies which used think-aloud to assess charts [@kulhavyCartographicExperienceThinking1992;@leeHowPeopleMake2016;@haiderStrategiesDetectingDifference2021], suggests that think-aloud methods are an often-overlooked but useful tool for assessing data visualizations.
 
@@
 Of course, in an online, asynchronous experiment, every user interaction with the testing materials (typically hosted on a web page) can also be recorded along with time stamps, mouse positions, browser size and screen resolution, and other information.
-While we have not used this type of information heavily in our experimental analyses thus far, in most experiments, we collect time stamp data in order to assess how long participants spend on each question.
+While we have not used this type of information heavily in our experimental analyses thus far, in most experiments we collect time stamp data in order to assess how long participants spend on each question.
 Typically, the first round of test questions takes the longest for participants to complete. 
@@
 This information provides real insight into how participants were looking at the data, what strategies they tried and discarded for reading the chart, and what visual estimation methods were used. 
-While systematic analysis and modeling of this messy data may be difficult, the insights provided can be extremely useful.
+While systematic analysis and modeling of this data may be difficult, as it is usually messy and often must be manually coded, the insights provided can be extremely useful. 
+However, unless participants are required to use these tools, it is difficult to gather comprehensive information - those participants that don't use supportive tools likely differ in meaningful ways from those who do.
+As a result, the information gathered from supportive tools likely does not generalize to the entire sample.
 
@@
 
-The choice of the tested space is crucial to gain insight from a study without putting too much burden on participants with overlong studies.
-Choosing an appropriate space for testing parameters is a well-known problem in psychometric testing [@shah]: the space considered should cover the area between 'only some activation' to 'almost full activation' of an appropriate psychometric function.
-When testing charts, visual assessment is obviously key, but researchers can make use of statistical indices related to the testing condition to narrow the parameter space.  
-These indices then serve as quantitative proxies of visual difficulty.
+The choice of the parameter space used in testing is crucial to gain insight from a study without putting too much burden on participants with overlong studies. 
+Choosing an appropriate space for testing parameters is a well-known problem in psychometric testing: the space considered should cover the area between 'only some activation' to 'almost full activation' of an appropriate psychometric function [@schuttPainfreeAccurateBayesian2016;@valentinDesigningOptimalBehavioral2024].
+When testing charts, visual assessment is obviously key, but researchers can make use of statistical indices related to the testing condition to narrow the parameter space to a reasonable and efficient subset from which maximal information can be acquired.
+These statistical indices may also serve as quantitative proxies for the difficulty of the visual task.
 
-If using model based sampling methods, it is important to iteratively assess the sampling procedure through simulation.
-Start with a sampling method and define a parameter space that is appropriate for the model, and then simulate many data sets from each possible parameter combinations using a grid search.
-When sampling from a large data set, it is still important to assess the sample size and any potential stratification methods employed in order to ensure that the visual effect is represented.
-
-In both cases, it can help to employ some general feature engineering - what numerical statistic provides an estimation of the basic concept which is being visually evaluated?
+To identify a statistical proxy for visual difficulty that may help with narrowing the parameter space, it can be useful to consider numerical measures used to estimate the same types of visual information that will be assessed in the experiment.
 For instance, we have used
@@
 
+While this method is certainly more critical for model-based sampling methods, it is also important when data are generated by sampling from a larger data set. 
+When sampling from a larger dataset, parameters are more often sample size and stratification methods, but it is still important to iteratively assess the data generating procedure through simulation.
+Using numerical proxies for visual characteristics of data displays such as curvature, linearity, scatter, dispersion can assist with identifying optimal parameter settings to use across different experimental conditions.
+Even with this strategy, it is still critical to fine-tune the parameter choices with visual calibration and pilot testing.
+
 ## Fine-Tuning Parameter Choices
@@
 We highly recommend performing several rounds of think-aloud pilot testing before deploying an experiment. 
-In support of this assessment, we offer up a cautionary tale of our own experience: that of [@clusters], where we designed an experiment to test which plot aesthetics promoted discovery of linear trends and/or clusters.
+In support of this assessment, we offer up a cautionary tale of our own experience: that of @clusters, where we designed an experiment to test which plot aesthetics promoted discovery of linear trends and/or clusters.
 
@@
 
-![A lineup from Vanderplas & Hofmann [@clusters].
+![A lineup from Vanderplas & Hofmann [-@clusters].
 Panel 10 shows the clustered target data and panel 17 shows the target data with a strong linear relationship; either of these target panels was the expected choice.
@@
 
+## Cognitive Load
+
 The primary human design factor to be aware of is that visual tasks and assessing statistical graphics can be extremely cognitively taxing. 
@@
 
+## Participant Instructions
+
 It can be extremely helpful to include "practice" demonstrations of the task to show the basic process, logic, and reasoning. 
@@
 In cases which require interactivity, gif animations of the task being carried out are useful, as are additional visual cues, such as the yellow box used in the `You Draw It' task[^1] to indicate that there were points which were not completed.
+
+## Accounting for Participant Variability
 
@@
 
+## Demographic Data
+
 It may be useful to ask a few more demographic questions about STEM education level for studies which ask more of participants from a mathematical standpoint; while lineup studies have not found strong associations with those variables, lineup studies also do not require participants to engage with the data presented in a chart in a way that requires higher-order mathematical reasoning.
-This has allowed us to make the argument to IRB that our research is exempt, as we do not collect enough demographic information to identify participants, however, this has occasionally come at some cost.
-Our recent study examining the use of log scales and exponential data was conducted using Prolific, which recruits participants from around the world; we required only that participants were fluent in English to participate.
+This has allowed us to make the argument to the ethics committee (institutional review board, or IRB) that our research is exempt, as we do not collect enough demographic information to identify participants, however, collecting reduced demographic information occasionally comes at a cost.
+One recent study examining the use of log scales and exponential data was conducted using Prolific, which recruits participants from around the world; we required only that participants were fluent in English to participate.
 It was only after the experiment was completed that we realized that different countries introduce logarithms as a concept at different points during primary and secondary education; it might be that individuals in some countries have much more experience with log scales than those educated in the United States.
 We mention this only to point out that while every experiment contains a few missed opportunities, it is worth giving careful thought to the demographic questions asked of participants and what information may be helpful during the analysis stage.
+
+## Trial Allocation to Participants
 
@@
 
-We have conducted visualization experiments using a wide variety of tools: custom web servers running experiments using PHP black magic, Qualtrics surveys for static graphics, and writing full Shiny applications that control every part of the experimental process from instructions to providing completion codes for payment.
+## Data Collection Platforms and Infrastructure
+
+We have conducted visualization experiments using a wide variety of tools: custom web PHP servers running interactive forms for data collection, generic web survey platforms (Qualtrics, Google Forms) for static graphics, and Shiny applications that control every part of the experimental process from instructions to providing completion codes for payment while rendering interactive graphics and generating fully randomized data for each participant.
 In our research, Shiny has provided the right balance between control over the experimental setting, procedure, etc.
 and the intricate details of web server management, however, this balance is likely different for every group and potentially for every experiment.
-Likewise, we have recruited participants using Amazon Mechanical Turk, Prolific, Reddit and other social media sites, and email; each recruitment method has trade-offs between cost, convenience, and control over demographic variables.
-@riceTestingPerceptualAccuracy2024 recruited participants using NORC panel samples and found that demographics in a fully representative sample are extremely important.
-While using representative panel-based sampling methods is considerably more expensive, there are generalizability concerns with online recruitment methods such as Prolific and Mechanical Turk. 
+Likewise, we have recruited participants using services such as [Amazon Mechanical Turk](https://www.mturk.com/) and [Prolific](https://www.prolific.com/), on social media sites ([Reddit](https://web.archive.org/web/20250228110244/https://www.reddit.com/r/SampleSize/), Twitter(X), Mastodon, and BlueSky), and by email; each recruitment method has trade-offs between cost, convenience, number of people in the sampling frame, and control over participant demographics.
+@riceTestingPerceptualAccuracy2024 recruited participants using nationally representative panel samples and found that conclusions from fully representative samples of the population can be very different from volunteer samples recruited using other services, as many people are unmotivated to engage with charts, don't know how to read charts, or impose pre-formed conclusions onto visual displays.
+These individuals are unlikely to participate in most ad-hoc volunteer-based sampling methods.
+While using representative panel sampling products offered by organizations such as NORC or Gallup is considerably more expensive than samples from Amazon MTurk or Prolific, these results suggest that results from sampling methods used online may not generalize to the broader population.
 People who participate in studies via online platforms are more technologically sophisticated and educated than the general population, and this bias may significantly impact the conclusions.
 
+If conducting an online experiment, we recommend consideration of panel-based sampling methods as well as online testing platforms. 
+While Amazon MTurk was once the platform of choice for this type of research [@heerCrowdsourcingGraphicalPerception2010], in 2025 Prolific, which is aimed at academic experiments, seems to provide better options that ensure that experiments are completed by people rather than AI than MTurk.
+It seems likely that the preferred platform for participant recruitment will continue to evolve over time, thus, we will recommend that experimenters consider available options, compare pricing structures (as these vary widely), and consider whether add-on fees for e.g.
+demographically representative samples are worth the additional cost.
+An additional consideration when choosing between panel samples and online recruitment platforms is the constraints placed on the types of questions and graphics which can be used. 
+In our experience, it is much easier to run an experiment using the online recruitment platforms, which place many fewer constraints on both the number and types of questions which can be asked and the types of stimuli which can be used in the experiment.
+
 [^1]: See a gif of testing with `You Draw It' [here](https://i.imgur.com/GM5YSen.gif)
+
+## Ethical Review of Human Subjects Experiments
+
+In almost any research environment, there are requirements that experiments involving human participants undergo review by an ethics board to ensure that participants are not harmed by participating in the study.
+We are most familiar with regulations in the United States and can only speak to that environment.
+Most graphics experiments conducted in the US fall under the "exempt" category of experiments which require only basic review and approval. 
+These experiments record only basic demographic information, ask participants to complete tasks that involve no risk, and experimenters record information which would not be embarrassing to participants if exposed.
+In order to ensure that our experiments fall into this category, we will often ensure that demographic information is collected and stored separately from any identifiable information, such as e.g.
+user IDs from the participant recruitment platform that allow us to monitor task completion and pay participants for their time.
+Researchers should also ensure that they comply with privacy laws such as the European [General Data Protection Regulation (GDPR)](https://gdpr-info.eu/chapter-3/); complying with this law while collecting data which is not identifiable can be complicated.
+We recommend consulting with your ethics board and institutional recommendations in order to maintain legal compliance and safeguard participant privacy.
 
@@
 The see-value approach [@niladriroychowdhurySeeValueApp2020;@vanderplasStatisticalSignificanceCalculations2021;@majumderValidationVisualStatistical2013] is extremely useful for single lineups, but when a series of lineups that are part of a designed experiment are used, generalized linear mixed models are a much simpler way to summarize the overall effect of various manipulated factors [@power;@clusters;@emily-diss].
-This approach also works extremely well for psychometric experiments [@sineillusion], as psychometric models can be easily fit into the framework of a generalized linear mixed model [@juOneModelThat2024] that has higher power than the Rasch models which were historically recommended for these experiments.
-In addition, similar model structures can be used to model accuracy, response time, and confidence, if all three types of information are collected from participants during the experiment.
+This approach also works extremely well for psychometric experiments [@sineillusion], as psychometric models can be easily fit into the framework of a generalized linear mixed model [@juOneModelThat2024] that has higher power than the Rasch models [@andrichRaschModelsMeasurement1988;@luModelingJustNoticeable2022] which were historically recommended for these experiments.
+In addition, similar model structures with different link functions can be used to model accuracy, response time, and confidence, if all three types of information are collected from participants during the experiment.
 
@@
 Anchoring and rounding cause participant responses to cluster in ways that can bias statistical estimators, requiring methods designed for these types of data [@ushakovStatisticalAnalysisRounded2017;@tourangeau_rips_rasinski_2000;@heitjanIgnorabilityCoarseData1991].
-An alternative approach is to analyze the data graphically, as shown in @fig-density-rug, which uses a density plot with rug annotations for point estimates.
-When individual response curves are generated, a spaghetti plot with annotations to show anchor points may be useful. 
-The advantage of this approach is that it can accommodate extremely messy data without requiring the extensive data cleaning and elimination of nonsensical responses that might be necessary to fit a statistical model.
+An alternative approach is to analyze the data graphically, as shown in @fig-density-rug, which uses a density plot with rug annotations to show individual participant point estimates. 
+Rounding effects can clearly be seen in the rug plot, but a kernel density calculated using an appropriately selected bandwidth shows clear visual differences between linear and log scale charts. 
+In addition, a smaller second mode can be seen in both linear and log conditions that corresponds to the underlying model value; this suggests that a minority of participants fit a mental regression model to the data and use that model to estimate rather than estimating based off of the closest point.
+When charts are carefully constructed to account for the experimental structure and participant estimation strategies, displaying both individual and aggregate responses, it is possible to demonstrate a measurable difference between conditions without the need for complicated statistical modeling which corrects for rounding and anchoring effects.
+
+While @fig-density-rug shows point estimates, the same approach can be modified to account for experimental methods which generate participant response curves. 
+In these cases, rather than adding a rug plot to a density plot showing aggregate estimates, it may be more helpful to create spaghetti plots of individual estimates with a superimposed consensus estimate and relevant annotations showing e.g.
+anchor points.
+The advantage of this approach is that it can accommodate extremely messy data without requiring the extensive data cleaning, modeling of heuristics like rounding and anchoring, and elimination of nonsensical responses that might be necessary to fit a statistical model.
+While statistical models are undoubtedly beneficial in many situations, we have often found that graphical displays of experimental results are at least as useful for analyzing and presenting the results of graphical experiments.
 
@@
 # Conclusion
+
 Testing features of visualization graphically using online platforms provides an incredibly powerful and efficient way to establish empirical guidelines for statistical graphics and visualization. 
@@
 
+In this paper, we have attempted to contextualize and motivate the logic behind the process we use to design empirical graphics experiments. 
+We've discussed model development, experimental design considerations, pilot testing, and data analysis methods that have been honed over many successful and less-than-optimal experiments. 
+While no experiment involving humans ever goes exactly to plan, following this process helps to avoid some of the most likely mishaps, ensuring that each new experiment's "bonus" findings have only minimal impacts on the study's overall utility.
+
+While it can be difficult to conduct empirical tests of different visualizations, the results of these experiments support guidelines for graphical design and communication of statistical findings in an accessible and explainable way.
+Many chart design guidelines and recommendations are based on heuristics, but as scientists, we should prefer guidelines which are based on empirical, experimentally derived results over opinions. 
+Testing statistical graphics and developing empirically supported guidelines for chart creation promises to support better scientific communication, which is critical for educating the public about topics like climate change, public health, the risk of severe weather and more.
+
 # References

### 2025-03-14 index.tex
Review modifications

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  10pt,
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
    \setmainfont[]{Helvetica}
    \setmonofont[]{Roboto}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{2}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\usepackage[dvipsnames]{xcolor} % colors
\newcommand{\ear}[1]{{\textcolor{blue}{#1}}}
\newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
\newcommand{\hh}[1]{{\textcolor{Green}{#1}}}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A Guide to Designing Experiments to Test Statistical Graphics},
  pdfauthor={Emily Robinson; Heike Hofmann; Susan Vanderplas},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{A Guide to Designing Experiments to Test Statistical Graphics}
\author{Emily Robinson \and Heike Hofmann \and Susan Vanderplas}
\date{2025-03-14}

\begin{document}
\maketitle
\begin{abstract}
In this paper, we discuss considerations and methods for experimentally
testing visualizations. We discuss levels of user engagement with
graphics, common issues when developing a sampling or data generation
model, the importance of pilot testing, and data analysis methods. Along
the way, we also provide recommendations of how to avoid some of the
unique pitfalls of human testing in statistical and visualization
research.
\end{abstract}

\section{Introduction}\label{introduction}

Data visualizations are a critically important tool for communicating
scientific information to the public in what creators hope is an
easy-to-digest, visually attractive form. There are many strategies for
creating charts and graphs, from Tufte-esque minimalism (Tufte, 1991) to
charts designed with extra imagery and aesthetic appeal that draw the
viewer's attention and persist in memory (Cairo, 2012). For a specific
type of data, there are also usually many different chart forms to
display that data: for instance, if we have a set of categorical data
and we wish to show the relative proportions of each category, we could
do so using a stacked bar chart or the polar equivalent, a pie chart.
There have been several attempts to list out all of the types of charts
(Ribecca, 2022), create a taxonomy of charts (Bertin \& Berg, 1983;
Desnoyers, 2011), and even to create charts using a domain-specific
grammar of graphics (Wilkinson, 1999) that is also useful for
classification. One extremely useful reference is from Data to Viz
(https://www.data-to-viz.com/), which uses a decision tree to show
different visualizations compatible with the data; R, python, D3.js, and
React code are provided to demonstrate how to create those
visualizations. With all of the different design choices available, how
are chart creators to know what is the best approach for communicating
data to the appropriate audience?

While there are heuristics, rules-of-thumb, and guidelines (Allen \&
Erhardt, 2016; Few, 2006; Haemer, 1948; {``Joint Committee on Standards
for Graphic Presentation,''} 1915; Kosslyn, 2006) for creating useful
and visually attractive data displays, the best way to establish the
efficacy of various design decisions is to test the visualization on
humans, evaluating different variants under controlled conditions
(Cleveland et al., 1988; Cleveland \& McGill, 1985). Empirical
assessments of visualizations, when carefully designed, allow
statisticians to determine which representation of the same data is most
effective along one or more dimension(s) of interest: estimation or
prediction accuracy, within or between group comparisons, response time,
and ability to draw real-world decisions are common goals for charts.

It is extremely challenging to design studies which strike the right
balance between experimental control (i.e.~internal validity) and
generalizability to a wider context (i.e.~external validity). Simply
asking people to read quantities off of a graph may not generalize
beyond the questions asked or the data used in the chart (Croxton, 1932;
Croxton \& Stryker, 1927; Eells, 1926; Huhn, 1927), but designing a
study that is sufficiently robust to those issues requires manipulation
or control of so many factors that the amount of participants and trials
quickly becomes daunting or unaffordable. In this paper, we attempt to
distill the experience gained from conducting several different types of
graphics experiments (Hofmann et al., 2012; Robinson, 2022; VanderPlas
et al., 2019; Vanderplas et al., 2024; VanderPlas \& Hofmann, 2015,
2017), discussing the use of different testing methods (Vanderplas et
al., 2020), the process of designing a graphical experiment, and
analysis of the resulting empirical data. It is our hope that this paper
will lower the barriers that exist for conducting empirical graphics
research and reduce the probability of costly mistakes.

Section~\ref{sec-testing-methods} discusses different methods for
testing graphics, and which methods best address different levels of
user engagement. In Section~\ref{sec-model-dev}, we discuss the process
of developing a data-generating model used to control the statistical
features of data in the tested visualizations. Model development is a
nuanced and iterative process that ultimately determines the success and
generalizability of the experimental results. We explore different
experimental design considerations in Section~\ref{sec-exp-design} and
then move to the importance of pilot testing in
Section~\ref{sec-pilot-test}. Finally, we provide some common analysis
strategies in Section~\ref{sec-analysis}, including strategies for
handling the unexpected data features which are so common in graphical
testing experiments.

\section{Testing Methods and User Engagement}\label{sec-testing-methods}

There are many different testing methods used to empirically assess
statistical graphics. This paper uses studies conducted online without
additional equipment as primary examples, though many of the same
considerations apply to in-person experiments conducted using additional
equipment, including 3D printed charts, eye-tracking equipment, and
interactive data displays. Online experiments have lower overhead, offer
relatively fast data collection, and provide useful results for
well-designed experiments. The toolkit used for these experiments is
R-based (R Core Team, 2022), and includes ggplot2 (Wickham, 2016) and
Shiny (Chang et al., 2021) as primary components. In many experiments,
we customized the Shiny interface with JavaScript and d3 (Bostock et
al., 2011), enabling interactive graphics, use of svgs, and other useful
extensions. While we prefer this set of tools, most of the observations
described here apply to a wide variety of different workflows for
graphical experimentation, including in-person experiments.

It is important to consider the level of user engagement which is
necessary to complete a particular visual or graphical task. For
instance, testing whether someone can detect an effect such as a linear
trend in noisy data is a perceptual question. Perceptual questions are
often examined experimentally using methods which allow the user to
interact with the data on a basic visual level: users are presented with
a visual stimulus and answer yes/no questions to indicate whether the
effect is detected. Numerical estimation is another common task when
testing graphics: in these experiments, the participant views a chart,
estimates the requested numerical quantity, and enters the estimate into
the application through a numerical input, slider, or other form
element. Sometimes, it is possible to set up a scenario where the user
adjusts the plot using a set of controls designed to provide a fixed set
of interactive operations. This type of user engagement was used to
assess the strength of the sine illusion (VanderPlas \& Hofmann, 2015):
users adjusted the strength of a transformation designed to correct the
illusion until the lines appeared to be the same length, as shown in
Figure~\ref{fig-sine-illusion}, providing a direct measure of the
magnitude of the sine illusion's effect. In other situations, it may be
preferable to have the user directly interact with the visual stimulus.
In Vanderplas et al. (2024), participants were asked to rotate and
interact with a 3D rendered bar chart; the application recorded user
interactions and corresponding rotation matrices, providing insight into
the visual comparisons the user may have been performing. This
information was used as a supplement to the explicitly provided
estimates, providing some contextual information as well as the ability
to identify the level of participant engagement with the questions. When
experiments are conducted as part of classroom experiential learning, it
is sometimes helpful to be able to separate the low-effort participants
from those who were fully intellectually engaged in the task.
Interactive graphics provide another level of user engagement that can
be much more open-ended. With interactive graphics, researchers can ask
participants to directly annotate plots, toggle aesthetics, and
highlight groups and plot features. Careful implementation of the
experiment application may allow for each of these interactions to be
recorded and analyzed, producing a rich, if messy, set of data that may
allow researchers to tease apart visual estimation error from common
shortcuts such as rounding used during direct numerical estimation.

\begin{figure}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{sine_illusion_screenshot.png}}

}

\caption{\label{fig-sine-illusion}Direct adjustment of a plot in a
perceptual task. In this experiment, designed to assess the strength of
the sine illusion, the user adjusts the plot using - and + buttons,
which control the strength of a transformation designed to correct the
effect of the sine illusion. When the user is satisfied that the lines
are of equal length, they select the `Finished' button to move to the
next task. The experiment used a psychophysics experimental design, the
method of adjustment, but leveraged the interactive Shiny interface to
record the entire sequence of adjustments made by the user for each
trial. A demo version of this application can be found at
https://shiny.srvanderplas.com/sine-illusion/.}

\end{figure}%

Visual inference (Buja et al., 2009; Wickham et al., 2010) is another
useful testing tool for perceptual questions such as ``which chart
displays this data more clearly'' (Hofmann et al., 2012) while
simultaneously assessing the statistical significance of the graphical
finding in a chart. Visual inference charts are often called ``lineups''
in analogy to the criminal procedure where the suspect is placed in a
line with several other individuals with similar characteristics. In a
graphical lineup procedure, there is a target plot containing the real
data, embedded in an array of innocent ``null'' plots generated through
resampling or simulation. If viewers consistently pick the target plot
at a higher rate than any of the null plots, the target plot is said to
be visually significant (Loy \& Hofmann, 2013; Majumder et al., 2013)
and a ``see'' value, the visual analogue of a \(p\)-value (Chowdhury et
al., 2020), can be calculated using the \texttt{vinference} R package or
the process described in VanderPlas et al. (2021). The details of this
calculation are beyond the scope of this broader discussion of how to
test charts, but more detail on visual inference is provided in
\textless insert citation to visual inference WIRE article under
development\textgreater.

In another variation of the statistical lineup procedure, data generated
from two models are compared, with target plots from each model embedded
in the array of \(K\) total plots. The \(K-2\) null plots are
constructed from a mixture model that blends the two competing models
(VanderPlas \& Hofmann, 2017). Viewers are asked to select the panel(s)
which are most different, and the primary source of information are
trials in which viewers selected the target from one model but not the
other, indicating that the display method used allowed viewers to
differentiate one model's data (but not the other) from the nulls
created through a mixture model. This variation allows the experimenter
to assess graphical design choices to determine whether they effectively
emphasize structural differences in the data (VanderPlas \& Hofmann,
2017).

One advantage of the visual inference technique is that the experimenter
can ask a very general question, such as ``which of these plots is the
most different?'', rather than a specific question about the displayed
data which may require more quantitative sophistication. All of the
necessary information to make the decision is embedded in the choice of
the model used to generate the null plots. This feature is extremely
convenient when conducting the experiment and even allows small children
to complete the task. The downside is that as a result, visual inference
experiments do not allow experimenters to assess the viewer's
understanding of the information shown in the chart. In most cases,
visual inference experiments remove any contextual information from the
charts, including axis labels and values, plot titles, and so on, in
order to encourage participants to make decisions based solely on the
graphical presentation. This lack of context is a double-edged sword:
visual inference can involve participants who do not have any
mathematical training or instincts (including children), but researchers
also cannot use this technique to assess higher levels of engagement
with a chart, such as estimation, prediction, or reasoning based on
displayed information.\\

To assess the viewer's \emph{understanding} of information shown in a
chart, we must ask questions and allow the user to provide feedback.
User feedback may be collected on a numerical scale or through the use
of written comments, recorded ``think-aloud'' processes, and other more
qualitative interaction methods. In some studies, asking users to
interpret a chart within a larger scenario can be effective, as in
Figure~\ref{fig-estimation-describe}, while in others it is more helpful
to ask users to explain answers. In visual inference studies, asking
users why a specific panel was chosen has been demonstrated to provide
rich insight into otherwise confusing numerical results (VanderPlas \&
Hofmann, 2017).

Think-aloud methods ask the viewer to narrate their internal thought
process, either during or after completing a task (Haak et al., 2003).
These recordings (or transcripts) can provide valuable insights into
conscious cognition, and are often used when conducting usability
studies. While we have not to date recorded users talking out loud about
what they are seeing during a study, think aloud methods could easily be
implemented within a Shiny application, with audio recordings saved to
the server for transcription and analysis (Dunbar, 1995; Kirschenbaum,
2003; Trafton et al., 2000). It is even possible that these recordings
could be automatically transcribed using speech-to-text models. We have
used think-aloud methods informally during pilot studies to ``harden''
graphical experiments and verify the selection of parameters used in an
experiment. The success of this approach, combined with the few studies
which used think-aloud to assess charts (Haider et al., 2021; Kulhavy et
al., 1992; Lee et al., 2016), suggests that think-aloud methods are an
often-overlooked but useful tool for assessing data visualizations.

\begin{figure}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{Estimation_describe_plot_crop.png}}

}

\caption{\label{fig-estimation-describe}This question asks users to
write out a description of how the population of Ewoks changes over
time, without any further cues, to determine whether participants
default to multiplicative or additive language descriptions.}

\end{figure}%

Of course, in an online, asynchronous experiment, every user interaction
with the testing materials (typically hosted on a web page) can also be
recorded along with time stamps, mouse positions, browser size and
screen resolution, and other information. While we have not used this
type of information heavily in our experimental analyses thus far, in
most experiments we collect time stamp data in order to assess how long
participants spend on each question. Typically, the first round of test
questions takes the longest for participants to complete. Additional
replicates do not usually affect accuracy (i.e.~there is no immediate
learning effect) until after `too many' tests cognitive fatigue proves
to be detrimental to accuracy (Chowdhury et al., 2018). This sweet spot
between replicates and fatigue depends on the cognitive burden in each
test and should factor into designing the experiment. In some
experiments, we have provided participants with supportive tools, such
as ``scratch pads'' and calculators built into the Shiny application to
support the complex calculations required to answer higher-level
numerical estimation questions (Figure~\ref{fig-estimation-calc}). In
order to be supportive, the tools must be easy to use, but assuming this
bar is met, the tools can reduce participant cognitive load while
recording a wealth of information. This information provides real
insight into how participants were looking at the data, what strategies
they tried and discarded for reading the chart, and what visual
estimation methods were used. While systematic analysis and modeling of
this data may be difficult, as it is usually messy and often must be
manually coded, the insights provided can be extremely useful. However,
unless participants are required to use these tools, it is difficult to
gather comprehensive information - those participants that don't use
supportive tools likely differ in meaningful ways from those who do. As
a result, the information gathered from supportive tools likely does not
generalize to the entire sample.

\begin{figure}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{Estimation_numerical_screenshot_crop.png}}

}

\caption{\label{fig-estimation-calc}This question asks participants for
a numerical estimate, but provides a basic calculator and scratchpad.
All user interactions with the calculator and scratchpad are logged,
providing insight into the user's thought process and estimation
strategy.}

\end{figure}%

One of the most difficult components of designing an experiment which
asks users to directly estimate information from a chart using a full
scenario (background information, etc. as well as contextual details
from the chart) is that the questions must be extremely carefully
constructed. Mathematics education researchers provide guidelines for
selecting different levels of questioning in order to assess graph
comprehension: literal reading of the data, reading between the data,
and reading beyond the data (Curcio, 1987; Friel et al., 2001; Glazer,
2011; Wood, 1968). In a recent study, we identified questions based on
this framework to evaluate direct estimates and extend those estimates
to make comparisons between two points.

Even when great care is taken with the construction of the question,
participant answer accuracy is fundamentally limited by the fact that
many participants do not read and interpret the question with the care
and precision that it was written. Questions that ask participants to
e.g.~estimate the multiplicative change in a quantity at two time points
may be misunderstood as asking for an estimate of the additive
difference, and the resulting estimates are then one or more orders of
magnitude off of the correct answer. This is one area where lineup
methods are convenient - they do not depend on participants to
understand the nuances of language or scenarios built around the chart
under investigation. However, in some situations it may be sufficient to
ask participants to estimate direct numerical quantities that have
little contextual information, as done in (VanderPlas et al., 2019) when
assessing the accuracy of framed plots re-created from the Statistical
Atlas.

Another useful measurement strategy is to require participants to
\emph{engage directly} with an interactive visualization. This is useful
in a directed task, where users are asked to interact with the chart in
a specific way and the result is recorded, but it is also possible to
use interactive visualizations in an open-ended task, recording how
users engage with the graphic in an exploratory (as opposed to
goal-directed) manner. In one recent experiment, we asked participants
to forecast an exponential trend, with data presented on either a linear
or log scale. Using JavaScript code modified from New York Times
interactive graphics ``You Draw It'' features (Katz, 2017), we had users
draw trend lines with their computer mouse and make forecasts directly
on interactive charts, with the data and user-drawn predictions recorded
to our database. With interactive graphics rendered using JavaScript (or
other web libraries), the only limit to the types of questions one can
ask in testing graphics is one's ability to write code to interact with
the visualization library. This type of testing method can be extremely
natural for participants, but it also is hard to generalize when
discussing testing methods because of the potential range of
applications where it might be employed.

Whichever testing method is chosen should be appropriate to the type of
question under investigation and the level of visual and cognitive
engagement required to answer that question. While lineups are excellent
tools for assessing perceptual questions, they cannot address questions
aimed at understanding how people use charts within the wider context of
a story or practical task; this requires more direct methods with higher
ecological validity.

All of the testing methods described here require significant work to
develop a strategy for data generation appropriate for testing the
underlying question. For instance, when testing the perception of
exponential growth, we had to develop a model which would generate data
with varying growth rates, but where the data had a pre-specified domain
and range. The data generating model is particularly critical when using
lineups, as the null sampling model must replicate the important visual
features in the data. Each testing method has specific requirements, but
it is important to carefully calibrate the model parameters to allow for
some variability, but not too much, and to ensure that participants can
succeed at the task and do not feel like they are being made to analyze
random noise. This Goldilocks-style problem is the focus of the next
section.

\section{Developing a Model}\label{sec-model-dev}

Once the graphical task has been identified, it is necessary to develop
a model which can be used to explore the graphical features of interest
in a precise manner. This is the single longest part of the entire
experimental design and execution process, in part because choosing a
model that replicates important visual features of the data is extremely
complex (Cook et al., 2021; Hullman \& Gelman, 2021; VanderPlas, 2021).

There are two main options when developing a statistical model for
graphical testing: start with a large data set and sample from that data
set (Hofmann et al., 2012), or start from a model and sample data from
that model generating process (Robinson, 2022; VanderPlas \& Hofmann,
2015, 2017). This decision is largely determined by the availability of
a large data set containing the requisite features of interest and the
qualities being manipulated in the experiment. For instance, Hofmann et
al. (2012) used samples of different sizes from a pre-existing data set
to manipulate the amount of signal in each comparison; with a small
sample, there is less signal and the same amount of noise, making the
true plot harder to spot. In many situations, though, a convenient data
set with the right properties is harder to acquire, and it becomes
necessary to develop a sampling model to generate data for user
evaluation.

The tools we discuss in the remainder of this section can be applied
both to pre-existing data sets and to model-based sampling methods.

\subsection{Screening Parameters with
Simulation}\label{screening-parameters-with-simulation}

The choice of the parameter space used in testing is crucial to gain
insight from a study without putting too much burden on participants
with overlong studies. Choosing an appropriate space for testing
parameters is a well-known problem in psychometric testing: the space
considered should cover the area between `only some activation' to
`almost full activation' of an appropriate psychometric function (Schütt
et al., 2016; Valentin et al., 2024). When testing charts, visual
assessment is obviously key, but researchers can make use of statistical
indices related to the testing condition to narrow the parameter space
to a reasonable and efficient subset from which maximal information can
be acquired.\\
These statistical indices may also serve as quantitative proxies for the
difficulty of the visual task.

To identify a statistical proxy for visual difficulty that may help with
narrowing the parameter space, it can be useful to consider numerical
measures used to estimate the same types of visual information that will
be assessed in the experiment. For instance, we have used

\begin{itemize}
\tightlist
\item
  \(R^2\) as a measure of the strength of a linear relationship
\item
  Gini inequality as a measure of the strength of clustering
\item
  lack-of-fit statistics to assess the amount of curvature in an
  exponential relationship (shown in
  Figure~\ref{fig-lof-density-curves})
\end{itemize}

Then, a wide range of potential combinations of parameter values or
sampling strategies can be explored and summarized graphically; if the
numerical statistic cannot differentiate between the null and target
under a condition, it is reasonable to expect that a visual inspection
of the data may also not show significant results. As with any measure,
it is important that difficulty levels span a range from easy to hard;
we do not learn anything from finding out that everyone can distinguish
all of the combinations.

\begin{figure}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/fig-lof-density-curves-1.pdf}}

}

\caption{\label{fig-lof-density-curves}Density plot of the lack of fit
statistic showing separation of selected difficulty levels: High
(obvious curvature), Medium (noticeable curvature), and Low (almost
linear). Each density plot is the result of 1000 simulations from a
model \(y_i = \alpha\cdot e^{\beta\cdot x_i + \epsilon_i} + \theta\),
where \(\epsilon \sim N(0, \sigma^2)\). \(\alpha\) and \(\theta\) were
selected after manipulation of \(\beta\) and \(\sigma\) to ensure that
all data generated had similar \(y\) ranges so as not to provide visual
cues about model differences outside of the plot curvature.}

\end{figure}%

While this method is certainly more critical for model-based sampling
methods, it is also important when data are generated by sampling from a
larger data set. When sampling from a larger dataset, parameters are
more often sample size and stratification methods, but it is still
important to iteratively assess the data generating procedure through
simulation. Using numerical proxies for visual characteristics of data
displays such as curvature, linearity, scatter, dispersion can assist
with identifying optimal parameter settings to use across different
experimental conditions. Even with this strategy, it is still critical
to fine-tune the parameter choices with visual calibration and pilot
testing.

\subsection{Fine-Tuning Parameter
Choices}\label{fine-tuning-parameter-choices}

Once an appropriate set of parameters are identified using the numerical
screening method, it is important to calibrate these parameter
selections visually. No numerical statistic is a perfect measure of what
we actually see: at best, they are approximations of what we might
potentially see. We have found it to be useful to have one experimenter
calibrate the model parameters at a gross level, and then have another
experimenter narrow in on the parameters which are visually reasonable
within the selected range. Then, both examiners visually inspect a large
number of plots generated using those parameters to get a sense for how
difficult the task at hand is (this strategy is also described by Lu et
al. (2022)). At some point, all experimenters become so visually
saturated with the nuances of the data generating mechanism that it may
become necessary to ``sanity check'' the protocol with family members,
friends, and colleagues. These informal surveys provide extremely useful
feedback and can help to counteract the visual saturation of being
immersed in the design of a visualization experiment for months at a
time.

\subsection{Visual Assessment is
Critical}\label{visual-assessment-is-critical}

We cannot overstate the importance of visual assessment of your model
stimuli, preferably with fresh eyes. We highly recommend performing
several rounds of think-aloud pilot testing before deploying an
experiment. In support of this assessment, we offer up a cautionary tale
of our own experience: that of VanderPlas \& Hofmann (2017), where we
designed an experiment to test which plot aesthetics promoted discovery
of linear trends and/or clusters.

The experiment was a full 2x3x3 factorial exploration of three data
generating parameters, with 3 replicates at each parameter combination
(54 data sets) and 10 aesthetic combinations (for a total of 540
lineups). Each lineup had 20 different sub-panels, so we should have
carefully visually inspected some 10,800 different panels. As is evident
from the fact that we're telling this story as a cautionary tale, we
missed a critical problem with our data-generating mechanism: when
clusters were assigned to randomly generated data after the fact, we
didn't control the cluster size, leading to clusters of one or two
points in relatively few sub-panels. This became particularly noticeable
when bounding ellipses were added to the plot, as the method used to
generate those ellipses required at least 3 points in the cluster. The
missing boundary ellipse in the corresponding sub-panels escaped our
notice during the stimuli proof-reading phase of the experiment, but did
not escape the notice of our participants, who only needed to examine
about 10 lineups each (around 200 panels). An example of one of the
problematic lineups is shown in Figure~\ref{fig-lineup-problems}: many
participants selected panel 16 because of the missing ellipse; not a
wrong choice, but certainly not the effect we intended to test.

\begin{figure}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{lineup-missing-ellipse.png}}

}

\caption{\label{fig-lineup-problems}A lineup from Vanderplas \& Hofmann
(2017). Panel 10 shows the clustered target data and panel 17 shows the
target data with a strong linear relationship; either of these target
panels was the expected choice. Unfortunately, panel 16 has only two
bounding ellipses shown, which is an unintentional difference that
resulted from a faulty method for assigning clusters to null plots; many
participants selected this panel instead of one of the target panels.}

\end{figure}%

One reason why it is so difficult to generate sampling models for visual
explorations is that our visual system is optimized for identifying
differences between groups. This ability can interfere with the natural
to use the null sampling models that might be used in equivalent
numerical tests when running experiments that use visualizations. We
re-ran the experiment using a different clustering method that
controlled the number of points in each group. Instead of noticing the
number of ellipses, participants instead used the differences in size
and shape of the ellipses formed when clustering after the data
generating procedure. That is, participants could still detect the
artificial nature of the induced clusters using other features. While it
can be difficult to get the data generating method right, it is
essential to conducting visual experiments that generalize well beyond
the effects shown in a single data set or phenomenon. The time and
effort invested in this step at the outset of the experiment pays
dividends when it allows for clear generalization of the experimental
results to an entire statistical concept rather than a single data set.

\section{Experimental Design Considerations}\label{sec-exp-design}

It would be difficult to develop a full data generating model without
some idea of the experimental design: the basic structure of the
parameters which are to be manipulated, how the users will be tested,
and so on. These experimental design factors are fairly natural for
scientists to accumulate over the course of imagining and planning an
experiment. When conducting graphical tests, however, there are
additional considerations beyond those taught in a standard experimental
design course.

\subsection{Cognitive Load}\label{cognitive-load}

The primary human design factor to be aware of is that visual tasks and
assessing statistical graphics can be extremely cognitively taxing. In
our experience, it is difficult to expect participants to evaluate more
than about 15 charts in one sitting. If users are asked to deeply engage
and answer multiple questions about each chart, this limit may be lower
(8-10), but even with a relatively simple task, as in lineup methods, it
is hard for participants to accurately evaluate more than about 15
charts. Tasks which are more interactive, such as `You Draw It', may be
somewhat easier for participants, but it is unlikely that participants
would be willing to complete more than about 20 tasks in one sitting
even with tasks that require fewer decisions and more engagement. At
some point, participants' interest in accuracy will decline, and the
researcher is better off using a larger number of participants
completing fewer tasks within the window where participants are
attentive and engaged.

\subsection{Participant Instructions}\label{participant-instructions}

It can be extremely helpful to include ``practice'' demonstrations of
the task to show the basic process, logic, and reasoning. While it is
tempting to make these tasks fully representative of the type of
judgement which will be required of participants, practice tasks which
are too close to the experimental task may bias participants; we have
found that it works well to have a relatively easy practice task which
utilizes a slightly different type of plot and/or type of data than what
will be tested in the experiment. In cases which require interactivity,
gif animations of the task being carried out are useful, as are
additional visual cues, such as the yellow box used in the `You Draw It'
task\footnote{See a gif of testing with `You Draw It'
  \href{https://i.imgur.com/GM5YSen.gif}{here}} to indicate that there
were points which were not completed.

\subsection{Accounting for Participant
Variability}\label{accounting-for-participant-variability}

While studies have found some relationship between lineup performance
and demographic factors (VanderPlas \& Hofmann, 2016), these differences
are relatively small when participants are recruited from online testing
platforms like Amazon Mechanical Turk or Prolific, or when participants
are recruited from a university student population. Studies have also
not found a strong relationship between visual task performance and
participant recruitment method (VanderPlas et al., 2019), perhaps in
part because demographics which use social media sites and demographics
completing online research tasks for pay overlap heavily. However, when
participants are recruited using a statistically representative sample
of the wider population, education becomes an extremely useful predictor
of ability to effectively read and draw conclusions about a chart (Rice
et al., 2024).\\
In almost every situation, though, we find that some participants are
very good at evaluating graphics and some participants are not; as a
result, it can be extremely helpful to use random effects models with an
effect for individual participants.

\subsection{Demographic Data}\label{demographic-data}

It may be useful to ask a few more demographic questions about STEM
education level for studies which ask more of participants from a
mathematical standpoint; while lineup studies have not found strong
associations with those variables, lineup studies also do not require
participants to engage with the data presented in a chart in a way that
requires higher-order mathematical reasoning. This has allowed us to
make the argument to the ethics committee (institutional review board,
or IRB) that our research is exempt, as we do not collect enough
demographic information to identify participants, however, collecting
reduced demographic information occasionally comes at a cost. One recent
study examining the use of log scales and exponential data was conducted
using Prolific, which recruits participants from around the world; we
required only that participants were fluent in English to participate.
It was only after the experiment was completed that we realized that
different countries introduce logarithms as a concept at different
points during primary and secondary education; it might be that
individuals in some countries have much more experience with log scales
than those educated in the United States. We mention this only to point
out that while every experiment contains a few missed opportunities, it
is worth giving careful thought to the demographic questions asked of
participants and what information may be helpful during the analysis
stage.

\subsection{Trial Allocation to
Participants}\label{trial-allocation-to-participants}

In statistical design terms, most of our studies involve some type of
balanced incomplete block design, where participants are assigned to a
subset of experimental conditions which allow for estimation of the full
range of effects specified in the model. The particular structure of
these designs depends heavily on the factorial structure of the study,
but we typically arrange participants' trials to ensure that
participants see the same data set only once (where possible) and see as
many different experimental conditions as possible. It is also important
to reduce the impact of order effects using either e.g.~Latin square
designs or randomization where possible, but we recognize that this is
not always feasible due to the need to maintain participant naivete
during some portions of the experiment.

\subsection{Data Collection Platforms and
Infrastructure}\label{data-collection-platforms-and-infrastructure}

We have conducted visualization experiments using a wide variety of
tools: custom web PHP servers running interactive forms for data
collection, generic web survey platforms (Qualtrics, Google Forms) for
static graphics, and Shiny applications that control every part of the
experimental process from instructions to providing completion codes for
payment while rendering interactive graphics and generating fully
randomized data for each participant. In our research, Shiny has
provided the right balance between control over the experimental
setting, procedure, etc. and the intricate details of web server
management, however, this balance is likely different for every group
and potentially for every experiment. Likewise, we have recruited
participants using services such as \href{https://www.mturk.com/}{Amazon
Mechanical Turk} and \href{https://www.prolific.com/}{Prolific}, on
social media sites
(\href{https://web.archive.org/web/20250228110244/https://www.reddit.com/r/SampleSize/}{Reddit},
Twitter(X), Mastodon, and BlueSky), and by email; each recruitment
method has trade-offs between cost, convenience, number of people in the
sampling frame, and control over participant demographics. Rice et al.
(2024) recruited participants using nationally representative panel
samples and found that conclusions from fully representative samples of
the population can be very different from volunteer samples recruited
using other services, as many people are unmotivated to engage with
charts, don't know how to read charts, or impose pre-formed conclusions
onto visual displays. These individuals are unlikely to participate in
most ad-hoc volunteer-based sampling methods. While using representative
panel sampling products offered by organizations such as NORC or Gallup
is considerably more expensive than samples from Amazon MTurk or
Prolific, these results suggest that results from sampling methods used
online may not generalize to the broader population. People who
participate in studies via online platforms are more technologically
sophisticated and educated than the general population, and this bias
may significantly impact the conclusions.

If conducting an online experiment, we recommend consideration of
panel-based sampling methods as well as online testing platforms. While
Amazon MTurk was once the platform of choice for this type of research
(Heer \& Bostock, 2010), in 2025 Prolific, which is aimed at academic
experiments, seems to provide better options that ensure that
experiments are completed by people rather than AI than MTurk. It seems
likely that the preferred platform for participant recruitment will
continue to evolve over time, thus, we will recommend that experimenters
consider available options, compare pricing structures (as these vary
widely), and consider whether add-on fees for e.g.~demographically
representative samples are worth the additional cost. An additional
consideration when choosing between panel samples and online recruitment
platforms is the constraints placed on the types of questions and
graphics which can be used. In our experience, it is much easier to run
an experiment using the online recruitment platforms, which place many
fewer constraints on both the number and types of questions which can be
asked and the types of stimuli which can be used in the experiment.

\subsection{Ethical Review of Human Subjects
Experiments}\label{ethical-review-of-human-subjects-experiments}

In almost any research environment, there are requirements that
experiments involving human participants undergo review by an ethics
board to ensure that participants are not harmed by participating in the
study. We are most familiar with regulations in the United States and
can only speak to that environment. Most graphics experiments conducted
in the US fall under the ``exempt'' category of experiments which
require only basic review and approval. These experiments record only
basic demographic information, ask participants to complete tasks that
involve no risk, and experimenters record information which would not be
embarrassing to participants if exposed. In order to ensure that our
experiments fall into this category, we will often ensure that
demographic information is collected and stored separately from any
identifiable information, such as e.g.~user IDs from the participant
recruitment platform that allow us to monitor task completion and pay
participants for their time. Researchers should also ensure that they
comply with privacy laws such as the European
\href{https://gdpr-info.eu/chapter-3/}{General Data Protection
Regulation (GDPR)}; complying with this law while collecting data which
is not identifiable can be complicated. We recommend consulting with
your ethics board and institutional recommendations in order to maintain
legal compliance and safeguard participant privacy.

\section{Pilot Testing and Quality Assurance}\label{sec-pilot-test}

Once the data generating model is set, the experiment is designed, and
the charts have been developed, the next step is an extensive round of
pilot testing. The goal of pilot testing is to ensure that the
experiment is set up properly and that no issues have been overlooked.
Pilot testing also provides an opportunity to ensure that directions are
clear, participants know what they are supposed to be doing, and to
estimate task completion time (which determines how much participant
will be paid on many online testing platforms). Our studies usually go
through 2-3 rounds of pilot testing, with at least one of those rounds
involving any relatives and friends who are less technologically savvy.
We also purposely include talented individuals who can accidentally
crash any testing applications. A final round of testing usually
includes any and all coworkers and friends who may be available for
10-15 minutes during the work day. Pilot study samples do not need to be
representative of any particular population, though we would caution
against using exclusively visualization researchers or statisticians in
your pilot sample because some issues are less likely to show up with
knowledgeable participants. In some studies, each participant may see a
new set of data, depending on how the study is designed. In these cases,
we highly recommend saving all generated data to a database as well, so
that it is possible to go back and examine exactly what happened and/or
how things went wrong. Hard drive space is extremely cheap relative to
almost any other cost in an experiment; saving all of the data is a
sensible measure.

One highly useful (but not strictly essential) component of an
experiment that can be set up during the pilot testing stage is a basic
analysis script which summarizes all data collected to date visually. We
have used such scripts in the past to produce automatically updating
dashboards or web pages, allowing for real time or near-real time
monitoring of data collection efforts. This provides an easy way to
summarize completion of the experiment so that individuals can receive
credit (if using services like MTurk or Prolific), but also allows
interested participants to see the data if recruiting participants who
are more likely to be interested in the results, as sometimes happens on
social media sites. As data collection online can also happen extremely
quickly (300 participants in \textless2h in our most recent Prolific
experiment), this can provide an illusion of control over the deluge of
data, allowing any issues to be spotted and resolved relatively quickly.
If server load is a potential issue, it may also help to release batches
of trials over a longer period of time in order to minimize the chance
of having to make participants wait for others to complete the task
before the server can handle additional connections\footnote{This is the
  one major drawback to our preferred solution of self-hosting a Shiny
  server to handle data collection: the free version of Shiny server is
  limited to about 15 connections at any given time. Prolific has
  recently added rate-limiting functions to the experimental control
  platform, which makes controlling the number of active jobs much
  easier.}.

\section{Analyzing the Data}\label{sec-analysis}

One constant with data analysis of these types of experiments is that no
matter how carefully the experiment is planned, designed, and executed,
there will be surprises. This is the dual curse and blessing of studying
human perception: the visual system never quite works the way that we
expect that it will, which provides endless fodder for science and
occasionally complicates the data analysis.

\subsection{Generalized Linear Mixed
Models}\label{generalized-linear-mixed-models}

The see-value approach (Chowdhury et al., 2020; Majumder et al., 2013;
VanderPlas et al., 2021) is extremely useful for single lineups, but
when a series of lineups that are part of a designed experiment are
used, generalized linear mixed models are a much simpler way to
summarize the overall effect of various manipulated factors (Hofmann et
al., 2012; Robinson, 2022; VanderPlas \& Hofmann, 2017). This approach
also works extremely well for psychometric experiments (VanderPlas \&
Hofmann, 2015), as psychometric models can be easily fit into the
framework of a generalized linear mixed model (Ju et al., 2024) that has
higher power than the Rasch models (Andrich, 1988; Lu et al., 2022)
which were historically recommended for these experiments. In addition,
similar model structures with different link functions can be used to
model accuracy, response time, and confidence, if all three types of
information are collected from participants during the experiment.

\subsection{Numerical Estimation}\label{numerical-estimation}

There are additional considerations that should be expected when asking
participants to estimate numerical quantities. Anchoring and rounding
cause participant responses to cluster in ways that can bias statistical
estimators, requiring methods designed for these types of data (Heitjan
\& Rubin, 1991; Tourangeau et al., 2000; Ushakov \& Ushakov, 2017). An
alternative approach is to analyze the data graphically, as shown in
Figure~\ref{fig-density-rug}, which uses a density plot with rug
annotations to show individual participant point estimates. Rounding
effects can clearly be seen in the rug plot, but a kernel density
calculated using an appropriately selected bandwidth shows clear visual
differences between linear and log scale charts. In addition, a smaller
second mode can be seen in both linear and log conditions that
corresponds to the underlying model value; this suggests that a minority
of participants fit a mental regression model to the data and use that
model to estimate rather than estimating based off of the closest point.
When charts are carefully constructed to account for the experimental
structure and participant estimation strategies, displaying both
individual and aggregate responses, it is possible to demonstrate a
measurable difference between conditions without the need for
complicated statistical modeling which corrects for rounding and
anchoring effects.

While Figure~\ref{fig-density-rug} shows point estimates, the same
approach can be modified to account for experimental methods which
generate participant response curves. In these cases, rather than adding
a rug plot to a density plot showing aggregate estimates, it may be more
helpful to create spaghetti plots of individual estimates with a
superimposed consensus estimate and relevant annotations showing
e.g.~anchor points.\\
The advantage of this approach is that it can accommodate extremely
messy data without requiring the extensive data cleaning, modeling of
heuristics like rounding and anchoring, and elimination of nonsensical
responses that might be necessary to fit a statistical model. While
statistical models are undoubtedly beneficial in many situations, we
have often found that graphical displays of experimental results are at
least as useful for analyzing and presenting the results of graphical
experiments.

\begin{figure}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/fig-density-rug-1.pdf}}

}

\caption{\label{fig-density-rug}Density of participant estimates for the
year in which the population reaches 4000. Colors are associated to
scale - linear (blue) and log (orange) - and vertical lines indicate the
true value based on the underlying model equation (black solid) and
closest point value based on the simulated data set (black dashed). A
jittered rug plot along the \(x\)-axis shows where participant estimates
were made. The plot shows anchoring occurred to the closest point as
shown by an increase in density around the dashed line. Density peaks
occurred at whole values indicating rounding errors.}

\end{figure}%

\subsection{Direct Interactions}\label{direct-interactions}

If participants are making predictions and/or fitting visual statistics,
we have had success analyzing these responses by comparing the responses
to results from a statistical model to determine how visual statistics
differ from the numerical quantities derived mathematically. For
instance, we calculate the deviation between participant responses and
the linear regression in `You Draw It' experiments, then fitted
generalized additive mixed models to summarize the results across
different experimental conditions to assess how user-drawn predictions
deviated from the statistical estimates. In other direct interactions,
it may be useful to compare participant selections or annotations to
closest points on the chart to assess anchoring behavior; for discrete
selections, methods discussed in numerical estimation may also be
useful.

\subsection{Qualitative Responses}\label{qualitative-responses}

In many cases, it is helpful to combine participants' qualitative
reasoning with their quantitative responses to designed graphical
experiments. This approach provides useful context as to what
participants use to make their decisions, and can be useful when
assessing why unexpected responses occurred. We have used word clouds to
show overall themes in participant responses successfully in (VanderPlas
\& Hofmann, 2017); when paired with an appropriate linear model it
became clear that participants were fixating on unequal cluster size as
a visual cue. In cases where participants are provided with additional
utilities such as calculators and scratchpads, it can be useful to
select responses from individual participants which illustrate the
different types of calculations performed (but analyzing this data
quantitatively can be difficult).

\section{Conclusion}\label{conclusion}

Testing features of visualization graphically using online platforms
provides an incredibly powerful and efficient way to establish empirical
guidelines for statistical graphics and visualization. There are nearly
endless ways to combine web graphics, user interactions, and data
collection to get insight into perception and use of graphics in
practical settings. We have been continually surprised at the richness
of the data collected in these experiments and the ability to combine
qualitative and quantitative assessment to support conclusions that are
both nuanced and of practical use when deciding how to design and
present data using visualizations.

In this paper, we have attempted to contextualize and motivate the logic
behind the process we use to design empirical graphics experiments.
We've discussed model development, experimental design considerations,
pilot testing, and data analysis methods that have been honed over many
successful and less-than-optimal experiments. While no experiment
involving humans ever goes exactly to plan, following this process helps
to avoid some of the most likely mishaps, ensuring that each new
experiment's ``bonus'' findings have only minimal impacts on the study's
overall utility.

While it can be difficult to conduct empirical tests of different
visualizations, the results of these experiments support guidelines for
graphical design and communication of statistical findings in an
accessible and explainable way. Many chart design guidelines and
recommendations are based on heuristics, but as scientists, we should
prefer guidelines which are based on empirical, experimentally derived
results over opinions. Testing statistical graphics and developing
empirically supported guidelines for chart creation promises to support
better scientific communication, which is critical for educating the
public about topics like climate change, public health, the risk of
severe weather and more.

\section{References}\label{references}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-allenVisualizingScientificData2016}
Allen, E. A., \& Erhardt, E. B. (2016). Visualizing {Scientific} {Data}.
In J. T. Cacioppo, L. G. Tassinary, \& G. G. Berntson (Eds.),
\emph{Handbook of {Psychophysiology}} (4th ed., pp. 679--697). Cambridge
University Press.
https://doi.org/\url{https://doi.org/10.1017/9781107415782.031}

\bibitem[\citeproctext]{ref-andrichRaschModelsMeasurement1988}
Andrich, D. (1988). \emph{Rasch models for measurement} (1st ed.).
{SAGE} Publications.

\bibitem[\citeproctext]{ref-bertin1983semiology}
Bertin, J., \& Berg, W. J. (1983). \emph{Semiology of graphics:
Diagrams, networks, maps} (Vol. 1). University of Wisconsin press
Madison.

\bibitem[\citeproctext]{ref-d3}
Bostock, M., Ogievetsky, V., \& Heer, J. (2011). D3 {Data}-{Driven}
{Documents}. \emph{IEEE Transactions on Visualization and Computer
Graphics}, \emph{17}(12), 2301--2309.
https://doi.org/\url{https://doi.org/10/b7bhhf}

\bibitem[\citeproctext]{ref-bujaStatisticalInferenceExploratory2009}
Buja, A., Cook, D., Hofmann, H., Lawrence, M., Lee, E.-K., Swayne, D.
F., \& Wickham, H. (2009). Statistical inference for exploratory data
analysis and model diagnostics. \emph{Philosophical Transactions of the
Royal Society of London A: Mathematical, Physical and Engineering
Sciences}, \emph{367}(1906), 4361--4383.

\bibitem[\citeproctext]{ref-cairoFunctionalArtIntroduction2012}
Cairo, A. (2012). \emph{The {Functional} {Art}: {An} introduction to
information graphics and visualization}. New Riders.

\bibitem[\citeproctext]{ref-shiny}
Chang, W., Cheng, J., Allaire, J., Sievert, C., Schloerke, B., Xie, Y.,
Allen, J., McPherson, J., Dipert, A., \& Borges, B. (2021). \emph{Shiny:
Web application framework for r}.
\url{https://CRAN.R-project.org/package=shiny}

\bibitem[\citeproctext]{ref-chowdhury2018}
Chowdhury, N. R., Cook, D., Hofmann, H., \& Majumder, M. (2018).
Measuring Lineup Difficulty By Matching Distance Metrics With Subject
Choices in Crowd-Sourced Data. \emph{Journal of Computational and
Graphical Statistics}, \emph{27}(1), 132--145.
https://doi.org/\url{https://doi.org/10.1080/10618600.2017.1356323}

\bibitem[\citeproctext]{ref-niladriroychowdhurySeeValueApp2020}
Chowdhury, N. R., Diehl, H., Broderick, T., \& Stein, A. (2020).
\emph{The {See} {Value} {App}: {Visual} {Decision} {Making} for {Drug}
{Development}}. https://rinpharma.com/publication/rinpharma\_183/.

\bibitem[\citeproctext]{ref-clevelandShapeParameterTwoVariable1988}
Cleveland, W. S., McGill, M. E., \& McGill, R. (1988). The {Shape}
{Parameter} of a {Two}-{Variable} {Graph}. \emph{Journal of the American
Statistical Association}, \emph{83}(402), 289--300.
https://doi.org/\url{https://doi.org/10.1080/01621459.1988.10478598}

\bibitem[\citeproctext]{ref-clevelandGraphicalPerceptionGraphical1985}
Cleveland, W. S., \& McGill, R. (1985). Graphical {Perception} and
{Graphical} {Methods} for {Analyzing} {Scientific} {Data}.
\emph{Science}, \emph{229}(4716), 828--833.
https://doi.org/\url{https://doi.org/10.1126/science.229.4716.828}

\bibitem[\citeproctext]{ref-cookFoundationAvailableThinking2021}
Cook, D., Reid, N., \& Tanaka, E. (2021). The {Foundation} {Is}
{Available} for {Thinking} {About} {Data} {Visualization}
{Inferentially}. \emph{Harvard Data Science Review}, \emph{3}(3).
https://doi.org/\url{https://doi.org/10.1162/99608f92.8453435d}

\bibitem[\citeproctext]{ref-croxtonGraphicComparisonsBars1932}
Croxton, F. E. (1932). Graphic {Comparisons} by {Bars}, {Squares},
{Circles}, and {Cubes}. \emph{Journal of the American Statistical
Association}, \emph{27}(177), 54--60.

\bibitem[\citeproctext]{ref-croxtonBarChartsCircle1927}
Croxton, F. E., \& Stryker, R. E. (1927). Bar {Charts} {Versus} {Circle}
{Diagrams}. \emph{Journal of the American Statistical Association},
\emph{22}(160), 473--482.
https://doi.org/\url{https://doi.org/10.2307/2276829}

\bibitem[\citeproctext]{ref-curcio1987comprehension}
Curcio, F. (1987). Comprehension of mathematical relationships expressed
in graphs. \emph{Journal for Research in Mathematics Education},
\emph{18}(5), 382--393.

\bibitem[\citeproctext]{ref-desnoyersTaxonomyVisualsScience2011}
Desnoyers, L. (2011). \emph{Toward a {Taxonomy} of {Visuals} in
{Science} {Communication}}. \emph{58}(2), 16.

\bibitem[\citeproctext]{ref-dunbar1995scientists}
Dunbar, K. (1995). How scientists really reason: {Scientific} reasoning
in real-world laboratories. \emph{The Nature of Insight}, \emph{18},
365--395.

\bibitem[\citeproctext]{ref-eellsRelativeMeritsCircles1926}
Eells, W. C. (1926). The {Relative} {Merits} of {Circles} and {Bars} for
{Representing} {Component} {Parts}. \emph{Journal of the American
Statistical Association}, \emph{21}(154), 119--132.
https://doi.org/\url{https://doi.org/10.2307/2277140}

\bibitem[\citeproctext]{ref-fewInformationDashboardDesign2006}
Few, S. (2006). \emph{Information {Dashboard} {Design}: {The}
{Effective} {Visual} {Communication} of {Data}}. O'Reilly Media,
Incorporated.

\bibitem[\citeproctext]{ref-friel2001making}
Friel, S., Curcio, F., \& Bright, G. (2001). Making sense of graphs:
Critical factors influencing comprehension and instructional
implications. \emph{Journal for Research in Mathematics Education},
\emph{32}(2), 124--158.

\bibitem[\citeproctext]{ref-glazer2011challenges}
Glazer, N. (2011). Challenges with graph interpretation: A review of the
literature. \emph{Studies in Science Education}, \emph{47}(2), 183--210.

\bibitem[\citeproctext]{ref-thinkaloud}
Haak, M., De Jong, M., \& Schellens, P. (2003). Retrospective vs.
Concurrent think-aloud protocols: Testing the usability of an online
library catalogue. \emph{Behaviour \& {IT}}, \emph{22}, 339--351.
\url{https://doi.org/10.1080/0044929031000}

\bibitem[\citeproctext]{ref-haemerDoubleScalesAre1948}
Haemer, K. W. (1948). Double {Scales} are {Dangerous}. \emph{The
American Statistician}, \emph{2}(3), 24--24.
https://doi.org/\url{https://doi.org/10.1080/00031305.1948.10501588}

\bibitem[\citeproctext]{ref-haiderStrategiesDetectingDifference2021}
Haider, J. D., Pohl, M., Beecham, R., \& Dykes, J. (2021). Strategies
for detecting difference in~map line-up tasks. In C. Ardito, R.
Lanzilotti, A. Malizia, H. Petrie, A. Piccinno, G. Desolda, \& K. Inkpen
(Eds.), \emph{Human-computer interaction -- {INTERACT} 2021} (pp.
558--578). Springer International Publishing.
\url{https://doi.org/10.1007/978-3-030-85613-7_36}

\bibitem[\citeproctext]{ref-heerCrowdsourcingGraphicalPerception2010}
Heer, J., \& Bostock, M. (2010). Crowdsourcing graphical perception:
Using mechanical turk to assess visualization design. \emph{Proceedings
of the {SIGCHI} Conference on Human Factors in Computing Systems},
203--212. \url{https://doi.org/10.1145/1753326.1753357}

\bibitem[\citeproctext]{ref-heitjanIgnorabilityCoarseData1991}
Heitjan, D. F., \& Rubin, D. B. (1991). Ignorability and {Coarse}
{Data}. \emph{The Annals of Statistics}, \emph{19}(4), 2244--2253.
https://doi.org/\url{https://doi.org/10.1214/aos/1176348396}

\bibitem[\citeproctext]{ref-power}
Hofmann, H., Follett, L., Majumder, M., \& Cook, D. (2012). Graphical
tests for power comparison of competing designs. \emph{IEEE Transactions
on Visualization and Computer Graphics}, \emph{18}(12), 2441--2448.
\url{https://doi.org/10/f4fwkz}

\bibitem[\citeproctext]{ref-vonhuhnFurtherStudiesGraphic1927}
Huhn, R. von. (1927). Further {Studies} in the {Graphic} {Use} of
{Circles} and {Bars}: {A} {Discussion} of the {Eell}'s {Experiment}.
\emph{Journal of the American Statistical Association}, \emph{22}(157),
31--39. https://doi.org/\url{https://doi.org/10.2307/2277346}

\bibitem[\citeproctext]{ref-hullmanDesigningInteractiveExploratory2021}
Hullman, J., \& Gelman, A. (2021). Designing for {Interactive}
{Exploratory} {Data} {Analysis} {Requires} {Theories} of {Graphical}
{Inference}. \emph{Harvard Data Science Review}, \emph{3}(3).
https://doi.org/\url{https://doi.org/10.1162/99608f92.3ab8a587}

\bibitem[\citeproctext]{ref-asme-standards-graphics}
Joint committee on standards for graphic presentation. (1915).
\emph{Publications of the American Statistical Association},
\emph{14}(112), 790--797. \url{http://www.jstor.org/stable/2965153}

\bibitem[\citeproctext]{ref-juOneModelThat2024}
Ju, W. (Will)., Vanderplas, S., \& Hofmann, H. (2024). One {Model} that
fits them {All}: {Psychometrics} with {Generalized Linear Mixed Effects
Models}. \emph{Electronic Imaging}, 1--9.
\url{https://doi.org/10.2352/EI.2023.35.1.VDA-A01}

\bibitem[\citeproctext]{ref-katzYouDrawIt2017}
Katz, J. (2017). You {Draw} {It}: {Just} {How} {Bad} {Is} the {Drug}
{Overdose} {Epidemic}? \emph{The New York Times}.
\url{https://www.nytimes.com/interactive/2017/04/14/upshot/drug-overdose-epidemic-you-draw-it.html}

\bibitem[\citeproctext]{ref-kirschenbaum2003comparative}
Kirschenbaum, S. S. (2003). Comparative {Cognitive} {Task} {Analysis}:
{The} {Cognition} of {Weather} {Forecasting}. \emph{Proceedings of the
{Human} {Factors} and {Ergonomics} {Society} {Annual} {Meeting}},
\emph{47}, 473--477.
https://doi.org/\url{https://doi.org/10.1177/154193120304700347}

\bibitem[\citeproctext]{ref-kosslynGraphDesignEye2006}
Kosslyn, S. M. (2006). \emph{Graph {Design} for the {Eye} and {Mind}}.
Oxford University Press.

\bibitem[\citeproctext]{ref-kulhavyCartographicExperienceThinking1992}
Kulhavy, Pridemore, \& Stock. (1992). Cartographic experience and
thinking aloud about thematic maps. \emph{Cartographica}, \emph{29}(1),
1--9. \url{https://doi.org/10.3138/H61J-VX35-J6WW-8111}

\bibitem[\citeproctext]{ref-leeHowPeopleMake2016}
Lee, S., Kim, S.-H., Hung, Y.-H., Lam, H., Kang, Y.-A., \& Yi, J. S.
(2016). How do people make sense of unfamiliar visualizations?: A
grounded model of novice's information visualization sensemaking.
\emph{{IEEE} Transactions on Visualization and Computer Graphics},
\emph{22}(1), 499--508. \url{https://doi.org/10.1109/TVCG.2015.2467195}

\bibitem[\citeproctext]{ref-loyDiagnosticToolsHierarchical2013}
Loy, A., \& Hofmann, H. (2013). Diagnostic tools for hierarchical linear
models. \emph{Wiley Interdisciplinary Reviews: Computational
Statistics}, \emph{5}(1), 48--61.
https://doi.org/\url{https://doi.org/10.1002/wics.1238}

\bibitem[\citeproctext]{ref-luModelingJustNoticeable2022}
Lu, M., Lanir, J., Wang, C., Yao, Y., Zhang, W., Deussen, O., \& Huang,
H. (2022). Modeling {Just} {Noticeable} {Differences} in {Charts}.
\emph{IEEE Transactions on Visualization and Computer Graphics},
\emph{28}(1), 718--726.
https://doi.org/\url{https://doi.org/10.1109/TVCG.2021.3114874}

\bibitem[\citeproctext]{ref-majumderValidationVisualStatistical2013}
Majumder, M., Hofmann, H., \& Cook, D. (2013). Validation of visual
statistical inference, applied to linear models. \emph{Journal of the
American Statistical Association}, \emph{108}(503), 942--956.
https://doi.org/\url{https://doi.org/10/f5gntt}

\bibitem[\citeproctext]{ref-R}
R Core Team. (2022). \emph{R: A language and environment for statistical
computing}. R Foundation for Statistical Computing.
\url{https://www.R-project.org/}

\bibitem[\citeproctext]{ref-ribeccaSearchChartsData}
Ribecca, S. (2022). \emph{The data visualisation catalogue}.
https://datavizcatalogue.com.

\bibitem[\citeproctext]{ref-riceTestingPerceptualAccuracy2024}
Rice, K., Hofmann, H., du Toit, N., \& Mulrow, E. (2024). Testing
{Perceptual Accuracy} in a {U}.{S}. {General Population Survey Using
Stacked Bar Charts}. \emph{Journal of Data Science}, 1--18.
\url{https://doi.org/10.6339/24-JDS1121}

\bibitem[\citeproctext]{ref-emily-diss}
Robinson, E. A. (2022). \emph{Human {Perception} of {Exponentially}
{Increasing} {Data} {Displayed} on a {Log} {Scale} {Evaluated} {Through}
{Experimental} {Graphics} {Tasks}} {[}Doctoral, University of Nebraska,
Lincoln{]}.
\url{https://github.com/earobinson95/EmilyARobinson-UNL-dissertation/raw/main/EmilyRobinson-final-dissertation.pdf}

\bibitem[\citeproctext]{ref-schuttPainfreeAccurateBayesian2016}
Schütt, H. H., Harmeling, S., Macke, J. H., \& Wichmann, F. A. (2016).
Painfree and accurate bayesian estimation of psychometric functions for
(potentially) overdispersed data. \emph{Vision Research}, \emph{122},
105--123. \url{https://doi.org/10.1016/j.visres.2016.02.002}

\bibitem[\citeproctext]{ref-tourangeau_rips_rasinski_2000}
Tourangeau, R., Rips, L. J., \& Rasinski, K. (2000). \emph{The
psychology of survey response}. Cambridge University Press.
https://doi.org/\url{https://doi.org/10.1017/CBO9780511819322}

\bibitem[\citeproctext]{ref-traftonTurningPicturesNumbers2000}
Trafton, G. J., Kirschenbaum, S. S., Tsui, T. L., Miyamoto, R. T.,
Ballas, J. A., \& Raymond, P. D. (2000). Turning pictures into numbers:
Extracting and generating information from complex visualizations.
\emph{International Journal of Human-Computer Studies}, \emph{53}(5),
827--850. \url{https://doi.org/10.1006/ijhc.2000.0419}

\bibitem[\citeproctext]{ref-tufte}
Tufte, E. (1991). \emph{The {Visual} {Display} of {Quantitative}
{Information}} (2nd ed.). Graphics Press.

\bibitem[\citeproctext]{ref-ushakovStatisticalAnalysisRounded2017}
Ushakov, N. G., \& Ushakov, V. G. (2017). Statistical analysis of
rounded data: {Recovering} of information lost due to rounding.
\emph{Journal of the Korean Statistical Society}, \emph{46}(3),
426--437.
https://doi.org/\url{https://doi.org/10.1016/j.jkss.2017.01.003}

\bibitem[\citeproctext]{ref-valentinDesigningOptimalBehavioral2024}
Valentin, S., Kleinegesse, S., Bramley, N. R., Seriès, P., Gutmann, M.
U., \& Lucas, C. G. (2024). Designing optimal behavioral experiments
using machine learning. \emph{{eLife}}, \emph{13}.
\url{https://doi.org/10.7554/eLife.86224}

\bibitem[\citeproctext]{ref-vanderplasDesigningGraphicsRequires2021}
VanderPlas, S. (2021). Designing {Graphics} {Requires} {Useful}
{Experimental} {Testing} {Frameworks} and {Graphics} {Derived} {From}
{Empirical} {Results}. \emph{Harvard Data Science Review}, \emph{3}(3).
https://doi.org/\url{https://doi.org/10.1162/99608f92.7d099fd0}

\bibitem[\citeproctext]{ref-vanderplasEscapingFlatlandGraphics2024}
Vanderplas, S., Blankenship, E., \& Wiederich, T. (2024). Escaping
{Flatland}: {Graphics}, {Dimensionality}, and~{Human Perception}. In H.
Mori \& Y. Asahi (Eds.), \emph{Human {Interface} and the {Management} of
{Information}} (pp. 140--156). Springer Nature Switzerland.
\url{https://doi.org/10.1007/978-3-031-60114-9_11}

\bibitem[\citeproctext]{ref-vanderplasTestingStatisticalCharts2020}
Vanderplas, S., Cook, D., \& Hofmann, H. (2020). Testing {Statistical}
{Charts}: {What} {Makes} a {Good} {Graph}? \emph{Annual Review of
Statistics and Its Application}, \emph{7}(1).
https://doi.org/\url{https://doi.org/10.1146/annurev-statistics-031219-041252}

\bibitem[\citeproctext]{ref-vanderplasFramedReproducingRevisiting2019}
VanderPlas, S., Goluch, R., \& Hofmann, H. (2019). Framed! {Reproducing}
and {Revisiting} 150 year old charts. \emph{Journal of Computational and
Graphical Statistics}.
https://doi.org/\url{https://doi.org/10.1080/10618600.2018.1562937}

\bibitem[\citeproctext]{ref-sineillusion}
VanderPlas, S., \& Hofmann, H. (2015). Signs of the {Sine}
{Illusion}---{Why} {We} {Need} to {Care}. \emph{Journal of Computational
and Graphical Statistics}, \emph{24}(4), 1170--1190.
https://doi.org/\url{https://doi.org/10.1080/10618600.2014.951547}

\bibitem[\citeproctext]{ref-vanderplasSpatialReasoningData2016a}
VanderPlas, S., \& Hofmann, H. (2016). Spatial {Reasoning} and {Data}
{Displays}. \emph{IEEE Transactions on Visualization and Computer
Graphics}, \emph{22}(1), 459--468.
https://doi.org/\url{https://doi.org/10.1109/TVCG.2015.2469125}

\bibitem[\citeproctext]{ref-clusters}
VanderPlas, S., \& Hofmann, H. (2017). Clusters {Beat} {Trend}!?
{Testing} {Feature} {Hierarchy} in {Statistical} {Graphics}.
\emph{Journal of Computational and Graphical Statistics}, \emph{26}(2),
231--242. \url{https://doi.org/10.1080/10618600.2016.1209116}

\bibitem[\citeproctext]{ref-vanderplasStatisticalSignificanceCalculations2021}
VanderPlas, S., Röttger, C., Cook, D., \& Hofmann, H. (2021).
Statistical significance calculations for scenarios in visual inference.
\emph{Stat}, \emph{10}(1).
https://doi.org/\url{https://doi.org/10.1002/sta4.337}

\bibitem[\citeproctext]{ref-ggplot2}
Wickham, H. (2016). \emph{ggplot2: Elegant graphics for data analysis}.
Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}

\bibitem[\citeproctext]{ref-wickhamGraphicalInferenceInfovis2010}
Wickham, H., Cook, D., Hofmann, H., \& Buja, A. (2010). Graphical
inference for infovis. \emph{IEEE Transactions on Visualization and
Computer Graphics}, \emph{16}(6), 973--979.
https://doi.org/\url{https://doi.org/10.1109/TVCG.2010.161}

\bibitem[\citeproctext]{ref-wilkinsonGrammarGraphics1999}
Wilkinson, L. (1999). \emph{The grammar of graphics}. Springer.
\url{http://public.ebookcentral.proquest.com/choice/publicfullrecord.aspx?p=3085765}

\bibitem[\citeproctext]{ref-wood1968objectives}
Wood, R. (1968). Objectives in the teaching of mathematics.
\emph{Educational Research}, \emph{10}(2), 83--98.

\end{CSLReferences}

\end{document}

### 2025-03-14 submitted-index.tex
Review modifications

--- 2025-03-14 index.tex
@@
 \makeatletter
-\newsavebox\pandoc@box
-\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
-  \sbox\pandoc@box{#1}%
-  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
-  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
-  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
-  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
-  \else\usebox{\pandoc@box}%
-  \fi%
-}
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
 % Set default figure placement to htbp
+\makeatletter
 \def\fps@figure{htbp}
@@
 
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
 \usepackage{bookmark}
@@
 \author{Emily Robinson \and Heike Hofmann \and Susan Vanderplas}
-\date{2025-03-14}
+\date{2024-11-01}
 
@@
 
-Data visualizations are a critically important tool for communicating
+Charts and data visualizations are primary methods of communicating
 scientific information to the public in what creators hope is an
@@
 (Ribecca, 2022), create a taxonomy of charts (Bertin \& Berg, 1983;
-Desnoyers, 2011), and even to create charts using a domain-specific
-grammar of graphics (Wilkinson, 1999) that is also useful for
-classification. One extremely useful reference is from Data to Viz
-(https://www.data-to-viz.com/), which uses a decision tree to show
-different visualizations compatible with the data; R, python, D3.js, and
-React code are provided to demonstrate how to create those
-visualizations. With all of the different design choices available, how
-are chart creators to know what is the best approach for communicating
-data to the appropriate audience?
-
-While there are heuristics, rules-of-thumb, and guidelines (Allen \&
-Erhardt, 2016; Few, 2006; Haemer, 1948; {``Joint Committee on Standards
-for Graphic Presentation,''} 1915; Kosslyn, 2006) for creating useful
-and visually attractive data displays, the best way to establish the
-efficacy of various design decisions is to test the visualization on
-humans, evaluating different variants under controlled conditions
-(Cleveland et al., 1988; Cleveland \& McGill, 1985). Empirical
-assessments of visualizations, when carefully designed, allow
-statisticians to determine which representation of the same data is most
-effective along one or more dimension(s) of interest: estimation or
-prediction accuracy, within or between group comparisons, response time,
-and ability to draw real-world decisions are common goals for charts.
-
-It is extremely challenging to design studies which strike the right
-balance between experimental control (i.e.~internal validity) and
-generalizability to a wider context (i.e.~external validity). Simply
-asking people to read quantities off of a graph may not generalize
-beyond the questions asked or the data used in the chart (Croxton, 1932;
-Croxton \& Stryker, 1927; Eells, 1926; Huhn, 1927), but designing a
-study that is sufficiently robust to those issues requires manipulation
-or control of so many factors that the amount of participants and trials
-quickly becomes daunting or unaffordable. In this paper, we attempt to
-distill the experience gained from conducting several different types of
-graphics experiments (Hofmann et al., 2012; Robinson, 2022; VanderPlas
-et al., 2019; Vanderplas et al., 2024; VanderPlas \& Hofmann, 2015,
-2017), discussing the use of different testing methods (Vanderplas et
-al., 2020), the process of designing a graphical experiment, and
-analysis of the resulting empirical data. It is our hope that this paper
-will lower the barriers that exist for conducting empirical graphics
-research and reduce the probability of costly mistakes.
+Desnoyers, 2011), and even to classify charts using a domain-specific
+grammar of graphics (Wilkinson, 1999). With all of the different design
+choices available, how are chart creators to know what is the best
+approach for communicating data to the appropriate audience?
+
+While there are heuristics, rules-of-thumb, and various guidelines
+(Allen \& Erhardt, 2016; Few, 2006; Haemer, 1948; {``Joint Committee on
+Standards for Graphic Presentation,''} 1915; Kosslyn, 2006) for creating
+useful and visually attractive charts, the best way to establish the
+efficacy of various design decisions is to test charts: evaluate them
+under controlled conditions, and gather appropriate data (Cleveland et
+al., 1988; Cleveland \& McGill, 1985) to compare different
+representations of the same data. It is extremely challenging to design
+studies which strike the right combination of internal and external
+validity. Simply asking people to read quantities off of a graph may not
+generalize beyond the questions asked or the data used in the chart
+(Croxton, 1932; Croxton \& Stryker, 1927; Eells, 1926; Huhn, 1927), but
+designing a study that is sufficiently robust to those issues requires
+manipulation of so many factors as to be functionally close to
+impossible to control or manipulate in a single experiment. In this
+paper, we attempt to distill the experience gained from conducting
+several different types of graphics experiments (Hofmann et al., 2012;
+Robinson, 2022; VanderPlas et al., 2019; Vanderplas et al., 2024;
+VanderPlas \& Hofmann, 2015, 2017), discussing the use of different
+testing methods (Vanderplas et al., 2020), the process of designing a
+graphical experiment, and analysis of the resulting empirical data. It
+is our hope that this paper will lower the barriers that exist for
+conducting empirical graphics research and reduce the probability of
+costly mistakes.
 
@@
 There are many different testing methods used to empirically assess
-statistical graphics. This paper uses studies conducted online without
-additional equipment as primary examples, though many of the same
-considerations apply to in-person experiments conducted using additional
-equipment, including 3D printed charts, eye-tracking equipment, and
-interactive data displays. Online experiments have lower overhead, offer
-relatively fast data collection, and provide useful results for
-well-designed experiments. The toolkit used for these experiments is
-R-based (R Core Team, 2022), and includes ggplot2 (Wickham, 2016) and
-Shiny (Chang et al., 2021) as primary components. In many experiments,
-we customized the Shiny interface with JavaScript and d3 (Bostock et
-al., 2011), enabling interactive graphics, use of svgs, and other useful
-extensions. While we prefer this set of tools, most of the observations
-described here apply to a wide variety of different workflows for
-graphical experimentation, including in-person experiments.
-
-It is important to consider the level of user engagement which is
-necessary to complete a particular visual or graphical task. For
-instance, testing whether someone can detect an effect such as a linear
-trend in noisy data is a perceptual question. Perceptual questions are
-often examined experimentally using methods which allow the user to
-interact with the data on a basic visual level: users are presented with
-a visual stimulus and answer yes/no questions to indicate whether the
-effect is detected. Numerical estimation is another common task when
-testing graphics: in these experiments, the participant views a chart,
-estimates the requested numerical quantity, and enters the estimate into
-the application through a numerical input, slider, or other form
-element. Sometimes, it is possible to set up a scenario where the user
-adjusts the plot using a set of controls designed to provide a fixed set
-of interactive operations. This type of user engagement was used to
-assess the strength of the sine illusion (VanderPlas \& Hofmann, 2015):
-users adjusted the strength of a transformation designed to correct the
-illusion until the lines appeared to be the same length, as shown in
-Figure~\ref{fig-sine-illusion}, providing a direct measure of the
-magnitude of the sine illusion's effect. In other situations, it may be
-preferable to have the user directly interact with the visual stimulus.
-In Vanderplas et al. (2024), participants were asked to rotate and
-interact with a 3D rendered bar chart; the application recorded user
-interactions and corresponding rotation matrices, providing insight into
-the visual comparisons the user may have been performing. This
-information was used as a supplement to the explicitly provided
-estimates, providing some contextual information as well as the ability
-to identify the level of participant engagement with the questions. When
-experiments are conducted as part of classroom experiential learning, it
-is sometimes helpful to be able to separate the low-effort participants
-from those who were fully intellectually engaged in the task.
-Interactive graphics provide another level of user engagement that can
-be much more open-ended. With interactive graphics, researchers can ask
-participants to directly annotate plots, toggle aesthetics, and
-highlight groups and plot features. Careful implementation of the
-experiment application may allow for each of these interactions to be
-recorded and analyzed, producing a rich, if messy, set of data that may
-allow researchers to tease apart visual estimation error from common
-shortcuts such as rounding used during direct numerical estimation.
+graphics. While it is certainly possible to use equipment such as
+eye-tracking devices, this paper focuses primarily focus on studies
+conducted without additional equipment, and primarily online, because
+these experiments have the lowest overhead, offer relatively fast data
+collection, and provide useful results, for well-designed experiments.
+The toolkit used for these experiments is R-based (R Core Team, 2022),
+and includes ggplot2 (Wickham, 2016) and Shiny (Chang et al., 2021) as
+primary components, as well as JavaScript (including D3 (Bostock et al.,
+2011)) to customize the Shiny interface for many of the experiments.
+While we prefer this set of tools, most of the observations described
+here apply to a wide variety of different workflows for graphical
+experimentation.
+
+One primary concern when testing graphics is the level of user
+engagement required. For instance, testing whether someone can detect an
+effect such as a trend line in noisy data is a \emph{perceptual}
+question best addressed using methods which allow the user to interact
+with the data on a basic visual level. Sometimes, it is possible to set
+up a scenario where the user adjusts the plot directly, as in an
+experiment testing the strength of the sine illusion experiment
+(VanderPlas \& Hofmann, 2015): users adjusted the strength of the
+illusion correction until the lines appeared to be the same length,
+providing a direct measure of the magnitude of the sine illusion's
+effect, as in Figure~\ref{fig-sine-illusion}. In other situations, it
+may be possible to have the user directly interact with the plot, as in
+Vanderplas et al. (2024), where participants were asked to rotate and
+interact with a 3D rendered bar chart; the application then records all
+relevant user interactions to provide insight into the visual
+comparisons the user may have been performing as a supplement to the
+explicitly provided estimates.
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{sine_illusion_screenshot.png}}
+\includegraphics{sine_illusion_screenshot.png}
 
@@
 \caption{\label{fig-sine-illusion}Direct adjustment of a plot in a
-perceptual task. In this experiment, designed to assess the strength of
-the sine illusion, the user adjusts the plot using - and + buttons,
-which control the strength of a transformation designed to correct the
-effect of the sine illusion. When the user is satisfied that the lines
-are of equal length, they select the `Finished' button to move to the
-next task. The experiment used a psychophysics experimental design, the
-method of adjustment, but leveraged the interactive Shiny interface to
-record the entire sequence of adjustments made by the user for each
-trial. A demo version of this application can be found at
-https://shiny.srvanderplas.com/sine-illusion/.}
+perceptual task}
 
@@
 
-Visual inference (Buja et al., 2009; Wickham et al., 2010) is another
-useful testing tool for perceptual questions such as ``which chart
-displays this data more clearly'' (Hofmann et al., 2012) while
-simultaneously assessing the statistical significance of the graphical
-finding in a chart. Visual inference charts are often called ``lineups''
-in analogy to the criminal procedure where the suspect is placed in a
-line with several other individuals with similar characteristics. In a
-graphical lineup procedure, there is a target plot containing the real
-data, embedded in an array of innocent ``null'' plots generated through
-resampling or simulation. If viewers consistently pick the target plot
-at a higher rate than any of the null plots, the target plot is said to
-be visually significant (Loy \& Hofmann, 2013; Majumder et al., 2013)
-and a ``see'' value, the visual analogue of a \(p\)-value (Chowdhury et
-al., 2020), can be calculated using the \texttt{vinference} R package or
-the process described in VanderPlas et al. (2021). The details of this
-calculation are beyond the scope of this broader discussion of how to
-test charts, but more detail on visual inference is provided in
-\textless insert citation to visual inference WIRE article under
-development\textgreater.
-
-In another variation of the statistical lineup procedure, data generated
-from two models are compared, with target plots from each model embedded
-in the array of \(K\) total plots. The \(K-2\) null plots are
-constructed from a mixture model that blends the two competing models
-(VanderPlas \& Hofmann, 2017). Viewers are asked to select the panel(s)
-which are most different, and the primary source of information are
-trials in which viewers selected the target from one model but not the
-other, indicating that the display method used allowed viewers to
-differentiate one model's data (but not the other) from the nulls
-created through a mixture model. This variation allows the experimenter
-to assess graphical design choices to determine whether they effectively
-emphasize structural differences in the data (VanderPlas \& Hofmann,
-2017).
-
-One advantage of the visual inference technique is that the experimenter
-can ask a very general question, such as ``which of these plots is the
-most different?'', rather than a specific question about the displayed
-data which may require more quantitative sophistication. All of the
-necessary information to make the decision is embedded in the choice of
-the model used to generate the null plots. This feature is extremely
-convenient when conducting the experiment and even allows small children
-to complete the task. The downside is that as a result, visual inference
-experiments do not allow experimenters to assess the viewer's
-understanding of the information shown in the chart. In most cases,
-visual inference experiments remove any contextual information from the
-charts, including axis labels and values, plot titles, and so on, in
-order to encourage participants to make decisions based solely on the
-graphical presentation. This lack of context is a double-edged sword:
-visual inference can involve participants who do not have any
-mathematical training or instincts (including children), but researchers
-also cannot use this technique to assess higher levels of engagement
-with a chart, such as estimation, prediction, or reasoning based on
-displayed information.\\
+Statistical lineups (Buja et al., 2009; Wickham et al., 2010) are
+another useful testing tool for perceptual questions such as ``which
+chart displays this data more clearly'' (Hofmann et al., 2012) while
+simultaneously assessing the significance of the graphical finding in
+the chart. As in the criminal procedure, lineups embed a target plot
+with signal into an array of 19 other innocent ``null'' plots that show
+only noise. If viewers consistently pick the target plot at a higher
+rate than any of the 19 null plots, the target plot is said to be
+visually significant (Loy \& Hofmann, 2013; Majumder et al., 2013) and a
+``see'' value (the visual analogue of a \(p\)-value for a statistical
+test) can be calculated (Chowdhury et al., 2020; VanderPlas et al.,
+2021). In another variation of the statistical lineup procedure, two
+models are compared, with target plots from each model embedded in the
+array of 20 total plots; the remaining null plots are constructed from a
+mixture model that blends the two competing models. This variation
+allows the experimenter to assess graphical design choices to determine
+whether they effectively emphasize structural differences in the data
+and assign a hierarchy of visual attributes (VanderPlas \& Hofmann,
+2017). Statistical lineups do not require the experimenter to provide
+any contextual information for the data: all of the necessary
+information is embedded in the choice of a null statistical model, which
+is convenient for testing purposes, but does not allow for testing the
+viewer's understanding of the information shown in the chart.
 
@@
 Figure~\ref{fig-estimation-describe}, while in others it is more helpful
-to ask users to explain answers. In visual inference studies, asking
-users why a specific panel was chosen has been demonstrated to provide
-rich insight into otherwise confusing numerical results (VanderPlas \&
-Hofmann, 2017).
-
-Think-aloud methods ask the viewer to narrate their internal thought
-process, either during or after completing a task (Haak et al., 2003).
-These recordings (or transcripts) can provide valuable insights into
-conscious cognition, and are often used when conducting usability
-studies. While we have not to date recorded users talking out loud about
-what they are seeing during a study, think aloud methods could easily be
-implemented within a Shiny application, with audio recordings saved to
-the server for transcription and analysis (Dunbar, 1995; Kirschenbaum,
-2003; Trafton et al., 2000). It is even possible that these recordings
-could be automatically transcribed using speech-to-text models. We have
-used think-aloud methods informally during pilot studies to ``harden''
-graphical experiments and verify the selection of parameters used in an
-experiment. The success of this approach, combined with the few studies
-which used think-aloud to assess charts (Haider et al., 2021; Kulhavy et
-al., 1992; Lee et al., 2016), suggests that think-aloud methods are an
-often-overlooked but useful tool for assessing data visualizations.
+to ask users to explain answers. In lineup studies, asking users why a
+specific panel was chosen can provide rich insight into otherwise
+confusing numerical results. While we have not (to date) recorded users
+talking out loud about what they are seeing, think aloud methods could
+be implemented within a Shiny application, with audio recordings saved
+to the server for transcription and analysis (Dunbar, 1995;
+Kirschenbaum, 2003; Trafton et al., 2000); it is even possible that
+these recordings could be automatically transcribed using text-to-speech
+functions of large language models.
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{Estimation_describe_plot_crop.png}}
+\includegraphics{Estimation_describe_plot_crop.png}
 
@@
 type of information heavily in our experimental analyses thus far, in
-most experiments we collect time stamp data in order to assess how long
+most experiments, we collect time stamp data in order to assess how long
 participants spend on each question. Typically, the first round of test
@@
 estimation methods were used. While systematic analysis and modeling of
-this data may be difficult, as it is usually messy and often must be
-manually coded, the insights provided can be extremely useful. However,
-unless participants are required to use these tools, it is difficult to
-gather comprehensive information - those participants that don't use
-supportive tools likely differ in meaningful ways from those who do. As
-a result, the information gathered from supportive tools likely does not
-generalize to the entire sample.
+this messy data may be difficult, the insights provided can be extremely
+useful.
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{Estimation_numerical_screenshot_crop.png}}
+\includegraphics{Estimation_numerical_screenshot_crop.png}
 
@@
 
-The choice of the parameter space used in testing is crucial to gain
-insight from a study without putting too much burden on participants
-with overlong studies. Choosing an appropriate space for testing
-parameters is a well-known problem in psychometric testing: the space
-considered should cover the area between `only some activation' to
-`almost full activation' of an appropriate psychometric function (Schütt
-et al., 2016; Valentin et al., 2024). When testing charts, visual
-assessment is obviously key, but researchers can make use of statistical
-indices related to the testing condition to narrow the parameter space
-to a reasonable and efficient subset from which maximal information can
-be acquired.\\
-These statistical indices may also serve as quantitative proxies for the
-difficulty of the visual task.
-
-To identify a statistical proxy for visual difficulty that may help with
-narrowing the parameter space, it can be useful to consider numerical
-measures used to estimate the same types of visual information that will
-be assessed in the experiment. For instance, we have used
+The choice of the tested space is crucial to gain insight from a study
+without putting too much burden on participants with overlong studies.
+Choosing an appropriate space for testing parameters is a well-known
+problem in psychometric testing (\textbf{shah?}): the space considered
+should cover the area between `only some activation' to `almost full
+activation' of an appropriate psychometric function. When testing
+charts, visual assessment is obviously key, but researchers can make use
+of statistical indices related to the testing condition to narrow the
+parameter space.\\
+These indices then serve as quantitative proxies of visual difficulty.
+
+If using model based sampling methods, it is important to iteratively
+assess the sampling procedure through simulation. Start with a sampling
+method and define a parameter space that is appropriate for the model,
+and then simulate many data sets from each possible parameter
+combinations using a grid search. When sampling from a large data set,
+it is still important to assess the sample size and any potential
+stratification methods employed in order to ensure that the visual
+effect is represented.
+
+In both cases, it can help to employ some general feature engineering -
+what numerical statistic provides an estimation of the basic concept
+which is being visually evaluated? For instance, we have used
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/fig-lof-density-curves-1.pdf}}
+\includegraphics{index_files/figure-pdf/fig-lof-density-curves-1.pdf}
 
@@
 \end{figure}%
-
-While this method is certainly more critical for model-based sampling
-methods, it is also important when data are generated by sampling from a
-larger data set. When sampling from a larger dataset, parameters are
-more often sample size and stratification methods, but it is still
-important to iteratively assess the data generating procedure through
-simulation. Using numerical proxies for visual characteristics of data
-displays such as curvature, linearity, scatter, dispersion can assist
-with identifying optimal parameter settings to use across different
-experimental conditions. Even with this strategy, it is still critical
-to fine-tune the parameter choices with visual calibration and pilot
-testing.
 
@@
 difficult the task at hand is (this strategy is also described by Lu et
-al. (2022)). At some point, all experimenters become so visually
-saturated with the nuances of the data generating mechanism that it may
-become necessary to ``sanity check'' the protocol with family members,
-friends, and colleagues. These informal surveys provide extremely useful
-feedback and can help to counteract the visual saturation of being
-immersed in the design of a visualization experiment for months at a
-time.
+al. (Lu et al., 2022)). At some point, all experimenters become so
+visually saturated with the nuances of the data generating mechanism
+that it may become necessary to ``sanity check'' the protocol with
+family members, friends, and colleagues. These informal surveys provide
+extremely useful feedback and can help to counteract the visual
+saturation of being immersed in the design of a visualization experiment
+for months at a time.
 
@@
 experiment. In support of this assessment, we offer up a cautionary tale
-of our own experience: that of VanderPlas \& Hofmann (2017), where we
+of our own experience: that of (VanderPlas \& Hofmann, 2017), where we
 designed an experiment to test which plot aesthetics promoted discovery
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{lineup-missing-ellipse.png}}
+\includegraphics{lineup-missing-ellipse.png}
 
@@
 \caption{\label{fig-lineup-problems}A lineup from Vanderplas \& Hofmann
-(2017). Panel 10 shows the clustered target data and panel 17 shows the
-target data with a strong linear relationship; either of these target
-panels was the expected choice. Unfortunately, panel 16 has only two
-bounding ellipses shown, which is an unintentional difference that
-resulted from a faulty method for assigning clusters to null plots; many
-participants selected this panel instead of one of the target panels.}
+(VanderPlas \& Hofmann, 2017). Panel 10 shows the clustered target data
+and panel 17 shows the target data with a strong linear relationship;
+either of these target panels was the expected choice. Unfortunately,
+panel 16 has only two bounding ellipses shown, which is an unintentional
+difference that resulted from a faulty method for assigning clusters to
+null plots; many participants selected this panel instead of one of the
+target panels.}
 
@@
 One reason why it is so difficult to generate sampling models for visual
-explorations is that our visual system is optimized for identifying
-differences between groups. This ability can interfere with the natural
-to use the null sampling models that might be used in equivalent
-numerical tests when running experiments that use visualizations. We
-re-ran the experiment using a different clustering method that
-controlled the number of points in each group. Instead of noticing the
-number of ellipses, participants instead used the differences in size
-and shape of the ellipses formed when clustering after the data
-generating procedure. That is, participants could still detect the
-artificial nature of the induced clusters using other features. While it
-can be difficult to get the data generating method right, it is
-essential to conducting visual experiments that generalize well beyond
-the effects shown in a single data set or phenomenon. The time and
-effort invested in this step at the outset of the experiment pays
-dividends when it allows for clear generalization of the experimental
-results to an entire statistical concept rather than a single data set.
+explorations is that our visual system is very, very good at finding
+differences and unexpected results. We re-ran the experiment using a
+different clustering method, and found that instead of noticing the
+number of ellipses, when that variable was removed participants instead
+cued in on the differences in size and shape of the ellipses formed
+using k-means clustering after the data generating procedure. That is,
+participants could still detect the artificial nature of the induced
+clusters using other features. While it can be difficult to get the data
+generating method right, it is essential to conducting visual
+experiments that generalize well beyond the effects shown in a single
+data set or phenomenon. The time and effort invested in this step at the
+outset of the experiment pays dividends when it allows for clear
+generalization of the experimental results to an entire statistical
+concept rather than a single data set.
 
@@
 design course.
-
-\subsection{Cognitive Load}\label{cognitive-load}
 
@@
 
-\subsection{Participant Instructions}\label{participant-instructions}
-
 It can be extremely helpful to include ``practice'' demonstrations of
@@
 were points which were not completed.
-
-\subsection{Accounting for Participant
-Variability}\label{accounting-for-participant-variability}
 
@@
 
-\subsection{Demographic Data}\label{demographic-data}
-
 It may be useful to ask a few more demographic questions about STEM
@@
 requires higher-order mathematical reasoning. This has allowed us to
-make the argument to the ethics committee (institutional review board,
-or IRB) that our research is exempt, as we do not collect enough
-demographic information to identify participants, however, collecting
-reduced demographic information occasionally comes at a cost. One recent
-study examining the use of log scales and exponential data was conducted
-using Prolific, which recruits participants from around the world; we
-required only that participants were fluent in English to participate.
-It was only after the experiment was completed that we realized that
-different countries introduce logarithms as a concept at different
-points during primary and secondary education; it might be that
-individuals in some countries have much more experience with log scales
-than those educated in the United States. We mention this only to point
-out that while every experiment contains a few missed opportunities, it
-is worth giving careful thought to the demographic questions asked of
-participants and what information may be helpful during the analysis
-stage.
-
-\subsection{Trial Allocation to
-Participants}\label{trial-allocation-to-participants}
+make the argument to IRB that our research is exempt, as we do not
+collect enough demographic information to identify participants,
+however, this has occasionally come at some cost. Our recent study
+examining the use of log scales and exponential data was conducted using
+Prolific, which recruits participants from around the world; we required
+only that participants were fluent in English to participate. It was
+only after the experiment was completed that we realized that different
+countries introduce logarithms as a concept at different points during
+primary and secondary education; it might be that individuals in some
+countries have much more experience with log scales than those educated
+in the United States. We mention this only to point out that while every
+experiment contains a few missed opportunities, it is worth giving
+careful thought to the demographic questions asked of participants and
+what information may be helpful during the analysis stage.
 
@@
 
-\subsection{Data Collection Platforms and
-Infrastructure}\label{data-collection-platforms-and-infrastructure}
-
 We have conducted visualization experiments using a wide variety of
-tools: custom web PHP servers running interactive forms for data
-collection, generic web survey platforms (Qualtrics, Google Forms) for
-static graphics, and Shiny applications that control every part of the
-experimental process from instructions to providing completion codes for
-payment while rendering interactive graphics and generating fully
-randomized data for each participant. In our research, Shiny has
-provided the right balance between control over the experimental
-setting, procedure, etc. and the intricate details of web server
-management, however, this balance is likely different for every group
-and potentially for every experiment. Likewise, we have recruited
-participants using services such as \href{https://www.mturk.com/}{Amazon
-Mechanical Turk} and \href{https://www.prolific.com/}{Prolific}, on
-social media sites
-(\href{https://web.archive.org/web/20250228110244/https://www.reddit.com/r/SampleSize/}{Reddit},
-Twitter(X), Mastodon, and BlueSky), and by email; each recruitment
-method has trade-offs between cost, convenience, number of people in the
-sampling frame, and control over participant demographics. Rice et al.
-(2024) recruited participants using nationally representative panel
-samples and found that conclusions from fully representative samples of
-the population can be very different from volunteer samples recruited
-using other services, as many people are unmotivated to engage with
-charts, don't know how to read charts, or impose pre-formed conclusions
-onto visual displays. These individuals are unlikely to participate in
-most ad-hoc volunteer-based sampling methods. While using representative
-panel sampling products offered by organizations such as NORC or Gallup
-is considerably more expensive than samples from Amazon MTurk or
-Prolific, these results suggest that results from sampling methods used
-online may not generalize to the broader population. People who
-participate in studies via online platforms are more technologically
+tools: custom web servers running experiments using PHP black magic,
+Qualtrics surveys for static graphics, and writing full Shiny
+applications that control every part of the experimental process from
+instructions to providing completion codes for payment. In our research,
+Shiny has provided the right balance between control over the
+experimental setting, procedure, etc. and the intricate details of web
+server management, however, this balance is likely different for every
+group and potentially for every experiment. Likewise, we have recruited
+participants using Amazon Mechanical Turk, Prolific, Reddit and other
+social media sites, and email; each recruitment method has trade-offs
+between cost, convenience, and control over demographic variables. Rice
+et al. (2024) recruited participants using NORC panel samples and found
+that demographics in a fully representative sample are extremely
+important. While using representative panel-based sampling methods is
+considerably more expensive, there are generalizability concerns with
+online recruitment methods such as Prolific and Mechanical Turk. People
+who participate in studies via online platforms are more technologically
 sophisticated and educated than the general population, and this bias
 may significantly impact the conclusions.
-
-If conducting an online experiment, we recommend consideration of
-panel-based sampling methods as well as online testing platforms. While
-Amazon MTurk was once the platform of choice for this type of research
-(Heer \& Bostock, 2010), in 2025 Prolific, which is aimed at academic
-experiments, seems to provide better options that ensure that
-experiments are completed by people rather than AI than MTurk. It seems
-likely that the preferred platform for participant recruitment will
-continue to evolve over time, thus, we will recommend that experimenters
-consider available options, compare pricing structures (as these vary
-widely), and consider whether add-on fees for e.g.~demographically
-representative samples are worth the additional cost. An additional
-consideration when choosing between panel samples and online recruitment
-platforms is the constraints placed on the types of questions and
-graphics which can be used. In our experience, it is much easier to run
-an experiment using the online recruitment platforms, which place many
-fewer constraints on both the number and types of questions which can be
-asked and the types of stimuli which can be used in the experiment.
-
-\subsection{Ethical Review of Human Subjects
-Experiments}\label{ethical-review-of-human-subjects-experiments}
-
-In almost any research environment, there are requirements that
-experiments involving human participants undergo review by an ethics
-board to ensure that participants are not harmed by participating in the
-study. We are most familiar with regulations in the United States and
-can only speak to that environment. Most graphics experiments conducted
-in the US fall under the ``exempt'' category of experiments which
-require only basic review and approval. These experiments record only
-basic demographic information, ask participants to complete tasks that
-involve no risk, and experimenters record information which would not be
-embarrassing to participants if exposed. In order to ensure that our
-experiments fall into this category, we will often ensure that
-demographic information is collected and stored separately from any
-identifiable information, such as e.g.~user IDs from the participant
-recruitment platform that allow us to monitor task completion and pay
-participants for their time. Researchers should also ensure that they
-comply with privacy laws such as the European
-\href{https://gdpr-info.eu/chapter-3/}{General Data Protection
-Regulation (GDPR)}; complying with this law while collecting data which
-is not identifiable can be complicated. We recommend consulting with
-your ethics board and institutional recommendations in order to maintain
-legal compliance and safeguard participant privacy.
 
@@
 framework of a generalized linear mixed model (Ju et al., 2024) that has
-higher power than the Rasch models (Andrich, 1988; Lu et al., 2022)
-which were historically recommended for these experiments. In addition,
-similar model structures with different link functions can be used to
-model accuracy, response time, and confidence, if all three types of
+higher power than the Rasch models which were historically recommended
+for these experiments. In addition, similar model structures can be used
+to model accuracy, response time, and confidence, if all three types of
 information are collected from participants during the experiment.
@@
 \& Rubin, 1991; Tourangeau et al., 2000; Ushakov \& Ushakov, 2017). An
-alternative approach is to analyze the data graphically, as shown in
-Figure~\ref{fig-density-rug}, which uses a density plot with rug
-annotations to show individual participant point estimates. Rounding
-effects can clearly be seen in the rug plot, but a kernel density
-calculated using an appropriately selected bandwidth shows clear visual
-differences between linear and log scale charts. In addition, a smaller
-second mode can be seen in both linear and log conditions that
-corresponds to the underlying model value; this suggests that a minority
-of participants fit a mental regression model to the data and use that
-model to estimate rather than estimating based off of the closest point.
-When charts are carefully constructed to account for the experimental
-structure and participant estimation strategies, displaying both
-individual and aggregate responses, it is possible to demonstrate a
-measurable difference between conditions without the need for
-complicated statistical modeling which corrects for rounding and
-anchoring effects.
-
-While Figure~\ref{fig-density-rug} shows point estimates, the same
-approach can be modified to account for experimental methods which
-generate participant response curves. In these cases, rather than adding
-a rug plot to a density plot showing aggregate estimates, it may be more
-helpful to create spaghetti plots of individual estimates with a
-superimposed consensus estimate and relevant annotations showing
-e.g.~anchor points.\\
-The advantage of this approach is that it can accommodate extremely
-messy data without requiring the extensive data cleaning, modeling of
-heuristics like rounding and anchoring, and elimination of nonsensical
-responses that might be necessary to fit a statistical model. While
-statistical models are undoubtedly beneficial in many situations, we
-have often found that graphical displays of experimental results are at
-least as useful for analyzing and presenting the results of graphical
-experiments.
+alternative approach is to analyze the data graphically using e.g.~a
+combination of density and rug geoms for point estimates (as shown in
+Figure~\ref{fig-density-rug}), or spaghetti plots for individual
+response curves, with annotations to show anchor points. The advantage
+of this approach is that it can accommodate extremely messy data without
+requiring the extensive data cleaning and elimination of nonsensical
+responses that might be necessary to fit a statistical model.
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/fig-density-rug-1.pdf}}
+\includegraphics{index_files/figure-pdf/fig-density-rug-1.pdf}
 
@@
 
-In this paper, we have attempted to contextualize and motivate the logic
-behind the process we use to design empirical graphics experiments.
-We've discussed model development, experimental design considerations,
-pilot testing, and data analysis methods that have been honed over many
-successful and less-than-optimal experiments. While no experiment
-involving humans ever goes exactly to plan, following this process helps
-to avoid some of the most likely mishaps, ensuring that each new
-experiment's ``bonus'' findings have only minimal impacts on the study's
-overall utility.
-
-While it can be difficult to conduct empirical tests of different
-visualizations, the results of these experiments support guidelines for
-graphical design and communication of statistical findings in an
-accessible and explainable way. Many chart design guidelines and
-recommendations are based on heuristics, but as scientists, we should
-prefer guidelines which are based on empirical, experimentally derived
-results over opinions. Testing statistical graphics and developing
-empirically supported guidelines for chart creation promises to support
-better scientific communication, which is critical for educating the
-public about topics like climate change, public health, the risk of
-severe weather and more.
-
 \section{References}\label{references}
@@
 
-\bibitem[\citeproctext]{ref-andrichRaschModelsMeasurement1988}
-Andrich, D. (1988). \emph{Rasch models for measurement} (1st ed.).
-{SAGE} Publications.
-
 \bibitem[\citeproctext]{ref-bertin1983semiology}
@@
 \bibitem[\citeproctext]{ref-d3}
-Bostock, M., Ogievetsky, V., \& Heer, J. (2011). D3 {Data}-{Driven}
+Bostock, M., Ogievetsky, V., \& Heer, J. (2011). D³ {Data}-{Driven}
 {Documents}. \emph{IEEE Transactions on Visualization and Computer
@@
 
-\bibitem[\citeproctext]{ref-thinkaloud}
-Haak, M., De Jong, M., \& Schellens, P. (2003). Retrospective vs.
-Concurrent think-aloud protocols: Testing the usability of an online
-library catalogue. \emph{Behaviour \& {IT}}, \emph{22}, 339--351.
-\url{https://doi.org/10.1080/0044929031000}
-
 \bibitem[\citeproctext]{ref-haemerDoubleScalesAre1948}
@@
 
-\bibitem[\citeproctext]{ref-haiderStrategiesDetectingDifference2021}
-Haider, J. D., Pohl, M., Beecham, R., \& Dykes, J. (2021). Strategies
-for detecting difference in~map line-up tasks. In C. Ardito, R.
-Lanzilotti, A. Malizia, H. Petrie, A. Piccinno, G. Desolda, \& K. Inkpen
-(Eds.), \emph{Human-computer interaction -- {INTERACT} 2021} (pp.
-558--578). Springer International Publishing.
-\url{https://doi.org/10.1007/978-3-030-85613-7_36}
-
-\bibitem[\citeproctext]{ref-heerCrowdsourcingGraphicalPerception2010}
-Heer, J., \& Bostock, M. (2010). Crowdsourcing graphical perception:
-Using mechanical turk to assess visualization design. \emph{Proceedings
-of the {SIGCHI} Conference on Human Factors in Computing Systems},
-203--212. \url{https://doi.org/10.1145/1753326.1753357}
-
 \bibitem[\citeproctext]{ref-heitjanIgnorabilityCoarseData1991}
@@
 on Visualization and Computer Graphics}, \emph{18}(12), 2441--2448.
-\url{https://doi.org/10/f4fwkz}
+\url{https://doi.org/f4fwkz}
 
@@
 
-\bibitem[\citeproctext]{ref-kulhavyCartographicExperienceThinking1992}
-Kulhavy, Pridemore, \& Stock. (1992). Cartographic experience and
-thinking aloud about thematic maps. \emph{Cartographica}, \emph{29}(1),
-1--9. \url{https://doi.org/10.3138/H61J-VX35-J6WW-8111}
-
-\bibitem[\citeproctext]{ref-leeHowPeopleMake2016}
-Lee, S., Kim, S.-H., Hung, Y.-H., Lam, H., Kang, Y.-A., \& Yi, J. S.
-(2016). How do people make sense of unfamiliar visualizations?: A
-grounded model of novice's information visualization sensemaking.
-\emph{{IEEE} Transactions on Visualization and Computer Graphics},
-\emph{22}(1), 499--508. \url{https://doi.org/10.1109/TVCG.2015.2467195}
-
 \bibitem[\citeproctext]{ref-loyDiagnosticToolsHierarchical2013}
@@
 
-\bibitem[\citeproctext]{ref-schuttPainfreeAccurateBayesian2016}
-Schütt, H. H., Harmeling, S., Macke, J. H., \& Wichmann, F. A. (2016).
-Painfree and accurate bayesian estimation of psychometric functions for
-(potentially) overdispersed data. \emph{Vision Research}, \emph{122},
-105--123. \url{https://doi.org/10.1016/j.visres.2016.02.002}
-
 \bibitem[\citeproctext]{ref-tourangeau_rips_rasinski_2000}
@@
 
-\bibitem[\citeproctext]{ref-valentinDesigningOptimalBehavioral2024}
-Valentin, S., Kleinegesse, S., Bramley, N. R., Seriès, P., Gutmann, M.
-U., \& Lucas, C. G. (2024). Designing optimal behavioral experiments
-using machine learning. \emph{{eLife}}, \emph{13}.
-\url{https://doi.org/10.7554/eLife.86224}
-
 \bibitem[\citeproctext]{ref-vanderplasDesigningGraphicsRequires2021}

### 2025-03-15 index.qmd
More changes

@@
 When conducting graphical tests, however, there are additional considerations beyond those taught in a standard experimental design course.
+Experimenters must carefully consider how much information participants should have about the experiment, what platform to use to recruit participants, and the experimental infrastructure underlying data collection.
+At the same time, more standard design considerations, like blocking, randomization, sample size, and analysis methodology are also important. 
+In this section, we broadly consider three different components of the experimental design process: infrastructure for data collection and participant recruitment, experimental design, and ethics panel review.
 
-## Cognitive Load
+## Data Collection
 
-The primary human design factor to be aware of is that visual tasks and assessing statistical graphics can be extremely cognitively taxing. 
-In our experience, it is difficult to expect participants to evaluate more than about 15 charts in one sitting. 
-If users are asked to deeply engage and answer multiple questions about each chart, this limit may be lower (8-10), but even with a relatively simple task, as in lineup methods, it is hard for participants to accurately evaluate more than about 15 charts.
-Tasks which are more interactive, such as `You Draw It', may be somewhat easier for participants, but it is unlikely that participants would be willing to complete more than about 20 tasks in one sitting even with tasks that require fewer decisions and more engagement.
-At some point, participants' interest in accuracy will decline, and the researcher is better off using a larger number of participants completing fewer tasks within the window where participants are attentive and engaged.
-
-## Participant Instructions
+### Participant Instructions
 
 It can be extremely helpful to include "practice" demonstrations of the task to show the basic process, logic, and reasoning. 
-While it is tempting to make these tasks fully representative of the type of judgement which will be required of participants, practice tasks which are too close to the experimental task may bias participants; we have found that it works well to have a relatively easy practice task which utilizes a slightly different type of plot and/or type of data than what will be tested in the experiment.
+While it is tempting to make these tasks fully representative of the type of judgement which will be required of participants, practice tasks which are too close to the experimental task may bias participants.
+We have found that it works well to have a relatively easy practice task which utilizes a slightly different type of plot and/or type of data than what will be tested in the experiment. 
 In cases which require interactivity, gif animations of the task being carried out are useful, as are additional visual cues, such as the yellow box used in the `You Draw It' task[^1] to indicate that there were points which were not completed.
+Demonstrations can reduce cognitive load, but it is often critical not to prime participants to focus on the specific effect manipulated during the experiment. 
+Finding the right examples and instructions to use in an experiment is a delicate process: priming participants can reduce the generalizability or relevance of the experiment, but if participants are confused about what question to answer or how to complete the task, response variability and participant cognitive load may be too high to detect an effect.
+Pilot testing with a small group of volunteers who think aloud while completing the experiment can be extremely helpful in identifying problems with participant instructions. 
+However, it is important that some pilot participants have lower levels of mathematical and statistical training that are similar to the participant population.
+One technique that we have used is to wrap pilot testing into a presentation about the experiment to a general audience, perhaps as part of an undergraduate course or undergraduate research experience recruitment activity.
+The experiential component increases participant understanding of the research before the presenter explains the scientific goals behind the project, and it is often easy to get feedback on the experiment at the same time.
 
-## Accounting for Participant Variability
-
-While studies have found some relationship between lineup performance and demographic factors [@vanderplasSpatialReasoningData2016a], these differences are relatively small when participants are recruited from online testing platforms like Amazon Mechanical Turk or Prolific, or when participants are recruited from a university student population.
-Studies have also not found a strong relationship between visual task performance and participant recruitment method [@vanderplasFramedReproducingRevisiting2019], perhaps in part because demographics which use social media sites and demographics completing online research tasks for pay overlap heavily.
-However, when participants are recruited using a statistically representative sample of the wider population, education becomes an extremely useful predictor of ability to effectively read and draw conclusions about a chart [@riceTestingPerceptualAccuracy2024].
-In almost every situation, though, we find that some participants are very good at evaluating graphics and some participants are not; as a result, it can be extremely helpful to use random effects models with an effect for individual participants.
-
-## Demographic Data
+### Demographic Data
 
@@
 
+### Infrastructure
+
+We have conducted visualization experiments using a wide variety of tools: custom web servers running interactive, PHP-based forms, generic web survey platforms (Qualtrics, Google Forms) for static graphics, and Shiny applications that control every part of the experiment interface (instructions, generating completion codes for participant payment, rendering interactive graphics, and generating fully randomized data for each participant).
+In our research, Shiny has provided the right balance between control over the experimental setting, procedure, etc.
+and the intricate details of web server administration and management, however, this balance is likely different for every lab and potentially for every experiment.
+
+### Participant Recruitment
+
+Visualization experiments tend to opt for one of the following sampling strategies, ordered by generalizability to the wider population:
+
+- Validated, representative panels of participants offered by specialized polling groups.
+In the US, this includes [NORC's AmeriSpeak](https://amerispeak.norc.org/us/en/amerispeak/about-amerispeak/panel-design.html) and [Gallup Panels](https://www.gallup.com/analytics/318911/us-social-research.aspx)
+
+- Recruitment platforms designed for crowdsourcing research tasks, including  [Amazon Mechanical Turk](https://www.mturk.com/) and [Prolific](https://www.prolific.com/).
+Researchers may be able to obtain representative samples measured against census data along variables such as race, age, and sex, but participants are often more technologically sophisticated and educated than the general population on these platforms.
+
+- In-person studies often recruit from undergraduate psychology or statistics students because researchers often have access to these populations and participation in studies may be part of a course's experiential learning activities, as in @vanderplasEscapingFlatlandGraphics2024.
+While the data generated from required classroom experiments may have higher variability, and it may be hard to generalize findings beyond undergraduate students, these studies can be conducted much more cheaply than studies conducted through online platforms or panels.
+
+- Social media and email-based recruitment.
+Researchers may post directly to social media sites such as Twitter/X, Mastodon, and BlueSky, forums and discussion sites such as Reddit, or general email lists that may be purchased from universities and other marketing organizations.
+If the goal is to recruit a specific population, such as meterologists or forensic examiners, forums may be an extremely effective way to recruit participants; we have used the [SampleSize subreddit](https://web.archive.org/web/20250228110244/https://www.reddit.com/r/SampleSize/) successfully and obtained participants that, while younger and more highly educated than those recruited from a platform, had similar results on visual estimation tasks [@vanderplasFramedReproducingRevisiting2019].
+
+@riceTestingPerceptualAccuracy2024 recruited participants using nationally representative panel samples and found that conclusions from fully representative samples of the population can be very different from volunteer samples recruited using other services, as many people are unmotivated to engage with charts, don't know how to read charts, or impose pre-formed conclusions onto visual displays.
+While using representative panel sampling products offered by organizations such as NORC or Gallup is considerably more expensive than samples from Amazon MTurk or Prolific, these results suggest that results from sampling methods used online may not generalize to the broader population.
+People who participate in studies via online platforms are more technologically sophisticated and educated than the general population, and this bias may significantly impact the conclusions. 
+However, it is also reasonable to argue that this group of people may be less likely to use graphics at all in daily life. 
+Researchers designing a study should carefully consider whether the goal of the study is to generalize results to the adult population or to a subset of that population who make decisions based on data.
+
+If conducting an online experiment, we recommend consideration of panel-based sampling methods as well as online testing platforms. 
+While Amazon MTurk was once the only large platform for this type of research [@heerCrowdsourcingGraphicalPerception2010], many researchers now prefer Prolific because it has stronger limitations on use of automation tools to complete tasks.
+We recommend that experimenters consider available options, compare pricing structures (as these vary widely), and consider whether add-on fees for e.g.
+demographically representative samples are worth the additional cost.
+It is also important to ensure that the platform supports the type of user engagement required for the study. 
+In our experience, online recruitment platforms are more flexible and allow experimenters to use a much wider range of stimuli and experimental designs, but this sacrifices some ability to generalize results to a wider population.
+
+[^1]: See a gif of testing with `You Draw It' [here](https://i.imgur.com/GM5YSen.gif)
+
+## Experimental Design
+
+Data collection process and infrastructure considerations can be complex in graphics experiments, but they are critically important for ensuring that data is successfully collected and that results can be generalized to the wider population.
+However, there are also several considerations that are related to more standard experimental design concerns, but with twists unique to cognitive science experiments. 
+Experimenters must consider factors like cognitive load when determining how to block and randomize factors and allocate trials to participants. 
+These factors are also extremely important when deciding on the number of participants necessary for an experiment.
+
+## Cognitive Load
+
+In our experience, it is difficult to expect participants to evaluate more than about 15 charts in one sitting. 
+If users are asked to deeply engage and answer multiple questions about each chart, this limit may be lower (8-10), but even with a relatively simple task, as in lineup methods, it is hard for participants to accurately evaluate more than about 15 charts.
+Tasks which are more interactive, such as `You Draw It', may be somewhat easier for participants, but it is unlikely that participants would be willing to complete more than about 20 tasks in one sitting even with tasks that require fewer decisions because these tasks often require more engagement.
+At some point, participants' interest in accuracy or attention to detail will decline, and the researcher is better off using a larger number of participants completing fewer tasks within the window where participants are attentive and engaged.
+As a result, constraints on cognitive load are perhaps the most important factor in determining the sample size necessary for a study.
+
+## Sample Size
+
+It is difficult to offer specific advice on the number of participants to include in the study, because the experimental design, participant engagement level, and the specific factors under investigation have such a large effect on the number of trials one participant can reasonably complete before fatigue effects increase response variance.
+Because there are so many factors which have an effect on the required sample size, explicit power calculations for these experiments can be extremely difficult - experimenters would have to run each experiment once to get the necessary information for a power calculation and then run the **exact same** experiment a second time with the calculated $n$ participants.
+Often, after the pilot experiment is completed, more significant changes are necessary that invalidate the estimates which might contribute to the power calculation.
+Some experts have begun to recommend that instead of pilot-study informed power calculations, experimenters use a more general approach based on the intended analysis method [@brysbaertHowManyParticipants2019].
+
+When designing graphics experiments, we typically optimize sample size according to constraints identified during the iterative development of the experiment's structure.
+First, we optimize the model and data generation method.
+Next, we revise the graphics used in the experiment, removing any extraneous information and ensuring that sufficient cues (ticks, grid lines, labels, colors, aesthetics) are provided across experimental conditions to complete the required graphical task.
+Then, we develop the data collection application and determine what types of questions the user will answer during the experiment.
+This includes both the primary questions of interest and any secondary measures, such as completion time, participant reasoning, confidence level, and so on. 
+As the number of tasks required of the participant increases, the number of trials we can ask participants to complete during a session decreases. 
+We pilot test the application, often using a pool of graduate students, colleagues, social media acquaintances, and conscripted family members. 
+During the pilot test stage, we pick an optimistic number of trials per participant -- for instance, we might ask pilot participants to evaluate 20 graphical lineups (400 total plots), even though we expect to reduce the number to 10 or 15 during the main experiment.
+After pilot participants complete the study, we ask questions about fatigue to determine what range of trials per participant is reasonable; in order to find this point, pilot testing needs to include more trials than we expect people to complete without fatigue effects.
+
+At the end of our pilot study, the power calculation becomes an optimization of constraints imposed by the experimental design (e.g.
+how many blocks in an incomplete block design can we have with the factors under investigation), anticipated analysis method, and budget for the experiment.
+The number of trials a single participant can complete without fatigue affects the block size and strategy by which we allocate trials to each participant.
+In addition, we must consider how long participants take to complete the required number of trials.
+Completion time is used to determine participant compensation (if using a participant recruiting platform). 
+Ethics boards and some recruitment platforms require that participants are paid a reasonable wage for their time (currently, around \$15 US per hour), and platforms may ask for median completion time and automatically reject submissions from participants who are too far under or over the specified time limits; they may also require additional participant payments if the median time estimate is too far below the actual average completion time during the experiment.
+Platforms may also calculate fees based on both the participant payment and number of participants recruited, with additional fees to recruit e.g.
+demographically representative samples; as a result, it can be advantageous to balance cognitive load concerns with the fee structure used by the selected recruitment platform.
+This complex calculus is often not considered in sample size calculations taught in statistics courses, but experimenters do not often have budgets of infinite size to complete experiments, so some financial optimization is necessary in practice.
+This pragmatic sample size calculation may boil down to designing the experiment in a way that allows for collection of data using as many participants as the budget will support.
+
+In psychophysical graphics experiments, such as @luModelingJustNoticeable2022, there are often fewer participants with many more trials per participant (28 participants, 250 trials each); the stimuli in these experiments are often much simpler (e.g.
+one or two plots instead of 20 in a lineup) and engagement is limited to detection rather than estimation or prediction.
+Psychophysics experiments, particularly those designed for analysis using Rasch models, require that each participant assesses a full factorial set of stimuli, whereas analysis with generalized linear mixed models allows for use of incomplete blocks and other strategies that spread the cognitive burden across several participants.
+Online experiments which require more complex engagement with statistical graphics often have between 300 and 600 participants [@robinsonPerceptionCognitiveImplications,], but this varies widely with the experimental design, type of stimuli used, and level of participant engagement required; @loyVariationsQQPlots2016 and @clusters had more than 1300 participants,  while @sineillusion and @vanderplasFramedReproducingRevisiting2019 had fewer than 150 participants.
+In-person experiments [e.g.
+@vanderplasSpatialReasoningData2016] often use fewer participants than online equivalents, but these experiments may require more tasks of each participant, which reflects the increased logistical costs of scheduling participants, supervising the experiment, and entering the data, if tasks are completed on paper.
+In some cases, in-person experiments may combat cognitive fatigue by scheduling multiple visits per participant; this leaves experiments vulnerable to participant drop-out effects, but can effectively balance cognitive load and participant recruitment costs in some situations.
+
 ## Trial Allocation to Participants
@@
 Latin square designs or randomization where possible, but we recognize that this is not always feasible due to the need to maintain participant naivete during some portions of the experiment.
-
-## Data Collection Platforms and Infrastructure
-
-We have conducted visualization experiments using a wide variety of tools: custom web PHP servers running interactive forms for data collection, generic web survey platforms (Qualtrics, Google Forms) for static graphics, and Shiny applications that control every part of the experimental process from instructions to providing completion codes for payment while rendering interactive graphics and generating fully randomized data for each participant.
-In our research, Shiny has provided the right balance between control over the experimental setting, procedure, etc.
-and the intricate details of web server management, however, this balance is likely different for every group and potentially for every experiment.
-Likewise, we have recruited participants using services such as [Amazon Mechanical Turk](https://www.mturk.com/) and [Prolific](https://www.prolific.com/), on social media sites ([Reddit](https://web.archive.org/web/20250228110244/https://www.reddit.com/r/SampleSize/), Twitter(X), Mastodon, and BlueSky), and by email; each recruitment method has trade-offs between cost, convenience, number of people in the sampling frame, and control over participant demographics.
-@riceTestingPerceptualAccuracy2024 recruited participants using nationally representative panel samples and found that conclusions from fully representative samples of the population can be very different from volunteer samples recruited using other services, as many people are unmotivated to engage with charts, don't know how to read charts, or impose pre-formed conclusions onto visual displays.
-These individuals are unlikely to participate in most ad-hoc volunteer-based sampling methods.
-While using representative panel sampling products offered by organizations such as NORC or Gallup is considerably more expensive than samples from Amazon MTurk or Prolific, these results suggest that results from sampling methods used online may not generalize to the broader population.
-People who participate in studies via online platforms are more technologically sophisticated and educated than the general population, and this bias may significantly impact the conclusions.
-
-If conducting an online experiment, we recommend consideration of panel-based sampling methods as well as online testing platforms. 
-While Amazon MTurk was once the platform of choice for this type of research [@heerCrowdsourcingGraphicalPerception2010], in 2025 Prolific, which is aimed at academic experiments, seems to provide better options that ensure that experiments are completed by people rather than AI than MTurk.
-It seems likely that the preferred platform for participant recruitment will continue to evolve over time, thus, we will recommend that experimenters consider available options, compare pricing structures (as these vary widely), and consider whether add-on fees for e.g.
-demographically representative samples are worth the additional cost.
-An additional consideration when choosing between panel samples and online recruitment platforms is the constraints placed on the types of questions and graphics which can be used. 
-In our experience, it is much easier to run an experiment using the online recruitment platforms, which place many fewer constraints on both the number and types of questions which can be asked and the types of stimuli which can be used in the experiment.
-
-[^1]: See a gif of testing with `You Draw It' [here](https://i.imgur.com/GM5YSen.gif)
 

### 2025-03-16 index.qmd
Rearranging, more refs, etc.

@@
 Simply asking people to read quantities off of a graph may not generalize beyond the questions asked or the data used in the chart [@eellsRelativeMeritsCircles1926;@croxtonBarChartsCircle1927;@vonhuhnFurtherStudiesGraphic1927;@croxtonGraphicComparisonsBars1932], but designing a study that is sufficiently robust to those issues requires manipulation or control of so many factors that the amount of participants and trials quickly becomes daunting or unaffordable.
+In addition, when conducting graphics experiments, researchers are in the unusual position of being both the subject matter expert and the statistician, providing an unusual amount of control over not just the experimental design but also the specific treatments, levels, and experimental protocols.
+The amount of choices required to develop, pilot, and run an experiment can be overwhelming. 
 In this paper, we attempt to distill the experience gained from conducting several different types of graphics experiments [@clusters;@sineillusion;@emily-diss;@power;@vanderplasFramedReproducingRevisiting2019;@vanderplasEscapingFlatlandGraphics2024], discussing the use of different testing methods [@vanderplasTestingStatisticalCharts2020], the process of designing a graphical experiment, and analysis of the resulting empirical data.
@@
 @sec-testing-methods discusses different methods for testing graphics, and which methods best address different levels of user engagement. 
-In @sec-model-dev, we discuss the process of developing a data-generating model used to control the statistical features of data in the tested visualizations. 
+In @sec-model-dev, we discuss the process of developing the data-generating model used to control the statistical features of data in the tested visualizations. 
 Model development is a nuanced and iterative process that ultimately determines the success and generalizability of the experimental results. 
-We explore different experimental design considerations in @sec-exp-design and then move to the importance of pilot testing in @sec-pilot-test. 
+In @sec-exp-dev, we discuss the design of the experimental protocol - the choice of platform, number and type of trials, and flow of the experiment. 
+We briefly consider different experimental design considerations in @sec-exp-design, but focus primarily on factors specific to graphics experiments, and then move to the importance of pilot testing in @sec-pilot-test.
 Finally, we  provide some common analysis strategies in @sec-analysis, including strategies for handling the unexpected data features which are so common in graphical testing experiments.
@@
 These statistical indices may also serve as quantitative proxies for the difficulty of the visual task. 
-
 To identify a statistical proxy for visual difficulty that may help with narrowing the parameter space, it can be useful to consider numerical measures used to estimate the same types of visual information that will be assessed in the experiment.
@@
 
-Then, a wide range of potential combinations of parameter values or sampling strategies can be explored and summarized graphically; if the numerical statistic cannot differentiate between the null and target under a condition, it is reasonable to expect that a visual inspection of the data may also not show significant results.
+Then, a wide range of potential combinations of parameter values or sampling strategies can be explored and summarized graphically; if the numerical statistic cannot differentiate between the null and target under a condition, it is reasonable to think that a visual inspection of the data may also not show significant results.
 As with any measure, it is important that difficulty levels span a range from easy to hard; we do not learn anything from finding out that everyone can distinguish all of the combinations.
+This portion of the design is somewhat analogous to selecting a range of doses of a chemical in a dose-response experiment.
 
@@
 At some point, all experimenters become so visually saturated with the nuances of the data generating mechanism that it may become necessary to "sanity check" the protocol with family members, friends, and colleagues.
-These informal surveys provide extremely useful feedback and can help to counteract the visual saturation of being immersed in the design of a visualization experiment for months at a time.
+These informal focus groups  provide extremely useful feedback and can help to counteract the visual saturation of being immersed in the design of a visualization experiment for months at a time.
 
@@
 We cannot overstate the importance of visual assessment of your model stimuli, preferably with fresh eyes. 
-We highly recommend performing several rounds of think-aloud pilot testing before deploying an experiment. 
+We highly recommend performing several rounds of think-aloud pilot testing (focus) before deploying an experiment. 
 In support of this assessment, we offer up a cautionary tale of our own experience: that of @clusters, where we designed an experiment to test which plot aesthetics promoted discovery of linear trends and/or clusters.
 
-The experiment was a full 2x3x3 factorial exploration of three data generating parameters, with 3 replicates at each parameter combination (54 data sets) and 10 aesthetic combinations (for a total of 540 lineups).
+The experiment was a 2x3x3 factorial exploration of three data generating parameters, with 3 replicates at each parameter combination (54 data sets) and 10 aesthetic combinations (for a total of 540 lineups).
 Each lineup had 20 different sub-panels, so we should have carefully visually inspected some 10,800 different panels. 
@@
 
-# Experimental Design Considerations {#sec-exp-design}
+# Experiment Development {#sec-exp-dev}
 
-It would be difficult to develop a full data generating model without some idea of the experimental design: the basic structure of the parameters which are to be manipulated, how the users will be tested, and so on.
+It would be difficult to develop a full data generating model without some idea of the experimental protocol: the basic equipment required for the experiment, some idea of what questions users will answer, where and how data will be collecteed, and so on.
 These experimental design factors are fairly natural for scientists to accumulate over the course of imagining and planning an experiment. 
@@
 Experimenters must carefully consider how much information participants should have about the experiment, what platform to use to recruit participants, and the experimental infrastructure underlying data collection.
-At the same time, more standard design considerations, like blocking, randomization, sample size, and analysis methodology are also important. 
-In this section, we broadly consider three different components of the experimental design process: infrastructure for data collection and participant recruitment, experimental design, and ethics panel review.
+In addition to our brief overview of considerations in our research, @kochariConductingWebBasedExperiments2019 has many helpful suggestions for conducting web-based cognitive and perceptual experiments that also apply to studies of statistical graphics.
 
-## Data Collection
+Of course, more standard statistical design considerations, like blocking, randomization, sample size, and analysis methodology are also important; we will discuss these briefly in @sec-exp-design. 
+Here, we focus primarily on the experiment development process, with critical decisions in a roughly temporal order.
+
+### Infrastructure
+
+We have conducted visualization experiments using a wide variety of tools: custom web servers running interactive, PHP-based forms, generic web survey platforms (Qualtrics, Google Forms) for static graphics, and Shiny applications that control every part of the experiment interface (instructions, generating completion codes for participant payment, rendering interactive graphics, and generating fully randomized data for each participant).
+In our research, Shiny has provided the right balance between control over the experimental setting, procedure, etc.
+and the intricate details of web server administration and management, however, this balance is likely different for every lab and potentially for every experiment.
+
+### Participant Recruitment
+
+There are several different, commonly used methods for recruiting participants for visualization experiments and cognitive experiments more broadly.
+The selection of participant recruitment method depends on the infrastructure which will be used in the experiment, but the choice of participant pool may be more critical to the experiment results than the infrastructure and modality [@uittenhoveLabTestingWebTestingCognitive2023].
+
+We have used each of the following strategies in experiments:
+
+- Validated, representative panels of participants offered by specialized polling groups.
+In the US, this includes [NORC's AmeriSpeak](https://amerispeak.norc.org/us/en/amerispeak/about-amerispeak/panel-design.html) and [Gallup Panels](https://www.gallup.com/analytics/318911/us-social-research.aspx)
+
+- Recruitment platforms designed for crowdsourcing research tasks, including  [Amazon Mechanical Turk](https://www.mturk.com/) and [Prolific](https://www.prolific.com/).
+Researchers may be able to obtain representative samples measured against census data along variables such as race, age, and sex, but participants are often more technologically sophisticated and educated than the general population on these platforms.
+
+- In-person studies.
+In person participants are commonly recruited from undergraduate students, but in some experiments it may be preferable to recruit participants from the local community.
+Students are often convenient for academic studies, as participation in experiments is often a component of introductory course experiential learning activities, as in @vanderplasEscapingFlatlandGraphics2024.
+While the data generated from required classroom experiments may have higher variability, and it may be hard to generalize findings beyond undergraduate students, these studies can be conducted much more cheaply than studies conducted through online platforms or panels.
+Recruiting participants from the community for in-person research is also viable, but can be much more complicated, however, some topics, including research involving persons with specific disabilities, the elderly, or subject-matter experts may require recruiting participants outside the university.
+
+- Social media and email-based recruitment.
+Researchers may post directly to social media sites such as Twitter/X, Mastodon, and BlueSky, forums and discussion sites such as Reddit, or general email lists that may be purchased from universities and other marketing organizations.
+If the goal is to recruit a specific population, such as meterologists or forensic examiners, forums and email lists may be an extremely effective way to recruit participants.
+We have used the [SampleSize subreddit](https://web.archive.org/web/20250228110244/https://www.reddit.com/r/SampleSize/) successfully and obtained participants that, while younger and more highly educated than those recruited from a platform, had similar results on visual estimation tasks [@vanderplasFramedReproducingRevisiting2019].
+When using samples obtained from acquaintances, researchers should try to obtain more diversity than was present in historical papers, such as  @clevelandGraphicalPerceptionTheory1984, where the authors sampled their colleagues (and their wives).
+
+@riceTestingPerceptualAccuracy2024 recruited participants using nationally representative panel samples and found that conclusions from fully representative samples of the population can be very different from volunteer samples recruited using other services, as many people are unmotivated to engage with charts, don't know how to read charts, or impose pre-formed conclusions onto visual displays.
+While using representative panel sampling products offered by organizations such as NORC or Gallup is more expensive than samples from Amazon MTurk or Prolific, these results may imply that results from less representative participant recruitment methods may not generalize to the broader population.
+People who participate in studies via online platforms are more technologically sophisticated and educated than the general population, and this bias may significantly impact the conclusions. 
+However, it is also reasonable to argue that higher education might make someone more likely to use charts and data for decision-making.  
+Researchers designing a study should carefully consider whether the goal of the study is to generalize results to the adult population or to a subset of that population who make decisions based on data.
+
+If data will be collected online, then we recommend considering the pros and cons of panel-based sampling methods and  online recruitment platforms. 
+While Amazon MTurk was once the only large platform for this type of research [@heerCrowdsourcingGraphicalPerception2010], many researchers now prefer Prolific because its participants appear to be more attentive to tasks [@albertComparingAttentionalDisengagement2023;@peerDataQualityPlatforms2022].
+Prolific's focus on academic research rather than developing training data for machine learning and AI means that its policies are tailored for these types of projects and its users may be more interested in science than those on other platforms.
+Experimenters should consider available options, compare pricing structures (as these vary widely), and consider whether add-on fees for e.g.
+demographically representative samples are worth the additional cost.
+It is also important to ensure that the platform supports the type of user engagement required for the study. 
+Online recruitment platforms are more flexible than many panel survey platforms and allow experimenters to use a much wider range of stimuli and experimental designs, but this sacrifices some ability to generalize results to a wider population.
 
@@
 
+[^1]: See a gif of testing with `You Draw It' [here](https://i.imgur.com/GM5YSen.gif)
+
+### Attention Checks and Answer Validation
+
+In a perfect world, all participants would be fully engaged in the experiment, focusing only on that task from start to finish without interruption or distraction. 
+In this world, participants would attentively read the unambiguous instructions for completing the experiment and executing these instructions flawlessly. 
+Unfortunately, we do not live in this perfect world. 
+Here, we focus primarily on mechanisms which can be used to exclude data based on participant noncompliance, rather than mechanisms which are used to automatically withhold participant compensation.  
+There are several mechanisms that can be used to guard against or at least identify participants who are not fully engaged in the experiment or have severely misunderstood the instructions:
+
+- Attention checks - questions inserted into the experiment which appear similarly to actual trials but which instruct participants to select a specific response or complete a trivial task (e.g.
+select the result of 2+2).
+More guidance on attention checks can be found in @muszynskiAttentionChecksHow2023.
+
+- User input validation - mechanisms which reduce the potential parameter space of user inputs to a valid or reasonable set of parameters.
+Examples include not allowing participants to select negative variance values, requiring continuous user-drawn lines on an interactive chart, and even checking to ensure that the user's input is of the correct type.
+
+-
+
 ### Demographic Data
@@
 
-### Infrastructure
+# Experimental Design Considerations {#sec-exp-design}
 
-We have conducted visualization experiments using a wide variety of tools: custom web servers running interactive, PHP-based forms, generic web survey platforms (Qualtrics, Google Forms) for static graphics, and Shiny applications that control every part of the experiment interface (instructions, generating completion codes for participant payment, rendering interactive graphics, and generating fully randomized data for each participant).
-In our research, Shiny has provided the right balance between control over the experimental setting, procedure, etc.
-and the intricate details of web server administration and management, however, this balance is likely different for every lab and potentially for every experiment.
+In this section, we broadly consider three different components of the experimental design process: infrastructure for data collection and participant recruitment, experimental design, and ethics panel review.
 
-### Participant Recruitment
-
-Visualization experiments tend to opt for one of the following sampling strategies, ordered by generalizability to the wider population:
-
-- Validated, representative panels of participants offered by specialized polling groups.
-In the US, this includes [NORC's AmeriSpeak](https://amerispeak.norc.org/us/en/amerispeak/about-amerispeak/panel-design.html) and [Gallup Panels](https://www.gallup.com/analytics/318911/us-social-research.aspx)
-
-- Recruitment platforms designed for crowdsourcing research tasks, including  [Amazon Mechanical Turk](https://www.mturk.com/) and [Prolific](https://www.prolific.com/).
-Researchers may be able to obtain representative samples measured against census data along variables such as race, age, and sex, but participants are often more technologically sophisticated and educated than the general population on these platforms.
-
-- In-person studies often recruit from undergraduate psychology or statistics students because researchers often have access to these populations and participation in studies may be part of a course's experiential learning activities, as in @vanderplasEscapingFlatlandGraphics2024.
-While the data generated from required classroom experiments may have higher variability, and it may be hard to generalize findings beyond undergraduate students, these studies can be conducted much more cheaply than studies conducted through online platforms or panels.
-
-- Social media and email-based recruitment.
-Researchers may post directly to social media sites such as Twitter/X, Mastodon, and BlueSky, forums and discussion sites such as Reddit, or general email lists that may be purchased from universities and other marketing organizations.
-If the goal is to recruit a specific population, such as meterologists or forensic examiners, forums may be an extremely effective way to recruit participants; we have used the [SampleSize subreddit](https://web.archive.org/web/20250228110244/https://www.reddit.com/r/SampleSize/) successfully and obtained participants that, while younger and more highly educated than those recruited from a platform, had similar results on visual estimation tasks [@vanderplasFramedReproducingRevisiting2019].
-
-@riceTestingPerceptualAccuracy2024 recruited participants using nationally representative panel samples and found that conclusions from fully representative samples of the population can be very different from volunteer samples recruited using other services, as many people are unmotivated to engage with charts, don't know how to read charts, or impose pre-formed conclusions onto visual displays.
-While using representative panel sampling products offered by organizations such as NORC or Gallup is considerably more expensive than samples from Amazon MTurk or Prolific, these results suggest that results from sampling methods used online may not generalize to the broader population.
-People who participate in studies via online platforms are more technologically sophisticated and educated than the general population, and this bias may significantly impact the conclusions. 
-However, it is also reasonable to argue that this group of people may be less likely to use graphics at all in daily life. 
-Researchers designing a study should carefully consider whether the goal of the study is to generalize results to the adult population or to a subset of that population who make decisions based on data.
-
-If conducting an online experiment, we recommend consideration of panel-based sampling methods as well as online testing platforms. 
-While Amazon MTurk was once the only large platform for this type of research [@heerCrowdsourcingGraphicalPerception2010], many researchers now prefer Prolific because it has stronger limitations on use of automation tools to complete tasks.
-We recommend that experimenters consider available options, compare pricing structures (as these vary widely), and consider whether add-on fees for e.g.
-demographically representative samples are worth the additional cost.
-It is also important to ensure that the platform supports the type of user engagement required for the study. 
-In our experience, online recruitment platforms are more flexible and allow experimenters to use a much wider range of stimuli and experimental designs, but this sacrifices some ability to generalize results to a wider population.
-
-[^1]: See a gif of testing with `You Draw It' [here](https://i.imgur.com/GM5YSen.gif)
+## Data Collection
 

### 2025-03-17 response.qmd
Updating refs and finishing response

---
title: "Response to Review"
format: pdf
bibliography: refs.bib
---

We thank the reviewer for their careful notes about our manuscript. 
We have addressed many of the issues identified; in some cases, we defer to Dr. Jurgen Symanczyk, who asked that we e.g. not focus as much on statistical lineups because there is at least one other paper slated for WIRE on that topic which will provide much more detail.

Below, please find the point-by-point responses.

- The title refers to “statistical graphics”, however in text authors use “charts”. In fact
on Page 1, line 27: Authors write “charts and data visualization” but why the
distinction?

> In our experience, graphics, charts, and visualizations are used nearly interchangeably, and we often rotate between them to ensure that no single word becomes semantically satiated.

- The authors call the visualization lineup protocol as “statistical lineups” but is this really a good name for it? It has not been called as such before and I don’t think it is a good idea to give unintuitive names (it will confuse future research in the area). If you must give it a name, I would prefer “visualization lineups” but it should really be defined, e.g. “In this paper, visualization lineups refer to ...".

> The term statistical lineups have been used throughout several papers, including @loyDiagnosticToolsHierarchical2013,@vanderplasSpatialReasoningData2016,@loyModelChoiceDiagnostics2017, @vanderplasTestingStatisticalCharts2020, @vanderplasStatisticalSignificanceCalculations2021, @liPlotWorthThousand2024, @robinsonPerceptionCognitiveImplications. 
> We have added some text explaining the origin of the term.

- The tone is sometimes casual, e.g. Page 8, Line 33: “system is very, very good”, Page 13, Line 5: “a combination of density and rug geoms”. These sound colloquial and it is not of an academic convention to write things like “very, very” and “rug geoms”.

> We have addressed these language issues, but note that 'rug geom' is a convention from the grammar of graphics as implemented in ggplot2, describing the type of chart. We've adjusted how we refer to this to clarify that it is a composition of two different plots which show both the aggregate density and the individual observations.

- Citation style issues. E.g., page 8, Line 3 “Lu et al (Lu e al, 2022)” should be just “Lu
et al (2022)”.

> We've fixed these (whoops) and also the citation issue noted further below ([@shah]).

- Being specific, e.g. Page 1 line 43: be explicit that “test charts _with human
evaluation_”.

> We've used slightly different language (test the visualization on humans), but have added this specificity.

- I would appreciate a more informative caption for Figure 1.

>  The caption now reads: "Direct adjustment of a plot in a perceptual task. In this experiment designed to assess the strength of the sine illusion, the user adjusts the plot using - and + buttons, which control the strength of a transformation designed to correct the effect of the sine illusion. When the user is satisfied that the lines are of equal length, they select the ‘Finished’ button to move to the next task. The experiment used a psychophysics experimental design, the method of adjustment, but leveraged the interactive Shiny interface to record the entire sequence of adjustments made by the user for each trial. A demo version of this application can be found at https://shiny.srvanderplas.com/sine-illusion/."

- The description of the lineup itself and its role in visual testing is somewhat abstract. Figure 5 might need to be introduced earlier for better clarity. Similarly, no explanation to how the see-value is calculated – if this is expected knowledge, could authors give recommendation of what literature readers should read to know the appropriate background, otherwise I think it is useful to make this guide self-contained and describe it briefly.

> We specifically were asked to not go into depth about lineups because of a pair of other articles which are intended to be published alongside this article. We have added the appropriate references so that readers can find the calculation details there. We have also included some additional language (which can be removed easily if not desireable) referring readers to the companion article about lineups.

- Page 1 line 46: the concept of internal and external validity is not explained. This might be common knowledge in psychometrics or psychology but I don’t think it is in statistics.

> Well, that is a pity, as the concepts are broadly applicable to many different areas of statistics and were part of my statistical education. In any case, the language has been rephrased to refer to generalizability (external validity) and experimental control (internal validity), which hopefully will capture not only statisticians but also the InfoVis crowd made up of e.g. computer scientists and psychologists.

- Is there a guideline or reference for “think-aloud” processes? It’s a concept that not a lot of statisticians would be familiar with so it would be nice to know more about this and the benefits of this approach.

> We've added several additional citations for think-aloud processes and explained both the informal use in experiment testing and the formal use in data collection.

- Should you assume your audience knows what a “PHP” (page 11, line 18) is? Same goes for Qualtric surveys, Amazon Mechanical Turk, Prolific and Reddit. These have no reference/link and little to no explanation. Given technology platforms can evolve rapidly, perhaps authors would like to give more context so that their paper can be understood by future readers. E.g. hypothetically suppose if Reddit disappears today then researchers in 20 years may not know what Reddit is.

> I have attempted to explain these sites in context wherever possible and link to relevant information. I have emphasized that Reddit is a social media site alongside X, Mastodon, and Bluesky -- while we have not yet attempted to pilot studies on Mastodon/Bluesky, those are the natural replacements for X now that it is becoming less popular. I have screenshots of a recruiting message on X as well as the reddit.com/r/samplesize homepage, which could be included in the paper but seem slightly distracting. I have linked to an Internet Archive of the samplesize homepage to preserve this information for the future, but X blocks archival sites, so a similar archive could not be created for that platform. If the goal is that the paper is still comprehensible in 20 years, I think this should be sufficient - MySpace and Blogger, which have been obsolete for 15+ years, still have wikipedia pages which could be used to understand the references.

- What is IRB? Page 10, Line 43: “make the argument to IRB that our research is exempt”.

> We've added the acronym definition and some additional information clarifying the ethics review process for human subjects experiments, in response to an additional suggestion.

- Also NORC panel is not defined.

> This one is surprisingly tricky - NORC used to stand for National Opinion Research Center, but they got rid of the acronym and just named the organization NORC within the last few years. We've clarified that it's an organization like Gallup and that the basic goal is to have a panel survey rather than a sample recruited from a platform like Prolific/MTurk.

- Rasch models on Page 12, Line 39 not defined nor cited.

> We've added additional citations for Rasch models, but as they're suboptimal relative to mixed models for these types of experiments, we hope that this will be sufficient, as we don't want to spend the space explaining a less powerful model.

- how many participants should be recruited?

> This is so heavily dependent on the experimental design and factors under investigation that it is difficult to offer guidance on sample size. In attempting to add information about sample size, we refactored the paper somewhat, differentiating the protocol development process from the experimental design calculations. Graphics studies are sometimes overwhelming because statisticians do not usually control both the subject matter expert decisions and the statistical design decisions, and separating these two out helps to ensure that power calculations and statistical design decisions are isolated (as much as possible) from protocol decisions like the number of trials per participant, which is typically determined based on fatigue.

- what number of plots should be shown to participants in a lineup?

> We have attempted to not make this paper specific to lineup studies, so we will defer this discussion to the visual inference paper that is also to be submitted to WIRE.

- mention of ethic approval? A reminder that people do need to have ethics approval lined up and linked in their study to use results for publication?

> A section on ethics approval has been added, though we have no experience with this process outside of the United States and so have tried to balance specificity (e.g. exempt research and expedited review) with the fact that many readers will be outside of the US and subject to different regulations.

- can authors comment on financial costs related to participant recruitment?

> Every platform is different and even within platforms, pricing schemes vary by the week or month. We've included some mention of considerations like participant completion time vs. number of participants with regard to platform fees, and have also included some general comments about the relative expense of panel surveys vs. online platforms vs. in-person studies vs. social media recruitment.

- any guide to ensure representative sample in participant recruitment? Do authors adopt particular method or strategies?

> We have added some additional references on the effects of participant sampling on results of studies, as well as the different characteristics of different platforms, though at least one of the studies we cite are published by those platforms, so there are conflict of interest concerns as well.

- what about filtering out obviously non-serious participants?

> A section on response validation and attention checks has been added

- Authors mention usage of Amazon Mechanical Turk, Prolific, Reddit and social media. Could they provide advice on which online recruitment platform to use?

> Additional discussions of the advantages of categories of recruitment platforms has been added, but we have tried to remain as factual as possible about different competitors within those categories, in part because this space is evolving rapidly and if mentioning sites like Reddit is a concern (as above), then recommendations for specific platforms would be similarly problematic. Some discussion of the evolution of platforms over time is included, though, to explain why Amazon Turk is less popular now than it was in the 2010s.

- any advice for the design of interface that participants use?

> This is heavily dependent on the choice of data collection software. Bespoke solutions like Shiny and PHP web applications can be infinitely customized, but platforms like Qualtrics and Google Forms are much more rigid in design. Any advice we could give about the interface would have to handle many different configurations.

- Any examples of qualitative questions asked?

> We have added a few suggestions for qualitative questions, and reorganized the qualitative analysis section to better emphasize the citation to @clusters, which contains a qualitative analysis that complements the quantitative modeling.

- Page 1 line 36: grammar of graphics technically doesn’t “classify” charts but just is a means to specify one.

> Yes, this is true, but the grammar also provides a convenient way to classify charts that separates the mappings from the geometric representation. This is particularly powerful in graphics studies, where different geometrical representations of the same mappings can be compared head-to-head. We've modified the language slightly in the hopes of clarifying what we mean here.

- Page 2 line 46: “target plot with signal” — but there can be a target plot with no signal right?

> We've amended the language to avoid this issue.

-  “automatically transcribed using text-to-speech functions of large models” — why LLM particularly? What about the approaches that pre-dates LLM?

> It is only with the advent of LLMs that these features have reached an accuracy where speech-to-text might be a viable option for think-aloud recording, but we've removed the mention of LLMs in any case.

- Page 6, line 40: what is "(shah?)” supposed to be?

> That's a failed markdown reference that we have fixed. Thanks for catching it!

### 2025-03-17 index.qmd
Updating refs and finishing response

@@
 
+# Experiment Development Life Cycle
+
+Developing a graphics experiment is often a highly iterative process, but it can help to approach the design process by first optimizing the model and data generation method before spending time on optimizing the specific stimuli or customizing the data collection platform.
+This is important because the model parameters and data generation process inform the experiment structure and thus impact decisions made downstream.
+
+Once the model and data generating mechanism are set, it is useful to revisit the primary questions of interest and determine how to measure the responses effectively. 
+Secondary measures, such as response time, free responses, and confidence level should also be determined. 
+These choices will inform the choice of a data collection platform and may also inform the participant recruitment method.
+
+Next, we recommend developing a preliminary data analysis plan, specifying the general category of model which will be used (e.g.
+generalized linear mixed-effects model, t-test) and the contrasts which are most interesting.
+This sets up the experimental design decisions, but also ensures that as the data collection platform and process is developed, any design constraints are considered. 
+Development of the data collection application is the next step, using draft graphics and the set of participant response measures of interest.
+
+There are at least 3 stages of testing in a graphics experiment: informal tests, a pilot study, and the main experiment. 
+The informal tests are critical for identifying issues with the data collection application, but can also be used to calibrate the number of tasks required of each participant. 
+As the number and complexity of tasks increases, the number of trials we can ask participants to complete during a session decreases. 
+The informal testing stage allows researchers to consider the tradeoffs inherent in the decision to reduce the amount of information collected for each task, reduce the number of tasks, or mitigate participant fatigue in other ways.
+
+During preliminary testing, we use an optimistic number of trials per participant, so that we can determine when participants become overly fatigued. 
+For instance, we might ask test participants to evaluate 20 graphical lineups (400 total plots), even though we expect to reduce the number to 10 or 15 during the main experiment. 
+We test the application in individual or focus group sessions, often using graduate students, colleagues, social media acquaintances, and conscripted family members. 
+After these participants complete the study, we ask questions about fatigue to determine what range of trials per participant is reasonable. 
+At the end of our initial experiment tests, we have enough information to determine the basic parameters of the experimental design (e.g.
+how many blocks in an incomplete block design can we have with the factors under investigation).
+The number of trials a single participant can complete without excessive fatigue impacts the number of blocks and the strategy by which we allocate trials to each participant.
+
+In addition, we must consider how long participants take to complete the required number of trials.
+Completion time is used to determine participant compensation (if using a participant recruiting platform). 
+Ethics boards and some recruitment platforms require that participants are paid a reasonable wage for their time (currently, around \$15 US per hour), and platforms may ask for median completion time and automatically reject submissions from participants who are too far under or over the specified time limits; they may also require additional participant payments if the median time estimate is too far below the actual average completion time during the experiment.
+Platforms may also calculate fees based on both the participant payment and number of participants recruited, with additional fees to recruit e.g.
+demographically representative samples; as a result, it can be advantageous to balance cognitive load concerns with the fee structure used by the selected recruitment platform.
+
+The findings from the initial test of the experimental procedure are then used to revise the data collection procedure in preparation for one or more pilot studies.
+It is important to ensure that the software platform, trial allocation, and other components of the experiment are functioning as desired before a formal pilot study is conducted. 
+In some cases, the pilot study is as simple as a "soft launch" of the main experiment, where the total number of trials is pre-specified and only a few trials are released initially to ensure that data collection works as expected.
+In others, the pilot study is conducted first, and results from that study are used to determine the sample size for the main experiment. 
+At this point, data collection, analysis, and reporting proceed much as in any other experiment.
+
 # Developing a Model {#sec-model-dev}
@@
 
-# Experiment Development {#sec-exp-dev}
+# Protocol Development {#sec-exp-dev}
 
@@
 Experimenters must carefully consider how much information participants should have about the experiment, what platform to use to recruit participants, and the experimental infrastructure underlying data collection.
-In addition to our brief overview of considerations in our research, @kochariConductingWebBasedExperiments2019 has many helpful suggestions for conducting web-based cognitive and perceptual experiments that also apply to studies of statistical graphics.
+In addition to our brief overview of considerations in our research, @kochariConductingWebBasedExperiments2019 has many helpful suggestions for conducting web-based cognitive and perceptual studies that also apply to statistical graphics experiments.
 
@@
 
-### Infrastructure
+## Infrastructure
 
@@
 
-### Participant Recruitment
+When using Shiny to collect data, we store participant response data and experiment parameters using SQLite tables, which are automatically synchronized with cloud storage and tracked with version control.
+This ensures that we have incremental records of the tables during data collection, and that all data is stored across multiple locations, guarding against hardware failures. 
+In some studies, data is unique to each participant; in these cases, we highly recommend saving all generated data to a database as well, so that it is possible to go back and examine responses alongside the data used to generate the graphical stimuli.
+Hard drive space is extremely cheap relative to almost any other cost in an experiment; saving all of the data is a sensible measure.
+
+## Participant Recruitment
 
@@
 
-### Participant Instructions
+## Participant Instructions
 
@@
 
-### Attention Checks and Answer Validation
+## Attention Checks and Answer Validation
 
@@
 select the result of 2+2).
+Participants who do not successfully complete a certain percentage of attention check tasks may be excluded from the experiment.
 More guidance on attention checks can be found in @muszynskiAttentionChecksHow2023.
@@
 Examples include not allowing participants to select negative variance values, requiring continuous user-drawn lines on an interactive chart, and even checking to ensure that the user's input is of the correct type.
+Validation mechanisms may also ensure that participants answer all questions before proceeding to the next page.
 
--
+- Time-based checks - @kochariConductingWebBasedExperiments2019 suggests that in tasks with longer instructions, it may be useful to remove participants who did not spend a certain amount of time on the instructions page.
+Similarly, it may be justifiable to remove participants who completed the experiment in an unreasonably short or long amount of time.
+On the other end of the spectrum, participants who take several hours to complete an experiment with a median completion time of 15 minutes may have experienced technical issues or been distracted.
 
-### Demographic Data
+Pre-specifying conditions where participant data will be removed from the study before analysis is critical for predictable issues, in order to streamline the analysis and defend against p-hacking concerns.
+When participants provide anomalous responses that are identified after the experiment, it is much harder, but not impossible, to justify removing the data. 
+Justifications might include that no reasonable viewer would have produced a certain response, as in @sineillusion, or that the response most likely occurred due to a data entry error or a fundamental misunderstanding of the question, but this is only viable if the data is an egregious outlier.
+In less severe cases, it is often preferable to leave the data in, increasing estimate variability but avoiding the need to justify and defend data cleaning decisions.
+
+## Demographic Data
 
@@
 
-# Experimental Design Considerations {#sec-exp-design}
+## Ethics Review
+Once the protocol is developed, researchers generally have enough information to get approval from the ethics board to conduct an experiment on human participants. 
+We are most familiar with regulations in the United States and can speak to that general process. 
+Most graphics experiments conducted in the US fall under the "exempt" category of experiments which require only basic review and approval. 
+These experiments record demographic information at a level that is not individually identifiable, ask participants to complete tasks that involve no risk, and experimenters record information which would not be embarrassing to participants if exposed.
+In order to ensure that our experiments fall into this category, we will often ensure that demographic information and participant responses are collected and stored separately from any identifiable information, such as e.g.
+user IDs from the participant recruitment platform that allow us to monitor task completion and pay participants for their time.
+Researchers should also ensure that they comply with privacy laws such as the European [General Data Protection Regulation (GDPR)](https://gdpr-info.eu/chapter-3/); complying with this law while collecting data which is not identifiable can be complicated.
+We recommend consulting with your ethics board and institutional recommendations in order to maintain legal compliance and safeguard participant privacy.
 
-In this section, we broadly consider three different components of the experimental design process: infrastructure for data collection and participant recruitment, experimental design, and ethics panel review.
+# Experimental Design {#sec-exp-design}
 
-## Data Collection
-
-## Experimental Design
-
-Data collection process and infrastructure considerations can be complex in graphics experiments, but they are critically important for ensuring that data is successfully collected and that results can be generalized to the wider population.
-However, there are also several considerations that are related to more standard experimental design concerns, but with twists unique to cognitive science experiments. 
-Experimenters must consider factors like cognitive load when determining how to block and randomize factors and allocate trials to participants. 
-These factors are also extremely important when deciding on the number of participants necessary for an experiment.
-
-## Cognitive Load
-
-In our experience, it is difficult to expect participants to evaluate more than about 15 charts in one sitting. 
-If users are asked to deeply engage and answer multiple questions about each chart, this limit may be lower (8-10), but even with a relatively simple task, as in lineup methods, it is hard for participants to accurately evaluate more than about 15 charts.
-Tasks which are more interactive, such as `You Draw It', may be somewhat easier for participants, but it is unlikely that participants would be willing to complete more than about 20 tasks in one sitting even with tasks that require fewer decisions because these tasks often require more engagement.
-At some point, participants' interest in accuracy or attention to detail will decline, and the researcher is better off using a larger number of participants completing fewer tasks within the window where participants are attentive and engaged.
-As a result, constraints on cognitive load are perhaps the most important factor in determining the sample size necessary for a study.
-
-## Sample Size
+In statistical design terms, most of our studies involve some type of balanced incomplete block design, where participants are assigned to a subset of experimental conditions which allow for estimation of the full range of effects specified in the model.
+The particular structure of these designs depends heavily on the factorial structure of the study and the contrasts of interest. 
+In some experiments, it is important that participants see each set of data only once, while in others, it is critical that participants see the same set of data represented graphically under multiple conditions, at which point it becomes necessary to manipulate the order of trials to ensure that participants do not have back-to-back trials using the same data, because that might influence the responses.
+When randomization and blocking are used in the experiment, it is often beneficial to simulate the experiment before it is conducted to ensure that the data collection platform is correctly balancing and randomizing trials.
+Of particular concern is that in online experiments with demographic controls, some categories fill faster than others. 
+In extreme circumstances, this can result in imbalances in block allocation within subgroups, which may complicate analysis of demographic data. 
+It is preferable to iron out these issues before the main study begins, rather than trying to patch the data collection software in the middle of the study.
 
 It is difficult to offer specific advice on the number of participants to include in the study, because the experimental design, participant engagement level, and the specific factors under investigation have such a large effect on the number of trials one participant can reasonably complete before fatigue effects increase response variance.
-Because there are so many factors which have an effect on the required sample size, explicit power calculations for these experiments can be extremely difficult - experimenters would have to run each experiment once to get the necessary information for a power calculation and then run the **exact same** experiment a second time with the calculated $n$ participants.
-Often, after the pilot experiment is completed, more significant changes are necessary that invalidate the estimates which might contribute to the power calculation.
+The cognitive fatigue constraints on the experimental design are important, but otherwise, power calculations for graphics experiments are similar to comparable experiments in other disciplines, in that a pilot study provides the necessary inputs to the sample size calculation for the main experiment.
 Some experts have begun to recommend that instead of pilot-study informed power calculations, experimenters use a more general approach based on the intended analysis method [@brysbaertHowManyParticipants2019].
-
-When designing graphics experiments, we typically optimize sample size according to constraints identified during the iterative development of the experiment's structure.
-First, we optimize the model and data generation method.
-Next, we revise the graphics used in the experiment, removing any extraneous information and ensuring that sufficient cues (ticks, grid lines, labels, colors, aesthetics) are provided across experimental conditions to complete the required graphical task.
-Then, we develop the data collection application and determine what types of questions the user will answer during the experiment.
-This includes both the primary questions of interest and any secondary measures, such as completion time, participant reasoning, confidence level, and so on. 
-As the number of tasks required of the participant increases, the number of trials we can ask participants to complete during a session decreases. 
-We pilot test the application, often using a pool of graduate students, colleagues, social media acquaintances, and conscripted family members. 
-During the pilot test stage, we pick an optimistic number of trials per participant -- for instance, we might ask pilot participants to evaluate 20 graphical lineups (400 total plots), even though we expect to reduce the number to 10 or 15 during the main experiment.
-After pilot participants complete the study, we ask questions about fatigue to determine what range of trials per participant is reasonable; in order to find this point, pilot testing needs to include more trials than we expect people to complete without fatigue effects.
-
-At the end of our pilot study, the power calculation becomes an optimization of constraints imposed by the experimental design (e.g.
-how many blocks in an incomplete block design can we have with the factors under investigation), anticipated analysis method, and budget for the experiment.
-The number of trials a single participant can complete without fatigue affects the block size and strategy by which we allocate trials to each participant.
-In addition, we must consider how long participants take to complete the required number of trials.
-Completion time is used to determine participant compensation (if using a participant recruiting platform). 
-Ethics boards and some recruitment platforms require that participants are paid a reasonable wage for their time (currently, around \$15 US per hour), and platforms may ask for median completion time and automatically reject submissions from participants who are too far under or over the specified time limits; they may also require additional participant payments if the median time estimate is too far below the actual average completion time during the experiment.
-Platforms may also calculate fees based on both the participant payment and number of participants recruited, with additional fees to recruit e.g.
-demographically representative samples; as a result, it can be advantageous to balance cognitive load concerns with the fee structure used by the selected recruitment platform.
-This complex calculus is often not considered in sample size calculations taught in statistics courses, but experimenters do not often have budgets of infinite size to complete experiments, so some financial optimization is necessary in practice.
-This pragmatic sample size calculation may boil down to designing the experiment in a way that allows for collection of data using as many participants as the budget will support.
+As power calculations for various statistical methods can be easily found in any experimental design textbook (such as @easterlingFundamentalsStatisticalExperimental2015), here we provide some basic guidelines based on past experiments.
 
@@
 Psychophysics experiments, particularly those designed for analysis using Rasch models, require that each participant assesses a full factorial set of stimuli, whereas analysis with generalized linear mixed models allows for use of incomplete blocks and other strategies that spread the cognitive burden across several participants.
-Online experiments which require more complex engagement with statistical graphics often have between 300 and 600 participants [@robinsonPerceptionCognitiveImplications,], but this varies widely with the experimental design, type of stimuli used, and level of participant engagement required; @loyVariationsQQPlots2016 and @clusters had more than 1300 participants,  while @sineillusion and @vanderplasFramedReproducingRevisiting2019 had fewer than 150 participants.
+These experiments have also been traditionally completed in person, which may also explain the large number of trials each participant is asked to complete.
+
+Online graphics experiments often have between 300 and 600 participants [@robinsonPerceptionCognitiveImplications,], but this varies widely with the experimental design, type of stimuli used, and level of participant engagement required; @loyVariationsQQPlots2016 and @clusters had more than 1300 participants,  while @sineillusion and @vanderplasFramedReproducingRevisiting2019 had fewer than 150 participants.
 In-person experiments [e.g.
-@vanderplasSpatialReasoningData2016] often use fewer participants than online equivalents, but these experiments may require more tasks of each participant, which reflects the increased logistical costs of scheduling participants, supervising the experiment, and entering the data, if tasks are completed on paper.
-In some cases, in-person experiments may combat cognitive fatigue by scheduling multiple visits per participant; this leaves experiments vulnerable to participant drop-out effects, but can effectively balance cognitive load and participant recruitment costs in some situations.
-
-## Trial Allocation to Participants
-
-In statistical design terms, most of our studies involve some type of balanced incomplete block design, where participants are assigned to a subset of experimental conditions which allow for estimation of the full range of effects specified in the model.
-The particular structure of these designs depends heavily on the factorial structure of the study, but we typically arrange participants' trials to ensure that participants see the same data set only once (where possible) and see as many different experimental conditions as possible.
-It is also important to reduce the impact of order effects using either e.g.
-Latin square designs or randomization where possible, but we recognize that this is not always feasible due to the need to maintain participant naivete during some portions of the experiment.
-
-## Ethical Review of Human Subjects Experiments
-
-In almost any research environment, there are requirements that experiments involving human participants undergo review by an ethics board to ensure that participants are not harmed by participating in the study.
-We are most familiar with regulations in the United States and can only speak to that environment.
-Most graphics experiments conducted in the US fall under the "exempt" category of experiments which require only basic review and approval. 
-These experiments record only basic demographic information, ask participants to complete tasks that involve no risk, and experimenters record information which would not be embarrassing to participants if exposed.
-In order to ensure that our experiments fall into this category, we will often ensure that demographic information is collected and stored separately from any identifiable information, such as e.g.
-user IDs from the participant recruitment platform that allow us to monitor task completion and pay participants for their time.
-Researchers should also ensure that they comply with privacy laws such as the European [General Data Protection Regulation (GDPR)](https://gdpr-info.eu/chapter-3/); complying with this law while collecting data which is not identifiable can be complicated.
-We recommend consulting with your ethics board and institutional recommendations in order to maintain legal compliance and safeguard participant privacy.
+@vanderplasSpatialReasoningData2016] often use fewer participants than online equivalents, but these experiments may require more tasks of each participant, reflecting the increased logistical costs of scheduling participants, supervising the experiment, and entering the data, if tasks are completed on paper.
+In some cases, in-person experiments may combat cognitive fatigue by scheduling the experiment across multiple visits; this leaves experiments vulnerable to participant drop-out effects, but can effectively balance cognitive load and participant recruitment costs in some situations.
 
@@
 
-Once the data generating model is set, the experiment is designed, and the charts have been developed, the next step is an extensive round of pilot testing. 
-The goal of pilot testing is to ensure that the experiment is set up properly and that no issues have been overlooked.
-Pilot testing also provides an opportunity to ensure that directions are clear, participants know what they are supposed to be doing, and to estimate task completion time (which determines how much participant will be paid on many online testing platforms).
-Our studies usually go through 2-3 rounds of pilot testing, with at least one of those rounds involving any relatives and friends who are less technologically savvy. 
-We also purposely include talented individuals who can accidentally crash any testing applications. 
-A final round of testing usually includes any and all coworkers and friends who may be available for 10-15 minutes during the work day. 
-Pilot study samples do not need to be representative of any particular population, though we would caution against using exclusively visualization researchers or statisticians in your pilot sample because some issues are less likely to show up with knowledgeable participants.
-In some studies, each participant may see a new set of data, depending on how the study is designed. 
-In these cases, we highly recommend saving all generated data to a database as well, so that it is possible to go back and examine exactly what happened and/or how things went wrong. 
-Hard drive space is extremely cheap relative to almost any other cost in an experiment; saving all of the data is a sensible measure.
+Once the data generating model is set, the protocol is developed, the experiment is designed, and ethics paperwork has been submitted, the next step is another round of testing.
+The goal of the initial sequence of testing is to ensure that the experiment is set up properly and that no issues have been overlooked. 
+Pilot testing also provides an opportunity to ensure that directions are clear, participants know what they are supposed to be doing, and that the designed study has sufficient power to detect an effect.
+Our studies usually go through 2-3 rounds of preliminary testing, with at least one of those rounds involving any relatives and friends who are less technologically savvy. 
+We also purposely include talented individuals who can accidentally crash any testing applications, as a way to harden our data collection software before deployment.
+
+<!-- Pilot study samples do not need to be representative of any particular population, though we would caution against using exclusively visualization researchers or statisticians in your pilot sample because some issues are less likely to show up with knowledgeable participants.
+-->
 
@@
 We have used such scripts in the past to produce automatically updating dashboards or web pages, allowing for real time or near-real time monitoring of data collection efforts. 
-This provides an easy way to summarize completion of the experiment so that individuals can receive credit (if using services like MTurk or Prolific), but also allows interested participants to see the data if recruiting participants who are more likely to be interested in the results, as sometimes happens on social media sites.
-As data collection online can also happen extremely quickly (300 participants in <2h in our most recent Prolific experiment), this can provide an illusion of control over the deluge of data, allowing any issues to be spotted and resolved relatively quickly.
+This provides an easy way to summarize completion of the experiment so that individuals can receive credit (if using services like MTurk or Prolific), but also allows interested participants to see individual results, which can be a factor when recruiting on social media.
+As data collection online can also happen extremely quickly (300 participants in <2h in our most recent Prolific experiment), this can improve data monitoring, allowing any issues to be spotted and resolved quickly.
 If server load is a potential issue, it may also help to release batches of trials over a longer period of time in order to minimize the chance of having to make participants wait for others to complete the task before the server can handle additional connections^[This is the one major drawback to our preferred solution of self-hosting a Shiny server to handle data collection: the free version of Shiny server is limited to about 15 connections at any given time.
 Prolific has recently added rate-limiting functions to the experimental control  platform, which makes controlling the number of active jobs much easier.].
+Batch trial releases can also be used to ensure that participants are recruited across different time zones; in some studies this is beneficial, while in others it may be more important to control the release time to target individuals in e.g.
+North America rather than Europe.
 
@@
 This approach provides useful context as to what participants use to make their decisions, and can be useful when assessing why unexpected responses occurred. 
-We have used word clouds to show overall themes in participant responses successfully in [@clusters]; when paired with an appropriate linear model it became clear that participants were fixating on unequal cluster size as a visual cue.
-In cases where participants are provided with additional utilities such as calculators and scratchpads, it can be useful to select responses from individual participants which illustrate the different types of calculations performed (but analyzing this data quantitatively can be difficult).
+Qualitative questions can be as simple as asking for a free response explanation of the quantitative response(s), but it can also be effective to use qualitative questions to prompt participants about problem solving strategies, interesting features in the data, and more.
+
+We have used word clouds to show overall themes in participant explanations of why specific lineup panels were selected [@clusters]; when paired with an appropriate linear model it became clear that participants were fixating on unequal cluster size as a visual cue.
+In cases where participants are provided with additional utilities such as calculators and scratchpads, it can be useful to select responses from individual participants which illustrate the different types of calculations performed, but analyzing this data quantitatively can be difficult, as it may be incomplete or difficult to code systematically.
 

### 2025-03-17 index.tex
diff files

@@
 \author{Emily Robinson \and Heike Hofmann \and Susan Vanderplas}
-\date{2025-03-14}
+\date{2025-03-16}
 
@@
 or control of so many factors that the amount of participants and trials
-quickly becomes daunting or unaffordable. In this paper, we attempt to
-distill the experience gained from conducting several different types of
-graphics experiments (Hofmann et al., 2012; Robinson, 2022; VanderPlas
-et al., 2019; Vanderplas et al., 2024; VanderPlas \& Hofmann, 2015,
-2017), discussing the use of different testing methods (Vanderplas et
-al., 2020), the process of designing a graphical experiment, and
-analysis of the resulting empirical data. It is our hope that this paper
-will lower the barriers that exist for conducting empirical graphics
-research and reduce the probability of costly mistakes.
+quickly becomes daunting or unaffordable. In addition, when conducting
+graphics experiments, researchers are in the unusual position of being
+both the subject matter expert and the statistician, providing an
+unusual amount of control over not just the experimental design but also
+the specific treatments, levels, and experimental protocols. The amount
+of choices required to develop, pilot, and run an experiment can be
+overwhelming. In this paper, we attempt to distill the experience gained
+from conducting several different types of graphics experiments (Hofmann
+et al., 2012; Robinson, 2022; Vanderplas et al., 2019, 2024; Vanderplas
+\& Hofmann, 2015, 2017), discussing the use of different testing methods
+(Vanderplas et al., 2020), the process of designing a graphical
+experiment, and analysis of the resulting empirical data. It is our hope
+that this paper will lower the barriers that exist for conducting
+empirical graphics research and reduce the probability of costly
+mistakes.
 
@@
 user engagement. In Section~\ref{sec-model-dev}, we discuss the process
-of developing a data-generating model used to control the statistical
+of developing the data-generating model used to control the statistical
 features of data in the tested visualizations. Model development is a
 nuanced and iterative process that ultimately determines the success and
-generalizability of the experimental results. We explore different
-experimental design considerations in Section~\ref{sec-exp-design} and
-then move to the importance of pilot testing in
-Section~\ref{sec-pilot-test}. Finally, we provide some common analysis
-strategies in Section~\ref{sec-analysis}, including strategies for
-handling the unexpected data features which are so common in graphical
-testing experiments.
+generalizability of the experimental results. In
+Section~\ref{sec-exp-dev}, we discuss the design of the experimental
+protocol - the choice of platform, number and type of trials, and flow
+of the experiment. We briefly consider different experimental design
+considerations in Section~\ref{sec-exp-design}, but focus primarily on
+factors specific to graphics experiments, and then move to the
+importance of pilot testing in Section~\ref{sec-pilot-test}. Finally, we
+provide some common analysis strategies in Section~\ref{sec-analysis},
+including strategies for handling the unexpected data features which are
+so common in graphical testing experiments.
 
@@
 of interactive operations. This type of user engagement was used to
-assess the strength of the sine illusion (VanderPlas \& Hofmann, 2015):
+assess the strength of the sine illusion (Vanderplas \& Hofmann, 2015):
 users adjusted the strength of a transformation designed to correct the
@@
 al., 2020), can be calculated using the \texttt{vinference} R package or
-the process described in VanderPlas et al. (2021). The details of this
+the process described in Vanderplas et al. (2021). The details of this
 calculation are beyond the scope of this broader discussion of how to
@@
 constructed from a mixture model that blends the two competing models
-(VanderPlas \& Hofmann, 2017). Viewers are asked to select the panel(s)
+(Vanderplas \& Hofmann, 2017). Viewers are asked to select the panel(s)
 which are most different, and the primary source of information are
@@
 to assess graphical design choices to determine whether they effectively
-emphasize structural differences in the data (VanderPlas \& Hofmann,
+emphasize structural differences in the data (Vanderplas \& Hofmann,
 2017).
@@
 users why a specific panel was chosen has been demonstrated to provide
-rich insight into otherwise confusing numerical results (VanderPlas \&
+rich insight into otherwise confusing numerical results (Vanderplas \&
 Hofmann, 2017).
@@
 ask participants to estimate direct numerical quantities that have
-little contextual information, as done in (VanderPlas et al., 2019) when
+little contextual information, as done in (Vanderplas et al., 2019) when
 assessing the accuracy of framed plots re-created from the Statistical
@@
 
+\section{Experiment Development Life
+Cycle}\label{experiment-development-life-cycle}
+
+Developing a graphics experiment is often a highly iterative process,
+but it can help to approach the design process by first optimizing the
+model and data generation method before spending time on optimizing the
+specific stimuli or customizing the data collection platform. This is
+important because the model parameters and data generation process
+inform the experiment structure and thus impact decisions made
+downstream.
+
+Once the model and data generating mechanism are set, it is useful to
+revisit the primary questions of interest and determine how to measure
+the responses effectively. Secondary measures, such as response time,
+free responses, and confidence level should also be determined. These
+choices will inform the choice of a data collection platform and may
+also inform the participant recruitment method.
+
+Next, we recommend developing a preliminary data analysis plan,
+specifying the general category of model which will be used
+(e.g.~generalized linear mixed-effects model, t-test) and the contrasts
+which are most interesting. This sets up the experimental design
+decisions, but also ensures that as the data collection platform and
+process is developed, any design constraints are considered. Development
+of the data collection application is the next step, using draft
+graphics and the set of participant response measures of interest.
+
+There are at least 3 stages of testing in a graphics experiment:
+informal tests, a pilot study, and the main experiment. The informal
+tests are critical for identifying issues with the data collection
+application, but can also be used to calibrate the number of tasks
+required of each participant. As the number and complexity of tasks
+increases, the number of trials we can ask participants to complete
+during a session decreases. The informal testing stage allows
+researchers to consider the tradeoffs inherent in the decision to reduce
+the amount of information collected for each task, reduce the number of
+tasks, or mitigate participant fatigue in other ways.
+
+During preliminary testing, we use an optimistic number of trials per
+participant, so that we can determine when participants become overly
+fatigued. For instance, we might ask test participants to evaluate 20
+graphical lineups (400 total plots), even though we expect to reduce the
+number to 10 or 15 during the main experiment. We test the application
+in individual or focus group sessions, often using graduate students,
+colleagues, social media acquaintances, and conscripted family members.
+After these participants complete the study, we ask questions about
+fatigue to determine what range of trials per participant is reasonable.
+At the end of our initial experiment tests, we have enough information
+to determine the basic parameters of the experimental design (e.g.~how
+many blocks in an incomplete block design can we have with the factors
+under investigation). The number of trials a single participant can
+complete without excessive fatigue impacts the number of blocks and the
+strategy by which we allocate trials to each participant.
+
+In addition, we must consider how long participants take to complete the
+required number of trials. Completion time is used to determine
+participant compensation (if using a participant recruiting platform).
+Ethics boards and some recruitment platforms require that participants
+are paid a reasonable wage for their time (currently, around \$15 US per
+hour), and platforms may ask for median completion time and
+automatically reject submissions from participants who are too far under
+or over the specified time limits; they may also require additional
+participant payments if the median time estimate is too far below the
+actual average completion time during the experiment. Platforms may also
+calculate fees based on both the participant payment and number of
+participants recruited, with additional fees to recruit
+e.g.~demographically representative samples; as a result, it can be
+advantageous to balance cognitive load concerns with the fee structure
+used by the selected recruitment platform.
+
+The findings from the initial test of the experimental procedure are
+then used to revise the data collection procedure in preparation for one
+or more pilot studies. It is important to ensure that the software
+platform, trial allocation, and other components of the experiment are
+functioning as desired before a formal pilot study is conducted. In some
+cases, the pilot study is as simple as a ``soft launch'' of the main
+experiment, where the total number of trials is pre-specified and only a
+few trials are released initially to ensure that data collection works
+as expected. In others, the pilot study is conducted first, and results
+from that study are used to determine the sample size for the main
+experiment. At this point, data collection, analysis, and reporting
+proceed much as in any other experiment.
+
 \section{Developing a Model}\label{sec-model-dev}
@@
 model that replicates important visual features of the data is extremely
-complex (Cook et al., 2021; Hullman \& Gelman, 2021; VanderPlas, 2021).
+complex (Cook et al., 2021; Hullman \& Gelman, 2021; Vanderplas, 2021).
 
@@
 set (Hofmann et al., 2012), or start from a model and sample data from
-that model generating process (Robinson, 2022; VanderPlas \& Hofmann,
+that model generating process (Robinson, 2022; Vanderplas \& Hofmann,
 2015, 2017). This decision is largely determined by the availability of
@@
 These statistical indices may also serve as quantitative proxies for the
-difficulty of the visual task.
-
-To identify a statistical proxy for visual difficulty that may help with
-narrowing the parameter space, it can be useful to consider numerical
-measures used to estimate the same types of visual information that will
-be assessed in the experiment. For instance, we have used
+difficulty of the visual task. To identify a statistical proxy for
+visual difficulty that may help with narrowing the parameter space, it
+can be useful to consider numerical measures used to estimate the same
+types of visual information that will be assessed in the experiment. For
+instance, we have used
 
@@
 numerical statistic cannot differentiate between the null and target
-under a condition, it is reasonable to expect that a visual inspection
-of the data may also not show significant results. As with any measure,
-it is important that difficulty levels span a range from easy to hard;
-we do not learn anything from finding out that everyone can distinguish
-all of the combinations.
+under a condition, it is reasonable to think that a visual inspection of
+the data may also not show significant results. As with any measure, it
+is important that difficulty levels span a range from easy to hard; we
+do not learn anything from finding out that everyone can distinguish all
+of the combinations. This portion of the design is somewhat analogous to
+selecting a range of doses of a chemical in a dose-response experiment.
 
@@
 become necessary to ``sanity check'' the protocol with family members,
-friends, and colleagues. These informal surveys provide extremely useful
-feedback and can help to counteract the visual saturation of being
-immersed in the design of a visualization experiment for months at a
-time.
+friends, and colleagues. These informal focus groups provide extremely
+useful feedback and can help to counteract the visual saturation of
+being immersed in the design of a visualization experiment for months at
+a time.
 
@@
 stimuli, preferably with fresh eyes. We highly recommend performing
-several rounds of think-aloud pilot testing before deploying an
+several rounds of think-aloud pilot testing (focus) before deploying an
 experiment. In support of this assessment, we offer up a cautionary tale
-of our own experience: that of VanderPlas \& Hofmann (2017), where we
+of our own experience: that of Vanderplas \& Hofmann (2017), where we
 designed an experiment to test which plot aesthetics promoted discovery
@@
 
-The experiment was a full 2x3x3 factorial exploration of three data
+The experiment was a 2x3x3 factorial exploration of three data
 generating parameters, with 3 replicates at each parameter combination
@@
 
-\section{Experimental Design Considerations}\label{sec-exp-design}
+\section{Protocol Development}\label{sec-exp-dev}
 
 It would be difficult to develop a full data generating model without
-some idea of the experimental design: the basic structure of the
-parameters which are to be manipulated, how the users will be tested,
-and so on. These experimental design factors are fairly natural for
-scientists to accumulate over the course of imagining and planning an
-experiment. When conducting graphical tests, however, there are
-additional considerations beyond those taught in a standard experimental
-design course.
-
-\subsection{Cognitive Load}\label{cognitive-load}
-
-The primary human design factor to be aware of is that visual tasks and
-assessing statistical graphics can be extremely cognitively taxing. In
-our experience, it is difficult to expect participants to evaluate more
-than about 15 charts in one sitting. If users are asked to deeply engage
-and answer multiple questions about each chart, this limit may be lower
-(8-10), but even with a relatively simple task, as in lineup methods, it
-is hard for participants to accurately evaluate more than about 15
-charts. Tasks which are more interactive, such as `You Draw It', may be
-somewhat easier for participants, but it is unlikely that participants
-would be willing to complete more than about 20 tasks in one sitting
-even with tasks that require fewer decisions and more engagement. At
-some point, participants' interest in accuracy will decline, and the
-researcher is better off using a larger number of participants
-completing fewer tasks within the window where participants are
-attentive and engaged.
+some idea of the experimental protocol: the basic equipment required for
+the experiment, some idea of what questions users will answer, where and
+how data will be collecteed, and so on. These experimental design
+factors are fairly natural for scientists to accumulate over the course
+of imagining and planning an experiment. When conducting graphical
+tests, however, there are additional considerations beyond those taught
+in a standard experimental design course. Experimenters must carefully
+consider how much information participants should have about the
+experiment, what platform to use to recruit participants, and the
+experimental infrastructure underlying data collection. In addition to
+our brief overview of considerations in our research, Kochari (2019) has
+many helpful suggestions for conducting web-based cognitive and
+perceptual studies that also apply to statistical graphics experiments.
+
+Of course, more standard statistical design considerations, like
+blocking, randomization, sample size, and analysis methodology are also
+important; we will discuss these briefly in
+Section~\ref{sec-exp-design}. Here, we focus primarily on the experiment
+development process, with critical decisions in a roughly temporal
+order.
+
+\subsection{Infrastructure}\label{infrastructure}
+
+We have conducted visualization experiments using a wide variety of
+tools: custom web servers running interactive, PHP-based forms, generic
+web survey platforms (Qualtrics, Google Forms) for static graphics, and
+Shiny applications that control every part of the experiment interface
+(instructions, generating completion codes for participant payment,
+rendering interactive graphics, and generating fully randomized data for
+each participant). In our research, Shiny has provided the right balance
+between control over the experimental setting, procedure, etc. and the
+intricate details of web server administration and management, however,
+this balance is likely different for every lab and potentially for every
+experiment.
+
+When using Shiny to collect data, we store participant response data and
+experiment parameters using SQLite tables, which are automatically
+synchronized with cloud storage and tracked with version control. This
+ensures that we have incremental records of the tables during data
+collection, and that all data is stored across multiple locations,
+guarding against hardware failures. In some studies, data is unique to
+each participant; in these cases, we highly recommend saving all
+generated data to a database as well, so that it is possible to go back
+and examine responses alongside the data used to generate the graphical
+stimuli. Hard drive space is extremely cheap relative to almost any
+other cost in an experiment; saving all of the data is a sensible
+measure.
+
+\subsection{Participant Recruitment}\label{participant-recruitment}
+
+There are several different, commonly used methods for recruiting
+participants for visualization experiments and cognitive experiments
+more broadly. The selection of participant recruitment method depends on
+the infrastructure which will be used in the experiment, but the choice
+of participant pool may be more critical to the experiment results than
+the infrastructure and modality (Uittenhove et al., 2023).
+
+We have used each of the following strategies in experiments:
+
+\begin{itemize}
+\item
+  Validated, representative panels of participants offered by
+  specialized polling groups. In the US, this includes
+  \href{https://amerispeak.norc.org/us/en/amerispeak/about-amerispeak/panel-design.html}{NORC's
+  AmeriSpeak} and
+  \href{https://www.gallup.com/analytics/318911/us-social-research.aspx}{Gallup
+  Panels}
+\item
+  Recruitment platforms designed for crowdsourcing research tasks,
+  including \href{https://www.mturk.com/}{Amazon Mechanical Turk} and
+  \href{https://www.prolific.com/}{Prolific}. Researchers may be able to
+  obtain representative samples measured against census data along
+  variables such as race, age, and sex, but participants are often more
+  technologically sophisticated and educated than the general population
+  on these platforms.
+\item
+  In-person studies. In person participants are commonly recruited from
+  undergraduate students, but in some experiments it may be preferable
+  to recruit participants from the local community. Students are often
+  convenient for academic studies, as participation in experiments is
+  often a component of introductory course experiential learning
+  activities, as in Vanderplas et al. (2024). While the data generated
+  from required classroom experiments may have higher variability, and
+  it may be hard to generalize findings beyond undergraduate students,
+  these studies can be conducted much more cheaply than studies
+  conducted through online platforms or panels. Recruiting participants
+  from the community for in-person research is also viable, but can be
+  much more complicated, however, some topics, including research
+  involving persons with specific disabilities, the elderly, or
+  subject-matter experts may require recruiting participants outside the
+  university.
+\item
+  Social media and email-based recruitment. Researchers may post
+  directly to social media sites such as Twitter/X, Mastodon, and
+  BlueSky, forums and discussion sites such as Reddit, or general email
+  lists that may be purchased from universities and other marketing
+  organizations. If the goal is to recruit a specific population, such
+  as meterologists or forensic examiners, forums and email lists may be
+  an extremely effective way to recruit participants. We have used the
+  \href{https://web.archive.org/web/20250228110244/https://www.reddit.com/r/SampleSize/}{SampleSize
+  subreddit} successfully and obtained participants that, while younger
+  and more highly educated than those recruited from a platform, had
+  similar results on visual estimation tasks (Vanderplas et al., 2019).
+  When using samples obtained from acquaintances, researchers should try
+  to obtain more diversity than was present in historical papers, such
+  as Cleveland \& McGill (1984), where the authors sampled their
+  colleagues (and their wives).
+\end{itemize}
+
+Rice et al. (2024) recruited participants using nationally
+representative panel samples and found that conclusions from fully
+representative samples of the population can be very different from
+volunteer samples recruited using other services, as many people are
+unmotivated to engage with charts, don't know how to read charts, or
+impose pre-formed conclusions onto visual displays. While using
+representative panel sampling products offered by organizations such as
+NORC or Gallup is more expensive than samples from Amazon MTurk or
+Prolific, these results may imply that results from less representative
+participant recruitment methods may not generalize to the broader
+population. People who participate in studies via online platforms are
+more technologically sophisticated and educated than the general
+population, and this bias may significantly impact the conclusions.
+However, it is also reasonable to argue that higher education might make
+someone more likely to use charts and data for decision-making.\\
+Researchers designing a study should carefully consider whether the goal
+of the study is to generalize results to the adult population or to a
+subset of that population who make decisions based on data.
+
+If data will be collected online, then we recommend considering the pros
+and cons of panel-based sampling methods and online recruitment
+platforms. While Amazon MTurk was once the only large platform for this
+type of research (Heer \& Bostock, 2010), many researchers now prefer
+Prolific because its participants appear to be more attentive to tasks
+(Albert \& Smilek, 2023; Peer et al., 2022). Prolific's focus on
+academic research rather than developing training data for machine
+learning and AI means that its policies are tailored for these types of
+projects and its users may be more interested in science than those on
+other platforms. Experimenters should consider available options,
+compare pricing structures (as these vary widely), and consider whether
+add-on fees for e.g.~demographically representative samples are worth
+the additional cost. It is also important to ensure that the platform
+supports the type of user engagement required for the study. Online
+recruitment platforms are more flexible than many panel survey platforms
+and allow experimenters to use a much wider range of stimuli and
+experimental designs, but this sacrifices some ability to generalize
+results to a wider population.
 
@@
 judgement which will be required of participants, practice tasks which
-are too close to the experimental task may bias participants; we have
+are too close to the experimental task may bias participants. We have
 found that it works well to have a relatively easy practice task which
@@
   \href{https://i.imgur.com/GM5YSen.gif}{here}} to indicate that there
-were points which were not completed.
-
-\subsection{Accounting for Participant
-Variability}\label{accounting-for-participant-variability}
-
-While studies have found some relationship between lineup performance
-and demographic factors (VanderPlas \& Hofmann, 2016), these differences
-are relatively small when participants are recruited from online testing
-platforms like Amazon Mechanical Turk or Prolific, or when participants
-are recruited from a university student population. Studies have also
-not found a strong relationship between visual task performance and
-participant recruitment method (VanderPlas et al., 2019), perhaps in
-part because demographics which use social media sites and demographics
-completing online research tasks for pay overlap heavily. However, when
-participants are recruited using a statistically representative sample
-of the wider population, education becomes an extremely useful predictor
-of ability to effectively read and draw conclusions about a chart (Rice
-et al., 2024).\\
-In almost every situation, though, we find that some participants are
-very good at evaluating graphics and some participants are not; as a
-result, it can be extremely helpful to use random effects models with an
-effect for individual participants.
+were points which were not completed. Demonstrations can reduce
+cognitive load, but it is often critical not to prime participants to
+focus on the specific effect manipulated during the experiment. Finding
+the right examples and instructions to use in an experiment is a
+delicate process: priming participants can reduce the generalizability
+or relevance of the experiment, but if participants are confused about
+what question to answer or how to complete the task, response
+variability and participant cognitive load may be too high to detect an
+effect. Pilot testing with a small group of volunteers who think aloud
+while completing the experiment can be extremely helpful in identifying
+problems with participant instructions. However, it is important that
+some pilot participants have lower levels of mathematical and
+statistical training that are similar to the participant population. One
+technique that we have used is to wrap pilot testing into a presentation
+about the experiment to a general audience, perhaps as part of an
+undergraduate course or undergraduate research experience recruitment
+activity. The experiential component increases participant understanding
+of the research before the presenter explains the scientific goals
+behind the project, and it is often easy to get feedback on the
+experiment at the same time.
+
+\subsection{Attention Checks and Answer
+Validation}\label{attention-checks-and-answer-validation}
+
+In a perfect world, all participants would be fully engaged in the
+experiment, focusing only on that task from start to finish without
+interruption or distraction. In this world, participants would
+attentively read the unambiguous instructions for completing the
+experiment and executing these instructions flawlessly. Unfortunately,
+we do not live in this perfect world. Here, we focus primarily on
+mechanisms which can be used to exclude data based on participant
+noncompliance, rather than mechanisms which are used to automatically
+withhold participant compensation.\\
+There are several mechanisms that can be used to guard against or at
+least identify participants who are not fully engaged in the experiment
+or have severely misunderstood the instructions:
+
+\begin{itemize}
+\item
+  Attention checks - questions inserted into the experiment which appear
+  similarly to actual trials but which instruct participants to select a
+  specific response or complete a trivial task (e.g.~select the result
+  of 2+2). Participants who do not successfully complete a certain
+  percentage of attention check tasks may be excluded from the
+  experiment. More guidance on attention checks can be found in
+  Muszyński (2023).
+\item
+  User input validation - mechanisms which reduce the potential
+  parameter space of user inputs to a valid or reasonable set of
+  parameters. Examples include not allowing participants to select
+  negative variance values, requiring continuous user-drawn lines on an
+  interactive chart, and even checking to ensure that the user's input
+  is of the correct type. Validation mechanisms may also ensure that
+  participants answer all questions before proceeding to the next page.
+\item
+  Time-based checks - Kochari (2019) suggests that in tasks with longer
+  instructions, it may be useful to remove participants who did not
+  spend a certain amount of time on the instructions page. Similarly, it
+  may be justifiable to remove participants who completed the experiment
+  in an unreasonably short or long amount of time. On the other end of
+  the spectrum, participants who take several hours to complete an
+  experiment with a median completion time of 15 minutes may have
+  experienced technical issues or been distracted.
+\end{itemize}
+
+Pre-specifying conditions where participant data will be removed from
+the study before analysis is critical for predictable issues, in order
+to streamline the analysis and defend against p-hacking concerns. When
+participants provide anomalous responses that are identified after the
+experiment, it is much harder, but not impossible, to justify removing
+the data. Justifications might include that no reasonable viewer would
+have produced a certain response, as in Vanderplas \& Hofmann (2015), or
+that the response most likely occurred due to a data entry error or a
+fundamental misunderstanding of the question, but this is only viable if
+the data is an egregious outlier. In less severe cases, it is often
+preferable to leave the data in, increasing estimate variability but
+avoiding the need to justify and defend data cleaning decisions.
 
@@
 
-\subsection{Trial Allocation to
-Participants}\label{trial-allocation-to-participants}
-
-In statistical design terms, most of our studies involve some type of
-balanced incomplete block design, where participants are assigned to a
-subset of experimental conditions which allow for estimation of the full
-range of effects specified in the model. The particular structure of
-these designs depends heavily on the factorial structure of the study,
-but we typically arrange participants' trials to ensure that
-participants see the same data set only once (where possible) and see as
-many different experimental conditions as possible. It is also important
-to reduce the impact of order effects using either e.g.~Latin square
-designs or randomization where possible, but we recognize that this is
-not always feasible due to the need to maintain participant naivete
-during some portions of the experiment.
-
-\subsection{Data Collection Platforms and
-Infrastructure}\label{data-collection-platforms-and-infrastructure}
-
-We have conducted visualization experiments using a wide variety of
-tools: custom web PHP servers running interactive forms for data
-collection, generic web survey platforms (Qualtrics, Google Forms) for
-static graphics, and Shiny applications that control every part of the
-experimental process from instructions to providing completion codes for
-payment while rendering interactive graphics and generating fully
-randomized data for each participant. In our research, Shiny has
-provided the right balance between control over the experimental
-setting, procedure, etc. and the intricate details of web server
-management, however, this balance is likely different for every group
-and potentially for every experiment. Likewise, we have recruited
-participants using services such as \href{https://www.mturk.com/}{Amazon
-Mechanical Turk} and \href{https://www.prolific.com/}{Prolific}, on
-social media sites
-(\href{https://web.archive.org/web/20250228110244/https://www.reddit.com/r/SampleSize/}{Reddit},
-Twitter(X), Mastodon, and BlueSky), and by email; each recruitment
-method has trade-offs between cost, convenience, number of people in the
-sampling frame, and control over participant demographics. Rice et al.
-(2024) recruited participants using nationally representative panel
-samples and found that conclusions from fully representative samples of
-the population can be very different from volunteer samples recruited
-using other services, as many people are unmotivated to engage with
-charts, don't know how to read charts, or impose pre-formed conclusions
-onto visual displays. These individuals are unlikely to participate in
-most ad-hoc volunteer-based sampling methods. While using representative
-panel sampling products offered by organizations such as NORC or Gallup
-is considerably more expensive than samples from Amazon MTurk or
-Prolific, these results suggest that results from sampling methods used
-online may not generalize to the broader population. People who
-participate in studies via online platforms are more technologically
-sophisticated and educated than the general population, and this bias
-may significantly impact the conclusions.
-
-If conducting an online experiment, we recommend consideration of
-panel-based sampling methods as well as online testing platforms. While
-Amazon MTurk was once the platform of choice for this type of research
-(Heer \& Bostock, 2010), in 2025 Prolific, which is aimed at academic
-experiments, seems to provide better options that ensure that
-experiments are completed by people rather than AI than MTurk. It seems
-likely that the preferred platform for participant recruitment will
-continue to evolve over time, thus, we will recommend that experimenters
-consider available options, compare pricing structures (as these vary
-widely), and consider whether add-on fees for e.g.~demographically
-representative samples are worth the additional cost. An additional
-consideration when choosing between panel samples and online recruitment
-platforms is the constraints placed on the types of questions and
-graphics which can be used. In our experience, it is much easier to run
-an experiment using the online recruitment platforms, which place many
-fewer constraints on both the number and types of questions which can be
-asked and the types of stimuli which can be used in the experiment.
-
-\subsection{Ethical Review of Human Subjects
-Experiments}\label{ethical-review-of-human-subjects-experiments}
-
-In almost any research environment, there are requirements that
-experiments involving human participants undergo review by an ethics
-board to ensure that participants are not harmed by participating in the
-study. We are most familiar with regulations in the United States and
-can only speak to that environment. Most graphics experiments conducted
-in the US fall under the ``exempt'' category of experiments which
-require only basic review and approval. These experiments record only
-basic demographic information, ask participants to complete tasks that
+\subsection{Ethics Review}\label{ethics-review}
+
+Once the protocol is developed, researchers generally have enough
+information to get approval from the ethics board to conduct an
+experiment on human participants. We are most familiar with regulations
+in the United States and can speak to that general process. Most
+graphics experiments conducted in the US fall under the ``exempt''
+category of experiments which require only basic review and approval.
+These experiments record demographic information at a level that is not
+individually identifiable, ask participants to complete tasks that
 involve no risk, and experimenters record information which would not be
@@
 experiments fall into this category, we will often ensure that
-demographic information is collected and stored separately from any
-identifiable information, such as e.g.~user IDs from the participant
-recruitment platform that allow us to monitor task completion and pay
-participants for their time. Researchers should also ensure that they
-comply with privacy laws such as the European
+demographic information and participant responses are collected and
+stored separately from any identifiable information, such as e.g.~user
+IDs from the participant recruitment platform that allow us to monitor
+task completion and pay participants for their time. Researchers should
+also ensure that they comply with privacy laws such as the European
 \href{https://gdpr-info.eu/chapter-3/}{General Data Protection
@@
 
+\section{Experimental Design}\label{sec-exp-design}
+
+In statistical design terms, most of our studies involve some type of
+balanced incomplete block design, where participants are assigned to a
+subset of experimental conditions which allow for estimation of the full
+range of effects specified in the model. The particular structure of
+these designs depends heavily on the factorial structure of the study
+and the contrasts of interest. In some experiments, it is important that
+participants see each set of data only once, while in others, it is
+critical that participants see the same set of data represented
+graphically under multiple conditions, at which point it becomes
+necessary to manipulate the order of trials to ensure that participants
+do not have back-to-back trials using the same data, because that might
+influence the responses. When randomization and blocking are used in the
+experiment, it is often beneficial to simulate the experiment before it
+is conducted to ensure that the data collection platform is correctly
+balancing and randomizing trials. Of particular concern is that in
+online experiments with demographic controls, some categories fill
+faster than others. In extreme circumstances, this can result in
+imbalances in block allocation within subgroups, which may complicate
+analysis of demographic data. It is preferable to iron out these issues
+before the main study begins, rather than trying to patch the data
+collection software in the middle of the study.
+
+It is difficult to offer specific advice on the number of participants
+to include in the study, because the experimental design, participant
+engagement level, and the specific factors under investigation have such
+a large effect on the number of trials one participant can reasonably
+complete before fatigue effects increase response variance. The
+cognitive fatigue constraints on the experimental design are important,
+but otherwise, power calculations for graphics experiments are similar
+to comparable experiments in other disciplines, in that a pilot study
+provides the necessary inputs to the sample size calculation for the
+main experiment. Some experts have begun to recommend that instead of
+pilot-study informed power calculations, experimenters use a more
+general approach based on the intended analysis method (Brysbaert,
+2019). As power calculations for various statistical methods can be
+easily found in any experimental design textbook (such as Easterling
+(2015)), here we provide some basic guidelines based on past
+experiments.
+
+In psychophysical graphics experiments, such as Lu et al. (2022), there
+are often fewer participants with many more trials per participant (28
+participants, 250 trials each); the stimuli in these experiments are
+often much simpler (e.g.~one or two plots instead of 20 in a lineup) and
+engagement is limited to detection rather than estimation or prediction.
+Psychophysics experiments, particularly those designed for analysis
+using Rasch models, require that each participant assesses a full
+factorial set of stimuli, whereas analysis with generalized linear mixed
+models allows for use of incomplete blocks and other strategies that
+spread the cognitive burden across several participants. These
+experiments have also been traditionally completed in person, which may
+also explain the large number of trials each participant is asked to
+complete.
+
+Online graphics experiments often have between 300 and 600 participants
+(Robinson et al., 2025,), but this varies widely with the experimental
+design, type of stimuli used, and level of participant engagement
+required; Loy et al. (2016) and Vanderplas \& Hofmann (2017) had more
+than 1300 participants, while Vanderplas \& Hofmann (2015) and
+Vanderplas et al. (2019) had fewer than 150 participants. In-person
+experiments (e.g. Vanderplas \& Hofmann, 2016) often use fewer
+participants than online equivalents, but these experiments may require
+more tasks of each participant, reflecting the increased logistical
+costs of scheduling participants, supervising the experiment, and
+entering the data, if tasks are completed on paper. In some cases,
+in-person experiments may combat cognitive fatigue by scheduling the
+experiment across multiple visits; this leaves experiments vulnerable to
+participant drop-out effects, but can effectively balance cognitive load
+and participant recruitment costs in some situations.
+
 \section{Pilot Testing and Quality Assurance}\label{sec-pilot-test}
 
-Once the data generating model is set, the experiment is designed, and
-the charts have been developed, the next step is an extensive round of
-pilot testing. The goal of pilot testing is to ensure that the
-experiment is set up properly and that no issues have been overlooked.
-Pilot testing also provides an opportunity to ensure that directions are
-clear, participants know what they are supposed to be doing, and to
-estimate task completion time (which determines how much participant
-will be paid on many online testing platforms). Our studies usually go
-through 2-3 rounds of pilot testing, with at least one of those rounds
+Once the data generating model is set, the protocol is developed, the
+experiment is designed, and ethics paperwork has been submitted, the
+next step is another round of testing. The goal of the initial sequence
+of testing is to ensure that the experiment is set up properly and that
+no issues have been overlooked. Pilot testing also provides an
+opportunity to ensure that directions are clear, participants know what
+they are supposed to be doing, and that the designed study has
+sufficient power to detect an effect. Our studies usually go through 2-3
+rounds of preliminary testing, with at least one of those rounds
 involving any relatives and friends who are less technologically savvy.
 We also purposely include talented individuals who can accidentally
-crash any testing applications. A final round of testing usually
-includes any and all coworkers and friends who may be available for
-10-15 minutes during the work day. Pilot study samples do not need to be
-representative of any particular population, though we would caution
-against using exclusively visualization researchers or statisticians in
-your pilot sample because some issues are less likely to show up with
-knowledgeable participants. In some studies, each participant may see a
-new set of data, depending on how the study is designed. In these cases,
-we highly recommend saving all generated data to a database as well, so
-that it is possible to go back and examine exactly what happened and/or
-how things went wrong. Hard drive space is extremely cheap relative to
-almost any other cost in an experiment; saving all of the data is a
-sensible measure.
+crash any testing applications, as a way to harden our data collection
+software before deployment.
 
@@
 credit (if using services like MTurk or Prolific), but also allows
-interested participants to see the data if recruiting participants who
-are more likely to be interested in the results, as sometimes happens on
-social media sites. As data collection online can also happen extremely
-quickly (300 participants in \textless2h in our most recent Prolific
-experiment), this can provide an illusion of control over the deluge of
-data, allowing any issues to be spotted and resolved relatively quickly.
-If server load is a potential issue, it may also help to release batches
-of trials over a longer period of time in order to minimize the chance
-of having to make participants wait for others to complete the task
-before the server can handle additional connections\footnote{This is the
-  one major drawback to our preferred solution of self-hosting a Shiny
-  server to handle data collection: the free version of Shiny server is
-  limited to about 15 connections at any given time. Prolific has
-  recently added rate-limiting functions to the experimental control
-  platform, which makes controlling the number of active jobs much
-  easier.}.
+interested participants to see individual results, which can be a factor
+when recruiting on social media. As data collection online can also
+happen extremely quickly (300 participants in \textless2h in our most
+recent Prolific experiment), this can improve data monitoring, allowing
+any issues to be spotted and resolved quickly. If server load is a
+potential issue, it may also help to release batches of trials over a
+longer period of time in order to minimize the chance of having to make
+participants wait for others to complete the task before the server can
+handle additional connections\footnote{This is the one major drawback to
+  our preferred solution of self-hosting a Shiny server to handle data
+  collection: the free version of Shiny server is limited to about 15
+  connections at any given time. Prolific has recently added
+  rate-limiting functions to the experimental control platform, which
+  makes controlling the number of active jobs much easier.}. Batch trial
+releases can also be used to ensure that participants are recruited
+across different time zones; in some studies this is beneficial, while
+in others it may be more important to control the release time to target
+individuals in e.g.~North America rather than Europe.
 
@@
 The see-value approach (Chowdhury et al., 2020; Majumder et al., 2013;
-VanderPlas et al., 2021) is extremely useful for single lineups, but
+Vanderplas et al., 2021) is extremely useful for single lineups, but
 when a series of lineups that are part of a designed experiment are
@@
 summarize the overall effect of various manipulated factors (Hofmann et
-al., 2012; Robinson, 2022; VanderPlas \& Hofmann, 2017). This approach
-also works extremely well for psychometric experiments (VanderPlas \&
+al., 2012; Robinson, 2022; Vanderplas \& Hofmann, 2017). This approach
+also works extremely well for psychometric experiments (Vanderplas \&
 Hofmann, 2015), as psychometric models can be easily fit into the
@@
 participants use to make their decisions, and can be useful when
-assessing why unexpected responses occurred. We have used word clouds to
-show overall themes in participant responses successfully in (VanderPlas
-\& Hofmann, 2017); when paired with an appropriate linear model it
-became clear that participants were fixating on unequal cluster size as
-a visual cue. In cases where participants are provided with additional
+assessing why unexpected responses occurred. Qualitative questions can
+be as simple as asking for a free response explanation of the
+quantitative response(s), but it can also be effective to use
+qualitative questions to prompt participants about problem solving
+strategies, interesting features in the data, and more.
+
+We have used word clouds to show overall themes in participant
+explanations of why specific lineup panels were selected (Vanderplas \&
+Hofmann, 2017); when paired with an appropriate linear model it became
+clear that participants were fixating on unequal cluster size as a
+visual cue. In cases where participants are provided with additional
 utilities such as calculators and scratchpads, it can be useful to
 select responses from individual participants which illustrate the
-different types of calculations performed (but analyzing this data
-quantitatively can be difficult).
+different types of calculations performed, but analyzing this data
+quantitatively can be difficult, as it may be incomplete or difficult to
+code systematically.
 
@@
 \begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-albertComparingAttentionalDisengagement2023}
+Albert, D. A., \& Smilek, D. (2023). Comparing attentional disengagement
+between {Prolific} and {MTurk} samples. \emph{Scientific Reports},
+\emph{13}(1), 20574. \url{https://doi.org/10.1038/s41598-023-46048-5}
+
 \bibitem[\citeproctext]{ref-allenVisualizingScientificData2016}
@@
 
+\bibitem[\citeproctext]{ref-brysbaertHowManyParticipants2019}
+Brysbaert, M. (2019). How many participants do we have to include in
+properly powered experiments? {A} tutorial of power analysis with
+reference tables. \emph{Journal of Cognition}, \emph{2}(1).
+\url{https://doi.org/10.5334/joc.72}
+
 \bibitem[\citeproctext]{ref-bujaStatisticalInferenceExploratory2009}
@@
 
+\bibitem[\citeproctext]{ref-clevelandGraphicalPerceptionTheory1984}
+Cleveland, W. S., \& McGill, R. (1984). Graphical perception: {Theory},
+experimentation, and application to the development of graphical
+methods. \emph{Journal of the American Statistical Association},
+\emph{79}(387), 531--554.
+\url{https://doi.org/10.1080/01621459.1984.10478080}
+
 \bibitem[\citeproctext]{ref-clevelandGraphicalPerceptionGraphical1985}
@@
 365--395.
+
+\bibitem[\citeproctext]{ref-easterlingFundamentalsStatisticalExperimental2015}
+Easterling, R. G. (2015). \emph{Fundamentals of {Statistical
+Experimental Design} and {Analysis}}. John Wiley \& Sons.
 
@@
 
+\bibitem[\citeproctext]{ref-kochariConductingWebBasedExperiments2019}
+Kochari, A. R. (2019). Conducting {Web-Based Experiments} for {Numerical
+Cognition Research}. \emph{Journal of Cognition}, \emph{2}(1), 39.
+\url{https://doi.org/10.5334/joc.85}
+
 \bibitem[\citeproctext]{ref-kosslynGraphDesignEye2006}
@@
 
+\bibitem[\citeproctext]{ref-loyVariationsQQPlots2016}
+Loy, A., Follett, L., \& Hofmann, H. (2016). Variations of {Q-Q Plots}:
+{The Power} of {Our Eyes}! \emph{The American Statistician},
+\emph{70}(2), 202--214.
+\url{https://doi.org/10.1080/00031305.2015.1077728}
+
 \bibitem[\citeproctext]{ref-loyDiagnosticToolsHierarchical2013}
@@
 
+\bibitem[\citeproctext]{ref-muszynskiAttentionChecksHow2023}
+Muszyński, M. (2023). Attention checks and how to use them: {Review} and
+practical recommendations. \emph{Ask: Research and Methods},
+\emph{32}(1), 3--38. \url{https://doi.org/10.18061/ask.v32i1.0001}
+
+\bibitem[\citeproctext]{ref-peerDataQualityPlatforms2022}
+Peer, E., Rothschild, D., Gordon, A., Evernden, Z., \& Damer, E. (2022).
+Data quality of platforms and panels for online behavioral research.
+\emph{Behavior Research Methods}, \emph{54}(4), 1643--1662.
+\url{https://doi.org/10.3758/s13428-021-01694-3}
+
 \bibitem[\citeproctext]{ref-R}
@@
 
+\bibitem[\citeproctext]{ref-robinsonPerceptionCognitiveImplications}
+Robinson, E. A., Howard, R., \& Vanderplas, S. (2025). Perception and
+{Cognitive Implications} of {Logarithmic Scales} for {Exponentially
+Increasing Data}: {Perceptual Sensitivity Tested} with {Statistical
+Lineups}. \emph{Journal of Computational and Graphical Statistics},
+\emph{0}(ja), 1--14. \url{https://doi.org/10.1080/10618600.2025.2476097}
+
 \bibitem[\citeproctext]{ref-schuttPainfreeAccurateBayesian2016}
@@
 
+\bibitem[\citeproctext]{ref-uittenhoveLabTestingWebTestingCognitive2023}
+Uittenhove, D. K., Jeanneret, S., \& Vergauwe, E. (2023). From
+{Lab-Testing} to {Web-Testing} in {Cognitive Research}: {Who You Test}
+is {More Important} than how {You Test}. \emph{Journal of Cognition},
+\emph{6}(1), 13. \url{https://doi.org/10.5334/joc.259}
+
 \bibitem[\citeproctext]{ref-ushakovStatisticalAnalysisRounded2017}
@@
 \bibitem[\citeproctext]{ref-vanderplasDesigningGraphicsRequires2021}
-VanderPlas, S. (2021). Designing {Graphics} {Requires} {Useful}
+Vanderplas, S. (2021). Designing {Graphics} {Requires} {Useful}
 {Experimental} {Testing} {Frameworks} and {Graphics} {Derived} {From}
@@
 \bibitem[\citeproctext]{ref-vanderplasFramedReproducingRevisiting2019}
-VanderPlas, S., Goluch, R., \& Hofmann, H. (2019). Framed! {Reproducing}
+Vanderplas, S., Goluch, R., \& Hofmann, H. (2019). Framed! {Reproducing}
 and {Revisiting} 150 year old charts. \emph{Journal of Computational and
@@
 \bibitem[\citeproctext]{ref-sineillusion}
-VanderPlas, S., \& Hofmann, H. (2015). Signs of the {Sine}
+Vanderplas, S., \& Hofmann, H. (2015). Signs of the {Sine}
 {Illusion}---{Why} {We} {Need} to {Care}. \emph{Journal of Computational
@@
 
-\bibitem[\citeproctext]{ref-vanderplasSpatialReasoningData2016a}
-VanderPlas, S., \& Hofmann, H. (2016). Spatial {Reasoning} and {Data}
-{Displays}. \emph{IEEE Transactions on Visualization and Computer
+\bibitem[\citeproctext]{ref-vanderplasSpatialReasoningData2016}
+Vanderplas, S., \& Hofmann, H. (2016). Spatial {Reasoning} and {Data
+Displays}. \emph{IEEE Transactions on Visualization \& Computer
 Graphics}, \emph{22}(1), 459--468.
-https://doi.org/\url{https://doi.org/10.1109/TVCG.2015.2469125}
+\url{https://doi.org/10.1109/TVCG.2015.2469125}
 
 \bibitem[\citeproctext]{ref-clusters}
-VanderPlas, S., \& Hofmann, H. (2017). Clusters {Beat} {Trend}!?
+Vanderplas, S., \& Hofmann, H. (2017). Clusters {Beat} {Trend}!?
 {Testing} {Feature} {Hierarchy} in {Statistical} {Graphics}.
@@
 \bibitem[\citeproctext]{ref-vanderplasStatisticalSignificanceCalculations2021}
-VanderPlas, S., Röttger, C., Cook, D., \& Hofmann, H. (2021).
+Vanderplas, S., Röttger, C., Cook, D., \& Hofmann, H. (2021).
 Statistical significance calculations for scenarios in visual inference.

### 2025-03-26 index.qmd
minor updates

@@
 
-While there are heuristics, rules-of-thumb, and guidelines [@asme-standards-graphics;@allenVisualizingScientificData2016;@haemerDoubleScalesAre1948;@fewInformationDashboardDesign2006;@kosslynGraphDesignEye2006] for creating useful and visually attractive data displays, the best way to establish the efficacy of various design decisions is to test the visualization on humans, evaluating different variants under controlled conditions [@clevelandShapeParameterTwoVariable1988;@clevelandGraphicalPerceptionGraphical1985].
+While there are heuristics, general guidelines, and best practices [@asme-standards-graphics;@allenVisualizingScientificData2016;@haemerDoubleScalesAre1948;@fewInformationDashboardDesign2006;@kosslynGraphDesignEye2006] for creating useful and visually attractive data displays, the best way to establish the efficacy of various design decisions is to test the visualization on humans, evaluating different variants under controlled conditions [@clevelandShapeParameterTwoVariable1988;@clevelandGraphicalPerceptionGraphical1985].
 Empirical assessments of visualizations, when carefully designed, allow statisticians to determine which representation of the same data is most effective along one or more dimension(s) of interest: estimation or prediction accuracy, within or between group comparisons, response time, and ability to draw real-world decisions are common goals for charts.
@@
 The toolkit used for these experiments is R-based [@R], and includes ggplot2 [@ggplot2] and Shiny [@shiny] as primary components.
-In many experiments, we customized the Shiny interface with JavaScript and d3 [@d3], enabling interactive graphics, use of svgs, and other useful extensions.
+In many experiments, we customized the Shiny interface with JavaScript and D3 [@d3], enabling interactive graphics, use of svgs, and other useful extensions.
 While we prefer this set of tools, most of the observations described here apply to a wide variety of different workflows for graphical experimentation, including in-person experiments.
@@
 Visual inference charts are often called "lineups" in analogy to the criminal procedure where the suspect is placed in a line with several other individuals with similar characteristics. 
-In a graphical lineup procedure, there is a target plot containing the real data, embedded in an array of innocent "null" plots generated through resampling or simulation. 
+In a graphical lineup procedure, there is a target plot containing the real data, embedded in an array of (typically) 19 innocent "null" plots generated through resampling or simulation, for a total of 20 panels.
 If viewers consistently pick the target plot at a higher rate than any of the null plots, the target plot is said to be visually significant [@loyDiagnosticToolsHierarchical2013;@majumderValidationVisualStatistical2013] and a "see" value, the visual analogue of a $p$-value [@niladriroychowdhurySeeValueApp2020], can be calculated using the `vinference` R package or the process described in @vanderplasStatisticalSignificanceCalculations2021.
@@
 In one recent experiment, we asked participants to forecast an exponential trend, with data presented on either a linear or log scale. 
-Using JavaScript code modified from New York Times interactive graphics "You Draw It" features [@katzYouDrawIt2017], we had users draw trend lines with their computer mouse and make forecasts directly on interactive charts, with the data and user-drawn predictions recorded to our database.
+Using JavaScript code modified from New York Times interactive graphics "You Draw It" features [@katzYouDrawIt2017], we had users draw trend lines with their computer mouse and make forecasts directly on interactive charts, with the data and user-drawn predictions recorded to our database [@robinson2023you].
 With interactive graphics rendered using JavaScript (or other web libraries), the only limit to the types of questions one can ask in testing graphics is one's ability to write code to interact with the visualization library.
@@
 For instance, when testing the perception of exponential growth, we had to develop a model which would generate data with varying growth rates, but where the data had a pre-specified domain and range.
-The data generating model is particularly critical when using lineups, as the null sampling model must replicate the important visual features in the data. 
+<!-- The data generating model is particularly critical when using lineups, as the null sampling model must replicate the important visual features in the data.  -->
+If the null plots fail to capture the key visual characteristics - such as trend, spread, or clustering - then any standout visual differences may be attributed to those unintended features, rather than the perceptual cue being tested.
+In other words, if the nulls are too obviously different, participants might detect the real plot for the wrong reason.
 Each testing method has specific requirements, but it is important to carefully calibrate the model parameters to allow for some variability, but not too much, and to ensure that participants can succeed at the task and do not feel like they are being made to analyze random noise.
@@
 To identify a statistical proxy for visual difficulty that may help with narrowing the parameter space, it can be useful to consider numerical measures used to estimate the same types of visual information that will be assessed in the experiment.
-For instance, we have used
+For instance, we have used:
 
-- $R^2$ as a measure of the strength of a linear relationship
-- Gini inequality as a measure of the strength of clustering
-- lack-of-fit statistics to assess the amount of curvature in an exponential relationship (shown in @fig-lof-density-curves)
+- $R^2$ as a measure of the strength of a linear relationship,
+- Gini inequality as a measure of the strength of clustering, and
+- lack-of-fit statistics to assess the amount of curvature in an exponential relationship (shown in @fig-lof-density-curves).
 
@@
 We cannot overstate the importance of visual assessment of your model stimuli, preferably with fresh eyes. 
-We highly recommend performing several rounds of think-aloud pilot testing (focus) before deploying an experiment. 
+We highly recommend performing several rounds of think-aloud pilot testing (e.g., focus groups) before deploying an experiment. 
 In support of this assessment, we offer up a cautionary tale of our own experience: that of @clusters, where we designed an experiment to test which plot aesthetics promoted discovery of linear trends and/or clusters.
@@
 While it can be difficult to get the data generating method right, it is essential to conducting visual experiments that generalize well beyond the effects shown in a single data set or phenomenon.
+This is also why it is critical to include independent replications of the simulated parameters, so that the results reflect variability due to the data generation process, not just the specifics of a single simulated dataset.
 The time and effort invested in this step at the outset of the experiment pays dividends when it allows for clear generalization of the experimental results to an entire statistical concept rather than a single data set.
@@
 This ensures that we have incremental records of the tables during data collection, and that all data is stored across multiple locations, guarding against hardware failures. 
-In some studies, data is unique to each participant; in these cases, we highly recommend saving all generated data to a database as well, so that it is possible to go back and examine responses alongside the data used to generate the graphical stimuli.
+In some studies, generated data is unique to each participant; in these cases, we highly recommend saving all generated data to a database as well, so that it is possible to go back and examine responses alongside the data used to generate the graphical stimuli.
 Hard drive space is extremely cheap relative to almost any other cost in an experiment; saving all of the data is a sensible measure.
@@
 This provides an easy way to summarize completion of the experiment so that individuals can receive credit (if using services like MTurk or Prolific), but also allows interested participants to see individual results, which can be a factor when recruiting on social media.
-As data collection online can also happen extremely quickly (300 participants in <2h in our most recent Prolific experiment), this can improve data monitoring, allowing any issues to be spotted and resolved quickly.
+As data collection online can also happen extremely quickly (300 participants in <2 hours in our most recent Prolific experiment), this can improve data monitoring, allowing any issues to be spotted and resolved quickly.
 If server load is a potential issue, it may also help to release batches of trials over a longer period of time in order to minimize the chance of having to make participants wait for others to complete the task before the server can handle additional connections^[This is the one major drawback to our preferred solution of self-hosting a Shiny server to handle data collection: the free version of Shiny server is limited to about 15 connections at any given time.
@@
 ## Generalized Linear Mixed Models
-The see-value approach [@niladriroychowdhurySeeValueApp2020;@vanderplasStatisticalSignificanceCalculations2021;@majumderValidationVisualStatistical2013] is extremely useful for single lineups, but when a series of lineups that are part of a designed experiment are used, generalized linear mixed models are a much simpler way to summarize the overall effect of various manipulated factors [@power;@clusters;@emily-diss].
+The see-value approach [@niladriroychowdhurySeeValueApp2020;@vanderplasStatisticalSignificanceCalculations2021;@majumderValidationVisualStatistical2013] is extremely useful for single lineups, but when a series of lineups that are part of a designed experiment are used, generalized linear mixed models are a much simpler way to summarize the overall effect of various manipulated factors [@power;@clusters;@robinsonPerceptionCognitiveImplications].
 This approach also works extremely well for psychometric experiments [@sineillusion], as psychometric models can be easily fit into the framework of a generalized linear mixed model [@juOneModelThat2024] that has higher power than the Rasch models [@andrichRaschModelsMeasurement1988;@luModelingJustNoticeable2022] which were historically recommended for these experiments.
@@
 #| label: fig-density-rug
-#| fig-cap: "Density of participant estimates for the year in which the population reaches 4000.
+#| fig-cap: "Density of participant estimates for the year (x-value) in which the population reaches 4000 (y-value).
 Colors are associated to scale - linear (blue) and log (orange) - and vertical lines indicate the true value based on the underlying model equation (black solid) and closest point value based on the simulated data set (black dashed).
@@
 If participants are making predictions and/or fitting visual statistics, we have had success analyzing these responses by comparing the responses to results from a statistical model to determine how visual statistics differ from the numerical quantities derived mathematically.
-For instance, we calculate the deviation between participant responses and the linear regression in `You Draw It' experiments, then fitted generalized additive mixed models to summarize the results across different experimental conditions to assess how user-drawn predictions deviated from the statistical estimates.
+For instance, in @robinson2023eye, we calculate the deviation between participant responses and the linear regression in `You Draw It' experiments, then fitted generalized additive mixed models to summarize the results across different experimental conditions to assess how user-drawn predictions deviated from the statistical estimates.
 In other direct interactions, it may be useful to compare participant selections or annotations to closest points on the chart to assess anchoring behavior; for discrete selections, methods discussed in numerical estimation may also be useful.

### 2025-03-26 index.tex
Update index and diffs

@@
 \author{Emily Robinson \and Heike Hofmann \and Susan Vanderplas}
-\date{2025-03-16}
+\date{2025-03-26}
 
@@
 
-While there are heuristics, rules-of-thumb, and guidelines (Allen \&
-Erhardt, 2016; Few, 2006; Haemer, 1948; {``Joint Committee on Standards
-for Graphic Presentation,''} 1915; Kosslyn, 2006) for creating useful
-and visually attractive data displays, the best way to establish the
-efficacy of various design decisions is to test the visualization on
+While there are heuristics, general guidelines, and best practices
+(Allen \& Erhardt, 2016; Few, 2006; Haemer, 1948; {``Joint Committee on
+Standards for Graphic Presentation,''} 1915; Kosslyn, 2006) for creating
+useful and visually attractive data displays, the best way to establish
+the efficacy of various design decisions is to test the visualization on
 humans, evaluating different variants under controlled conditions
@@
 Shiny (Chang et al., 2021) as primary components. In many experiments,
-we customized the Shiny interface with JavaScript and d3 (Bostock et
+we customized the Shiny interface with JavaScript and D3 (Bostock et
 al., 2011), enabling interactive graphics, use of svgs, and other useful
@@
 graphical lineup procedure, there is a target plot containing the real
-data, embedded in an array of innocent ``null'' plots generated through
-resampling or simulation. If viewers consistently pick the target plot
-at a higher rate than any of the null plots, the target plot is said to
-be visually significant (Loy \& Hofmann, 2013; Majumder et al., 2013)
-and a ``see'' value, the visual analogue of a \(p\)-value (Chowdhury et
-al., 2020), can be calculated using the \texttt{vinference} R package or
-the process described in Vanderplas et al. (2021). The details of this
-calculation are beyond the scope of this broader discussion of how to
-test charts, but more detail on visual inference is provided in
-\textless insert citation to visual inference WIRE article under
-development\textgreater.
+data, embedded in an array of (typically) 19 innocent ``null'' plots
+generated through resampling or simulation, for a total of 20 panels. If
+viewers consistently pick the target plot at a higher rate than any of
+the null plots, the target plot is said to be visually significant (Loy
+\& Hofmann, 2013; Majumder et al., 2013) and a ``see'' value, the visual
+analogue of a \(p\)-value (Chowdhury et al., 2020), can be calculated
+using the \texttt{vinference} R package or the process described in
+Vanderplas et al. (2021). The details of this calculation are beyond the
+scope of this broader discussion of how to test charts, but more detail
+on visual inference is provided in \textless insert citation to visual
+inference WIRE article under development\textgreater.
 
@@
 on interactive charts, with the data and user-drawn predictions recorded
-to our database. With interactive graphics rendered using JavaScript (or
-other web libraries), the only limit to the types of questions one can
-ask in testing graphics is one's ability to write code to interact with
-the visualization library. This type of testing method can be extremely
-natural for participants, but it also is hard to generalize when
-discussing testing methods because of the potential range of
-applications where it might be employed.
+to our database (Robinson et al., 2023b). With interactive graphics
+rendered using JavaScript (or other web libraries), the only limit to
+the types of questions one can ask in testing graphics is one's ability
+to write code to interact with the visualization library. This type of
+testing method can be extremely natural for participants, but it also is
+hard to generalize when discussing testing methods because of the
+potential range of applications where it might be employed.
 
@@
 with varying growth rates, but where the data had a pre-specified domain
-and range. The data generating model is particularly critical when using
-lineups, as the null sampling model must replicate the important visual
-features in the data. Each testing method has specific requirements, but
-it is important to carefully calibrate the model parameters to allow for
-some variability, but not too much, and to ensure that participants can
-succeed at the task and do not feel like they are being made to analyze
-random noise. This Goldilocks-style problem is the focus of the next
-section.
+and range. If the null plots fail to capture the key visual
+characteristics - such as trend, spread, or clustering - then any
+standout visual differences may be attributed to those unintended
+features, rather than the perceptual cue being tested. In other words,
+if the nulls are too obviously different, participants might detect the
+real plot for the wrong reason. Each testing method has specific
+requirements, but it is important to carefully calibrate the model
+parameters to allow for some variability, but not too much, and to
+ensure that participants can succeed at the task and do not feel like
+they are being made to analyze random noise. This Goldilocks-style
+problem is the focus of the next section.
 
@@
 types of visual information that will be assessed in the experiment. For
-instance, we have used
+instance, we have used:
 
@@
 \item
-  \(R^2\) as a measure of the strength of a linear relationship
+  \(R^2\) as a measure of the strength of a linear relationship,
 \item
-  Gini inequality as a measure of the strength of clustering
+  Gini inequality as a measure of the strength of clustering, and
 \item
@@
   exponential relationship (shown in
-  Figure~\ref{fig-lof-density-curves})
+  Figure~\ref{fig-lof-density-curves}).
 \end{itemize}
@@
 stimuli, preferably with fresh eyes. We highly recommend performing
-several rounds of think-aloud pilot testing (focus) before deploying an
-experiment. In support of this assessment, we offer up a cautionary tale
-of our own experience: that of Vanderplas \& Hofmann (2017), where we
-designed an experiment to test which plot aesthetics promoted discovery
-of linear trends and/or clusters.
+several rounds of think-aloud pilot testing (e.g., focus groups) before
+deploying an experiment. In support of this assessment, we offer up a
+cautionary tale of our own experience: that of Vanderplas \& Hofmann
+(2017), where we designed an experiment to test which plot aesthetics
+promoted discovery of linear trends and/or clusters.
 
@@
 essential to conducting visual experiments that generalize well beyond
-the effects shown in a single data set or phenomenon. The time and
-effort invested in this step at the outset of the experiment pays
-dividends when it allows for clear generalization of the experimental
-results to an entire statistical concept rather than a single data set.
+the effects shown in a single data set or phenomenon. This is also why
+it is critical to include independent replications of the simulated
+parameters, so that the results reflect variability due to the data
+generation process, not just the specifics of a single simulated
+dataset. The time and effort invested in this step at the outset of the
+experiment pays dividends when it allows for clear generalization of the
+experimental results to an entire statistical concept rather than a
+single data set.
 
@@
 collection, and that all data is stored across multiple locations,
-guarding against hardware failures. In some studies, data is unique to
-each participant; in these cases, we highly recommend saving all
-generated data to a database as well, so that it is possible to go back
-and examine responses alongside the data used to generate the graphical
-stimuli. Hard drive space is extremely cheap relative to almost any
-other cost in an experiment; saving all of the data is a sensible
-measure.
+guarding against hardware failures. In some studies, generated data is
+unique to each participant; in these cases, we highly recommend saving
+all generated data to a database as well, so that it is possible to go
+back and examine responses alongside the data used to generate the
+graphical stimuli. Hard drive space is extremely cheap relative to
+almost any other cost in an experiment; saving all of the data is a
+sensible measure.
 
@@
 However, it is also reasonable to argue that higher education might make
-someone more likely to use charts and data for decision-making.\\
+someone more likely to use charts and data for decision-making.
 Researchers designing a study should carefully consider whether the goal
@@
 when recruiting on social media. As data collection online can also
-happen extremely quickly (300 participants in \textless2h in our most
-recent Prolific experiment), this can improve data monitoring, allowing
-any issues to be spotted and resolved quickly. If server load is a
-potential issue, it may also help to release batches of trials over a
-longer period of time in order to minimize the chance of having to make
-participants wait for others to complete the task before the server can
-handle additional connections\footnote{This is the one major drawback to
-  our preferred solution of self-hosting a Shiny server to handle data
-  collection: the free version of Shiny server is limited to about 15
-  connections at any given time. Prolific has recently added
+happen extremely quickly (300 participants in \textless2 hours in our
+most recent Prolific experiment), this can improve data monitoring,
+allowing any issues to be spotted and resolved quickly. If server load
+is a potential issue, it may also help to release batches of trials over
+a longer period of time in order to minimize the chance of having to
+make participants wait for others to complete the task before the server
+can handle additional connections\footnote{This is the one major
+  drawback to our preferred solution of self-hosting a Shiny server to
+  handle data collection: the free version of Shiny server is limited to
+  about 15 connections at any given time. Prolific has recently added
   rate-limiting functions to the experimental control platform, which
@@
 summarize the overall effect of various manipulated factors (Hofmann et
-al., 2012; Robinson, 2022; Vanderplas \& Hofmann, 2017). This approach
-also works extremely well for psychometric experiments (Vanderplas \&
-Hofmann, 2015), as psychometric models can be easily fit into the
-framework of a generalized linear mixed model (Ju et al., 2024) that has
-higher power than the Rasch models (Andrich, 1988; Lu et al., 2022)
-which were historically recommended for these experiments. In addition,
-similar model structures with different link functions can be used to
-model accuracy, response time, and confidence, if all three types of
-information are collected from participants during the experiment.
+al., 2012; Robinson et al., 2025; Vanderplas \& Hofmann, 2017). This
+approach also works extremely well for psychometric experiments
+(Vanderplas \& Hofmann, 2015), as psychometric models can be easily fit
+into the framework of a generalized linear mixed model (Ju et al., 2024)
+that has higher power than the Rasch models (Andrich, 1988; Lu et al.,
+2022) which were historically recommended for these experiments. In
+addition, similar model structures with different link functions can be
+used to model accuracy, response time, and confidence, if all three
+types of information are collected from participants during the
+experiment.
 
@@
 superimposed consensus estimate and relevant annotations showing
-e.g.~anchor points.\\
-The advantage of this approach is that it can accommodate extremely
-messy data without requiring the extensive data cleaning, modeling of
-heuristics like rounding and anchoring, and elimination of nonsensical
-responses that might be necessary to fit a statistical model. While
-statistical models are undoubtedly beneficial in many situations, we
-have often found that graphical displays of experimental results are at
-least as useful for analyzing and presenting the results of graphical
-experiments.
+e.g.~anchor points. The advantage of this approach is that it can
+accommodate extremely messy data without requiring the extensive data
+cleaning, modeling of heuristics like rounding and anchoring, and
+elimination of nonsensical responses that might be necessary to fit a
+statistical model. While statistical models are undoubtedly beneficial
+in many situations, we have often found that graphical displays of
+experimental results are at least as useful for analyzing and presenting
+the results of graphical experiments.
 
@@
 \caption{\label{fig-density-rug}Density of participant estimates for the
-year in which the population reaches 4000. Colors are associated to
-scale - linear (blue) and log (orange) - and vertical lines indicate the
-true value based on the underlying model equation (black solid) and
-closest point value based on the simulated data set (black dashed). A
-jittered rug plot along the \(x\)-axis shows where participant estimates
-were made. The plot shows anchoring occurred to the closest point as
-shown by an increase in density around the dashed line. Density peaks
-occurred at whole values indicating rounding errors.}
+year (x-value) in which the population reaches 4000 (y-value). Colors
+are associated to scale - linear (blue) and log (orange) - and vertical
+lines indicate the true value based on the underlying model equation
+(black solid) and closest point value based on the simulated data set
+(black dashed). A jittered rug plot along the \(x\)-axis shows where
+participant estimates were made. The plot shows anchoring occurred to
+the closest point as shown by an increase in density around the dashed
+line. Density peaks occurred at whole values indicating rounding
+errors.}
 
@@
 differ from the numerical quantities derived mathematically. For
-instance, we calculate the deviation between participant responses and
-the linear regression in `You Draw It' experiments, then fitted
-generalized additive mixed models to summarize the results across
-different experimental conditions to assess how user-drawn predictions
-deviated from the statistical estimates. In other direct interactions,
-it may be useful to compare participant selections or annotations to
-closest points on the chart to assess anchoring behavior; for discrete
-selections, methods discussed in numerical estimation may also be
-useful.
+instance, in Robinson et al. (2023a), we calculate the deviation between
+participant responses and the linear regression in `You Draw It'
+experiments, then fitted generalized additive mixed models to summarize
+the results across different experimental conditions to assess how
+user-drawn predictions deviated from the statistical estimates. In other
+direct interactions, it may be useful to compare participant selections
+or annotations to closest points on the chart to assess anchoring
+behavior; for discrete selections, methods discussed in numerical
+estimation may also be useful.
 
@@
 
+\bibitem[\citeproctext]{ref-robinson2023eye}
+Robinson, E. A., Howard, R., \& VanderPlas, S. (2023a). Eye fitting
+straight lines in the modern era. \emph{Journal of Computational and
+Graphical Statistics}, \emph{32}(4), 1537--1544.
+
+\bibitem[\citeproctext]{ref-robinson2023you}
+Robinson, E. A., Howard, R., \& VanderPlas, S. (2023b). {``You draw
+it''}: Implementation of visually fitted trends with r2d3. \emph{Journal
+of Data Science}, \emph{21}(2), 281--294.
+
 \bibitem[\citeproctext]{ref-robinsonPerceptionCognitiveImplications}

### 2025-04-19 index.qmd
Add COI statement
Cleaning up

@@
 title: A Guide to Designing Experiments to Test Statistical Graphics
-author:
-  - name: Emily Robinson
+authors:
+  - id: ear
+    name: Emily Robinson
     dept: Statistics Department
-    affiliation: 
     email: erobin17@calpoly.edu
@@
     affiliations: 
-      - name: Statistics Department, California Polytechnic State University
+    - name: California Polytechnic State University
+      department: Statistics Department
       address: Faculty Offices East, Building 25, Room 107D
@@
       postal-code: 93407-0405
-  - name: Heike Hofmann
+
+  - id: hh
+    name: Heike Hofmann
     orcid: 0000-0001-6216-5183
@@
     affiliations: 
-      - name: Statistics Department, University of Nebraska - Lincoln
+    - name: University of Nebraska - Lincoln
+      department: Statistics Department
       address: 340 Hardin Hall North Wing
@@
       postal-code: 68503
-  - name: Susan Vanderplas
+
+  - id: svp
+    name: Susan Vanderplas
     email: svanderplas2@unl.edu
     orcid: 0000-0002-3803-0972
+    corresponding: true
     affiliations: 
-      - name: Statistics Department, University of Nebraska - Lincoln
+    - name: University of Nebraska - Lincoln
+      department: Statistics Department
       address: 340 Hardin Hall North Wing
@@
       postal-code: 68503
+
+thanks: "The author(s) received no specific funding for this work."
 
@@
 bibliography: refs.bib
-csl: apa.csl
-
-mainfont: Helvetica
-monofont: Roboto
-number-depth: 2
-number-sections: true
 
@@
    - \usepackage[dvipsnames]{xcolor} % colors
+   - \renewcommand{\thefootnote}{\arabic{footnote}}
    - \newcommand{\ear}[1]{{\textcolor{blue}{#1}}}
@@
    - \newcommand{\hh}[1]{{\textcolor{Green}{#1}}}
-
 date: last-modified
@@
     keep-tex: true
-    fontsize: 10pt
+    template-partials:
+      - partials/title.tex
+      - partials/_print_author.tex
 ---
@@
 
-```{r lof-density-curves, echo = F, include = T, message = F, warning = F}
+```{r, echo = F, include = T, message = F, warning = F}
 #| label: fig-lof-density-curves
@@
 
-```{r, qe2-density-plot, message=FALSE, warning=FALSE, echo = F}
+```{r, message=FALSE, warning=FALSE, echo = F}
 #| label: fig-density-rug
@@
 
+# Author Statement
+The authors have no conflicts of interest or additional funding to declare in this manuscript.
+
 # References

### 2025-04-19 index.tex
Add COI statement
Cleaning up

@@
   10pt,
-  letterpaper,
-  DIV=11,
-  numbers=noendperiod]{scrartcl}
+]{article}
 
@@
 
-\KOMAoption{captions}{tableheading}
+\usepackage{orcidlink}
+\definecolor{mypink}{RGB}{219, 48, 122}
 \usepackage[dvipsnames]{xcolor} % colors
+\renewcommand{\thefootnote}{\arabic{footnote}}
 \newcommand{\ear}[1]{{\textcolor{blue}{#1}}}
@@
   pdftitle={A Guide to Designing Experiments to Test Statistical Graphics},
-  pdfauthor={Emily Robinson; Heike Hofmann; Susan Vanderplas},
   colorlinks=true,
@@
   citecolor={Blue},
-  urlcolor={Blue},
+  urlcolor={red},
   pdfcreator={LaTeX via pandoc}}
 
-\title{A Guide to Designing Experiments to Test Statistical Graphics}
-\author{Emily Robinson \and Heike Hofmann \and Susan Vanderplas}
-\date{2025-03-26}
-
+\title{A Guide to Designing Experiments to Test Statistical
+Graphics\thanks{The author(s) received no specific funding for this
+work.}}
+
+%% Author information
+\author{Emily
+Robinson\footnote{Email: erobin17@calpoly.edu}~\orcidlink{0000-0001-9800-7304}\\{\footnotesize Statistics
+Department}, {\footnotesize California Polytechnic State
+University}\\Heike
+Hofmann\footnote{Email: hhofmann4@unl.edu}~\orcidlink{0000-0001-6216-5183}\\{\footnotesize Statistics
+Department}, {\footnotesize University of Nebraska - Lincoln}\\Susan
+Vanderplas\footnote{Email: svanderplas2@unl.edu}~\orcidlink{0000-0002-3803-0972}\\{\footnotesize Statistics
+Department}, {\footnotesize University of Nebraska - Lincoln}\\}
 \begin{document}
@@
 
+\section{Author Statement}\label{author-statement}
+
+The authors have no conflicts of interest or additional funding to
+declare in this manuscript.
+
 \section{References}\label{references}

### 2025-04-19 partials/title.tex
Cleaning up

@@
+$-- Provides configuration of document metadata for writing the title block.
+$-- Note that in addition to these templates and partials, Quarto will also make normalized authors and affiliations data available to the template,
+$-- making is easy to write custom title blocks against a standard schema.
+$--
+$-- %%%% TODO %%%%%
+$-- Customize is needed, like below for printing the authors. Otherwise remove this partials to use Quarto default one.
+$-- %%%%%%%%%%%%%%%%
+
 $if(title)$
-  \title{$title$}
+\title{$title$$if(thanks)$\thanks{$thanks$}$endif$}
 $endif$
 $if(subtitle)$
+\usepackage{etoolbox}
+\makeatletter
+\providecommand{\subtitle}[1]{% add subtitle to \maketitle
+  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
+}
+\makeatother
 \subtitle{$subtitle$}
 $endif$
+$-- This supports Quarto's authors normalization
 
-\author{%
-$for(author-list)$%
-  $it.name$\thanks{email: $it.email$}\\\scriptsize $it.dept$\\\scriptsize $it.affiliation$%
-  $sep$\and%
-$endfor$%
-}
-
-$if(date)$
-  \date{$date$}
-$endif$
-
-$if(teaserimg)$
-\teaser{
-  \centering
-  \includegraphics[width=\linewidth]{$teaser-img$}
-  \caption{$teaser-caption$}
-  \label{fig:teaser}
-}
-$endif$
-
-$if(abstract)$
-\abstract{$abstract$}
-$endif$
-
-\vgtccategory{$category$}
-
-$if(ieee-keywords)$
-\CCScatlist{
-$for(ieee-keywords)$
-$it$
-$sep$;
-$endfor$
-}
-$endif$
+%% Author information
+\author{$for(by-author)$$_print_author.tex()$\\$endfor$}

### 2025-04-22 index.tex
fix corresponding author thing

@@
 %% Author information
-\author{Emily
-Robinson\footnote{Email: erobin17@calpoly.edu}~\orcidlink{0000-0001-9800-7304}\\{\footnotesize Statistics
-Department}, {\footnotesize California Polytechnic State
-University}\\Heike
-Hofmann\footnote{Email: hhofmann4@unl.edu}~\orcidlink{0000-0001-6216-5183}\\{\footnotesize Statistics
-Department}, {\footnotesize University of Nebraska - Lincoln}\\Susan
-Vanderplas\footnote{Email: svanderplas2@unl.edu}~\orcidlink{0000-0002-3803-0972}\\{\footnotesize Statistics
-Department}, {\footnotesize University of Nebraska - Lincoln}\\}
+\author{%
+%
+Emily Robinson%\footnote{Email: erobin17@calpoly.edu}
+~\orcidlink{0000-0001-9800-7304}%
+%
+\\{\footnotesize Statistics Department}, {\footnotesize California
+Polytechnic State University}\\%
+Heike Hofmann%\footnote{Email: hhofmann4@unl.edu}
+~\orcidlink{0000-0001-6216-5183}%
+%
+\\{\footnotesize Statistics Department}, {\footnotesize University of
+Nebraska - Lincoln}\\%
+Susan Vanderplas%\footnote{Email: svanderplas2@unl.edu}
+~\orcidlink{0000-0002-3803-0972}%
+\footnote{Corresponding author: Susan Vanderplas, svanderplas2@unl.edu}%
+\\{\footnotesize Statistics Department}, {\footnotesize University of
+Nebraska - Lincoln}\\
+}
 \begin{document}

### 2025-06-04 index.qmd
Add references and make some minor fixes to language. Convert refs.bib to ASCII.

@@
 
-Developing a graphics experiment is often a highly iterative process, but it can help to approach the design process by first optimizing the model and data generation method before spending time on optimizing the specific stimuli or customizing the data collection platform.
+Developing a graphics experiment is often a highly iterative process, but it can help to approach the design process by first optimizing the model and data generation method (@sec-model-dev) before spending time on optimizing the specific stimuli or customizing the data collection platform (@sec-protocol-dev).
 This is important because the model parameters and data generation process inform the experiment structure and thus impact decisions made downstream.
@@
 
-One reason why it is so difficult to generate sampling models for visual explorations is that our visual system is optimized for identifying differences between groups.
-This ability can interfere with the natural  to use the null sampling models that might be used in equivalent numerical tests when running experiments that use visualizations. 
+One reason why it is so difficult to generate sampling models for visual explorations is that our visual system is very good at creating groups and identifying differences between them [@zekiFunctionalSpecializationGeneralization2013;@lupyanConceptualGroupingEffect2008].
+This ability can interfere with the natural inclination to use the null sampling models that might be used in equivalent numerical tests when running experiments that use visualizations. 
+Numerical tests consider one facet of the data; the power of graphical testing is that it allows examination of a battery of hypotheses simultaneously. 
+Unfortunately, that makes designing an effective null sampling model much more difficult than in the numerical test case, as the researcher must effectively control for several different hypotheses when generating null data.
+
 We re-ran the experiment using a different clustering method that controlled the number of points in each group. 
@@
 
+Once the pilot data has been collected and the planned analysis method has been developed, it may be beneficial to pre-register the main study and proposed analysis. 
+This can be an effective way to separate exploratory research from confirmatory research involving hypothesis testing. 
+However, graphics studies (and other human subjects research) often produce unexpected results that would not be imagined during the pre-registration process (for an example, see @clusters); using only the pre-registered analysis method may disincentivise analyzing (or collecting) the rich, messy open-ended response data which can be incredibly useful but is hard to model and quantify.
+
 # Analyzing the Data {#sec-analysis}

### 2025-06-04 index-revised.qmd
Revision updates

--- 2025-04-19 index.qmd
@@
 A demo version of this application can be found at https://shiny.srvanderplas.com/sine-illusion/.
-](sine_illusion_screenshot.png){#fig-sine-illusion}
+](images/sine_illusion_screenshot.png){#fig-sine-illusion}
 
@@
 
-![This question asks users to write out a description of how the population of Ewoks changes over time, without any further cues, to determine whether participants default to multiplicative or additive language descriptions.](Estimation_describe_plot_crop.png){#fig-estimation-describe}
+![This question asks users to write out a description of how the population of Ewoks changes over time, without any further cues, to determine whether participants default to multiplicative or additive language descriptions.](images/Estimation_describe_plot_crop.png){#fig-estimation-describe}
 
@@
 ![This question asks participants for a numerical estimate, but provides a basic calculator and scratchpad.
-All user interactions with the calculator and scratchpad are logged, providing insight into the user's thought process and estimation strategy.](Estimation_numerical_screenshot_crop.png){#fig-estimation-calc}
+All user interactions with the calculator and scratchpad are logged, providing insight into the user's thought process and estimation strategy.](images/Estimation_numerical_screenshot_crop.png){#fig-estimation-calc}
 
@@
 
-```{r, echo = F, include = T, message = F, warning = F}
+```{r lof-density-curves, echo = F, include = T, message = F, warning = F}
 #| label: fig-lof-density-curves
@@
 Panel 10 shows the clustered target data and panel 17 shows the target data with a strong linear relationship; either of these target panels was the expected choice.
-Unfortunately, panel 16 has only two bounding ellipses shown, which is an unintentional difference that resulted from a faulty method for assigning clusters to null plots; many participants selected this panel instead of one of the target panels.](lineup-missing-ellipse.png){#fig-lineup-problems}
+Unfortunately, panel 16 has only two bounding ellipses shown, which is an unintentional difference that resulted from a faulty method for assigning clusters to null plots; many participants selected this panel instead of one of the target panels.](images/lineup-missing-ellipse.png){#fig-lineup-problems}
 
@@
 
-```{r, message=FALSE, warning=FALSE, echo = F}
+```{r, qe2-density-plot, message=FALSE, warning=FALSE, echo = F}
 #| label: fig-density-rug
@@
 
-# Author Statement
-The authors have no conflicts of interest or additional funding to declare in this manuscript.
+# Author Statements
+
+**Conflict of Interest** The authors have no conflicts of interest to declare in this manuscript.
+
+**Funding** No outside funding was used to create this manuscript.
+
+**Data Availability** This manuscript does not present any newly collected data.
+Simulated data and code to create many of the images in this paper can be found at https://github.com/srvanderplas/Guide-Testing-Statistical-Graphics.
 

### 2025-06-04 index-revised.tex
Revision updates

--- 2025-04-22 index.tex
@@
 While there are heuristics, general guidelines, and best practices
-(Allen \& Erhardt, 2016; Few, 2006; Haemer, 1948; {``Joint Committee on
-Standards for Graphic Presentation,''} 1915; Kosslyn, 2006) for creating
-useful and visually attractive data displays, the best way to establish
-the efficacy of various design decisions is to test the visualization on
-humans, evaluating different variants under controlled conditions
-(Cleveland et al., 1988; Cleveland \& McGill, 1985). Empirical
-assessments of visualizations, when carefully designed, allow
-statisticians to determine which representation of the same data is most
-effective along one or more dimension(s) of interest: estimation or
-prediction accuracy, within or between group comparisons, response time,
-and ability to draw real-world decisions are common goals for charts.
+(Allen \& Erhardt, 2016; Few, 2006; Haemer, 1948; Kosslyn, 2006;
+presentation, 1915) for creating useful and visually attractive data
+displays, the best way to establish the efficacy of various design
+decisions is to test the visualization on humans, evaluating different
+variants under controlled conditions (Cleveland et al., 1988; Cleveland
+\& McGill, 1985). Empirical assessments of visualizations, when
+carefully designed, allow statisticians to determine which
+representation of the same data is most effective along one or more
+dimension(s) of interest: estimation or prediction accuracy, within or
+between group comparisons, response time, and ability to draw real-world
+decisions are common goals for charts.
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{sine_illusion_screenshot.png}}
+\pandocbounded{\includegraphics[keepaspectratio]{images/sine_illusion_screenshot.png}}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{Estimation_describe_plot_crop.png}}
+\pandocbounded{\includegraphics[keepaspectratio]{images/Estimation_describe_plot_crop.png}}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{Estimation_numerical_screenshot_crop.png}}
+\pandocbounded{\includegraphics[keepaspectratio]{images/Estimation_numerical_screenshot_crop.png}}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/fig-lof-density-curves-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{index-revised_files/figure-pdf/fig-lof-density-curves-1.pdf}}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{lineup-missing-ellipse.png}}
+\pandocbounded{\includegraphics[keepaspectratio]{images/lineup-missing-ellipse.png}}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/fig-density-rug-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{index-revised_files/figure-pdf/fig-density-rug-1.pdf}}
 
@@
 
-\section{Author Statement}\label{author-statement}
-
-The authors have no conflicts of interest or additional funding to
-declare in this manuscript.
+\section{Author Statements}\label{author-statements}
+
+\textbf{Conflict of Interest} The authors have no conflicts of interest
+to declare in this manuscript.
+
+\textbf{Funding} No outside funding was used to create this manuscript.
+
+\textbf{Data Availability} This manuscript does not present any newly
+collected data. Simulated data and code to create many of the images in
+this paper can be found at
+https://github.com/srvanderplas/Guide-Testing-Statistical-Graphics.
 
@@
 \emph{Handbook of {Psychophysiology}} (4th ed., pp. 679--697). Cambridge
-University Press.
-https://doi.org/\url{https://doi.org/10.1017/9781107415782.031}
+University Press. \url{https://doi.org/10.1017/9781107415782.031}
 
@@
 {Documents}. \emph{IEEE Transactions on Visualization and Computer
-Graphics}, \emph{17}(12), 2301--2309.
-https://doi.org/\url{https://doi.org/10/b7bhhf}
+Graphics}, \emph{17}(12), 2301--2309. \url{https://doi.org/10/b7bhhf}
 
@@
 Graphical Statistics}, \emph{27}(1), 132--145.
-https://doi.org/\url{https://doi.org/10.1080/10618600.2017.1356323}
+\url{https://doi.org/10.1080/10618600.2017.1356323}
 
@@
 Statistical Association}, \emph{83}(402), 289--300.
-https://doi.org/\url{https://doi.org/10.1080/01621459.1988.10478598}
+\url{https://doi.org/10.1080/01621459.1988.10478598}
 
@@
 \emph{Science}, \emph{229}(4716), 828--833.
-https://doi.org/\url{https://doi.org/10.1126/science.229.4716.828}
+\url{https://doi.org/10.1126/science.229.4716.828}
 
@@
 {Inferentially}. \emph{Harvard Data Science Review}, \emph{3}(3).
-https://doi.org/\url{https://doi.org/10.1162/99608f92.8453435d}
+\url{https://doi.org/10.1162/99608f92.8453435d}
 
@@
 {Diagrams}. \emph{Journal of the American Statistical Association},
-\emph{22}(160), 473--482.
-https://doi.org/\url{https://doi.org/10.2307/2276829}
+\emph{22}(160), 473--482. \url{https://doi.org/10.2307/2276829}
 
@@
 \bibitem[\citeproctext]{ref-desnoyersTaxonomyVisualsScience2011}
-Desnoyers, L. (2011). \emph{Toward a {Taxonomy} of {Visuals} in
-{Science} {Communication}}. \emph{58}(2), 16.
+Desnoyers, L. (2011). Toward a {Taxonomy} of {Visuals} in {Science}
+{Communication}. \emph{Technical Communication}, \emph{58}(2), 16.
 
@@
 Statistical Association}, \emph{21}(154), 119--132.
-https://doi.org/\url{https://doi.org/10.2307/2277140}
+\url{https://doi.org/10.2307/2277140}
 
@@
 Concurrent think-aloud protocols: Testing the usability of an online
-library catalogue. \emph{Behaviour \& {IT}}, \emph{22}, 339--351.
-\url{https://doi.org/10.1080/0044929031000}
+library catalogue. \emph{Behaviour \& {Information Technology}},
+\emph{22}, 339--351. \url{https://doi.org/10.1080/0044929031000}
 
@@
 American Statistician}, \emph{2}(3), 24--24.
-https://doi.org/\url{https://doi.org/10.1080/00031305.1948.10501588}
+\url{https://doi.org/10.1080/00031305.1948.10501588}
 
@@
 Haider, J. D., Pohl, M., Beecham, R., \& Dykes, J. (2021). Strategies
-for detecting difference in~map line-up tasks. In C. Ardito, R.
+for detecting difference in map line-up tasks. In C. Ardito, R.
 Lanzilotti, A. Malizia, H. Petrie, A. Piccinno, G. Desolda, \& K. Inkpen
-(Eds.), \emph{Human-computer interaction -- {INTERACT} 2021} (pp.
+(Eds.), \emph{Human-computer interaction - {INTERACT} 2021} (pp.
 558--578). Springer International Publishing.
@@
 {Data}. \emph{The Annals of Statistics}, \emph{19}(4), 2244--2253.
-https://doi.org/\url{https://doi.org/10.1214/aos/1176348396}
+\url{https://doi.org/10.1214/aos/1176348396}
 
@@
 \emph{Journal of the American Statistical Association}, \emph{22}(157),
-31--39. https://doi.org/\url{https://doi.org/10.2307/2277346}
+31--39. \url{https://doi.org/10.2307/2277346}
 
@@
 {Inference}. \emph{Harvard Data Science Review}, \emph{3}(3).
-https://doi.org/\url{https://doi.org/10.1162/99608f92.3ab8a587}
-
-\bibitem[\citeproctext]{ref-asme-standards-graphics}
-Joint committee on standards for graphic presentation. (1915).
-\emph{Publications of the American Statistical Association},
-\emph{14}(112), 790--797. \url{http://www.jstor.org/stable/2965153}
+\url{https://doi.org/10.1162/99608f92.3ab8a587}
 
@@
 {Human} {Factors} and {Ergonomics} {Society} {Annual} {Meeting}},
-\emph{47}, 473--477.
-https://doi.org/\url{https://doi.org/10.1177/154193120304700347}
+\emph{47}, 473--477. \url{https://doi.org/10.1177/154193120304700347}
 
@@
 Statistics}, \emph{5}(1), 48--61.
-https://doi.org/\url{https://doi.org/10.1002/wics.1238}
+\url{https://doi.org/10.1002/wics.1238}
 
@@
 Lu, M., Lanir, J., Wang, C., Yao, Y., Zhang, W., Deussen, O., \& Huang,
-H. (2022). Modeling {Just} {Noticeable} {Differences} in {Charts}.
-\emph{IEEE Transactions on Visualization and Computer Graphics},
-\emph{28}(1), 718--726.
-https://doi.org/\url{https://doi.org/10.1109/TVCG.2021.3114874}
+H. (2022). Modeling just noticeable differences in charts. \emph{{IEEE}
+Transactions on Visualization and Computer Graphics}, \emph{28}(1),
+718--726. \url{https://doi.org/10.1109/TVCG.2021.3114874}
 
@@
 American Statistical Association}, \emph{108}(503), 942--956.
-https://doi.org/\url{https://doi.org/10/f5gntt}
+\url{https://doi.org/10/f5gntt}
 
@@
 \url{https://doi.org/10.3758/s13428-021-01694-3}
+
+\bibitem[\citeproctext]{ref-asme-standards-graphics}
+presentation, J. committee on standards for graphic. (1915). Preliminary
+report published for the purpose of inviting suggestions for the benefit
+of the committee. \emph{Publications of the American Statistical
+Association}, \emph{14}(112), 790--797.
+\url{http://www.jstor.org/stable/2965153}
 
@@
 \bibitem[\citeproctext]{ref-robinson2023you}
-Robinson, E. A., Howard, R., \& VanderPlas, S. (2023b). {``You draw
-it''}: Implementation of visually fitted trends with r2d3. \emph{Journal
-of Data Science}, \emph{21}(2), 281--294.
+Robinson, E. A., Howard, R., \& VanderPlas, S. (2023b). 'You draw it':
+Implementation of visually fitted trends with r2d3. \emph{Journal of
+Data Science}, \emph{21}(2), 281--294.
 
@@
 psychology of survey response}. Cambridge University Press.
-https://doi.org/\url{https://doi.org/10.1017/CBO9780511819322}
+\url{https://doi.org/10.1017/CBO9780511819322}
 
@@
 \emph{International Journal of Human-Computer Studies}, \emph{53}(5),
-827--850. \url{https://doi.org/10.1006/ijhc.2000.0419}
+827--850. \href{https://10.1006/ijhc.2000.0419}{10.1006/ijhc.2000.0419}
 
@@
 \emph{Journal of the Korean Statistical Society}, \emph{46}(3),
-426--437.
-https://doi.org/\url{https://doi.org/10.1016/j.jkss.2017.01.003}
+426--437. \url{https://doi.org/10.1016/j.jkss.2017.01.003}
 
 \bibitem[\citeproctext]{ref-valentinDesigningOptimalBehavioral2024}
-Valentin, S., Kleinegesse, S., Bramley, N. R., Seriès, P., Gutmann, M.
+Valentin, S., Kleinegesse, S., Bramley, N. R., Seriés, P., Gutmann, M.
 U., \& Lucas, C. G. (2024). Designing optimal behavioral experiments
@@
 {Empirical} {Results}. \emph{Harvard Data Science Review}, \emph{3}(3).
-https://doi.org/\url{https://doi.org/10.1162/99608f92.7d099fd0}
+\url{https://doi.org/10.1162/99608f92.7d099fd0}
 
@@
 Statistics and Its Application}, \emph{7}(1).
-https://doi.org/\url{https://doi.org/10.1146/annurev-statistics-031219-041252}
+\url{https://doi.org/10.1146/annurev-statistics-031219-041252}
 
@@
 Graphical Statistics}.
-https://doi.org/\url{https://doi.org/10.1080/10618600.2018.1562937}
+\url{https://doi.org/10.1080/10618600.2018.1562937}
 
 \bibitem[\citeproctext]{ref-sineillusion}
-Vanderplas, S., \& Hofmann, H. (2015). Signs of the {Sine}
-{Illusion}---{Why} {We} {Need} to {Care}. \emph{Journal of Computational
-and Graphical Statistics}, \emph{24}(4), 1170--1190.
-https://doi.org/\url{https://doi.org/10.1080/10618600.2014.951547}
+Vanderplas, S., \& Hofmann, H. (2015). Signs of the {Sine} {Illusion} --
+{Why} {We} {Need} to {Care}. \emph{Journal of Computational and
+Graphical Statistics}, \emph{24}(4), 1170--1190.
+\url{https://doi.org/10.1080/10618600.2014.951547}
 
@@
 Statistical significance calculations for scenarios in visual inference.
-\emph{Stat}, \emph{10}(1).
-https://doi.org/\url{https://doi.org/10.1002/sta4.337}
+\emph{Stat}, \emph{10}(1). \url{https://doi.org/10.1002/sta4.337}
 
@@
 Computer Graphics}, \emph{16}(6), 973--979.
-https://doi.org/\url{https://doi.org/10.1109/TVCG.2010.161}
+\url{https://doi.org/10.1109/TVCG.2010.161}
 

### 2025-06-04 correspondence/revision-review-response.qmd
Updates to correspondence

---
title: "Revision Review Responses"
format: pdf
---

We thank both reviewers for their comments, and in particular, Reviewer 2, who caught a number of things that were minor but embarrassing typos!

## Wanted fixes

- It should be noted that the on page 3 (in diff.pdf), the authors make reference to a yet to be published article. Please remember to fix this before publication, or change it to “provided in a future article” if yet not available at the time of publication.

> Yes, this is important -- but it will have to be added by the editorial staff, because we don't know the citation for the paper.

- Page 8 (in diff.pdf)-> “opimizing the model and generation method” -> more specificity needed — model and generation method for what? This looks like it’s the subject of Section 4, so just a mention that it’ll be covered in more detail in Section 4 would be helpful.

> Parenthetical references have been added to section 4 and 5 to clarify

- Authors don’t mention pre-registration. Author themselves don’t seem to have done pre-registration in the past. I think it warrants a small mention as part of the protocol.

> We haven't found preregistration to be particularly helpful (or feasible) in the past, but much of our work has been more exploratory in nature. We've added a paragraph with the pros and cons of preregistration to the section on pilot testing, as visual experiments so easily go sideways that preregistration before pilot testing seems overly confident and likely to fail.

- “our visual system is optimised for identifying differences between groups” -> any citation to back-up this claim?

> We've added citations to both neuroscience papers identifying specific regions of the brain which process groups and differences between them, as well as to the more phenomenological Gestalt psychology view of heuristics for similarity and grouping -- grouping necessarily involves clustering based on characteristics that are similar (within the group) and different (between groups).

- “This ability can interfere with the natural to use the null sampling models” -> looks like a grammar issue — I don’t understand what the authors mean by “with the natural to use the null sampling models”.

> Yes, thanks for catching that, we left out a word.

## Minor

- Page 8 (in diff.pdf) -> “experiment structure” -> “experimental structure"

> Fixed

- Page 11 (in diff.pdf) -> “2x3x3” replace x with $\times$.

> Fixed

- Some of the citations need fixing for DOI as it contains prefix duplicates: "https://urldefense.com/v3/__https://doi.org/https:/*doi.org__;Lw!!PvXuogZ4sRB2p-tU!Cq6ylkfZW13S760wtA3vZIM5ImtHcC0SuhnWAXTDntbx8_ats4DGZjOjbbdP8BD98BtxElGHT_hAj_aLXbZhtDkY_SMUV-ARig$ "

> I couldn't find any evidence of this one, but I've gone over the bibliography and made sure there aren't any transformed URLs in it.

### 2025-06-04 correspondence/revision-review.qmd
Updates to correspondence

---
title: "Revision Reviews"
format: html
---

# Reviewer: 1

## Comments to the Author(s)
I appreciate the thoughtful and thorough responses to the previous comments. I have no additional suggestions for this version.

# Reviewer: 2

## Comments to the Author(s)
The authors have largely addressed comments from the first review. The paper reads much stronger now. Well done to the authors.

## Wanted fixes

- It should be noted that the on page 3 (in diff.pdf), the authors make reference to a yet to be published article. Please remember to fix this before publication, or change it to “provided in a future article” if yet not available at the time of publication.
- Page 8 (in diff.pdf)-> “opimizing the model and generation method” -> more specificity needed — model and generation method for what? This looks like it’s the subject of Section 4, so just a mention that it’ll be covered in more detail in Section 4 would be helpful.
- Authors don’t mention pre-registration. Author themselves don’t seem to have done pre-registration in the past. I think it warrants a small mention as part of the protocol.
- “our visual system is optimised for identifying differences between groups” -> any citation to back-up this claim?
- “This ability can interfere with the natural to use the null sampling models” -> looks like a grammar issue — I don’t understand what the authors mean by “with the natural to use the null sampling models”.

## Minor

- Page 8 (in diff.pdf) -> “experiment structure” -> “experimental structure"
- Page 11 (in diff.pdf) -> “2x3x3” replace x with $\times$.
- Some of the citations need fixing for DOI as it contains prefix duplicates: "https://urldefense.com/v3/__https://doi.org/https:/*doi.org__;Lw!!PvXuogZ4sRB2p-tU!Cq6ylkfZW13S760wtA3vZIM5ImtHcC0SuhnWAXTDntbx8_ats4DGZjOjbbdP8BD98BtxElGHT_hAj_aLXbZhtDkY_SMUV-ARig$ "
